---
title: "Read single-cell and meta-data - AD"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# Plots directory
plot_dir <- file.path(params$output, "plots_AD/01_read_data")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot, dispatching on its object type
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # base R / other: assume already drawn
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  message("Saved: ", full_path)
}

# Global knitr chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Read in the data {#read-data}

We  will first read in single-cell data, previously extracted from processed multi-modal (IMC,IF) images in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP) or downloaded from [Zenodo](https://zenodo.org/records/13220635/files/MapMetIP_ProcessedDataset.zip). For reading the data, we use the [imcRtools](https://github.com/BodenmillerGroup/imcRtools) package. This package uses the `read_steinbock` function to read data into a `SpatialExperiment` (SPE). The [SpatialExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) class [@Righelli2021] is an extension of the `SingleCellExperiment` class. 
For more information, please refer to `?read_steinbock`. 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(imcRtools)
library(stringr)
library(RColorBrewer)
library(circlize)
library(paletteer)
library(dittoSeq)
library(readxl)
library(dplyr)
library(patchwork)
```
</details>

## Preprocess adrenal gland (AD) files

All AD samples share the input folder with the other cohorts; we read them
as-is and never modify the input CSVs. The supplementary channel
`IF4_NEGATIVE_MIX` is **retained**, matching the main analysis. To exclude any
channel, drop it from the SPE in memory after reading rather than editing the
inputs.
## Read single-cell measurement

Replace `path` with the path to the folder where you stored the single-cell data (e.g. `path/to/folder/MapMetIP_ProcessedDataset` in case processed data was downloaded from Zenodo). 
In [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP), we exported single-cell marker measurements based on non-dilated and dilated (by 1px) segmentation masks to capture membrane signals. As both are required in the downstream analysis, we will read in both and combine them into one SPE object saving each expression matrix into one `assays`slot. Measurements based on non-dilated masks will be in the `assays`slot `counts` and the measurements based on dilated masks will be stored in `counts_1px`. 


```{r read-steinbock}
path <- file.path(params$input)

# All cohorts (AD, TU, BM) share one folder; select AD files by filename pattern.
sample_pattern <- "AD"

spe_0px <- read_steinbock(path=path, image_file=NULL, panel_file = file.path(path, 'NB_Panel.csv'), intensities_folder = "intensities-0px",regionprops_folder = "regionprops", graphs_folder=NULL, pattern = sample_pattern)
spe_1px <- read_steinbock(path=path, image_file=NULL, panel_file = file.path(path, 'NB_Panel.csv'), intensities_folder = "intensities-1px",regionprops_folder = "regionprops", graphs_folder=NULL, pattern = sample_pattern)

# Combine into one SPE: counts = 0px masks, counts_1px = 1px masks
spe <- spe_0px
assay(spe, "counts_1px") <- assay(spe_1px, "counts")

rm(spe_0px)
rm(spe_1px)
```

## Set cell identifiers and sample IDs

We set the `colnames` of the object to generate unique identifiers per cell.

```{r set-colnames}
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)
```

For the adrenal gland (AD) samples there is no sample- or patient-specific metadata table to read in. We only derive `sample` and `fm_id` identifiers from the `sample_id` and assign these to the colData slot of our SPE object. Tissue, year, stainer, etc. are not relevant here, as they are identical for all AD samples.

```{r add-coldata}

spe$sample <- sub("_[^_]+$", "", spe$sample_id)
spe$fm_id <- str_match(spe$sample, "_.*_(.*)_")[,2]

```

Before we save the SPE object, we check the distribution of the two DNA channels across samples.

```{r plot by staining date}

marker <- 'DNA1_Ir191_mean'
p1 <- dittoRidgePlot(spe, var = marker, group.by = "sample", assay = "counts") +
    ggtitle(paste(marker,  "- per sample"))
marker <- 'DNA1_Ir193_mean'
p2 <- dittoRidgePlot(spe, var = marker, group.by = "sample", assay = "counts") +
    ggtitle(paste(marker,  "- per sample"))
p <- (p1 | p2)

p

save_plot(p, "DNA_markers_rawdata.pdf", width = 24, height = 12)

```
## Save object

Finally, the generated SPE object can be saved for further downstream processing and analysis.

```{r save-objects-read-data}

saveRDS(spe, file.path(params$output, "spe_01_read_data_AD.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
