---
title: "Read single-cell and meta-data - AD"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: "/home/rstudio/data/input_AD"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc_AD"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/01_read_data")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Read in the data {#read-data}

We  will first read in single-cell data, previously extracted from processed multi-modal (IMC,IF) images in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP) or downloaded from [Zenodo](https://zenodo.org/records/13220635/files/MapMetIP_ProcessedDataset.zip). For reading the data, we use the [imcRtools](https://github.com/BodenmillerGroup/imcRtools) package. This package uses the `read_steinbock` function to read data into a `SpatialExperiment` (SPE). The [SpatialExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) class [@Righelli2021] is an extension of the `SingleCellExperiment` class. 
For more information, please refer to `?read_steinbock`. 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(imcRtools)
library(stringr)
library(RColorBrewer)
library(circlize)
library(paletteer)
library(dittoSeq)
library(readxl)
library(writexl)
library(dplyr)
library(patchwork)
```
</details>

## Preprocess adrenal gland (AD) files to remove supplementary channels.This does not run in docker and needs to be an in parent folder of intensities.

```{r}

# Folders to process
folders <- c("intensities-0px/", "intensities-1px/")

# Columns to remove
cols_to_remove <- c("IF4_NEGATIVE_MIX_mean", "IF4_NEGATIVE_MIX_mean-80")

# Loop through each folder
for(folder in folders){
  # List CSVs containing "AD"
  ad_files <- list.files(folder, pattern="AD.*\\.csv$", full.names = TRUE)
  print(ad_files)

  # Loop through each file
  for(f in ad_files){
    df <- read.csv(f, check.names = FALSE)        # read CSV
    df <- df %>% select(-any_of(cols_to_remove)) # remove columns if present
    write.csv(df, f, row.names = FALSE)           # overwrite file
  }
}

```
## Read single-cell measurement

Replace `path` with the path to the folder where you stored the single-cell data (e.g. `path/to/folder/MapMetIP_ProcessedDataset` in case processed data was downloaded from Zenodo). 
In [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP), we exported single-cell marker measurements based on non-dilated and dilated (by 1px) segmentation masks to capture membrane signals. As both are required in the downstream analysis, we will read in both and combine them into one SPE object saving each expression matrix into one `assays`slot. Measurements based on non-dilated masks will be in the `assays`slot `counts` and the measurements based on dilated masks will be stored in `counts_1px`. 


```{r read-steinbock}
path <- file.path(params$input)

spe_0px <- read_steinbock(path=path, image_file=NULL, panel_file = file.path(path, 'NB_Panel.csv'), intensities_folder = "intensities-0px",regionprops_folder = "regionprops", graphs_folder=NULL) 
spe_1px <- read_steinbock(path=path, image_file=NULL, panel_file = file.path(path, 'NB_Panel.csv'), intensities_folder = "intensities-1px",regionprops_folder = "regionprops", graphs_folder=NULL) 

#Combine SPE objects into one object
spe <- spe_0px
assay(spe, "counts_1px") <- assay(spe_1px, "counts")

#remove single SPE objects
rm(spe_0px)
rm(spe_1px)
```

## Set cell identifiers and sample IDs

We set the `colnames` of the object to generate unique identifiers per cell.

```{r set-colnames}
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)
```

For the adrenal gland (AD) samples there is no sample- or patient-specific metadata table to read in. We only derive `sample` and `fm_id` identifiers from the `sample_id` and assign these to the colData slot of our SPE object. Tissue, year, stainer, etc. are not relevant here, as they are identical for all AD samples.

```{r add-coldata}

#spe$sample <- gsub(spe$sample_id, pattern="([UMD]).*", replacement="\\1")
spe$sample <- sub("_[^_]+$", "", spe$sample_id)
spe$fm_id <- str_match(spe$sample, "_.*_(.*)_")[,2]

```

Before we save the SPE object, we check the distribution of markers.

```{r plot by staining date}

marker <- 'DNA1_Ir191_mean'
p1 <- dittoRidgePlot(spe, var = marker, group.by = "sample", assay = "counts") +
    ggtitle(paste(marker,  "- per sample"))
marker <- 'DNA1_Ir193_mean'
p2 <- dittoRidgePlot(spe, var = marker, group.by = "sample", assay = "counts") +
    ggtitle(paste(marker,  "- per sample"))
p <- (p1 | p2)

p

save_plot(p, "DNA_markers_rawdata.pdf", width = 24, height = 12)

```
## Save object

Finally, the generated SPE object can be saved for further downstream processing and analysis.

```{r save-objects-read-data}

saveRDS(spe, file.path(params$output, "spe.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
