---
title: "Quality control - Part 2 - AD"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---


```{r setup, include=FALSE}
# Plots directory
plot_dir <- file.path(params$output, "plots_AD/03_QC_2")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot, dispatching on its object type
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # base R / other: assume already drawn
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  message("Saved: ", full_path)
}

# Global knitr chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Quality control - Exclusion of segmentation artifacts {#qc-2}
Despite the employment of a highly-accurate (fine-tuned) cellpose model for segmentation, segmentation artifacts can't be avoided completely. As a second quality control step, we will exclude these before we continue with downstream analysis. 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ggplot2)
library(cowplot)
library(cytomapper)
```
</details>

## Read SPE object

We will first read the spatial experiment object (SPE) that passed the [first quality check](#qc-1). Please make sure, `path` points to the folder where you stored the SPE object.


```{r read-steinbock}
path <- params$output
spe <- readRDS(file.path(path,"spe_02_QC_1_AD.rds"))
```

For the exclusion of segmentation artifacts, we will examine three morphological features (area, solidity and aspect ratio). All samples here are adrenal gland (AD) tissue, so — unlike the combined primary tumor / bone marrow analysis — we process all samples together with a single set of thresholds.

```{r morph-features}
morph_names <- c("area",
                 "solidity",
                 "aspect_ratio-0"
                 )
```

## Threshold identification for AD samples
Next, we plot the distributions of the morphological features (as density plots) to identify feature-specific thresholds for excluding segmentation artifacts.

```{r plot-histograms-ad}
morph_mat <- as.matrix(t(colData(spe)[,morph_names]))
rownames(morph_mat) <- gsub('-0','',rownames(morph_mat))
colnames(morph_mat) <- colnames(spe)

plot_list <- list()

df <- as.data.frame(t(morph_mat))

for(i in 1: ncol(df)){

  feature <- colnames(df)[i]

  plot_list[[i]] <- ggplot(df, aes_string(feature)) +
  geom_density(adjust = 1, color="darkblue", fill="lightblue")
}

plot <- plot_grid(plotlist = plot_list, nrow=ceiling(ncol(df)/3), ncol=3)


save_plot(plot, "morpho_features_AD.pdf", width = 12, height = 12)


plot
```

In the plots above, the distributions are skewed, indicating the presence of outliers. Let's examine these outlier cells. We will identify their cell ID and check their morphology in images and masks. We start with the largest cells.

```{r quality-control-outliers}
feat <- "area"

morph_mat_order <- morph_mat[feat,order(morph_mat[feat,], decreasing=T)]
outlier_cell <- names(morph_mat_order[2])
value <- morph_mat_order[outlier_cell]
id <- sub(".*_", "", outlier_cell)
image <- sub("_[^_]*$", "", outlier_cell)
print(paste("The outlier cell with ", feat, "=", value, " with ID", id, "is in image", image,"."))
```

Let's inspect this cell in the image and mask.

```{r visualize-outliers, fig.show='hide'}
masks <- loadImages(file.path(params$input, "masks", paste0(image, ".tif")), as.is = TRUE)
img <- loadImages(file.path(params$input, "img", paste0(image, ".tif")))

mcols(masks) <- DataFrame(sample_id = names(masks))
mcols(img) <- DataFrame(sample_id = names(img))

# Raw images may carry extra channels (e.g. IF4_NEGATIVE_MIX); name channels
# from the marker (_mean) names, padded to the image's actual channel count.
marker_names <- rownames(spe)[grepl("_mean$", rownames(spe))]
n_ch <- dim(img[[1]])[3]
channelNames(img) <- if (n_ch <= length(marker_names)) {
  marker_names[seq_len(n_ch)]
} else {
  c(marker_names, paste0("extra_channel_", seq_len(n_ch - length(marker_names))))
}

tmp <- spe[,spe$sample_id==image]
tmp$outlier <- "no"
tmp[,colnames(tmp)==outlier_cell]$outlier <- "yes"

dapi_plot <- plotPixels(img, 
                       colour_by = "IF1_DAPI_mean",
                       bcg = list(IF1_DAPI_mean = c(0, 1, 1)),
                       legend=NULL,
                       return_plot = TRUE)

mask_plot <- plotCells(masks,
                       object = tmp, 
                       cell_id = "ObjectNumber", 
                       img_id = "sample_id",
                       colour_by = "outlier",
                       colour = list(outlier = c("no" = "gray", "yes" = "red")),
                       missing_colour = "white",
                       legend=NULL,
                       return_plot = TRUE)
```
```{r plot-grid-1}
p1 <- ggdraw(dapi_plot$plot, clip = "on")
p2 <- ggdraw(mask_plot$plot, clip = "on")

p <- plot_grid(p1, p2)

save_plot(p, "example_large_cell.pdf", width = 12, height = 12)
p
```


In the images above, we can see that segmentation artifacts occur, where no clear cell borders can be identified. To examine nuclei with an area below the lower threshold, it is better to visualize images in napari and use the label and zoom function. Next, we inspect the smallest cells to identify a lower threshold.

```{r quality-control-outliers-2, message=FALSE}
feat <- "area"

morph_mat_order <- morph_mat[feat,order(morph_mat[feat,], decreasing=F)]
outlier_cell <- names(morph_mat_order[1])
print(paste("The outlier cell with ", feat, "=", morph_mat_order[outlier_cell], " with ID", sub(".*_", "", outlier_cell), "is in image", sub("_[^_]*$", "", outlier_cell),". Check, whether this cell is an artifact by inspecting the large segmentation masks in napari."))
```

Using the code above and cross-checking images of corresponding nuclei in napari, we identify an appropriate lower threshold for nuclear area. Adrenal gland is solid tissue, so — as for primary tumor — the largest cells are typically correctly segmented and no upper threshold is strictly required; inspect the distributions above to decide whether an upper threshold is needed.

## Exclusion of segmentation artifacts
Let's exclude outliers from the SPE object and the morph. feature matrix based on defined thresholds.

```{r exclude-outliers}
# AD segmentation-artifact thresholds; Inf = no upper limit (set a finite
# value if the distributions above reveal oversized artifacts).
area_lower_thresh <- 100
area_upper_thresh <- Inf

outliers <- (spe$area <= area_lower_thresh | spe$area >= area_upper_thresh)

spe <- spe[,!outliers]

# rebuild the morphology matrix for the kept cells
morph_mat <- as.matrix(t(colData(spe)[,morph_names]))
rownames(morph_mat) <- gsub('-0','',rownames(morph_mat))
```

We excluded `r sum(outliers)` cells based on these thresholds. 

## Clipping and scaling of morphological features

Before we save the SPE object, we will clip and scale morphological features for **visualization purposes**. 

```{r clip-morph}

norm_minmax <- function(x){(x- min(x)) /(max(x)-min(x))}

clipper <- function(data, lower, upper) {
  data[data > quantile(data, probs = c(upper))] <- quantile(data, probs = c(upper))
  data[data < quantile(data, probs = c(lower))] <- quantile(data, probs = c(lower))
  clipped_data <- norm_minmax(data)
  return(clipped_data)}

area_upper <- 0.99999
solidity_lower <- 0.001
aspect_ratio_upper <- 0.9995

clipped_area <- clipper(morph_mat[1,], 0, area_upper)

morph_mat_clipped <- rbind(clipped_area, clipper(morph_mat[2,], solidity_lower, 1), clipper(morph_mat[3,], 0, aspect_ratio_upper))

colnames(morph_mat_clipped) <- colnames(spe)
rownames(morph_mat_clipped) <- c("area", "solidity", "aspect_ratio")

spe$area_clipped <- unname(morph_mat_clipped["area",])
spe$solidity_clipped <- unname(morph_mat_clipped["solidity",])
spe$aspect_ratio_clipped <- unname(morph_mat_clipped["aspect_ratio",])
```

We plot the distribution of clipped morphological features...

```{r histograms-separate-x-axis}
plot_list <- list()

df <- as.data.frame(t(morph_mat_clipped))

for(i in 1: ncol(df)){

  feature <- colnames(df)[i]
  
  plot_list[[i]] <- ggplot(df, aes_string(feature)) +
  geom_density(adjust = 5, color="darkblue", fill="lightblue")
}

plot <- plot_grid(plotlist = plot_list, nrow=ceiling(ncol(df)/3), ncol=3)



save_plot(plot, "morph_.pdf", width = 12, height = 12)


plot
```

...and visualize the masks in one representative AD sample.

```{r visualize-clipped-morph}
morph_names_clipped <- c("area_clipped", "solidity_clipped", "aspect_ratio_clipped")

# one representative AD sample
sample_to_plot <- unique(spe$sample_id)[1]

masks <- loadImages(
  file.path(params$input, "masks", paste0(sample_to_plot, ".tif")),
  as.is = TRUE
)

names(masks) <- sample_to_plot
mcols(masks)$sample_id <- sample_to_plot

# one mask plot per clipped morphology feature
for (m in seq_along(morph_names_clipped)) {

  current_id <- sample_to_plot
  current_morph <- morph_names_clipped[m]

  message(paste0("Plotting: ", current_morph, " | ", current_id))

  p <- plotCells(
    masks,
    object = spe,
    cell_id = "ObjectNumber",
    img_id = "sample_id",
    colour_by = current_morph,
    return_plot = TRUE,
    return_images = TRUE,
    legend = NULL
  )

  save_plot(
    p,
    filename = paste0(current_morph, "_", current_id, ".pdf"),
    width = 12,
    height = 12
  )
}
```

## Save object {#qc2-save}

From now on, we will use intensity measurements computed based on the dilated mask (by 1px). To reduce the size of the SPE object, we remove intensity measurements obtained based on non-dilated masks and set those based on dilated masks as default. 

```{r save-objects-read-data}
assay(spe, "counts") <- assay(spe, "counts_1px")
assay(spe, "counts_1px") <- NULL

saveRDS(spe, file.path(params$output, "spe_03_QC_2_AD.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
