---
title: "Cell phenotyping - AD"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# Plots directory
plot_dir <- file.path(params$output, "plots_AD/04_phenotyping")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot, dispatching on its object type (optional subdir under plot_dir)
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300, subdir = NULL) {
  out_dir <- if (is.null(subdir)) plot_dir else file.path(plot_dir, subdir)
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  full_path <- file.path(out_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # base R / other: assume already drawn
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  message("Saved: ", full_path)
}

# Global knitr chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Cell phenotyping {#phenotyping}
In this Rmd file, we will
1. reduce dimensionality with UMAP (for visualization)
2. cluster cells with phenograph
3. annotate and merge clusters into cell types and metaclusters using an externally curated table
4. define color codes for all cell types and metaclusters

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ComplexHeatmap)
library(RColorBrewer)
library(viridis)
library(scater)
library(BiocParallel)
library(Rphenoannoy)
library(tidyr)
library(tidyHeatmap)
library(bruceR)
library(dittoSeq)
library(dplyr)
library(mclust)
library(SingleR)
library(circlize)
library(readxl)
library(ggplot2)
library(patchwork)
library(harmony)
library(batchelor)
library(cluster)
```
</details>

## Read data
We will first read the data that passed the [second quality control step](#qc-2) and save the count data as a matrix. Low-quality markers (low SNR in a lot of samples) as well as all IF and nuclear IMC markers will be excluded for clustering. Moreover, mean intensity will be used to represent marker expression, to average out the signals arising from lateral spillover or artifacts. As explained in [the previous chapter](#qc2-save), we are using counts measured with respect to dilated masks in downstream analysis to capture membrane signal.

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_03_QC_2_AD.rds"))

intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- colnames(spe)

# mean intensity (drop mean-80)
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]

# high-quality clustering markers: drop low-SNR, IF and nuclear markers
intensity_mat_hq <- intensity_mat[!grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(intensity_mat)),]

non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)

intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]

# drop all-zero cells and flag the clustering channels on the SPE
spe <- spe[,non_zeros]
rowData(spe)$use_channel <- !grepl("mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))

mat <- intensity_mat_hq_wo_zeros
```
We are excluding `r length(non_zeros[non_zeros==FALSE])` more cells here, as they are negative for all markers used in downstream analysis. 
To get a first grasp of the single-cell data and its structure, we compute a UMAP for visualization.


## UMAP (for visualization only)
```{r umap}
set.seed(221228)

Sys.setenv(OMP_NUM_THREADS = 1)
Sys.setenv(MKL_NUM_THREADS = 1)

BPPARAM <- SnowParam(workers = 30, type = "SOCK", RNGseed = 221228)

mat_red <- calculateUMAP(
  mat,
  n_neighbors = 30,
  BPPARAM = BPPARAM
)

rownames(mat_red) <- colnames(spe)
reducedDim(spe, "UMAP") <- mat_red

p <- dittoDimPlot(spe, var = "fm_id", 
             reduction.use = "UMAP", size = 1,
             do.label = TRUE,
             legend.show=F) +
  theme(legend.title = element_blank()) +
  ggtitle("Sample ID on UMAP")


save_plot(p, "UMAP-vis-sampleID.pdf", width = 8, height = 6)

p

```



## UMAPs for each marker
```{r UMAP-all-markers, fig.width=24, fig.height=32}
excluded_markers <- "mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3"
max_cells_for_umap <- 20000

marker_vars <- rownames(spe)[
  !grepl(excluded_markers, rownames(spe))
]

set.seed(221228)
marker_umap_ncells <- min(max_cells_for_umap, ncol(spe))
marker_umap_idx <- sample(seq_len(ncol(spe)), marker_umap_ncells)
spe_umap_markers <- spe[, marker_umap_idx]

p_markers <- multi_dittoDimPlot(
  spe_umap_markers,
  vars = marker_vars,
  reduction.use = "UMAP",
  size = 0.5,
  ncol = 5,
  color.panel = viridis(100, direction = -1)
)

save_plot(p_markers, "UMAP-all-markers.pdf", width = 24, height = 32)
p_markers
```

## Bubble plot: Mean marker expression per cluster
Let's plot the mean marker expressions per cluster in a bubble plot.

```{r vis-data-bubbleplot}
# plot_cluster_bubble() and its colour helpers live in code/bubbleplot.R, shared
# with the PT/BM pipeline. It auto-detects tissue; AD is single-tissue, so it
# draws one count barplot.
source(if (file.exists("code/bubbleplot.R")) "code/bubbleplot.R" else "../code/bubbleplot.R",
       local = TRUE)   # local=TRUE so the helpers see spe/mat
```

```{r phenograph-umap-bubbleplot}
set.seed(221228)

k <- 15

# Reproducible phenograph clustering: Rphenoannoy (Annoy + igraph Louvain) is not
# reproducible across igraph/compiler/CPU, so load a frozen k=15 labelling from
# R_intermediary when present, else compute it here and save it. Stored as pg_clusters.
pg_name <- "pg_clusters_AD.rds"
inter_dirs <- c(Sys.getenv("MAPMET_INTERMEDIARY", ""),
                file.path(params$input,  "R_intermediary"),
                file.path(dirname(params$input), "R_intermediary"),
                file.path(params$output, "R_intermediary"))
inter_dirs <- inter_dirs[nzchar(inter_dirs)]
pg_file <- Find(file.exists, file.path(inter_dirs, pg_name))

if (!is.null(pg_file)) {
  pg <- readRDS(pg_file)
  if (!all(colnames(spe) %in% names(pg)))
    stop("Frozen clustering ", pg_file, " does not cover all cells; ",
         "regenerate it (code/generate_phenograph_intermediary.R).")
  lab <- as.integer(pg[colnames(spe)])
  spe$pg_clusters <- factor(lab, levels = sort(unique(lab)))
  message("Loaded frozen phenograph clustering: ", pg_file)
} else {
  Rphenograph_out <- Rphenoannoy(t(mat), k = k)
  spe$pg_clusters <- factor(membership(Rphenograph_out[[2]]))
  pg_out <- file.path(params$output, "R_intermediary")
  dir.create(pg_out, showWarnings = FALSE, recursive = TRUE)
  saveRDS(setNames(as.character(spe$pg_clusters), colnames(spe)),
          file.path(pg_out, pg_name))
  message("Saved frozen phenograph clustering: ", file.path(pg_out, pg_name))
}

# Cluster colours (helpers from bubbleplot.R)
if (is.null(metadata(spe)$color_vectors)) {
  metadata(spe)$color_vectors <- list()
}
cluster_levels <- get_cluster_levels(spe$pg_clusters)
color_key <- "col_pg_clusters"
if (is.null(metadata(spe)$color_vectors[[color_key]]) ||
    !all(cluster_levels %in% names(metadata(spe)$color_vectors[[color_key]]))) {
  metadata(spe)$color_vectors[[color_key]] <- generate_cluster_colors(
    cluster_levels = cluster_levels
  )
}
col_clusters <- metadata(spe)$color_vectors[[color_key]][cluster_levels]

# UMAP coloured by pg_clusters
p <- dittoDimPlot(
  spe, var = "pg_clusters", reduction.use = "UMAP",
  size = 0.5, do.label = TRUE, legend.show = FALSE,
  color.panel = col_clusters
) +
  theme(legend.title = element_blank()) +
  ggtitle("pg_clusters (k = 15) on UMAP")

label_layer <- which(
  sapply(p$layers, function(x) class(x$geom)[1]) %in% c("GeomLabelRepel", "GeomLabel")
)
label_data <- layer_data(p, label_layer)
label_fills <- col_clusters[as.character(label_data$label)]
p$layers[[label_layer]]$aes_params$fill <- label_fills
p$layers[[label_layer]]$aes_params$colour <- "black"

save_plot(p, "umap_pg_clusters.pdf", width = 8, height = 6)
print(p)

# Bubble plot for pg_clusters
plot_cluster_bubble(
  cluster_col = "pg_clusters",
  save_name = "bubbleplot_pg_clusters.pdf",
  ncells_break_at = 100000
)
```


## Plot clustered heatmaps for each cluster

```{r}
library(dittoSeq)
library(ggplot2)

clusters <- unique(colData(spe)$pg_clusters)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]


# one marker heatmap per cluster (up to 8000 cells each)
for (cl in clusters) {

  spe_cl <- spe[, colData(spe)$pg_clusters == cl]
  if (ncol(spe_cl) == 0) next

  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)

  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )

  save_plot(
    p1,
    paste0("heatmap_cluster_", cl, ".pdf"),
    width = 14,
    height = 6,
    subdir = "heatmap_after_subclustering"
  )
}

```


## Write cluster proportions for manual annotation (Excel)

The chunk below regenerates the cluster-annotation template that is filled in manually and published on Zenodo as a Supplementary Table (read back in the next section). The whole block is disabled with `eval=FALSE` so the pipeline does not depend on `writexl`; enable it only if you need to regenerate the template.

```{r cluster-annotation-excel, eval=FALSE}
library(writexl)

spe$pg_clusters <- as.character(spe$pg_clusters)

# 1. counts per cluster
count_wide <- as.data.frame(table(cluster = spe$pg_clusters)) %>%
  rename(AD = Freq)

# 2. blank annotation columns to fill in manually
annotation_template <- count_wide %>%
  mutate(
    celltype = "",
    metacluster = "",
    comments = ""
  ) %>%
  arrange(as.numeric(cluster))

# 3. write template
write_xlsx(
  annotation_template,
  file.path(params$output, "cluster_annotation_template_AD.xlsx")
)
```


## Annotation and merging of clusters

Next, we read the assigned celltype and metacluster labels for the AD clusters from the Supplementary Table provided on Zenodo [(Supplementary_Tables/20260526_Suppl_Table3_AD.xlsx)](https://doi.org/10.5281/zenodo.13220634). These have been assigned based on thorough examination of images and co-expressions and using literature. The sheet must contain the columns `cluster`, `celltype` and `metacluster`.
Let's start with celltype assignments...

```{r annotations-celltypes}
# Supplementary_Tables may sit in the cohort dir or one level up; check both.
.suppl_cands <- file.path(c(params$input, dirname(params$input)), "Supplementary_Tables", "20260526_Suppl_Table3_AD.xlsx")
file_path <- .suppl_cands[file.exists(.suppl_cands)][1]
if (is.na(file_path))
  stop("AD cluster annotation not found. Looked in:\n  ", paste(.suppl_cands, collapse = "\n  "),
       "\nIt must be a table (sheet 'Sheet1', columns `cluster`, `celltype`, `metacluster`) ",
       "mapping this run's AD clusters to cell types.")
annotation <- read_excel(file_path, sheet = "Sheet1")

cluster_names <- setNames(annotation$celltype, annotation$cluster)
celltypes <- recode(spe$pg_clusters, !!!cluster_names)
spe$celltype <-  as.vector(celltypes)

metacluster_names <- setNames(annotation$metacluster, annotation$cluster)
metacelltypes <- recode(spe$pg_clusters, !!!metacluster_names)
spe$metacluster <-  as.vector(metacelltypes)

```

## Assign colors
Finally, we assign colors to all celltypes and metaclusters. Neural crest lineage celltypes reuse the green tones of the reference tumor palette (except Schwann cell progenitors, colored brown to match the Schwann cells in the main cohort), mesenchymal celltypes use browns, and the `other` celltypes use greyish tones.

```{r colors-celltypes raw_k30}
# Cell-type and metacluster colors are defined centrally in 00_MapMet_color_code.R
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }

col_celltype <- COL_CELLTYPE_AD

metadata(spe)$color_vectors$col_celltype <- col_celltype
```

```{r colors-metaclusters}
col_metacluster <- COL_METACLUSTER_AD
metadata(spe)$color_vectors$col_metacluster <- col_metacluster
```

## Visualizations
For visualization purposes, we give the celltypes a meaningful order, read from the adrenal-gland-specific `celltype_order_AD.csv` (shipped in `R_intermediary`).

```{r order-celltypes}
# celltype_order_AD.csv: search R_intermediary (inter_dirs), fall back to params$input.
order_dirs <- c(inter_dirs, params$input)
order_file <- Find(file.exists, file.path(order_dirs, "celltype_order_AD.csv"))
if (is.null(order_file))
  stop("celltype_order_AD.csv not found in: ", paste(order_dirs, collapse = ", "))
clusternames_ordered <- read.csv(order_file, sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)

# The "NN_" prefix only imposes display order via the alphabetical factor() sort;
# strip it from the levels so plots show clean names in that order.
strip_order_prefix   <- function(x) sub("^[0-9]+_", "", x)
celltype_prefixed    <- recode(spe$celltype, !!!codes)            # e.g. "01_CD44+ Tu"
ordered_levels       <- strip_order_prefix(sort(unique(celltype_prefixed)))
spe$celltype_ordered <- factor(strip_order_prefix(celltype_prefixed), levels = ordered_levels)

# col_celltype is keyed by unprefixed names, so reuse it for the ordered levels.
metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
```

As a last step, we generate UMAP and bubble plots for the final celltypes/metaclusters.

```{r umap-final-celltypes, fig.width=10}
p1 <- dittoDimPlot(spe, var = "celltype_ordered", 
             reduction.use = "UMAP", size = 1, 
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Celltypes on UMAP")

save_plot(p1, "umap-celltypes.pdf", width = 16, height = 8)

p1
```

```{r vis-metacluster}
p2 <- dittoDimPlot(spe, var = "metacluster", 
             reduction.use = "UMAP", size = 1, labels.size=4,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_metacluster) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Metaclusters on UMAP")

save_plot(p2, "umap-metaclusters.pdf", width = 13, height = 8)

p2
```


```{r bubbleplot-celltypes, fig.width=16, fig.height=14}

plot_cluster_bubble(
  cluster_col = "celltype",
  save_name = "bubbleplot-celltypes.pdf",
  annotation_level = "metacluster",
  ncells_break_at = 60000,
)

```


Let's plot heatmap of clustered cells as a validation:


```{r}
library(dittoSeq)
library(ggplot2)
library(viridis)

clusters <- unique(colData(spe)$celltype)

genes_use <- rownames(spe)[
  !grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3",
         rownames(spe))
]

for (cl in clusters) {
  
  # make the cluster name safe for a filename
  safe_name <- cl
  safe_name <- gsub("\\+", "plus", safe_name)
  safe_name <- gsub("-", "minus", safe_name)
  safe_name <- gsub("[^A-Za-z0-9_]", "_", safe_name)
  
  spe_cl <- spe[, colData(spe)$celltype == cl]
  
  if (ncol(spe_cl) == 0) next
  
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  filename <- paste0("heatmap_celltype_", safe_name, ".png")
  
  save_plot(p1, filename = filename, width = 14, height = 6, subdir = "heatmap_celltype")
}
```
## Save object

```{r save-objects}

saveRDS(spe, file.path(params$output, "spe_04_phenotyping_AD.rds"))

```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
