---
title: "Supplementary: AD region and marker images"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input:  !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
  # Image(s) to render. Give a vector of ids in `image_ids` to render several in
  # one pass; a single `image_id` is still accepted (back-compat). Override from
  # run_analysis.R (see the render_step call). If `image_ids` is set it wins.
  image_id:  "20250623_19700101_25-XDD397_AD_001"
  image_ids: !r NULL
  # Markers rendered individually and as a colour merge. Either full channel
  # names (as in rownames(spe)) or the short marker name (token before the first
  # "_"). Here: DNA2 = DNA1_Ir193_mean (Ir193), VIM = Vimentin, KI67 = Ki-67.
  markers: !r c("DNA1_Ir193_mean", "ELAVL4_Yb174_mean", "Vimentin_Pt196_mean", "Ki-67_Tm169_mean", "CXCR4_Tb159_mean", "CHGA_Dy164_mean", "GD2_Gd155_mean")
  # Colour per marker (individual image + merge), keyed by the short marker name
  # (token before the first "_"; "DNA1" keys the DNA channel). Markers without an
  # entry fall back to a spare-colour palette.
  marker_colours: !r c("DNA1" = "blue", "ELAVL4" = "green", "Vimentin" = "red", "Ki-67" = "darkorange", "CXCR4" = "cyan", "CHGA" = "yellow", "GD2" = "purple")
  # Short marker names drawn individually but left out of the colour merge (each
  # still gets its own panel). Ki-67 sits out to keep the merge within
  # cytomapper's 6-marker limit.
  merge_exclude: !r c("Ki-67")
  # Zoom window(s) c(xmin, xmax, ymin, ymax) in pixel coordinates of the image
  # array (x = first image dimension, y = second). Both the whole image AND any
  # zoom are rendered. `crops` is a list with one entry per `image_ids` id
  # (NULL = whole image only, the default for every image); `crop` is a single
  # window for the back-compat single-image path. NULL renders whole image only.
  crop:  !r NULL
  crops: !r NULL
---

```{r setup, include=FALSE}
# Output directory for plots
plot_dir <- file.path(params$output, "plots_AD/05_generate_images")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot to plot_dir (dispatches on plot type)
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList"))) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

knitr::opts_chunk$set(
  echo = TRUE, error = FALSE, message = FALSE, warning = FALSE,
  dev = "png", dpi = 150, fig.width = 8, fig.height = 8
)
```

# AD region and marker images

Adrenal-gland (AD) counterpart of `analysis/08_generate_images.Rmd`. For each
requested AD image - as the **whole image** and, when a zoom window is given, a
**zoomed region** - it renders:

* each requested **marker** as a separate pixel image,
* a colour **merge** of all requested markers, and
* the **cell types**, drawn on the segmentation **masks** (`plotCells`).

The AD cohort has no tumor-compartment or cellular-neighborhood annotation, so
those panels (present in the main-cohort script) are omitted here. Rendered from
`run_analysis.R` with a *customised* parameter list (image ids, zoom crops, marker
list). Cell types and colours come from `spe_04_phenotyping_AD.rds`, so the
chosen images must be among the AD images processed in
[`analysis_AD/04_phenotyping`](04_phenotyping.html).

## Load libraries
```{r load-libraries}
library(SpatialExperiment)
library(cytomapper)
library(cowplot)
library(RColorBrewer)
library(ggplot2)
```

## Load data and images
```{r load-data}
spe <- readRDS(file.path(params$output, "spe_04_phenotyping_AD.rds"))

# Images to render + optional crops; vector image_ids/crops win over single image_id/crop.
image_ids <- if (!is.null(params$image_ids)) params$image_ids else params$image_id
if (is.null(image_ids) || !length(image_ids))
  stop("No image id provided: set params$image_ids (or params$image_id).")
crops <- if (!is.null(params$crops)) params$crops else
         if (!is.null(params$crop)) rep(list(params$crop), length(image_ids)) else
         vector("list", length(image_ids))          # all NULL -> whole image only
if (length(crops) != length(image_ids))
  stop("params$crops must have one entry per image id (NULL = whole image only). ",
       "Got ", length(crops), " crop(s) for ", length(image_ids), " id(s).")

missing_ids <- setdiff(image_ids, unique(as.character(spe$sample_id)))
if (length(missing_ids))
  stop("image id(s) not among the AD images in spe_04_phenotyping_AD.rds: ",
       paste(missing_ids, collapse = ", "), ". Available images include: ",
       paste(head(unique(as.character(spe$sample_id))), collapse = ", "), " ...")

# Channel (_mean) marker names; requests resolve against these.
marker_names <- rownames(spe)[grepl("_mean$", rownames(spe))]

# Accept either a full channel name or a short marker name (token before "_").
resolve_markers <- function(requested, channels) {
  token <- sub("_.*$", "", channels)
  vapply(requested, function(r) {
    if (r %in% channels) return(r)
    hit <- channels[token == r]
    if (length(hit)) hit[1] else NA_character_
  }, character(1))
}

markers_full <- resolve_markers(params$markers, marker_names)
if (anyNA(markers_full))
  warning("Markers not found among the spe channels (dropped): ",
          paste(params$markers[is.na(markers_full)], collapse = ", "))
markers <- unname(markers_full[!is.na(markers_full)])
if (!length(markers)) stop("None of the requested markers match the spe channels.")
message("Rendering markers: ", paste(markers, collapse = ", "))

# Load one image + its mask + matching cells (tmp); channels named from marker_names.
load_image <- function(image_id) {
  tmp   <- spe[, spe$sample_id == image_id]
  img   <- loadImages(file.path(params$input, "img",   paste0(image_id, ".tif")))
  masks <- loadImages(file.path(params$input, "masks", paste0(image_id, ".tif")), as.is = TRUE)
  mcols(img)   <- DataFrame(sample_id = names(img))
  mcols(masks) <- DataFrame(sample_id = names(masks))
  message("Loaded '", image_id, "': image ", dim(img[[1]])[1], " x ", dim(img[[1]])[2],
          " px (", dim(img[[1]])[3], " channels); mask ",
          dim(masks[[1]])[1], " x ", dim(masks[[1]])[2], " px.")
  n_ch <- dim(img[[1]])[3]
  channelNames(img) <- if (n_ch <= length(marker_names)) {
    marker_names[seq_len(n_ch)]
  } else {
    c(marker_names, paste0("extra_channel_", seq_len(n_ch - length(marker_names))))
  }
  list(img = img, masks = masks, tmp = tmp)
}
```

## Colours and helpers
```{r helpers}
# Clamp/shift a requested [lo, hi] window to fit inside [1, n].
.clamp_axis <- function(lo, hi, n) {
  lo <- as.integer(round(lo)); hi <- as.integer(round(hi))
  if (lo > hi) { tmp <- lo; lo <- hi; hi <- tmp }
  w <- hi - lo
  if (w >= n - 1L) return(c(1L, n))
  lo <- max(1L, min(lo, n - w))
  c(lo, lo + w)
}

# Crop a CytoImageList to c(xmin, xmax, ymin, ymax). Rebuild each frame as an
# EBImage Image - plain `[` subsetting can drop the class and render black.
crop_cil <- function(cil, crop) {
  d0 <- dim(cil[[1]])
  xr <- .clamp_axis(crop[1], crop[2], d0[1])
  yr <- .clamp_axis(crop[3], crop[4], d0[2])
  if (!identical(c(xr, yr), as.integer(c(crop[1], crop[2], crop[3], crop[4]))))
    message("Crop x[", crop[1], "-", crop[2], "] y[", crop[3], "-", crop[4],
            "] adjusted to x[", xr[1], "-", xr[2], "] y[", yr[1], "-", yr[2],
            "] to fit the ", d0[1], " x ", d0[2], " px image.")
  xs <- seq(xr[1], xr[2]); ys <- seq(yr[1], yr[2])
  imgs <- lapply(seq_along(cil), function(i) {
    im  <- cil[[i]]
    dat <- EBImage::imageData(im)
    sub <- if (length(dim(im)) == 3L) dat[xs, ys, , drop = FALSE]
           else                       dat[xs, ys,    drop = FALSE]
    EBImage::Image(sub, colormode = EBImage::colorMode(im))
  })
  out <- CytoImageList(imgs)
  names(out) <- names(cil)
  mcols(out) <- mcols(cil)
  if (!is.null(channelNames(cil))) channelNames(out) <- channelNames(cil)
  out
}

# Rotate a CytoImageList 90 degrees counter-clockwise (transpose + flip, lossless
# for integer mask labels) to match the main-cohort plotSpatial orientation.
rotate90_ccw <- function(cil) {
  out <- CytoImageList(lapply(seq_along(cil), function(i)
    EBImage::flip(EBImage::transpose(cil[[i]]))))
  names(out) <- names(cil)
  mcols(out) <- mcols(cil)
  if (!is.null(channelNames(cil))) channelNames(out) <- channelNames(cil)
  out
}

# Scale bar (IMC pixel = 1 um), drawn only on the merge panel.
scale_bar_spec <- list(length = 100, label = "100 µm", cex = 1.5)

# Evaluate a plot call, then save + show it (TRUE), or report and skip on error
# (FALSE) so one bad panel - e.g. a flat channel - never aborts the render.
# plot_call is a promise, so plotting happens inside the tryCatch.
render_and_save <- function(plot_call, filename, width, height) {
  po <- tryCatch(plot_call,
                 error = function(e) {
                   message("Skipped ", filename, ": ", conditionMessage(e)); NULL
                 })
  if (is.null(po)) return(invisible(FALSE))
  p <- ggdraw(po$plot)
  save_plot(p, filename, width = width, height = height)
  print(p)
  invisible(TRUE)
}

# Per-marker colour (individual + merge), keyed by short name; unmapped get spare hues.
.marker_col_map <- local({
  fallback <- c("white", "orange", "yellow", "magenta", "pink", "brown")
  tok  <- sub("_.*$", "", markers)
  cols <- unname(params$marker_colours[tok])   # NA where the short name is unmapped
  miss <- which(is.na(cols))
  if (length(miss))
    cols[miss] <- fallback[((seq_along(miss) - 1) %% length(fallback)) + 1]
  setNames(cols, markers)
})

# Markers to draw individually but leave out of the merge (via params$merge_exclude).
.merge_exclude <- markers[sub("_.*$", "", markers) %in% params$merge_exclude]

# Render every panel (individual markers, merge, celltype) for one view; tag goes
# into the filename. No legends.
plot_region <- function(img_v, masks_v, tmp, markers, tag) {

  # --- individual markers, each in its assigned colour; keep the ones that render ---
  usable <- character(0)
  for (m in markers) {
    ok <- render_and_save(
      plotPixels(img_v, colour_by = m,
                 colour = setNames(list(c("black", .marker_col_map[[m]])), m),
                 bcg = setNames(list(c(0, 1, 1)), m),
                 image_title = NULL, scale_bar = NULL,
                 legend = NULL, return_plot = TRUE),
      paste0(tag, "_marker_", m, ".png"), 8, 8)
    if (isTRUE(ok)) usable <- c(usable, m)
  }

  # --- merge of the rendered markers, minus the individual-only ones; scale bar
  #     here only. plotPixels shows at most 6 markers, so cap the merge at 6. ---
  merge_markers <- setdiff(usable, .merge_exclude)
  if (length(merge_markers) > 6) {
    message("Merge capped at 6 markers (cytomapper::plotPixels max); leaving out: ",
            paste(sub("_.*$", "", merge_markers[-seq_len(6)]), collapse = ", "))
    merge_markers <- merge_markers[seq_len(6)]
  }
  if (length(merge_markers)) {
    colour_list <- setNames(lapply(merge_markers, function(mm) c("black", .marker_col_map[[mm]])), merge_markers)
    bcg_list    <- setNames(lapply(merge_markers, function(x) c(0, 1, 1)), merge_markers)
    render_and_save(
      plotPixels(img_v, colour_by = merge_markers, colour = colour_list,
                 bcg = bcg_list, image_title = NULL, scale_bar = scale_bar_spec,
                 legend = NULL, return_plot = TRUE),
      paste0(tag, "_markers_merge.png"), 8, 8)
  }

  # --- cell types (on the masks) ---
  render_and_save(
    plotCells(masks_v, object = tmp, cell_id = "ObjectNumber", img_id = "sample_id",
              colour_by = "celltype",
              colour = list(celltype = metadata(spe)$color_vectors$col_celltype),
              missing_colour = "white", image_title = NULL, scale_bar = NULL,
              legend = NULL, return_plot = TRUE),
    paste0(tag, "_celltype.png"), 10, 8)
}
```

## Whole image and zoom
Every image in `image_ids` is rendered as a whole image; when that image has a
zoom window in `crops`, the zoomed region is rendered too. Every output filename
carries the image id and `_whole` / `_zoom`.

```{r generate}
for (i in seq_along(image_ids)) {
  image_id <- image_ids[[i]]
  crop     <- crops[[i]]
  message("=== image: ", image_id, " (", i, "/", length(image_ids), ") ===")

  loaded <- load_image(image_id)
  img <- loaded$img; masks <- loaded$masks; tmp <- loaded$tmp

  # Always the whole image; add the zoom when this image has crop coordinates.
  views <- list(whole = NULL)
  if (!is.null(crop)) views[["zoom"]] <- crop

  for (vn in names(views)) {
    cr      <- views[[vn]]
    # Crop in native orientation (crop coords stay in raw-image space), then rotate.
    img_v   <- if (is.null(cr)) img   else crop_cil(img,   cr)
    masks_v <- if (is.null(cr)) masks else crop_cil(masks, cr)
    img_v   <- rotate90_ccw(img_v)
    masks_v <- rotate90_ccw(masks_v)
    message("=== view: ", vn, " (", paste(dim(img_v[[1]])[1:2], collapse = " x "), " px) ===")
    plot_region(img_v, masks_v, tmp, markers, paste0(image_id, "_", vn))
  }
}
```
