---
title: "Supplementary: AD region and marker images"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input:  !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
  # Image to render. Override from run_analysis.R (see the render_step call).
  image_id: "20250623_19700101_25-XDD397_AD_001"
  # Markers rendered individually and as a colour merge. Either full channel
  # names (as in rownames(spe)) or the short marker name (token before the first
  # "_"). Here: DNA2 = DNA1_Ir193_mean (Ir193), VIM = Vimentin, KI67 = Ki-67.
  markers: !r c("DNA1_Ir193_mean", "ELAVL4_Yb174_mean", "Vimentin_Pt196_mean", "Ki-67_Tm169_mean", "CXCR4_Tb159_mean", "CHGA_Dy164_mean", "GD2_Gd155_mean")
  # Colour per marker (individual image + merge), keyed by the short marker name
  # (token before the first "_"; "DNA1" keys the DNA channel). Markers without an
  # entry fall back to a spare-colour palette.
  marker_colours: !r c("DNA1" = "blue", "ELAVL4" = "green", "Vimentin" = "red", "Ki-67" = "darkorange", "CXCR4" = "cyan", "CHGA" = "yellow", "GD2" = "purple")
  # Short marker names to draw individually but leave OUT of the colour merge
  # (each still gets its own panel). Only Ki-67 sits out (per request); with 7
  # markers this keeps the merge at cytomapper's 6-marker maximum. GD2 is purple
  # (not orange) so that where it overlaps ELAVL4 the additive blend is pale
  # rather than the green -> yellow an orange channel produced.
  merge_exclude: !r c("Ki-67")
  # Zoom window c(xmin, xmax, ymin, ymax) in pixel coordinates of the image
  # array (x = first image dimension, y = second). Both the whole image AND this
  # zoom are rendered. Set to NULL to render the whole image only.
  crop: !r c(2000, 3000, 2000, 3000)
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots_AD/05_generate_images")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList"))) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

knitr::opts_chunk$set(
  echo = TRUE, error = FALSE, message = FALSE, warning = FALSE,
  dev = "png", dpi = 150, fig.width = 8, fig.height = 8
)
```

# AD region and marker images

Adrenal-gland (AD) counterpart of `analysis/08_generate_images.Rmd`. For a single
AD image - as **both the whole image and a zoomed region** - it renders:

* each requested **marker** as a separate pixel image,
* a colour **merge** of all requested markers, and
* the **cell types**, drawn on the segmentation **masks** (`plotCells`).

The AD cohort has no tumor-compartment or cellular-neighborhood annotation, so
those panels (present in the main-cohort script) are omitted here. Rendered from
`run_analysis.R` with a *customised* parameter list (image id, zoom crop, marker
list). Cell types and colours come from `spe_04_phenotyping_AD.rds`, so the
chosen image must be one of the AD images processed in
[`analysis_AD/04_phenotyping`](04_phenotyping.html).

## Load libraries
```{r load-libraries}
library(SpatialExperiment)
library(cytomapper)
library(cowplot)
library(RColorBrewer)
library(ggplot2)
```

## Load data and images
```{r load-data}
spe <- readRDS(file.path(params$output, "spe_04_phenotyping_AD.rds"))

image_id <- params$image_id
if (!image_id %in% spe$sample_id)
  stop("image_id '", image_id, "' is not among the AD images in ",
       "spe_04_phenotyping_AD.rds. Available images include: ",
       paste(head(unique(as.character(spe$sample_id))), collapse = ", "), " ...")

tmp <- spe[, spe$sample_id == image_id]

# Multichannel image (pixels) and segmentation mask (cells)
img   <- loadImages(file.path(params$input, "img",   paste0(image_id, ".tif")))
masks <- loadImages(file.path(params$input, "masks", paste0(image_id, ".tif")), as.is = TRUE)

mcols(img)   <- DataFrame(sample_id = names(img))
mcols(masks) <- DataFrame(sample_id = names(masks))

# Report the actual pixel dimensions so the crop window (params$crop) can be set
# within range - an out-of-range zoom is the usual cause of a black zoom panel.
message("Loaded '", image_id, "': image ", dim(img[[1]])[1], " x ", dim(img[[1]])[2],
        " px (", dim(img[[1]])[3], " channels); mask ",
        dim(masks[[1]])[1], " x ", dim(masks[[1]])[2], " px.")

# Name the image channels from the marker (_mean) names, using the image's
# actual channel count (same robust rule as analysis_AD/03_QC_2.Rmd, in case the
# raw image still carries channels dropped from the intensity tables).
marker_names <- rownames(spe)[grepl("_mean$", rownames(spe))]
n_ch <- dim(img[[1]])[3]
channelNames(img) <- if (n_ch <= length(marker_names)) {
  marker_names[seq_len(n_ch)]
} else {
  c(marker_names, paste0("extra_channel_", seq_len(n_ch - length(marker_names))))
}

# Resolve requested markers: accept either a full channel name or a short marker
# name (the token before the first "_").
resolve_markers <- function(requested, channels) {
  token <- sub("_.*$", "", channels)
  vapply(requested, function(r) {
    if (r %in% channels) return(r)
    hit <- channels[token == r]
    if (length(hit)) hit[1] else NA_character_
  }, character(1))
}

markers_full <- resolve_markers(params$markers, channelNames(img))
if (anyNA(markers_full))
  warning("Markers not found as image channels (dropped): ",
          paste(params$markers[is.na(markers_full)], collapse = ", "))
markers <- unname(markers_full[!is.na(markers_full)])
if (!length(markers)) stop("None of the requested markers match the image channels.")
message("Rendering markers: ", paste(markers, collapse = ", "))
```

## Colours and helpers
```{r helpers}
# Clamp/shift a requested [lo, hi] window so it fits inside [1, n]. If the window
# is at least as wide as the axis it falls back to the whole axis; otherwise it
# is shifted inward to stay in range. Returns integer c(lo, hi).
.clamp_axis <- function(lo, hi, n) {
  lo <- as.integer(round(lo)); hi <- as.integer(round(hi))
  if (lo > hi) { tmp <- lo; lo <- hi; hi <- tmp }
  w <- hi - lo
  if (w >= n - 1L) return(c(1L, n))
  lo <- max(1L, min(lo, n - w))
  c(lo, lo + w)
}

# Crop a CytoImageList to c(xmin, xmax, ymin, ymax). The window is clamped to the
# image bounds (with a message), and every cropped frame is rebuilt as a proper
# EBImage Image - plain `[` subsetting can drop the Image class and yield a blank
# CytoImageList that plotPixels/plotCells then render black.
crop_cil <- function(cil, crop) {
  d0 <- dim(cil[[1]])
  xr <- .clamp_axis(crop[1], crop[2], d0[1])
  yr <- .clamp_axis(crop[3], crop[4], d0[2])
  if (!identical(c(xr, yr), as.integer(c(crop[1], crop[2], crop[3], crop[4]))))
    message("Crop x[", crop[1], "-", crop[2], "] y[", crop[3], "-", crop[4],
            "] adjusted to x[", xr[1], "-", xr[2], "] y[", yr[1], "-", yr[2],
            "] to fit the ", d0[1], " x ", d0[2], " px image.")
  xs <- seq(xr[1], xr[2]); ys <- seq(yr[1], yr[2])
  imgs <- lapply(seq_along(cil), function(i) {
    im  <- cil[[i]]
    dat <- EBImage::imageData(im)
    sub <- if (length(dim(im)) == 3L) dat[xs, ys, , drop = FALSE]
           else                       dat[xs, ys,    drop = FALSE]
    EBImage::Image(sub, colormode = EBImage::colorMode(im))
  })
  out <- CytoImageList(imgs)
  names(out) <- names(cil)
  mcols(out) <- mcols(cil)
  if (!is.null(channelNames(cil))) channelNames(out) <- channelNames(cil)
  out
}

# Rotate a CytoImageList 90 degrees counter-clockwise (lossless: transpose then
# vertical flip, so integer mask labels are preserved), matching the plotSpatial
# orientation used for the main-cohort images (same imaging platform).
rotate90_ccw <- function(cil) {
  out <- CytoImageList(lapply(seq_along(cil), function(i)
    EBImage::flip(EBImage::transpose(cil[[i]]))))
  names(out) <- names(cil)
  mcols(out) <- mcols(cil)
  if (!is.null(channelNames(cil))) channelNames(out) <- channelNames(cil)
  out
}

# Scale bar (IMC pixel size is 1 um, so a length in px is a length in um). It is
# drawn ONLY on the merge panel; cex sizes the "100 um" label - small enough to
# stay inside the frame (the default was cropped), large enough to read. The
# image name / title is removed from every panel (image_title = NULL).
scale_bar_spec <- list(length = 100, label = "100 µm", cex = 1.5)

# Evaluate a plotPixels/plotCells call lazily; on success save + show it and
# return TRUE, on any error report it, skip, and return FALSE - so one bad panel
# (e.g. a flat channel with no signal in this view) never aborts the whole
# render. `plot_call` is a promise, so plotting runs inside the tryCatch below.
render_and_save <- function(plot_call, filename, width, height) {
  po <- tryCatch(plot_call,
                 error = function(e) {
                   message("Skipped ", filename, ": ", conditionMessage(e)); NULL
                 })
  if (is.null(po)) return(invisible(FALSE))
  p <- ggdraw(po$plot)
  save_plot(p, filename, width = width, height = height)
  print(p)
  invisible(TRUE)
}

# Colour for each marker (used for BOTH its individual image and the merge),
# keyed by the short marker name. Markers without an explicit colour fall back to
# spare hues so they stay distinct in a merge.
.marker_col_map <- local({
  fallback <- c("white", "orange", "yellow", "magenta", "pink", "brown")
  tok  <- sub("_.*$", "", markers)
  cols <- unname(params$marker_colours[tok])   # NA where the short name is unmapped
  miss <- which(is.na(cols))
  if (length(miss))
    cols[miss] <- fallback[((seq_along(miss) - 1) %% length(fallback)) + 1]
  setNames(cols, markers)
})

# Marker channels to draw individually but leave OUT of the colour merge,
# matched by short marker name via params$merge_exclude.
.merge_exclude <- markers[sub("_.*$", "", markers) %in% params$merge_exclude]

# Render every panel (individual markers, merge, celltype) for one view (whole
# image or a crop). tag goes into the filename. No legends are drawn - each panel
# is just the image.
plot_region <- function(img_v, masks_v, tmp, markers, tag) {

  # --- individual markers, each in its assigned merge colour; keep the ones
  #     that render (a marker absent from this view is a flat channel, skipped) ---
  usable <- character(0)
  for (m in markers) {
    ok <- render_and_save(
      plotPixels(img_v, colour_by = m,
                 colour = setNames(list(c("black", .marker_col_map[[m]])), m),
                 bcg = setNames(list(c(0, 1, 1)), m),
                 image_title = NULL, scale_bar = NULL,
                 legend = NULL, return_plot = TRUE),
      paste0(tag, "_marker_", m, ".png"), 8, 8)
    if (isTRUE(ok)) usable <- c(usable, m)
  }

  # --- merge of the markers that rendered, minus any flagged individual-only;
  #     same per-marker colours. The scale bar is drawn on this panel only.
  #     cytomapper::plotPixels composites at most 6 markers, so cap the merge at
  #     6 (dropping the extras, which still have their own individual panels). ---
  merge_markers <- setdiff(usable, .merge_exclude)
  if (length(merge_markers) > 6) {
    message("Merge capped at 6 markers (cytomapper::plotPixels max); leaving out: ",
            paste(sub("_.*$", "", merge_markers[-seq_len(6)]), collapse = ", "))
    merge_markers <- merge_markers[seq_len(6)]
  }
  if (length(merge_markers)) {
    colour_list <- setNames(lapply(merge_markers, function(mm) c("black", .marker_col_map[[mm]])), merge_markers)
    bcg_list    <- setNames(lapply(merge_markers, function(x) c(0, 1, 1)), merge_markers)
    render_and_save(
      plotPixels(img_v, colour_by = merge_markers, colour = colour_list,
                 bcg = bcg_list, image_title = NULL, scale_bar = scale_bar_spec,
                 legend = NULL, return_plot = TRUE),
      paste0(tag, "_markers_merge.png"), 8, 8)
  }

  # --- cell types (on the masks) ---
  render_and_save(
    plotCells(masks_v, object = tmp, cell_id = "ObjectNumber", img_id = "sample_id",
              colour_by = "celltype",
              colour = list(celltype = metadata(spe)$color_vectors$col_celltype),
              missing_colour = "white", image_title = NULL, scale_bar = NULL,
              legend = NULL, return_plot = TRUE),
    paste0(tag, "_celltype.png"), 10, 8)
}
```

## Whole image and zoom
Both the whole image and (when `params$crop` is set) the zoomed region are
rendered; every output filename carries the image id and `_whole` / `_zoom`.

```{r generate}
# Always the whole image; add the zoom when crop coordinates are provided.
views <- list(whole = NULL)
if (!is.null(params$crop)) views[["zoom"]] <- params$crop

for (vn in names(views)) {
  cr      <- views[[vn]]
  # Crop in the native image orientation (so crop coordinates stay in raw-image
  # pixel space), then rotate the view to the plotSpatial orientation.
  img_v   <- if (is.null(cr)) img   else crop_cil(img,   cr)
  masks_v <- if (is.null(cr)) masks else crop_cil(masks, cr)
  img_v   <- rotate90_ccw(img_v)
  masks_v <- rotate90_ccw(masks_v)
  message("=== view: ", vn, " (", paste(dim(img_v[[1]])[1:2], collapse = " x "), " px) ===")
  plot_region(img_v, masks_v, tmp, markers, paste0(image_id, "_", vn))
}
```
