---
title: "Patient cohort overview"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/01.1_cohort_overview")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {

    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)

  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {

    png(full_path, width = width * dpi, height = height * dpi, res = dpi)
    draw(plot_obj)
    dev.off()

  } else {

    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }

  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 10,
  fig.height = 5
)
```

# Patient cohort overview {#cohort-overview}

We summarise the diagnostic primary tumor (PT, DE) cohort in a single overview
figure: patient age at diagnosis as a bar plot, together with sex and the key
genetic / clinical tracks (MYCN amplification, 1p loss and progression) drawn as
colored tiles — one column per patient. The tile tracks reuse the same
`geom_tile` technique as the metadata strips in
[chapter 6](#pt-dx); age is shown as a bar plot instead of a tile.

The table is built directly from the sample- and patient-level metadata (the
same Supplementary Table 1 sheets read in [chapter 1](#read-data)), so this
figure does not depend on any processed single-cell object.

## Load libraries
<details>
   <summary>Load libraries</summary>

```{r load-libraries}
library(readxl)
library(dplyr)
library(ggplot2)
library(patchwork)
library(grid)  # for unit()
```
</details>

## Read data

```{r read-data}
# Read Supplementary Table 1 (the same sheets used in 01_read_data) and merge the
# sample-level metadata with the patient-level clinical and genetic tables.
file_path <- file.path(params$input, "Supplementary_Tables/20260526_Suppl_Table1.xlsx")

sample_spec_metadata  <- read_excel(file_path, sheet = "sample_spec_metadata")
patient_spec_clinical <- read_excel(file_path, sheet = "patient_spec_clinical")
patient_spec_genetic  <- read_excel(file_path, sheet = "patient_spec_genetic")

meta <- sample_spec_metadata %>%
  left_join(patient_spec_clinical, by = "Patient_ID") %>%
  left_join(patient_spec_genetic,  by = "Patient_ID")
```

## Build the per-patient cohort table

We keep one row per patient with a diagnostic primary tumor sample
(`Timepoint == "DE"`, primary-tumor tissue), mirroring the cohort defined in
[chapter 1](#read-data) (Patient_018 is excluded there as a lymph-node, not a PT,
sample). Age and sex come from the clinical table (`age_at_diag`, `sex_aab`); MYCN
amplification, 1p loss and progression from the genetic / clinical tables
(`MYCN_amp`, `X1p_loss`, `efs`).

```{r cohort-table}
# Diagnostic primary tumor cohort: one row per patient. Tissue is coded "PT" in
# the metadata ("TU" is also accepted for robustness); Patient_018 is dropped to
# match 01_read_data (lymph-node, not a PT sample).
cohort <- meta %>%
  filter(Timepoint == "DE",
         Tissue %in% c("PT", "TU"),
         Patient_ID != "Patient_018") %>%
  distinct(Patient_ID, .keep_all = TRUE) %>%  # collapse multiple ROIs -> one row / patient
  transmute(
    patient_id = as.character(Patient_ID),
    age        = as.numeric(age_at_diag),  # age at diagnosis (months)
    sex        = as.character(sex_aab),    # sex (AAB)
    mycn       = as.character(MYCN_amp),
    loss1p     = as.character(X1p_loss),
    prog       = as.character(efs)         # event-free survival event: 0 = no, 1 = yes
  ) %>%
  arrange(patient_id)

# ---- Recode tracks ----
# MYCN amp. and 1p loss share a yes / no / het / unknown coding. We recode both
# to a common yes / no / het scheme (unknown -> NA -> white tile) so that the
# three genetic / clinical tracks can share a single legend, as in the reference
# figure. The mappings cover both the raw numeric codes in the metadata and the
# character labels used elsewhere in the analysis.
recode_ynh <- function(x) dplyr::recode(as.character(x),
  "1" = "yes", "0" = "no", "2" = "het", "9" = NA_character_,
  "MNA" = "yes", "nMNA" = "no", "het" = "het", "unknown" = NA_character_,
  "yes" = "yes", "no" = "no",
  .default = NA_character_)

cohort$mycn   <- recode_ynh(cohort$mycn)
cohort$loss1p <- recode_ynh(cohort$loss1p)

# progression is coded 0 (no) / 1 (yes)
cohort$prog <- dplyr::recode(as.character(cohort$prog),
  "1" = "yes", "0" = "no", "9" = NA_character_,
  "Prog" = "yes", "NoProg" = "no", "yes" = "yes", "no" = "no",
  .default = NA_character_)

# sex -> f / m (handle the most common encodings, otherwise keep as-is)
cohort$sex <- dplyr::recode(as.character(cohort$sex),
  "1" = "f", "2" = "m", "female" = "f", "male" = "m", "F" = "f", "M" = "m",
  .default = as.character(cohort$sex))

# Fix column (patient) order and factor levels
lvl_ynh <- c("yes", "no", "het")
cohort$patient_id <- factor(cohort$patient_id, levels = cohort$patient_id)
cohort$mycn   <- factor(cohort$mycn,   levels = lvl_ynh)
cohort$loss1p <- factor(cohort$loss1p, levels = lvl_ynh)
cohort$prog   <- factor(cohort$prog,   levels = lvl_ynh)
cohort$sex    <- factor(cohort$sex,    levels = c("f", "m"))
```

## Assemble the figure

The panels share the same discrete x-axis (one column per patient) and are
stacked with `patchwork`. Legends are collected into a single vertical column on
the right (displayed normally, not horizontally). The three genetic / clinical
tracks use an identical fill scale, so `guides = "collect"` merges them into one
`MNA / 1p loss / prog.` legend; sex keeps its own `f / m` legend.

```{r cohort-overview-figure, fig.width=10, fig.height=5}
# ---- Colors (shared with the rest of the analysis) ----
col_ynh <- c("yes" = rgb(202,   0,  32, maxColorValue = 255),  # red
             "no"  = rgb(  5, 113, 176, maxColorValue = 255),  # blue
             "het" = rgb(146, 197, 222, maxColorValue = 255))  # light blue
col_sex <- c("f"   = rgb(202,   0,  32, maxColorValue = 255),  # red
             "m"   = rgb(  5, 113, 176, maxColorValue = 255))  # blue

# Shared discrete x-axis expansion so every panel aligns column-by-column.
X_EXPAND <- expansion(add = 0.5)

# ---- Age bar plot (top panel) ----
max_age <- max(cohort$age, na.rm = TRUE)
p_age <- ggplot(cohort, aes(x = patient_id, y = age)) +
  geom_col(fill = "grey35", width = 0.8) +
  scale_x_discrete(expand = X_EXPAND) +
  scale_y_continuous(breaks = c(0, max_age),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(y = "age[months]") +
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size = 8),
    axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 1, size = 9),
    panel.grid   = element_blank(),
    plot.margin  = margin(2, 2, 0, 2)
  )

# ---- Tile track helper ----
# breaks controls which keys appear in the legend (NA / unknown is drawn white
# via na.value but kept out of the legend); drop = FALSE keeps every level so the
# three genetic / clinical tracks share an identical, mergeable legend.
tile_track <- function(var, fill_vals, breaks, legend_name, y_lab) {
  ggplot(cohort, aes(x = patient_id, y = 1, fill = .data[[var]])) +
    geom_tile(color = "white", linewidth = 0.4) +
    scale_x_discrete(expand = X_EXPAND) +
    scale_fill_manual(values = fill_vals, breaks = breaks, name = legend_name,
                      na.value = "white", drop = FALSE) +
    labs(y = y_lab) +
    theme_void() +
    theme(
      legend.position  = "right",
      legend.direction = "vertical",
      legend.key.size  = unit(0.4, "cm"),
      legend.text      = element_text(size = 9),
      legend.title     = element_text(size = 10, face = "bold"),
      axis.title.y     = element_text(angle = 0, vjust = 0.5, hjust = 1, size = 9),
      plot.margin      = margin(0, 2, 0, 2)
    )
}

p_sex  <- tile_track("sex",    col_sex, c("f", "m"), "sex (AAB)",             "sex (AAB)")
p_mycn <- tile_track("mycn",   col_ynh, lvl_ynh,     "MNA / 1p loss / prog.", "MYCN amp.")
p_1p   <- tile_track("loss1p", col_ynh, lvl_ynh,     "MNA / 1p loss / prog.", "1p loss")
p_prog <- tile_track("prog",   col_ynh, lvl_ynh,     "MNA / 1p loss / prog.", "progression")

# ---- Stack panels; collect legends into a single vertical column on the right ----
p_cohort <- (p_age / p_sex / p_mycn / p_1p / p_prog) +
  plot_layout(heights = c(2, 0.5, 0.5, 0.5, 0.5), guides = "collect") +
  plot_annotation(
    title = "a) Patient cohort overview",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  ) &
  theme(legend.direction = "vertical", legend.box = "vertical",
        legend.justification = "top")

p_cohort

save_plot(p_cohort, "Fig1a_patient_cohort_overview.png", width = 10, height = 5)
```

## Session Info

<details>
   <summary>SessionInfo</summary>

```{r, echo = FALSE}
sessionInfo()
```
</details>
