---
title: "Read single-cell and meta-data"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/data/input_all"
  output: "/mnt/data/output"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/01_read_data")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Read in the data {#read-data}

We  will first read in single-cell data, previously extracted from processed multi-modal (IMC,IF) images in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP) or downloaded from [Zenodo](https://zenodo.org/records/13220635/files/MapMetIP_ProcessedDataset.zip). For reading the data, we use the [imcRtools](https://github.com/BodenmillerGroup/imcRtools) package. This package uses the `read_steinbock` function to read data into a `SpatialExperiment` (SPE). The [SpatialExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) class [@Righelli2021] is an extension of the `SingleCellExperiment` class. 
For more information, please refer to `?read_steinbock`. 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(imcRtools)
library(stringr)
library(RColorBrewer)
library(circlize)
library(paletteer)
library(dittoSeq)
library(readxl)
library(dplyr)
library(patchwork)
library(ggplot2)
library(grid)  # for unit() in the cohort overview figure
```
</details>

## Read single-cell measurement

Replace `path` with the path to the folder where you stored the single-cell data (e.g. `path/to/folder/MapMetIP_ProcessedDataset` in case processed data was downloaded from Zenodo). 
In [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP), we exported single-cell marker measurements based on non-dilated and dilated (by 1px) segmentation masks to capture membrane signals. As both are required in the downstream analysis, we will read in both and combine them into one SPE object saving each expression matrix into one `assays`slot. Measurements based on non-dilated masks will be in the `assays`slot `counts` and the measurements based on dilated masks will be stored in `counts_1px`. 


```{r read-steinbock}
path <- file.path(params$input)

spe_0px <- read_steinbock(path=path, image_file=NULL, panel_file = file.path(path, 'NB_Panel.csv'), intensities_folder = "intensities-0px",regionprops_folder = "regionprops", graphs_folder=NULL) 
spe_1px <- read_steinbock(path=path, image_file=NULL, panel_file = file.path(path, 'NB_Panel.csv'), intensities_folder = "intensities-1px",regionprops_folder = "regionprops", graphs_folder=NULL) 

#Combine SPE objects into one object
spe <- spe_0px
assay(spe, "counts_1px") <- assay(spe_1px, "counts")

#remove single SPE objects
rm(spe_0px)
rm(spe_1px)
```

## Read single-cell metadata

We set the `colnames` of the object to generate unique identifiers per cell.

```{r set-colnames}
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)
```

When then read sample and patient-specific metadata from the metadata file provided on Zenodo [(Supplementary_Tables/20260526_Suppl_Table1.xlsx)](https://doi.org/10.5281/zenodo.13220634).

```{r read-metadata}

file_path <- file.path(params$input, 'Supplementary_Tables/20260526_Suppl_Table1.xlsx')
sample_spec_metadata <- read_excel(file_path, sheet = "sample_spec_metadata")
patient_spec_clinical <- read_excel(file_path, sheet = "patient_spec_clinical")
patient_spec_genetic <- read_excel(file_path, sheet = "patient_spec_genetic")

# Merge patient and sample-specific data
merged_data <- sample_spec_metadata %>%
  left_join(patient_spec_clinical, by = "Patient_ID")

meta <- merged_data %>%
  left_join(patient_spec_genetic, by = "Patient_ID")
```

and assign it to the colData slot in our SPE object.

```{r add-coldata}

#spe$sample <- gsub(spe$sample_id, pattern="([UMD]).*", replacement="\\1")
spe$sample <- sub("_[^_]+$", "", spe$sample_id)
spe$fm_id <- str_match(spe$sample, "_.*_(.*)_")[,2]  
spe$tissue <- str_match(spe$sample, "^.*_.*_(.*)")[,2] 
spe$tissue[spe$tissue=="TU"] = "PT"

spe$puncture_side <- meta$Puncture_side[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$imc_score <- meta$IMC_score[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$staining_date <- meta$Staining_date[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$stainer <- meta$Stainer[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$imc_date <- meta$IMC_date[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$control <- meta$control[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$patient_id <- meta$Patient_ID[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$timepoint <- meta$Timepoint[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]

# Keeping only primary tumor (PT) diagnosis (DE) samples + bone marrow (BM) control samples
spe <- spe[, (spe$timepoint == "DE" & spe$tissue == "PT") | spe$control == "yes"]

# Removing Patient_018 (after investigation, it's not a PT sample but a lymph node sample):
spe <- spe[, spe$patient_id != "Patient_018"]

spe$stage <- meta$INSS_stage[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$progression <- meta$efs[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$death <- meta$death[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$efs_dur <- meta$efs_dur[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$death_dur <- meta$death_dur[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$X1p_loss <- meta$X1p_loss[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$X11q_loss <- meta$X11q_loss[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$X17q_gain <- meta$X17q_gain[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$MYCN_amp <- meta$MYCN_amp[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$age_diag <- meta$age_at_diag[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$therapy <- meta$therapy[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]
spe$sex_aab <- meta$sex_aab[match(paste(spe$fm_id, spe$tissue), paste(meta$Sample_ID, meta$Tissue))]

spe$ROI <- str_match(spe$sample_id, "^.*_.*_.*_.*_(.*)")[,2]
spe$year <- str_c("20", str_extract(spe$fm_id, "^[0-9]+"))

```

Before we save the SPE object, we check the distribution of markers.

```{r plot by staining date}

tissue <- unique(spe$tissue)

marker <- 'DNA1_Ir191_mean'
p1 <- dittoRidgePlot(spe[,spe$tissue%in%tissue], var = marker, group.by = "tissue", assay = "counts") +
    ggtitle(paste(marker,  "- per tissue"))
marker <- 'DNA1_Ir193_mean'
p2 <- dittoRidgePlot(spe[,spe$tissue%in%tissue], var = marker, group.by = "tissue", assay = "counts") +
    ggtitle(paste(marker,  "- per tissue"))
p <- (p1 | p2)

p

save_plot(p, "DNA_markers_rawdata.png", width = 24, height = 12)

```

# Patient cohort overview {#cohort-overview}

We summarise the diagnostic primary tumor (PT, DE) cohort in a single overview figure: patient age at diagnosis as a bar plot, together with sex and the key genetic / clinical tracks (MYCN amplification, 1p loss and progression) drawn as colored tiles — one column per patient. The table is built directly from the sample- and patient-level metadata (`meta`, read above), so this figure does not depend on any processed single-cell object.

```{r cohort-table}
# Diagnostic primary tumor cohort: one row per patient. Tissue is coded "PT" in
# the metadata ("TU" is also accepted for robustness); Patient_018 is dropped to
# match the single-cell object (lymph-node, not a PT sample).
cohort <- meta %>%
  filter(Timepoint == "DE",
         Tissue %in% c("PT", "TU"),
         Patient_ID != "Patient_018") %>%
  distinct(Patient_ID, .keep_all = TRUE) %>%  # collapse multiple ROIs -> one row / patient
  transmute(
    patient_id = as.character(Patient_ID),
    age        = round(as.numeric(age_at_diag)),  # age at diagnosis (months), rounded
    sex        = as.character(sex_aab),    # sex (AAB)
    mycn       = as.character(MYCN_amp),
    loss1p     = as.character(X1p_loss),
    prog       = as.character(efs)         # event-free survival event: 0 = no, 1 = yes
  ) %>%
  arrange(patient_id)

# ---- Recode tracks ----
# MYCN amp. and 1p loss share a yes / no / het / unknown coding, recoded to a
# common yes / no / het scheme (unknown -> NA -> white tile) so the three
# genetic / clinical tracks can share a single legend.
recode_ynh <- function(x) dplyr::recode(as.character(x),
  "1" = "yes", "0" = "no", "2" = "het", "9" = NA_character_,
  "MNA" = "yes", "nMNA" = "no", "het" = "het", "unknown" = NA_character_,
  "yes" = "yes", "no" = "no",
  .default = NA_character_)

cohort$mycn   <- recode_ynh(cohort$mycn)
cohort$loss1p <- recode_ynh(cohort$loss1p)

# progression is coded 0 (no) / 1 (yes)
cohort$prog <- dplyr::recode(as.character(cohort$prog),
  "1" = "yes", "0" = "no", "9" = NA_character_,
  "Prog" = "yes", "NoProg" = "no", "yes" = "yes", "no" = "no",
  .default = NA_character_)

# sex -> f / m (handle the most common encodings, otherwise keep as-is)
cohort$sex <- dplyr::recode(as.character(cohort$sex),
  "1" = "f", "2" = "m", "female" = "f", "male" = "m", "F" = "f", "M" = "m",
  .default = as.character(cohort$sex))

# Fix column (patient) order and factor levels
lvl_ynh <- c("yes", "no", "het")
cohort$patient_id <- factor(cohort$patient_id, levels = cohort$patient_id)
cohort$mycn   <- factor(cohort$mycn,   levels = lvl_ynh)
cohort$loss1p <- factor(cohort$loss1p, levels = lvl_ynh)
cohort$prog   <- factor(cohort$prog,   levels = lvl_ynh)
cohort$sex    <- factor(cohort$sex,    levels = c("f", "m"))
```

The panels share the same discrete x-axis (one column per patient) and are stacked with `patchwork`, with the genetic / clinical tracks collected into a single legend.

```{r cohort-overview-figure, fig.width=7, fig.height=2.5}
# Cohort-track colors are defined centrally in 00_MapMet_color_code.R
for (p in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(p)) { source(p); break }
col_ynh <- COL_YNH
col_sex <- COL_SEX

# Shared discrete x-axis expansion so every panel aligns column-by-column.
X_EXPAND <- expansion(add = 0.5)

# ---- Age bar plot (top panel) ----
max_age <- max(cohort$age, na.rm = TRUE)
p_age <- ggplot(cohort, aes(x = patient_id, y = age)) +
  geom_col(fill = "grey35", width = 0.8) +
  scale_x_discrete(expand = X_EXPAND) +
  scale_y_continuous(breaks = c(0, max_age),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(y = "age[months]") +
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size = 7),
    axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 1, size = 8),
    panel.grid   = element_blank(),
    plot.margin  = margin(2, 2, 0, 2)
  )

# ---- Tile track helper ----
tile_track <- function(var, fill_vals, breaks, legend_name, y_lab) {
  ggplot(cohort, aes(x = patient_id, y = 1, fill = .data[[var]])) +
    geom_tile(color = "white", linewidth = 0.4) +
    scale_x_discrete(expand = X_EXPAND) +
    scale_fill_manual(values = fill_vals, breaks = breaks, name = legend_name,
                      na.value = "white", drop = FALSE) +
    labs(y = y_lab) +
    theme_void() +
    theme(
      legend.position  = "right",
      legend.direction = "vertical",
      legend.key.size  = unit(0.3, "cm"),
      legend.text      = element_text(size = 7),
      legend.title     = element_text(size = 8, face = "bold"),
      axis.title.y     = element_text(angle = 0, vjust = 0.5, hjust = 1, size = 8),
      plot.margin      = margin(0, 2, 0, 2)
    )
}

p_sex  <- tile_track("sex",    col_sex, c("f", "m"), "sex (AAB)",             "sex (AAB)")
p_mycn <- tile_track("mycn",   col_ynh, lvl_ynh,     "MNA / 1p loss / prog.", "MYCN amp.")
p_1p   <- tile_track("loss1p", col_ynh, lvl_ynh,     "MNA / 1p loss / prog.", "1p loss")
p_prog <- tile_track("prog",   col_ynh, lvl_ynh,     "MNA / 1p loss / prog.", "progression")

# ---- Stack panels; collect legends into a single vertical column on the right ----
p_cohort <- (p_age / p_sex / p_mycn / p_1p / p_prog) +
  plot_layout(heights = c(2, 0.5, 0.5, 0.5, 0.5), guides = "collect") &
  theme(legend.direction = "vertical", legend.box = "vertical",
        legend.justification = "top")

p_cohort

save_plot(p_cohort, "Fig1a_patient_cohort_overview.pdf", width = 7, height = 2.5)
```

## Save object

Finally, the generated SPE object can be saved for further downstream processing and analysis.

```{r save-objects-read-data}

saveRDS(spe, file.path(params$output, "spe_01_read_data.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
