---
title: "Quality control - Part 1"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
plot_dir <- file.path(params$output, "plots/02_QC_1")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot (ggplot / Heatmap / base R)
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap -> PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Quality control - Exclusion of lost cells {#qc-1}
Segmentation in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP) is performed based on high-resolution IF images. However, due to the washing step between IF microscopy and IMC staining, some cells might be lost. These have to be excluded before downstream analysis.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ggplot2)
library(bruceR)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(RColorBrewer)
library(paletteer)
library(Rphenoannoy)
library(scuttle)
library(cytomapper)
library(cowplot)
library(gridGraphics)
```
</details>

## Read SPE object

We will first read the spatial experiment object (SPE) generated in the [read data section](#read-data). Please make sure, `path` points to the folder where you stored the SPE object.

```{r read-steinbock}

path <- params$output
spe <- readRDS(file.path(path,"spe_01_read_data.rds"))
```

## Exclude lost cells

### Quality control step 1
In a first step, we exclude cells, that have a mean intensity of zero for all IMC channels. For that, we will use the intensities measured based on the non-dilated masks (stored in `counts`) to avoid capturing signals from neighboring cells. To correctly identify lost cells in the clustering step below, we use the mean of the 20% highest pixel intensities as a measure for marker intensity here, as this feature is less dependent on cell size.

```{r cell-exclusion-1}

# non-dilated masks (counts assay)
intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- spe$sample_id

# mean-80 (top-20% pixels): less dependent on cell size
intensity_mat_mean_80 <- intensity_mat[grepl("mean-80", rownames(intensity_mat)),]

# drop cells that are all-zero across IMC channels
intensity_mat_imc <- intensity_mat_mean_80[!grepl("IF2_GD2|IF3_CD45|DAPI", rownames(intensity_mat_mean_80)),]
non_zeros <- as.logical(colSums(intensity_mat_imc)!=0)

spe <- spe[, non_zeros]
mat <- intensity_mat_mean_80[, non_zeros]
```

We excluded `r length(non_zeros[non_zeros==FALSE])` cells in this first quality control step.

For the next quality control step, we will only consider IMC and nuclear IF channels. 
```{r cell-exclusion-2}
mat <- mat[!grepl("IF2_GD2|IF3_CD45", rownames(mat)),]
```

### Quality control step 2
Although, we excluded all cells with zero mean intensities in all IMC channels, there might still be some cells, which actually swam off between IF and IMC and have IMC counts due to background noise or lateral spillover. In the heatmap below, we can clearly observe a cluster of cells expressing the IF marker DAPI,  but no IMC markers.

```{r data-visualization, fig.width=11, fig.height=9}
# qualitative palette from 00_MapMet_color_code.R
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }
col_vector_433 <- COL_VECTOR_433

sample_names <- sub("_[^_]+$", "", colnames(mat))
col_sample <- setNames(sample(col_vector_433, length(unique(sample_names))), unique(sample_names))

set.seed(221228)
rand_cells <- sample(seq_len(ncol(mat)), 2000)

column_ha <- HeatmapAnnotation(sample = sample_names[rand_cells], 
                              col=list(sample=col_sample), 
                              show_legend=F)

p1 <- Heatmap(mat[,rand_cells], 
        column_title = "Single-cell heatmap",
        top_annotation = column_ha, 
        show_column_dend = F,
        show_column_names=F,
        col=viridis(100)
        )

save_plot(p1, "heatmap_mean80.png", width = 8, height = 6)

p1
```

We will cluster the data to identify cells that swam off.

```{r phenograph-clustering}
# Illustrative clustering only: Rphenoannoy is not reproducible across igraph
# versions / compilers / CPUs, so the actual lost-cell exclusion uses the frozen
# kept-cells list loaded below, not these clusters.
set.seed(230619)

mat <- t(mat)
Rphenograph_out <- Rphenoannoy(mat, k = 45)
clusters <- factor(membership(Rphenograph_out[[2]]))

spe$pg_clusters_lostcells <- clusters
```

Next, we will have a look at the mean marker expression per cluster in a heatmap.

```{r data-visualization-2, fig.width=11, fig.height=9}
mean <- aggregate(mat, list(clusters), mean)
mean_t <- t(mean[,2:ncol(mean)])
colnames(mean_t) <- mean[,"Group.1"]

n <- length(unique(clusters))
col_clusters <- setNames(sample(col_vector_433, n), unique(clusters))

column_ha = HeatmapAnnotation(cluster = mean[,"Group.1"], 
                              ncells=anno_barplot(as.data.frame(table(clusters))[,"Freq"], height = unit(3, "cm")),
                              col=list(cluster=col_clusters), 
                              show_legend=T)

p <- Heatmap(mean_t, 
        column_title = "Cluster-means heatmap",
        top_annotation = column_ha, 
        cluster_columns=T,
        show_column_dend = F,
        show_column_names=T,
        show_row_names=T,
        col=viridis(100)
        )

save_plot(p, "heatmap_clusters.png", width = 8, height = 6)

p
```

The cluster with DAPI signal but no IMC markers in the heatmap above is the
"lost cells" cluster - cells that swam off between IF and IMC. The clustering is
shown as an illustrative example only: because `Rphenoannoy` is not reproducible
across igraph versions, compilers or CPUs (see above), the exact clusters - and
therefore which cells are lost - can differ between machines. Instead of
excluding by cluster number, we load the frozen list of kept cells produced in
the reference environment and continue with those.

```{r load-kept-cells, message=TRUE}
kept_name <- "kept_cells_PTBM.rds"
inter_dirs <- c(Sys.getenv("MAPMET_INTERMEDIARY", ""),
                file.path(params$input,  "R_intermediary"),
                file.path(dirname(params$input), "R_intermediary"),
                file.path(params$output, "R_intermediary"))
inter_dirs <- inter_dirs[nzchar(inter_dirs)]
kept_file <- Find(file.exists, file.path(inter_dirs, kept_name))
if (is.null(kept_file))
  stop("Frozen kept-cells list '", kept_name, "' not found in any of: ",
       paste(inter_dirs, collapse = ", "), ". Generate it once in the reference ",
       "environment with code/generate_phenograph_intermediary.R, or set ",
       "MAPMET_INTERMEDIARY to the folder that holds it.")

kept_raw <- readRDS(kept_file)
# ids are the vector itself, or its names (older format)
kept_ids <- if (!is.null(names(kept_raw))) names(kept_raw) else as.character(kept_raw)
n_before <- ncol(spe)
keep <- colnames(spe) %in% kept_ids
n_missing <- length(setdiff(kept_ids, colnames(spe)))
if (!any(keep))
  stop("None of the ", length(kept_ids), " cells in ", kept_file, " match this run's ",
       "cells - wrong file or cell-id scheme (kept e.g. '", head(kept_ids, 1),
       "', current e.g. '", head(colnames(spe), 1), "').")
if (n_missing > 0)
  message(n_missing, " of ", length(kept_ids), " kept-list cells are not in this run's ",
          "post-QC1 set (they are already excluded here by cell-exclusion-1, e.g. as ",
          "all-zero cells) - ignoring them. A large count means the list does not ",
          "match this input.")
# keep only the frozen kept-cells
spe <- spe[, keep]
message("Lost-cell exclusion via frozen list: kept ", ncol(spe), " of ", n_before,
        " post-QC1 cells.")
```

### DNA-intensity check

As a further check on the surviving cells, we compare each cell's DNA intensity
against its total (non-DNA) protein intensity and quantify the fraction of
low-DNA cells per sample. The per-cell DNA intensity is stored on the SPE object
for downstream use; no cells are excluded at this step.

```{r}

library(SpatialExperiment)
library(ggplot2)

counts_matrix <- assay(spe, "counts")

# DNA channels, and channels excluded from the protein total
dna_idx <- grepl("DNA|Ir191|Ir193|DAPI", rownames(counts_matrix), ignore.case = TRUE)
unused_idx <- grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro",  rownames(counts_matrix))

dna_intensity <- colSums(counts_matrix[dna_idx, , drop = FALSE])
total_intensity_noDNA <- colSums(counts_matrix[!unused_idx, , drop = FALSE])

colData(spe)$DNAIntensity <- dna_intensity
colData(spe)$TotalIntensity_noDNA <- total_intensity_noDNA

df <- as.data.frame(colData(spe))

# linear fit + Spearman for the plot annotation
fit <- lm(TotalIntensity_noDNA ~ DNAIntensity, data = df)
r2 <- summary(fit)$r.squared
spearman_cor <- cor(df$DNAIntensity, df$TotalIntensity_noDNA, method = "spearman")

library(ggplot2)
library(ggrastr)

p <- ggplot(df, aes(x = DNAIntensity, y = TotalIntensity_noDNA)) +
  geom_point_rast(alpha = 0.05, size = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(
    title = paste0(
      "DNA vs Total intensity (non-DNA)\nR² = ",
      round(r2, 3),
      " | Spearman = ",
      round(spearman_cor, 3)
    ),
    x = "DNA intensity",
    y = "Total protein intensity"
  )

p

save_plot(p, "DNA_protein_correlation_scatter.png", width = 12, height = 12, dpi = 600)
```

```{r}

library(SpatialExperiment)
library(ggplot2)
library(dplyr)
library(hexbin)

spe_sub <- spe

counts_matrix <- assay(spe_sub, "counts")

# DNA channels (no DAPI here) and channels excluded from the protein total
dna_idx <- grepl("DNA|Ir191|Ir193", rownames(counts_matrix), ignore.case = TRUE)

unused_idx <- grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro",  
                   rownames(counts_matrix), ignore.case = TRUE)

dna_intensity <- colSums(counts_matrix[dna_idx, , drop = FALSE])
total_intensity_noDNA <- colSums(counts_matrix[!unused_idx, , drop = FALSE])

colData(spe_sub)$DNAIntensity <- dna_intensity
colData(spe_sub)$TotalIntensity_noDNA <- total_intensity_noDNA

df <- as.data.frame(colData(spe_sub))

# low-DNA threshold (23rd percentile)
low_dna_threshold <- quantile(df$DNAIntensity, 0.23)

# Spearman in the low-DNA subset (reported only)
df_low <- df %>% filter(DNAIntensity < low_dna_threshold)
low_dna_cor <- cor(df_low$DNAIntensity, df_low$TotalIntensity_noDNA, method = "spearman")
message("Spearman correlation in low DNA cells: ", round(low_dna_cor, 3))

p1 <- ggplot(df, aes(x = DNAIntensity, y = TotalIntensity_noDNA)) +
  geom_hex(bins = 100) +
  scale_fill_viridis_c() +
  geom_vline(xintercept = low_dna_threshold, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "DNA vs Total (non-DNA, filtered) intensity (density plot)",
       subtitle = paste("Low DNA cutoff at", round(low_dna_threshold, 3)),
       x = "DNA intensity",
       y = "Total protein intensity")



# median + mean protein intensity per DNA bin
df_summary <- df %>%
  mutate(bin = cut(DNAIntensity, breaks = 50)) %>%
  group_by(bin) %>%
  summarise(
    DNA = mean(DNAIntensity),
    MedianTotal = median(TotalIntensity_noDNA),
    MeanTotal = mean(TotalIntensity_noDNA)
  )

p_combined <- ggplot(df_summary, aes(x = DNA)) +
  geom_line(aes(y = MedianTotal, color = "Median"), size = 1) +
  geom_line(aes(y = MeanTotal, color = "Mean"), size = 1, linetype = "dashed") +
  geom_vline(xintercept = low_dna_threshold, color = "red", linetype = "dashed") +
  theme_minimal() +
  scale_color_manual(name = "Statistic", values = c("Median" = "blue", "Mean" = "darkgreen")) +
  labs(title = "Total protein intensity per DNA bin",
       subtitle = paste("Low DNA cutoff at", round(low_dna_threshold, 3)),
       x = "DNA (DAPI+DNA1/2) intensity",
       y = "Total protein intensity")


save_plot(p_combined, "DNA_protein_correlation_bin_mean.png", width = 12, height = 12)
save_plot(p1, "DNA_protein_correlation_density.png", width = 12, height = 12)

print(p_combined)
print(p1)

```


```{r, fig.width=12, fig.height=24}
library(dplyr)
library(ggplot2)

df <- as.data.frame(colData(spe_sub))
df <- df %>% mutate(LowDNA = DNAIntensity < low_dna_threshold)

# % low-DNA cells per sample, BM and PT treated separately
df_summary <- df %>%
  group_by(fm_id, tissue) %>%
  summarise(
    num_cells = n(),
    percent_low_DNA = 100 * sum(LowDNA) / n(),
    year = unique(year)
  ) %>%
  ungroup() %>%
  arrange(desc(percent_low_DNA)) %>%
  mutate(sample_label = paste(fm_id, tissue, sep = "_"))

p <- ggplot(df_summary, aes(x = percent_low_DNA, y = reorder(sample_label, percent_low_DNA), fill = tissue)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%% (%d, %s)", percent_low_DNA, num_cells, year)),
            hjust = -0.1, size = 3) +
  geom_vline(xintercept = 20, color = "red", linetype = "dashed", size = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = c("BM" = "skyblue", "PT" = "orange")) +
  labs(title = "% low-DNA cells per sample, colored by tissue (ncells, sample year)",
       x = "% low-DNA cells",
       y = "Sample ID + tissue",
       fill = "Tissue") +
  expand_limits(x = max(df_summary$percent_low_DNA) * 1.25)

p

save_plot(p, "low_DNA_samples.png", width = 9, height = 18)
```

## Save object

Finally, we save the SPE object with the lost cells removed.

```{r save-objects-read-data}

saveRDS(spe, file.path(params$output, "spe_02_QC_1.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
