---
title: "Quality control - Part 1"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/02_QC_1")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Quality control - Exclusion of lost cells {#qc-1}
Segmentation in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP) is performed based on high-resolution IF images. However, due to the washing step between IF microscopy and IMC staining, some cells might be lost. These have to be excluded before downstream analysis.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ggplot2)
library(bruceR)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(RColorBrewer)
library(paletteer)
library(Rphenoannoy)
library(scuttle)
library(cytomapper)
library(cowplot)
library(gridGraphics)
```
</details>

## Read SPE object

We will first read the spatial experiment object (SPE) generated in the [read data section](#read-data). Please make sure, `path` points to the folder where you stored the SPE object.

```{r read-steinbock}

path <- params$output
spe <- readRDS(file.path(path,"spe_01_read_data.rds"))
```

## Exclude lost cells

### Quality control step 1
In a first step, we exclude cells, that have a mean intensity of zero for all IMC channels. For that, we will use the intensities measured based on the non-dilated masks (stored in `counts`) to avoid capturing signals from neighboring cells. To correctly identify lost cells in the clustering step below, we use the mean of the 20% highest pixel intensities as a measure for marker intensity here, as this feature is less dependent on cell size.

```{r cell-exclusion-1}

# Use measurements based on non-dilated mask stored in counts
intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- spe$sample_id

#use mean-80 intensity to avoid losing big cells
intensity_mat_mean_80 <- intensity_mat[grepl("mean-80", rownames(intensity_mat)),]

#exclusion of all-zero cells
intensity_mat_imc <- intensity_mat_mean_80[!grepl("IF2_GD2|IF3_CD45|DAPI", rownames(intensity_mat_mean_80)),]
non_zeros <- as.logical(colSums(intensity_mat_imc)!=0)

spe <- spe[, non_zeros]
mat <- intensity_mat_mean_80[, non_zeros]
```

We excluded `r length(non_zeros[non_zeros==FALSE])` cells in this first quality control step.

For the next quality control step, we will only consider IMC and nuclear IF channels. 
```{r cell-exclusion-2}
mat <- mat[!grepl("IF2_GD2|IF3_CD45", rownames(mat)),]
```

### Quality control step 2
Although, we excluded all cells with zero mean intensities in all IMC channels, there might still be some cells, which actually swam off between IF and IMC and have IMC counts due to background noise or lateral spillover. In the heatmap below, we can clearly observe a cluster of cells expressing the IF marker DAPI,  but no IMC markers.

```{r data-visualization, fig.width=11, fig.height=9}
#named color vectors (generic qualitative palette defined centrally in 00_MapMet_color_code.R)
for (p in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(p)) { source(p); break }
qual_col_pals <- brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_433 <- COL_VECTOR_433

sample_names <- sub("_[^_]+$", "", colnames(mat))
col_sample <- setNames(sample(col_vector_433, length(unique(sample_names))), unique(sample_names))

set.seed(221228)
rand_cells <- sample(seq_len(ncol(mat)), 2000)

column_ha <- HeatmapAnnotation(sample = sample_names[rand_cells], 
                              col=list(sample=col_sample), 
                              show_legend=F)

p1 <- Heatmap(mat[,rand_cells], 
        column_title = "Single-cell heatmap",
        top_annotation = column_ha, 
        show_column_dend = F,
        show_column_names=F,
        col=viridis(100)
        )

save_plot(p1, "heatmap_mean80.png", width = 8, height = 6)

p1
```

We will cluster the data to identify cells that swam off.

```{r phenograph-clustering}
# Illustrative clustering only. Rphenoannoy (Annoy approximate-NN + igraph
# Louvain) is not reproducible across igraph versions, compilers or CPUs (the
# same caveat uwot documents for UMAP), so this clustering - and the heatmap
# below - is shown as an example. The actual lost-cell exclusion is driven by
# the frozen kept-cells list loaded further down, not by a cluster number.
set.seed(230619)

mat <- t(mat)
Rphenograph_out <- Rphenoannoy(mat, k = 45)
clusters <- factor(membership(Rphenograph_out[[2]]))

spe$pg_clusters_lostcells <- clusters
```

Next, we will have a look at the mean marker expression per cluster in a heatmap.

```{r data-visualization-2, fig.width=11, fig.height=9}
mean <- aggregate(mat, list(clusters), mean)
mean_t <- t(mean[,2:ncol(mean)])
colnames(mean_t) <- mean[,"Group.1"]

n <- length(unique(clusters))
col_clusters <- setNames(sample(col_vector_433, n), unique(clusters))

column_ha = HeatmapAnnotation(cluster = mean[,"Group.1"], 
                              ncells=anno_barplot(as.data.frame(table(clusters))[,"Freq"], height = unit(3, "cm")),
                              col=list(cluster=col_clusters), 
                              show_legend=T)

p <- Heatmap(mean_t, 
        column_title = "Cluster-means heatmap",
        top_annotation = column_ha, 
        cluster_columns=T,
        show_column_dend = F,
        show_column_names=T,
        show_row_names=T,
        col=viridis(100)
        )

save_plot(p, "heatmap_clusters.png", width = 8, height = 6)

p
```

The cluster with DAPI signal but no IMC markers in the heatmap above is the
"lost cells" cluster - cells that swam off between IF and IMC. The clustering is
shown as an illustrative example only: because `Rphenoannoy` is not reproducible
across igraph versions, compilers or CPUs (see above), the exact clusters - and
therefore which cells are lost - can differ between machines. Instead of
excluding by cluster number, we load the frozen list of kept cells produced in
the reference environment and continue with those.

```{r load-kept-cells}
kept_name <- "kept_cells_PTBM.rds"
inter_dirs <- c(Sys.getenv("MAPMET_INTERMEDIARY", ""),
                file.path(params$input,  "R_intermediary"),
                file.path(dirname(params$input), "R_intermediary"),
                file.path(params$output, "R_intermediary"))
inter_dirs <- inter_dirs[nzchar(inter_dirs)]
kept_file <- Find(file.exists, file.path(inter_dirs, kept_name))
if (is.null(kept_file))
  stop("Frozen kept-cells list '", kept_name, "' not found in any of: ",
       paste(inter_dirs, collapse = ", "), ". Generate it once in the reference ",
       "environment with code/generate_phenograph_intermediary.R, or set ",
       "MAPMET_INTERMEDIARY to the folder that holds it.")

kept_ids <- readRDS(kept_file)
if (!all(kept_ids %in% colnames(spe)))
  stop("Kept-cells list ", kept_file, " contains cells absent from this input; ",
       "regenerate it for this dataset.")

n_before <- ncol(spe)
spe <- spe[, colnames(spe) %in% kept_ids]
message("Kept ", ncol(spe), " of ", n_before, " cells from the frozen list ", kept_file)
```


```{r}

library(SpatialExperiment)
library(ggplot2)

# Extract counts (already normalized / arcsinh)
counts_matrix <- assay(spe, "counts")

# Identify DNA channels
dna_idx <- grepl("DNA|Ir191|Ir193|DAPI", rownames(counts_matrix), ignore.case = TRUE)
unused_idx <- grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro",  rownames(counts_matrix))

# Compute intensities
dna_intensity <- colSums(counts_matrix[dna_idx, , drop = FALSE])
total_intensity_noDNA <- colSums(counts_matrix[!unused_idx, , drop = FALSE])

# Store
colData(spe)$DNAIntensity <- dna_intensity
colData(spe)$TotalIntensity_noDNA <- total_intensity_noDNA

# Prepare dataframe
df <- as.data.frame(colData(spe))

# Fit linear model (no log)
fit <- lm(TotalIntensity_noDNA ~ DNAIntensity, data = df)
r2 <- summary(fit)$r.squared

# Spearman correlation (recommended for robustness)
spearman_cor <- cor(df$DNAIntensity, df$TotalIntensity_noDNA, method = "spearman")

library(ggplot2)
library(ggrastr)

p <- ggplot(df, aes(x = DNAIntensity, y = TotalIntensity_noDNA)) +
  geom_point_rast(alpha = 0.05, size = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(
    title = paste0(
      "DNA vs Total intensity (non-DNA)\nR² = ",
      round(r2, 3),
      " | Spearman = ",
      round(spearman_cor, 3)
    ),
    x = "DNA intensity",
    y = "Total protein intensity"
  )

p

save_plot(p, "DNA_protein_correlation_scatter.png", width = 12, height = 12, dpi = 600)
```

```{r}

library(SpatialExperiment)
library(ggplot2)
library(dplyr)
library(hexbin)

spe_sub <- spe

# Extract counts assay (already normalized)
counts_matrix <- assay(spe_sub, "counts")

# Identify DNA channels
dna_idx <- grepl("DNA|Ir191|Ir193", rownames(counts_matrix), ignore.case = TRUE)

# Identify channels to exclude from total intensity
unused_idx <- grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro",  
                   rownames(counts_matrix), ignore.case = TRUE)

# Compute DNA intensity (only DNA channels)
dna_intensity <- colSums(counts_matrix[dna_idx, , drop = FALSE])

# Compute total intensity excluding unused channels (including DNA and others)
total_intensity_noDNA <- colSums(counts_matrix[!unused_idx, , drop = FALSE])

# Add to colData
colData(spe_sub)$DNAIntensity <- dna_intensity
colData(spe_sub)$TotalIntensity_noDNA <- total_intensity_noDNA

# Prepare dataframe
df <- as.data.frame(colData(spe_sub))

# Define low DNA threshold (20th percentile)
low_dna_threshold <- quantile(df$DNAIntensity, 0.23)

# Subset low DNA cells
df_low <- df %>% filter(DNAIntensity < low_dna_threshold)

# Compute Spearman correlation in low DNA subset
low_dna_cor <- cor(df_low$DNAIntensity, df_low$TotalIntensity_noDNA, method = "spearman")
message("Spearman correlation in low DNA cells: ", round(low_dna_cor, 3))

# Hexbin plot of all data with cutoff line
p1 <- ggplot(df, aes(x = DNAIntensity, y = TotalIntensity_noDNA)) +
  geom_hex(bins = 100) +
  scale_fill_viridis_c() +
  geom_vline(xintercept = low_dna_threshold, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "DNA vs Total (non-DNA, filtered) intensity (density plot)",
       subtitle = paste("Low DNA cutoff at", round(low_dna_threshold, 3)),
       x = "DNA intensity",
       y = "Total protein intensity")



# Combine median and mean into one summary dataframe
df_summary <- df %>%
  mutate(bin = cut(DNAIntensity, breaks = 50)) %>%
  group_by(bin) %>%
  summarise(
    DNA = mean(DNAIntensity),
    MedianTotal = median(TotalIntensity_noDNA),
    MeanTotal = mean(TotalIntensity_noDNA)
  )

# Plot both median and mean on the same axes
p_combined <- ggplot(df_summary, aes(x = DNA)) +
  geom_line(aes(y = MedianTotal, color = "Median"), size = 1) +
  geom_line(aes(y = MeanTotal, color = "Mean"), size = 1, linetype = "dashed") +
  geom_vline(xintercept = low_dna_threshold, color = "red", linetype = "dashed") +
  theme_minimal() +
  scale_color_manual(name = "Statistic", values = c("Median" = "blue", "Mean" = "darkgreen")) +
  labs(title = "Total protein intensity per DNA bin",
       subtitle = paste("Low DNA cutoff at", round(low_dna_threshold, 3)),
       x = "DNA (DAPI+DNA1/2) intensity",
       y = "Total protein intensity")


save_plot(p_combined, "DNA_protein_correlation_bin_mean.png", width = 12, height = 12)
save_plot(p1, "DNA_protein_correlation_density.png", width = 12, height = 12)

# Print the combined plot
print(p_combined)


# Print plots
print(p1)

```


```{r, fig.width=12, fig.height=24}
library(dplyr)
library(ggplot2)

# Convert colData to dataframe
df <- as.data.frame(colData(spe_sub))

# Create LowDNA flag
df <- df %>% mutate(LowDNA = DNAIntensity < low_dna_threshold)

# Summarize per unique sample (fm_id + tissue)
df_summary <- df %>%
  group_by(fm_id, tissue) %>%   # treat BM and PT separately
  summarise(
    num_cells = n(),
    percent_low_DNA = 100 * sum(LowDNA) / n(),
    year = unique(year)  # assume all cells in a sample have same year
  ) %>%
  ungroup() %>%
  arrange(desc(percent_low_DNA)) %>%
  mutate(sample_label = paste(fm_id, tissue, sep = "_"))  # unique label for plotting

# Plot: x-axis = % low-DNA, color by tissue, 20% threshold line, text shows % + num cells + year
p <- ggplot(df_summary, aes(x = percent_low_DNA, y = reorder(sample_label, percent_low_DNA), fill = tissue)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%% (%d, %s)", percent_low_DNA, num_cells, year)),
            hjust = -0.1, size = 3) +
  geom_vline(xintercept = 20, color = "red", linetype = "dashed", size = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = c("BM" = "skyblue", "PT" = "orange")) +
  labs(title = "% low-DNA cells per sample, colored by tissue (ncells, sample year)",
       x = "% low-DNA cells",
       y = "Sample ID + tissue",
       fill = "Tissue") +
  expand_limits(x = max(df_summary$percent_low_DNA) * 1.25)  # space for text labels

p

save_plot(p, "low_DNA_samples.png", width = 9, height = 18)
```

## Save object

Finally, we exclude the lost cells from the SPE object and save the latter.

```{r save-objects-read-data}

saveRDS(spe, file.path(params$output, "spe_02_QC_1.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
