---
title: "Cell phenotyping"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/03.2_QC_3")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Cell phenotyping {#phenotyping}
In this Rmd file, we will perform 
1. dimensionality reduction
2. phenograph clustering
3. annotate and merge clusters & metaclusters externally 
4. retrieve tumor cells from GD2-MPO double positive cells (caused by spillover) 
5. re-classify granulocyte metaclusters into maturity states based on morphology
6. define color codes for all celltypes and metaclusters

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ComplexHeatmap)
library(RColorBrewer)
library(viridis)
library(scater)
library(BiocParallel)
library(Rphenoannoy)
library(tidyr)
library(tidyHeatmap)
library(bruceR)
library(dittoSeq)
library(dplyr)
library(mclust)
library(SingleR)
library(circlize)
library(readxl)
library(writexl)
library(ggplot2)
library(patchwork)
library(harmony)
library(cluster)
```
</details>

## Read data
We will first read the data that passed the [second quality control step](#qc-2) and save the count data as a matrix. Low-quality markers (low SNR in a lot of samples) as well as all IF and nuclear IMC markers will be excluded for clustering. Moreover, mean intensity will be used to represent marker expression, to average out the signals arising from lateral spillover or artifacts. As explained in [the previous chapter](#qc2-save), we are using counts measured with respect to dilated masks in downstream analysis do capture membrane signal. 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_qc2.rds"))

# Adding AD samples metadata for Harmony batch correction
colData(spe)$stainer[colData(spe)$tissue == "AD"] <- "Stainer_1"
colData(spe)$imc_date[colData(spe)$tissue == "AD"] <- "2025-06-23"
# typo correction, as IMC obviously didn't exist in 2002
colData(spe)$imc_date[colData(spe)$imc_date == "2002-10-18"] <- "2022-10-18"

# Subset data

tissue_vec <- colData(spe)$tissue
idx <- tissue_vec %in% c("PT", "BM")

spe <- spe[, idx]

intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- colnames(spe)

#use mean intensity
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]

#exclude low-quality markers and functional marker cl.PARP
intensity_mat_hq <- intensity_mat[!grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(intensity_mat)),]
intensity_mat_lq <- intensity_mat[grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(intensity_mat)),]

non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)

intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]
intensity_mat_lq_wo_zeros <- intensity_mat_lq[,non_zeros]

#Apply changes to SPE object
spe <- spe[,non_zeros]
rowData(spe)$use_channel <- !grepl("mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(spe))

##Matrix with morph. features
morph_mat_clipped <- (cbind("area"=spe$area_clipped, "solidity"=spe$solidity_clipped, "aspect_ratio"=spe$aspect_ratio_clipped))

tissue <- spe$tissue
cell_id <- colnames(spe)

mat <- intensity_mat_hq_wo_zeros

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_sample = setNames(sample(col_vector_433, length(unique(spe$sample))), unique(spe$sample))
```
We are excluding `r length(non_zeros[non_zeros==FALSE])` more cells here, as they are negative for all markers used in downstream analysis. 
To get a first grasp of the single-cell data and potential clusters, we will plot it in a single-cell heatmap.


## Dimensionality reduction via PCA prior to clustering

```{r}
set.seed(221228)


# Transpose: cells as rows
mat_t <- t(mat)

# Compute variance per marker (columns)
marker_var <- apply(mat_t, 2, var)

# Identify non-constant and removed markers
non_constant <- marker_var > 0
removed_markers <- colnames(mat_t)[!non_constant]

# Print summary
cat("Number of removed markers:", length(removed_markers), "\n")
cat("Removed markers:\n")
print(removed_markers)

# Filter matrix
mat_t_filt <- mat_t[, non_constant, drop = FALSE]

# Run PCA
pca_res <- prcomp(mat_t_filt, center = TRUE, scale. = TRUE)

# Variance explained
pca_var <- pca_res$sdev^2
pca_var_explained <- pca_var / sum(pca_var)
pca_var_cum <- cumsum(pca_var_explained)

# --- Elbow detection (2nd derivative method) ---
diff1 <- diff(pca_var_explained)
diff2 <- diff(diff1)
elbow_pc <- which.max(-diff2) + 1

# --- Variance thresholds ---
pc_80 <- which(pca_var_cum >= 0.8)[1]
pc_90 <- which(pca_var_cum >= 0.9)[1]



cat("Elbow PC:", elbow_pc, "\n")
cat("PCs for 80% variance:", pc_80, "\n")
cat("PCs for 90% variance:", pc_90, "\n")

# --- Plots ---
par(mfrow = c(1, 2))

# Scree plot
plot(pca_var_explained, type = "b", pch = 16,
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot")
abline(v = elbow_pc, col = "red", lty = 2)

# Cumulative variance
plot(pca_var_cum, type = "b", pch = 16,
     xlab = "Principal Component",
     ylab = "Cumulative Variance",
     main = "Cumulative Variance")
abline(h = 0.8, col = "red", lty = 2)
abline(h = 0.9, col = "blue", lty = 2)

```

```{r}
# --- Final choice ---
n_pcs_use <- 25
# --- Subset PCs ---
pca_mat <- pca_res$x[, 1:n_pcs_use]
rownames(pca_mat) <- colnames(spe)
```

## UMAP

```{r umap-post-harmony}
set.seed(221228)

Sys.setenv(OMP_NUM_THREADS = 1)
Sys.setenv(MKL_NUM_THREADS = 1)

BPPARAM <- SnowParam(workers = 30, type = "SOCK", RNGseed = 221228)

mat_red <- calculateUMAP(
  t(pca_mat),
  n_neighbors = 30,
  BPPARAM = BPPARAM
)

rownames(mat_red) <- colnames(spe)
reducedDim(spe, "UMAP") <- mat_red
```

## Phenograph clustering on PCA space after harmony

```{r phenograph-clustering}
set.seed(221228)

Rphenograph_out <- Rphenoannoy(pca_mat, k = 45)
clusters <- factor(membership(Rphenograph_out[[2]]))

saveRDS(clusters, file.path(params$output, "pg_clusters_45.rds"))

spe$pg_clusters <- clusters
```


## UMAP: Phenograph clusters
Lets visualize phenograph clusters and tissue types in UMAP space. Here `dittoDimPlot` can be used directly on the SPE object.

```{r UMAP-vis-clusters}
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
col_clusters = setNames(sample(col_vector_433, length(unique(spe$pg_clusters))), unique(spe$pg_clusters_))

p1 <- dittoDimPlot(spe, var = "pg_clusters", 
             reduction.use = "UMAP", size = 1,
             do.label = TRUE,
             legend.show=F) +
  scale_color_manual(values = col_clusters) + 
  theme(legend.title = element_blank()) +
  ggtitle("PG clusters on UMAP")

# Save ggplot
save_plot(p1, "UMAP-clusters.png", width = 8, height = 6)
p1
```

## Bubble plot: Mean marker expression per cluster
Let's plot the mean marker expressions per cluster in a bubble plot.

```{r vis-data-bubbleplot, fig.height=14, fig.width=16}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(viridis)
library(cowplot)

stopifnot(exists("mat"), exists("spe"))

cluster_vec <- spe$pg_clusters
tissue_vec  <- colData(spe)$tissue

message("Number of cells: ", ncol(mat))
message("Number of markers: ", nrow(mat))

mat_combined <- mat #rbind(mat, intensity_mat_lq_wo_zeros)

mat_df <- as.data.frame(t(mat_combined))

exp_mat_ <- aggregate(
  mat_df,
  by = list(cluster = cluster_vec),
  FUN = function(x) {
    if (all(x == 0)) return(0)
    median(log1p(x[x > 0]))
  }
)

cluster_names <- exp_mat_$cluster
exp_mat <- exp_mat_[, -1, drop = FALSE]
rownames(exp_mat) <- cluster_names
exp_mat <- t(as.matrix(exp_mat))

rmin <- apply(exp_mat, 1, min, na.rm = TRUE)
rmax <- apply(exp_mat, 1, max, na.rm = TRUE)
den  <- rmax - rmin

exp_mat <- (exp_mat - rmin) / den
exp_mat[is.nan(exp_mat)] <- 0
exp_mat[is.infinite(exp_mat)] <- 0

clusters <- colnames(exp_mat)
message("Dimensions of exp_mat: ", paste(dim(exp_mat), collapse = " x "))

percent_mat <- sapply(seq_len(nrow(mat_combined)), function(i) {
  row_vals <- mat_combined[i, ]
  pos <- row_vals[row_vals > 0]

  if (length(pos) == 0) {
    out <- rep(0, length(clusters))
    names(out) <- clusters
    return(out)
  }

  thresh <- quantile(pos, 0.2, na.rm = TRUE)
  row_bin <- ifelse(row_vals > thresh, 1, 0)

  tapply(row_bin, factor(cluster_vec, levels = clusters), mean)
})

percent_mat <- t(percent_mat)
percent_mat <- percent_mat[, clusters, drop = FALSE]
percent_mat[is.na(percent_mat)] <- 0
percent_mat[percent_mat < 0.1] <- 0.1

message("Dimensions percent_mat: ", paste(dim(percent_mat), collapse = " x "))

bm_count <- table(cluster_vec[tissue_vec == "BM"])
pt_count <- table(cluster_vec[tissue_vec == "PT"])

bm_count <- bm_count[clusters]; bm_count[is.na(bm_count)] <- 0
pt_count <- pt_count[clusters]; pt_count[is.na(pt_count)] <- 0

max_pt <- max(pt_count, na.rm = TRUE)
max_bm <- max(bm_count, na.rm = TRUE)

mean_morph <- aggregate(
  morph_mat_clipped,
  list(cluster = cluster_vec),
  function(x) mean(x, na.rm = TRUE)
)

area_mean <- mean_morph[match(clusters, mean_morph$cluster), "area"]

col_mean_area <- colorRamp2(
  c(min(area_mean), median(area_mean), max(area_mean)),
  c("blue", "white", "red")
)

if (!exists("col_clusters")) {
  col_clusters <- setNames(rainbow(length(clusters)), clusters)
} else {
  missing_clusters <- setdiff(clusters, names(col_clusters))
  if (length(missing_clusters) > 0) {
    col_clusters[missing_clusters] <- rainbow(length(missing_clusters))
  }
}

# Barplot axis: show only the max tick (no 0 tick)
axis_maxtick <- function(v) {
  vmax <- max(as.numeric(v), na.rm = TRUE)
  list(
    at = vmax,
    labels = as.character(vmax),
    gp = gpar(fontsize = 12)
  )
}


column_ha <- HeatmapAnnotation(
  pg_clusters = clusters,
  
  ncells_PT = anno_barplot(
    as.numeric(pt_count),
    height = unit(0.7, "cm"),
    gp = gpar(fill = "brown"),
    axis_param = axis_maxtick(pt_count)
  ),


  ncells_BM = anno_barplot(
    as.numeric(bm_count),
    height = unit(0.7, "cm"),
    gp = gpar(fill = "aquamarine4"),
    axis_param = axis_maxtick(bm_count)
  ),

  col = list(
    pg_clusters = col_clusters,
    roundness = col_mean_solidity,
    area = col_mean_area
  ),

  show_legend = TRUE
)

col_fun <- circlize::colorRamp2(
  c(0, 0.5, 1),
  viridis(100)[c(1, 50, 100)]
)

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid.rect(x, y, w, h, gp = gpar(col = NA, fill = NA))
  grid.circle(
    x = x,
    y = y,
    r = percent_mat[i, j] / 2 * min(unit.c(w, h)),
    gp = gpar(fill = col_fun(exp_mat[i, j]), col = NA)
  )
}

lgd_list <- list(
  Legend(
    labels = c(0, 0.25, 0.5, 0.75, 1),
    title = "percentage expressed",
    graphics = lapply(c(0, 0.25, 0.5, 0.75, 1), function(r) {
      function(x, y, w, h) grid.circle(
        x = x, y = y,
        r = r * unit(2, "mm"),
        gp = gpar(fill = "black")
      )
    })
  )
)

hp <- Heatmap(
  exp_mat,
  heatmap_legend_param = list(title = "expression"),
  column_title = "Bubble plot - normalized mean expression and percent expressed",
  top_annotation = column_ha,
  col = col_fun,
  rect_gp = gpar(type = "none"),
  clustering_distance_columns = "pearson",
  clustering_method_columns = "average",
  row_labels = sapply(strsplit(rownames(exp_mat), "_"), "[[", 1),
  cell_fun = cell_fun,
  show_row_dend = FALSE,
  column_dend_height = unit(1.5, "cm"),
  row_names_gp = gpar(fontsize = 15),
  column_names_gp = gpar(fontsize = 15),
  column_names_rot = 45,
  border = "black"
)

hp_drawn <- draw(hp, annotation_legend_list = lgd_list)
save_plot(hp_drawn, "vis-data-bubbleplot.pdf", width = 16, height = 14)

hp_drawn
```

```{r}
p1 <- dittoDimPlot(spe, var = "DNA1_Ir191_mean", assay = "counts",
             reduction.use = "UMAP", size = 0.2, 
             colors = viridis(100), do.label = TRUE)

p1
```


```{r, filtering lowqual clusters}

bad_quality_clusters <- c(1,2,3)

spe_clean <- spe[, !spe$pg_clusters %in% bad_quality_clusters]

```

## Save object

```{r save-objects}

saveRDS(spe_clean, file.path(params$output, "spe_QC.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>