---
title: "Quality control - Part 2"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---


```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/03_QC_2")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Quality control - Exclusion of segmentation artifacts {#qc-2}
Despite the employment of a highly-accurate (fine-tuned) cellpose model for segmentation, segmentation artifacts can't be avoided completely. As a second quality control step, we will exclude these before we continue with downstream analysis. 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ggplot2)
library(cowplot)
library(cytomapper)
```
</details>

## Read SPE object

We will first read the spatial experiment object (SPE) that passed the [first quality check](#qc-1). Please make sure, `path` points to the folder where you stored the SPE object.


```{r read-steinbock}
path <- params$output
spe <- readRDS(file.path(path,"spe_02_QC_1.rds"))
```

For the exclusion of segmentation artifacts, we will examine three morphological features (area, solidity and aspect ratio). This step will be done for primary tumor and bone marrow samples separately, since these two tissue types have distinct morphological appearances. Bone marrow samples were collected in the form of cytospin samples from bone marrow aspirates (cells in suspension) and hence present inherently different nuclear morphologies.

```{r sep-tissues}
tissue <- spe$tissue
spe_pt <- spe[,tissue=="PT"]
spe_bm <- spe[,tissue=="BM"]

morph_names <- c("area",
                 "solidity",
                 "aspect_ratio-0"
                 )
```

## Threshold identification for BM samples
Next, we plot distributions of morphological features in histograms to identify feature-individual thresholds for the exclusion of segmentation artifacts. We start with BM samples.

```{r plot-histograms-bm}
morph_mat_bm <- as.matrix(t(colData(spe_bm)[,morph_names]))
rownames(morph_mat_bm) <- gsub('-0','',rownames(morph_mat_bm))
colnames(morph_mat_bm) <- colnames(spe_bm)

plot_list <- list()

df <- as.data.frame(t(morph_mat_bm))

for(i in 1: ncol(df)){

  feature <- colnames(df)[i]
  
  plot_list[[i]] <- ggplot(df, aes_string(feature)) +
  geom_density(adjust = 1, color="darkblue", fill="lightblue")
}

plot <- plot_grid(plotlist = plot_list, nrow=ceiling(ncol(df)/3), ncol=3) #change this to the number of morphological features


save_plot(plot, "morpho_features_BM.png", width = 12, height = 12)


plot
```

In the plots above, the distributions are skewed indicating the presence of outliers. Let's examine these outliers cells. We will identify their cell ID and check their morphology in images and masks.

```{r quality-control-outliers}
feat <- "area"

morph_mat_order <- morph_mat_bm[feat,order(morph_mat_bm[feat,], decreasing=T)]
outlier_cell <- names(morph_mat_order[2])
value <- morph_mat_order[outlier_cell]
id <- sub(".*_", "", outlier_cell)
image <- sub("_[^_]*$", "", outlier_cell)
print(paste("The outlier cell with ", feat, "=", value, " with ID", id, "is in image", image,"."))
```

Let's inspect this cell in the image and mask.

```{r visualize-outliers, fig.show='hide'}
masks <- loadImages(file.path(params$input, "masks", paste0(image, ".tif")), as.is = TRUE)
img <- loadImages(file.path(params$input, "img", paste0(image, ".tif")))

mcols(masks) <- DataFrame(sample_id = names(masks))
mcols(img) <- DataFrame(sample_id = names(img))
channelNames(img) <- rownames(spe)[1:44]

tmp <- spe[,spe$sample_id==image]
tmp$outlier <- "no"
tmp[,colnames(tmp)==outlier_cell]$outlier <- "yes"

dapi_plot <- plotPixels(img, 
                       colour_by = "IF1_DAPI_mean",
                       bcg = list(IF1_DAPI_mean = c(0, 1, 1)),
                       legend=NULL,
                       return_plot = TRUE)

mask_plot <- plotCells(masks,
                       object = tmp, 
                       cell_id = "ObjectNumber", 
                       img_id = "sample_id",
                       colour_by = "outlier",
                       colour = list(outlier = c("no" = "gray", "yes" = "red")),
                       missing_colour = "white",
                       legend=NULL,
                       return_plot = TRUE)
```
```{r plot-grid-1}
p1 <- ggdraw(dapi_plot$plot, clip = "on")
p2 <- ggdraw(mask_plot$plot, clip = "on")

p <- plot_grid(p1, p2)

save_plot(p, "example_large_cell.png", width = 12, height = 12)
p
```


In the images above, we can see that segmentation artifacts occur, where no clear cell borders can be identified. For BM, we determined an upper threshold of **15,600** and a lower threshold of **333** for both PT and BM samples. To examine nuclei with an area below the lower threshold, it is better to visualize images in napari and use the label and zoom function.

## Threshold identification for PT samples
We will now repeat the same worfklow for PT samples.

```{r plot-histograms-pt}
morph_mat_pt <- as.matrix(t(colData(spe_pt)[,morph_names]))
rownames(morph_mat_pt) <- gsub('-0','',rownames(morph_mat_pt))
colnames(morph_mat_pt) <- colnames(spe_pt)

plot_list <- list()

df <- as.data.frame(t(morph_mat_pt))

for(i in 1: ncol(df)){

  feature <- colnames(df)[i]
  
  plot_list[[i]] <- ggplot(df, aes_string(feature)) +
  geom_density(adjust = 1, color="darkblue", fill="lightblue")
}

plot <- plot_grid(plotlist = plot_list, nrow=ceiling(ncol(df)/3), ncol=3) #change this to the number of morphological features

save_plot(plot, "morpho_features_PT.png", width = 12, height = 12)

plot
```

As described, PT-derived nuclei have an inherently different morphology. In the histograms above we can see that they are smaller and less round than BM-derived nuclei. Again, we explore outlier cells to identify appropriate thresholds.

```{r quality-control-outliers-2, message=FALSE}
feat <- "area"

morph_mat_order <- morph_mat_bm[feat,order(morph_mat_bm[feat,], decreasing=F)]
outlier_cell <- names(morph_mat_order[1])
print(paste("The outlier cell with ", feat, "=", morph_mat_order[outlier_cell], " with ID", sub(".*_", "", outlier_cell), "is in image", sub("_[^_]*$", "", outlier_cell),". Check, whether this cell is an artifact by inspecting the large segmentation masks in napari."))
```

Using the code above and cross-checking images of corresponding nuclei in napari, we can identify a lower threshold of **333** for nuclear area for PT samples. There is however no upper threshold, since the largest cells are also correctly segmented.

## Exclusion of segmentation artifacts
Let's exclude outliers from the SPE object and the morph. feature matrix based on defined thresholds.

```{r exclude-outliers}
outliers <- (spe$area <= 333 & spe$tissue == "PT" | spe$area <= 333 & spe$tissue == "BM" | spe$area >= 15600 & spe$tissue == "BM")

spe <- spe[,!outliers]

#Rebuild the matrix of morphological features of the remaining cells
morph_mat <- as.matrix(t(colData(spe)[,morph_names]))
rownames(morph_mat) <- gsub('-0','',rownames(morph_mat))
```

We excluded `r sum(outliers)` cells based on these thresholds. 

## Clipping and scaling of morphological features

Before we save the SPE object, we will clip and scale morphological features for **visualization purposes**. 

```{r clip-morph}

norm_minmax <- function(x){(x- min(x)) /(max(x)-min(x))}

clipper <- function(data, lower, upper) {
  data[data > quantile(data, probs = c(upper))] <- quantile(data, probs = c(upper))
  data[data < quantile(data, probs = c(lower))] <- quantile(data, probs = c(lower))
  clipped_data <- norm_minmax(data)
  return(clipped_data)}

tissue <- spe$tissue

area_upper <- 0.99999
solidity_lower <- 0.001
aspect_ratio_upper <- 0.9995 

clipped_area <- c(clipper(morph_mat[1,tissue=="BM"], 0, area_upper), clipper(morph_mat[1,tissue=="PT"], 0, area_upper))
clipped_area <- clipped_area[colnames(spe)]

morph_mat_clipped <- rbind(clipped_area, clipper(morph_mat[2,], solidity_lower, 1), clipper(morph_mat[3,], 0, aspect_ratio_upper))

colnames(morph_mat_clipped) <- colnames(spe)
rownames(morph_mat_clipped) <- c("area", "solidity", "aspect_ratio")

spe$area_clipped <- unname(morph_mat_clipped["area",])
spe$solidity_clipped <- unname(morph_mat_clipped["solidity",])
spe$aspect_ratio_clipped <- unname(morph_mat_clipped["aspect_ratio",])
```

We plot the distribution of clipped morphological features...

```{r histograms-separate-x-axis}
plot_list <- list()

df <- as.data.frame(t(morph_mat_clipped))

for(i in 1: ncol(df)){

  feature <- colnames(df)[i]
  
  plot_list[[i]] <- ggplot(df, aes_string(feature)) +
  geom_density(adjust = 5, color="darkblue", fill="lightblue")
}

plot <- plot_grid(plotlist = plot_list, nrow=ceiling(ncol(df)/3), ncol=3) #change this to the number of morphological features



save_plot(plot, "morph_.png", width = 12, height = 12)


plot
```

...and visualize the masks in one PT and one BM sample.

```{r visualize-clipped-morph}
# Define morphs
morph_names_clipped <- c("area_clipped", "solidity_clipped", "aspect_ratio_clipped")

# Load masks
masks <- loadImages(
  c(
    file.path(params$input, "masks/20220517_20220309_20-154_BM_003.tif"),
    file.path(params$input, "masks/20220926_20220809_16-005_TU_001.tif")
  ),
  as.is = TRUE
)

# Map mask files to the corresponding SPE sample IDs
# You need to adjust this vector to match your SPE object
mask_to_spe_ids <- c(
  "20220517_20220309_20-154_BM_003",  # maps to first mask
  "20220926_20220809_16-005_TU_001"   # maps to second mask
)

names(masks) <- mask_to_spe_ids
mcols(masks)$sample_id <- mask_to_spe_ids

# Store plots
plot_list <- list()

# Loop over morphs and masks
for (m in seq_along(morph_names_clipped)) {
  for (i in seq_along(masks)) {
    
    current_id <- names(masks)[i]
    current_morph <- morph_names_clipped[m]
    
    message(paste0("Plotting: ", current_morph, " | ", current_id))
    
    # Generate plot for a single mask/image
    p <- plotCells(
      masks[i],
      object = spe,
      cell_id = "ObjectNumber",
      img_id = "sample_id",
      colour_by = current_morph,
      return_plot = TRUE,
      return_images = TRUE,
      legend = NULL
    )
    
    # Save plot with morph + image ID
    save_plot(
      p,
      filename = paste0(current_morph, "_", current_id, ".png"),
      width = 12,
      height = 12
    )
    
    # Store plot in list (optional)
    plot_list[[paste0(current_morph, "_", current_id)]] <- p
  }
}
```

## Save object {#qc2-save}

From now on, we will use intensity measurements computed based on the dilated mask (by 1px). To reduce the size of the SPE object, we remove intensity measurements obtained based on non-dilated masks and set those based on dilated masks as default. 

```{r save-objects-read-data}
assay(spe, "counts") <- assay(spe, "counts_1px")
assay(spe, "counts_1px") <- NULL

saveRDS(spe, file.path(params$output, "spe_03_QC_2.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
