---
title: "Cell phenotyping"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/04.1_phenotyping_AD")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Cell phenotyping {#phenotyping}
In this Rmd file, we will perform 
1. dimensionality reduction
2. phenograph clustering
3. annotate and merge clusters & metaclusters externally 
4. retrieve tumor cells from GD2-MPO double positive cells (caused by spillover) 
5. re-classify granulocyte metaclusters into maturity states based on morphology
6. define color codes for all celltypes and metaclusters

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ComplexHeatmap)
library(RColorBrewer)
library(viridis)
library(scater)
library(BiocParallel)
library(Rphenoannoy)
library(tidyr)
library(tidyHeatmap)
library(bruceR)
library(dittoSeq)
library(dplyr)
library(mclust)
library(SingleR)
library(circlize)
library(readxl)
library(writexl)
library(ggplot2)
library(patchwork)
library(harmony)
library(cluster)
```
</details>

## Read data
We will first read the data that passed the [second quality control step](#qc-2) and save the count data as a matrix. Low-quality markers (low SNR in a lot of samples) as well as all IF and nuclear IMC markers will be excluded for clustering. Moreover, mean intensity will be used to represent marker expression, to average out the signals arising from lateral spillover or artifacts. As explained in [the previous chapter](#qc2-save), we are using counts measured with respect to dilated masks in downstream analysis do capture membrane signal. 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_qc2.rds"))

# Adding AD samples metadata for Harmony batch correction
colData(spe)$stainer[colData(spe)$tissue == "AD"] <- "Stainer_1"
colData(spe)$imc_date[colData(spe)$tissue == "AD"] <- "2025-06-23"
# typo correction, as IMC obviously didn't exist in 2022
colData(spe)$imc_date[colData(spe)$imc_date == "2002-10-18"] <- "2022-10-18"

# Subset data
idx_AD <- colData(spe)$tissue == "AD"
spe <- spe[, idx_AD]

intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- colnames(spe)

#use mean intensity
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]

#exclude low-quality markers and functional marker cl.PARP
intensity_mat_hq <- intensity_mat[!grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(intensity_mat)),]
intensity_mat_lq <- intensity_mat[grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(intensity_mat)),]

non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)

intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]
intensity_mat_lq_wo_zeros <- intensity_mat_lq[,non_zeros]

#Apply changes to SPE object
spe <- spe[,non_zeros]
rowData(spe)$use_channel <- !grepl("mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(spe))

##Matrix with morph. features
morph_mat_clipped <- (cbind("area"=spe$area_clipped, "solidity"=spe$solidity_clipped, "aspect_ratio"=spe$aspect_ratio_clipped))

tissue <- spe$tissue
cell_id <- colnames(spe)

mat <- intensity_mat_hq_wo_zeros

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_sample = setNames(sample(col_vector_433, length(unique(spe$sample))), unique(spe$sample))
```
We are excluding `r length(non_zeros[non_zeros==FALSE])` more cells here, as they are negative for all markers used in downstream analysis. 
To get a first grasp of the single-cell data and potential clusters, we will plot it in a single-cell heatmap.


## Dimensionality reduction via PCA prior to clustering

```{r}
set.seed(221228)


# Transpose: cells as rows
mat_t <- t(mat)

# Compute variance per marker (columns)
marker_var <- apply(mat_t, 2, var)

# Identify non-constant and removed markers
non_constant <- marker_var > 0
removed_markers <- colnames(mat_t)[!non_constant]

# Print summary
cat("Number of removed markers:", length(removed_markers), "\n")
cat("Removed markers:\n")
print(removed_markers)

# Filter matrix
mat_t_filt <- mat_t[, non_constant, drop = FALSE]

# Run PCA
pca_res <- prcomp(mat_t_filt, center = TRUE, scale. = TRUE)

# Variance explained
pca_var <- pca_res$sdev^2
pca_var_explained <- pca_var / sum(pca_var)
pca_var_cum <- cumsum(pca_var_explained)

# --- Elbow detection (2nd derivative method) ---
diff1 <- diff(pca_var_explained)
diff2 <- diff(diff1)
elbow_pc <- which.max(-diff2) + 1

# --- Variance thresholds ---
pc_80 <- which(pca_var_cum >= 0.8)[1]
pc_90 <- which(pca_var_cum >= 0.9)[1]



cat("Elbow PC:", elbow_pc, "\n")
cat("PCs for 80% variance:", pc_80, "\n")
cat("PCs for 90% variance:", pc_90, "\n")

# --- Plots ---
par(mfrow = c(1, 2))

# Scree plot
plot(pca_var_explained, type = "b", pch = 16,
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot")
abline(v = elbow_pc, col = "red", lty = 2)

# Cumulative variance
plot(pca_var_cum, type = "b", pch = 16,
     xlab = "Principal Component",
     ylab = "Cumulative Variance",
     main = "Cumulative Variance")
abline(h = 0.8, col = "red", lty = 2)
abline(h = 0.9, col = "blue", lty = 2)

```

```{r}
# --- Final choice ---
n_pcs_use <- 16
# --- Subset PCs ---
pca_mat <- pca_res$x[, 1:n_pcs_use]
rownames(pca_mat) <- colnames(spe)
```

## Pre-Harmony UMAP for batch effect visualization
Before applying Harmony batch correction, we compute a UMAP embedding from the raw PCA space to make the batch structure visible prior to correction.

```{r umap-pre-harmony}
set.seed(221228)

# Keep raw PCA for pre-Harmony comparison
pca_mat_raw <- pca_mat

# Compute UMAP on pre-Harmony PCA embeddings
mat_red_pre <- calculateUMAP(
  t(pca_mat_raw),
  n_neighbors = 30,
  BPPARAM = MulticoreParam(RNGseed = 221228)
)

rownames(mat_red_pre) <- colnames(spe)
reducedDim(spe, "UMAP_preharmony") <- mat_red_pre
```

## Phenograph before Harmony

```{r phenograph-clustering}
set.seed(221228)

Rphenograph_out <- Rphenoannoy(pca_mat, k = 100)
clusters <- factor(membership(Rphenograph_out[[2]]))

spe$pg_clusters_noharmony <- clusters
```

## Harmony batch correction
Harmony is applied to the PCA matrix to remove batch effects across samples before clustering and UMAP visualization.

```{r harmony-batch-correction}
# Apply Harmony batch correction to the PCA embeddings to account for
# batch effects across fm_id, tissue, stainer, and imc_date. The corrected matrix retains
# the same dimensions (cells x 32 PCs) and replaces pca_mat for all downstream steps.
pca_mat <- HarmonyMatrix(
  data_mat  = pca_mat,
  meta_data = as.data.frame(colData(spe)),
  vars_use  = c("fm_id"),
  do_pca    = FALSE
)

rownames(pca_mat) <- colnames(spe)
reducedDim(spe, "PCA_clustering") <- pca_mat
```

## UMAP after Harmony (for visualization only)

```{r umap-post-harmony}
set.seed(221228)

mat_red <- calculateUMAP(
  t(pca_mat),
  n_neighbors = 30,
  BPPARAM = MulticoreParam(RNGseed = 221228)
)

rownames(mat_red) <- colnames(spe)
reducedDim(spe, "UMAP") <- mat_red
```


## Phenograph clustering on PCA space after harmony

```{r phenograph-clustering}
set.seed(221228)

Rphenograph_out <- Rphenoannoy(pca_mat, k = 100)
clusters <- factor(membership(Rphenograph_out[[2]]))

saveRDS(clusters, file.path(params$output, "pg_clusters_AD.rds"))

spe$pg_clusters <- clusters
```

#

## UMAP: Phenograph clusters
Let's visualize phenograph clusters and tissue types in UMAP space. Here `dittoDimPlot` can be used directly on the SPE object.

```{r UMAP-vis-clusters}
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
col_clusters = setNames(sample(col_vector_433, length(unique(spe$pg_clusters))), unique(spe$pg_clusters))

p1 <- dittoDimPlot(spe, var = "pg_clusters", 
             reduction.use = "UMAP", size = 1,
             do.label = TRUE,
             legend.show=F) +
  scale_color_manual(values = col_clusters) + 
  theme(legend.title = element_blank()) +
  ggtitle("PG clusters on UMAP")

# Save ggplot
save_plot(p1, "UMAP-vis-clusters.png", width = 8, height = 6)
p1
```

## Bubble plot: Mean marker expression per cluster
Let's plot the mean marker expressions per cluster in a bubble plot.

```{r vis-data-bubbleplot, fig.height=14, fig.width=16}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(viridis)
library(cowplot)

stopifnot(exists("mat"), exists("spe"))

cluster_vec <- spe$pg_clusters
tissue_vec  <- colData(spe)$tissue

message("Number of cells: ", ncol(mat))
message("Number of markers: ", nrow(mat))

mat_df <- as.data.frame(t(mat))

exp_mat_ <- aggregate(
  mat_df,
  by = list(cluster = cluster_vec),
  FUN = function(x) {
    if (all(x == 0)) return(0)
    median(log1p(x[x > 0]))
  }
)

cluster_names <- exp_mat_$cluster
exp_mat <- exp_mat_[, -1, drop = FALSE]
rownames(exp_mat) <- cluster_names
exp_mat <- t(as.matrix(exp_mat))

rmin <- apply(exp_mat, 1, min, na.rm = TRUE)
rmax <- apply(exp_mat, 1, max, na.rm = TRUE)
den  <- rmax - rmin

exp_mat <- (exp_mat - rmin) / den
exp_mat[is.nan(exp_mat)] <- 0
exp_mat[is.infinite(exp_mat)] <- 0

clusters <- colnames(exp_mat)
message("Dimensions of exp_mat: ", paste(dim(exp_mat), collapse = " x "))

percent_mat <- sapply(seq_len(nrow(mat)), function(i) {
  row_vals <- mat[i, ]
  pos <- row_vals[row_vals > 0]

  if (length(pos) == 0) {
    out <- rep(0, length(clusters))
    names(out) <- clusters
    return(out)
  }

  thresh <- quantile(pos, 0.2, na.rm = TRUE)
  row_bin <- ifelse(row_vals > thresh, 1, 0)

  tapply(row_bin, factor(cluster_vec, levels = clusters), mean)
})

percent_mat <- t(percent_mat)
percent_mat <- percent_mat[, clusters, drop = FALSE]
percent_mat[is.na(percent_mat)] <- 0
percent_mat[percent_mat < 0.1] <- 0.1

message("Dimensions percent_mat: ", paste(dim(percent_mat), collapse = " x "))

ad_count <- table(cluster_vec[tissue_vec == "AD"])
ad_count <- ad_count[clusters]; ad_count[is.na(ad_count)] <- 0

mean_morph <- aggregate(
  morph_mat_clipped,
  list(cluster = cluster_vec),
  function(x) mean(x, na.rm = TRUE)
)

area_mean <- mean_morph[match(clusters, mean_morph$cluster), "area"]
solidity_mean <- mean_morph[match(clusters, mean_morph$cluster), "solidity"]

col_mean_area <- colorRamp2(
  c(min(area_mean), median(area_mean), max(area_mean)),
  c("blue", "white", "red")
)

col_mean_solidity <- colorRamp2(
  c(min(solidity_mean), median(solidity_mean), max(solidity_mean)),
  c("blue", "white", "red")
)

if (!exists("col_clusters")) {
  col_clusters <- setNames(rainbow(length(clusters)), clusters)
} else {
  missing_clusters <- setdiff(clusters, names(col_clusters))
  if (length(missing_clusters) > 0) {
    col_clusters[missing_clusters] <- rainbow(length(missing_clusters))
  }
}

# Barplot axis: show only the max tick (no 0 tick)
axis_maxtick <- function(v) {
  vmax <- max(as.numeric(v), na.rm = TRUE)
  list(
    at = vmax,
    labels = as.character(vmax),
    gp = gpar(fontsize = 12)
  )
}

column_ha <- HeatmapAnnotation(
  pg_clusters = clusters,
  roundness = solidity_mean,

  ncells_AD = anno_barplot(
    as.numeric(ad_count),
    height = unit(0.7, "cm"),
    gp = gpar(fill = "darkorchid3"),
    axis_param = axis_maxtick(ad_count)
  ),

  col = list(
    pg_clusters = col_clusters,
    roundness = col_mean_solidity,
    area = col_mean_area
  ),

  show_legend = TRUE
)

col_fun <- circlize::colorRamp2(
  c(0, 0.5, 1),
  viridis(100)[c(1, 50, 100)]
)

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid.rect(x, y, w, h, gp = gpar(col = NA, fill = NA))
  grid.circle(
    x = x,
    y = y,
    r = percent_mat[i, j] / 2 * min(unit.c(w, h)),
    gp = gpar(fill = col_fun(exp_mat[i, j]), col = NA)
  )
}

lgd_list <- list(
  Legend(
    labels = c(0, 0.25, 0.5, 0.75, 1),
    title = "percentage expressed",
    graphics = lapply(c(0, 0.25, 0.5, 0.75, 1), function(r) {
      function(x, y, w, h) grid.circle(
        x = x, y = y,
        r = r * unit(2, "mm"),
        gp = gpar(fill = "black")
      )
    })
  )
)

hp <- Heatmap(
  exp_mat,
  heatmap_legend_param = list(title = "expression"),
  column_title = "Bubble plot - normalized mean expression and percent expressed",
  top_annotation = column_ha,
  col = col_fun,
  rect_gp = gpar(type = "none"),
  clustering_distance_columns = "pearson",
  clustering_method_columns = "average",
  row_labels = sapply(strsplit(rownames(exp_mat), "_"), "[[", 1),
  cell_fun = cell_fun,
  show_row_dend = FALSE,
  column_dend_height = unit(1.5, "cm"),
  row_names_gp = gpar(fontsize = 15),
  column_names_gp = gpar(fontsize = 15),
  column_names_rot = 45,
  border = "black"
)

hp_drawn <- draw(hp, annotation_legend_list = lgd_list)
save_plot(hp_drawn, "vis-data-bubbleplot.pdf", width = 16, height = 14)

hp_drawn
```





We can see that PT samples are primarily comprised of tumor cells (GD2+) and bone marrow samples include a large proprotion of granulocytes (CD15+).

## Writing cluster proportions for manual annotation (Excel)

```{r cluster-annotation-excel}
# Ensure cluster is character
spe$pg_clusters <- as.character(spe$pg_clusters)

# -----------------------------
# 1. Absolute counts
# -----------------------------
count_table <- as.data.frame(table(
  cluster = spe$pg_clusters,
  tissue  = spe$tissue
))

count_wide <- count_table %>%
  pivot_wider(
    names_from = tissue,
    values_from = Freq,
    values_fill = 0
  )

# Ensure all tissues exist
for(t in c("AD")){
  if(!t %in% colnames(count_wide)){
    count_wide[[t]] <- 0
  }
}

count_wide <- count_wide %>%
  select(cluster, AD)

# -----------------------------
# 2. Percentages WITHIN cluster
# -----------------------------
percent_within_cluster <- count_wide %>%
  rowwise() %>%
  mutate(
    total = AD,
    AD_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * AD / total), "0.0%")
  ) %>%
  ungroup() %>%
  select(-total)

# -----------------------------
# 3. Add annotation columns
# -----------------------------
annotation_template <- percent_within_cluster %>%
  mutate(
    celltype_AD = "",
    metacluster_AD = "",
    literature = ""
  ) %>%
  arrange(as.numeric(cluster))

# -----------------------------
# 4. Write Excel
# -----------------------------
write_xlsx(
  annotation_template,
  file.path(params$output, "cluster_annotation_template_AD.xlsx")
)
```

```{r stop-here}
knitr::opts_chunk$set(eval = FALSE)
```

## Annotation and merging of clusters

Next, we read assigned celltype and metacluster-labels for clusters from the metadata file provided on Zenodo [(Supplementary_Tables/20240808_Suppl_Table3.xlsx)](https://doi.org/10.5281/zenodo.13220634 ). These have been assigned based on thorough examination of images and co-expressions and using literature.
Let's start with celltype assignments...

```{r annotations-celltypes}
file_path <- '/home/rstudio/data/output/harmony_sample_only/completed_cluster_annotation_template_AD.xlsx'
annotation <- read_excel(file_path, sheet = "annotation_v1")

cluster_names <- setNames(annotation$celltype, annotation$cluster)
celltypes <- recode(spe$pg_clusters, !!!cluster_names)
spe$celltype <-  as.vector(celltypes)

metacluster_names <- setNames(annotation$metacluster, annotation$cluster)
metacelltypes <- recode(spe$pg_clusters, !!!metacluster_names)
spe$metacluster <-  as.vector(metacelltypes)

```


## Assign colors 
Finally, we assign colors to all celltypes and metaclusters.

```{r colors-celltypes}
celltype_AD_colors <- c(
  "Ki67+ proliferating cells" = rgb(149,144,89, maxColorValue = 255),      # Ki67hi TC (proliferating)
  "SOX10+ S100B+ SCPs" = rgb(156,76,24, maxColorValue = 255),              # schwann cell (SCP)
  "CXCR4+ cells" = rgb(210,205,126, maxColorValue = 255),                  # CXCR4hi TC
  "other" = "goldenrod1",                         # AD (other)
  "CD15+ CHGA+ Chromaffin cells" = rgb(65,208,57, maxColorValue = 255),   # CHGAhi TC (chromaffin)
  "GATA3+ Early progenitors" = rgb(198,255,72, maxColorValue = 255),       # HSPC (early progenitors)
  "CD34- stromal  cells" = rgb(212,89,67, maxColorValue = 255),            # fibroblast/endothel (stromal CD34-)
  "CD15+ CHGAlo" = rgb(230,245,201, maxColorValue = 255),                  # marker-lo TC (low CHGA)
  "ELAVL4+ GD2+ sympathoblasts" = rgb(27,158,119, maxColorValue = 255),    # GD2hi TC (sympathoblasts)
  "CD34+ stromal  cells" = rgb(212,89,67, maxColorValue = 255),            # fibroblast/endothel (stromal CD34+)
  "CD45+ cells" = rgb(212,143,75, maxColorValue = 255),                    # HPC (immune)
  "CD44+ cells" = rgb(27,158,119, maxColorValue = 255)                     # GD2hi TC reused (CD44+)
)

metadata(spe)$color_vectors$col_celltype <- celltype_AD_colors
```

```{r colors-metaclusters}
col_metacluster <- c("neural crest lineage"="palegreen4",
                  "immune"="indianred",
                  "other"="goldenrod1",
                  "stromal"="tan4",
                  "proliferating"="grey",
                  "CXCR4+ cells"="blue")
                
metadata(spe)$color_vectors$col_metacluster <- col_metacluster
```

## Visualizations
For visualization purpose, we will give the celltypes a meaningful order. This order can be downloaded from a csv file from Zenodo [(R_intermediary/celltype_order.csv)](https://doi.org/10.5281/zenodo.13220634 )

```{r order-celltypes}
clusternames_ordered <- read.csv(file.path(params$input, "celltype_order.csv"), sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)
spe$celltype_ordered <- recode(spe$celltype, !!!codes)

metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
names(metadata(spe)$color_vectors$col_celltype_ordered) <- recode(names(metadata(spe)$color_vectors$col_celltype), !!!codes)
```

As a last step, we generate UMAP and bubble plots for the final celltypes/metaclusters.

```{r umap-final-celltypes, fig.width=10}
p1 <- dittoDimPlot(spe, var = "celltype_ordered", 
             reduction.use = "UMAP", size = 0.3, 
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(legend.title = element_blank()) +
  ggtitle("Celltypes on UMAP")

p2 <- dittoDimPlot(spe, var = "metacluster", 
             reduction.use = "UMAP", size = 0.3, labels.size=4,
             do.label = T,
             order = "randomize",
             legend.show=F) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_metacluster) +
  theme(legend.title = element_blank()) +
  ggtitle("Metaclusters on UMAP")

# Save both
save_plot(p1, "umap-final-celltypes_celltypes.png", width = 10, height = 8)
save_plot(p2, "umap-final-celltypes_metaclusters.png", width = 10, height = 8)

p1
```

```{r vis-metacluster}
p2
```


```{r bubbleplot-final-celltypes, fig.height=10, fig.width=12}
# -----------------------------
# Expression matrix
# -----------------------------

mat_df <- as.data.frame(t(mat))

exp_mat_ <- aggregate(
  mat_df,
  by = list(cluster = spe$celltype),
  FUN = function(x) {
    if (all(x == 0)) return(0)
    median(log1p(x[x > 0]))
  }
)

cluster_names <- exp_mat_$cluster
exp_mat <- exp_mat_[, -1, drop = FALSE]
rownames(exp_mat) <- cluster_names
exp_mat <- t(as.matrix(exp_mat))

# Row-wise scaling (per marker)
rmin <- apply(exp_mat, 1, min, na.rm = TRUE)
rmax <- apply(exp_mat, 1, max, na.rm = TRUE)
den  <- rmax - rmin

exp_mat <- (exp_mat - rmin) / den
exp_mat[is.nan(exp_mat)] <- 0
exp_mat[is.infinite(exp_mat)] <- 0


# -----------------------------
# Percentage expressed
# -----------------------------

clusters <- colnames(exp_mat)

percent_mat <- sapply(seq_len(nrow(mat)), function(i) {
  row_vals <- mat[i, ]
  pos <- row_vals[row_vals > 0]

  if (length(pos) == 0) {
    out <- rep(0, length(clusters))
    names(out) <- clusters
    return(out)
  }

  thresh <- quantile(pos, 0.2, na.rm = TRUE)
  row_bin <- ifelse(row_vals > thresh, 1, 0)

  tapply(row_bin, factor(spe$celltype, levels = clusters), mean)
})

percent_mat <- t(percent_mat)
percent_mat <- percent_mat[, clusters, drop = FALSE]
percent_mat[is.na(percent_mat)] <- 0

# Optional visualization floor
percent_mat[percent_mat < 0.1] <- 0.1

# Annotation bars
ad_count <- table(spe[, spe$tissue == "AD"]$celltype)
ad_count[setdiff(colnames(exp_mat), names(ad_count))] <- 0
ad_count <- ad_count[colnames(exp_mat)]


mean_metaclust <- aggregate(spe$metacluster, list(spe$celltype), max)

# Barplot axis: show only the max tick (no 0 tick)
axis_maxtick <- function(v) {
  vmax <- max(as.numeric(v), na.rm = TRUE)
  list(
    at = vmax,
    labels = as.character(vmax),
    gp = gpar(fontsize = 12)
  )
}

column_ha = HeatmapAnnotation(
  metacluster = mean_metaclust$x,
  
  ncells_AD = anno_barplot(
    unname(ad_count),
    add_numbers = FALSE,
    height = unit(0.7, "cm"),
    gp = gpar(fill = "darkorchid3"),
    axis_param = axis_maxtick(ad_count)
  ),
  
  col = list(
    metacluster = metadata(spe)$color_vectors$col_metacluster,
    roundness = col_mean_solidity
  ),
  
  show_legend = TRUE
)

#Function to set dot size based on percentage expressed
cell_fun = function(j, i, x, y, w, h, fill){
          grid.rect(x = x, y = y, width = w, height = h, 
                    gp = gpar(col = NA, fill = NA))
          grid.circle(x=x,y=y,r= percent_mat[i, j]/2 * min(unit.c(w, h)),
                      gp = gpar(fill = col_fun(exp_mat[i, j]), col = NA))}

#Function to set dot color based on mean expression
col_fun = circlize::colorRamp2(c(min(exp_mat),(max(exp_mat)-min(exp_mat))/2, max(exp_mat)), viridis(100)[c(1,50,100)])

#Function to set legend size according to percentage expressed
layer_fun = function(j, i, x, y, w, h, fill){
          grid.rect(x = x, y = y, width = w, height = h, 
                    gp = gpar(col = NA, fill = NA))
          grid.circle(x=x,y=y,r= pindex(percent_mat, i, j)/1 * unit(1.5, "mm"),
                      gp = gpar(fill = col_fun(pindex(exp_mat, i, j)), col = NA))}

#Generate legend for percentage expressed
lgd_list = list(
    Legend( labels = c(0,0.25,0.5,0.75,1), title = "percentage expressed",
            graphics = list(
              function(x, y, w, h) grid.circle(x = x, y = y, r = 0 * unit(2, "mm"),
                                               gp = gpar(fill = "black")),
              function(x, y, w, h) grid.circle(x = x, y = y, r = 0.25 * unit(2, "mm"),
                                               gp = gpar(fill = "black")),
              function(x, y, w, h) grid.circle(x = x, y = y, r = 0.5 * unit(2, "mm"),
                                               gp = gpar(fill = "black")),
              function(x, y, w, h) grid.circle(x = x, y = y, r = 0.75 * unit(2, "mm"),
                                               gp = gpar(fill = "black")),
              function(x, y, w, h) grid.circle(x = x, y = y, r = 1 * unit(2, "mm"),
                                               gp = gpar(fill = "black")))
            ))

#Heatmap
hp <- Heatmap(exp_mat,
        heatmap_legend_param=list(title="expression"),
        column_title = "Bubble plot - normalized mean expression and %expressed", 
        top_annotation = column_ha, 
        col=col_fun,
        rect_gp = gpar(type = "none"),
        clustering_distance_columns="pearson", 
        clustering_method_columns="average", 
        layer_fun = layer_fun,
        row_labels=sapply(strsplit(rownames(exp_mat), "_"), "[[", 1),
        cell_fun = cell_fun,
        show_row_dend = F,
        column_dend_height = unit(1.5, "cm"),
        row_names_gp = gpar(fontsize = 10),
        column_names_gp = gpar(fontsize = 10),
        column_names_rot = 45,
        border = "black")

# Save and draw
hp_drawn <- draw(hp, annotation_legend_list = lgd_list)
save_plot(hp_drawn, "bubbleplot-final-celltypes.pdf", width = 12, height = 10)
hp_drawn
```

## Save object

```{r save-objects}

saveRDS(spe, file.path(params$output, "spe_final.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>