---
title: "Cell phenotyping"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/04_phenotyping")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# create phenograph heatmap folder
outdir <- paste0(plot_dir,"/pg_heatmap")
dir.create(outdir, showWarnings = FALSE)


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Cell phenotyping {#phenotyping}
In this Rmd file, we will perform 
1. dimensionality reduction
2. phenograph clustering
3. annotate and merge clusters & metaclusters externally 
4. retrieve tumor cells from GD2-MPO double positive cells (caused by spillover) 
5. re-classify granulocyte metaclusters into maturity states based on morphology
6. define color codes for all celltypes and metaclusters

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ComplexHeatmap)
library(RColorBrewer)
library(viridis)
library(scater)
library(BiocParallel)
library(Rphenoannoy)
library(tidyr)
library(tidyHeatmap)
library(bruceR)
library(dittoSeq)
library(dplyr)
library(mclust)
library(SingleR)
library(circlize)
library(readxl)
library(ggplot2)
library(patchwork)
library(harmony)
library(batchelor)
library(cluster)
```
</details>

## Read data
We will first read the data that passed the [second quality control step](#qc-2) and save the count data as a matrix. Low-quality markers (low SNR in a lot of samples) as well as all IF and nuclear IMC markers will be excluded for clustering. Moreover, mean intensity will be used to represent marker expression, to average out the signals arising from lateral spillover or artifacts. As explained in [the previous chapter](#qc2-save), we are using counts measured with respect to dilated masks in downstream analysis do capture membrane signal. 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_03_QC_2.rds"))

#computing matrix
intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- colnames(spe)

#use mean intensity
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]

#exclude low-quality markers and functional marker cl.PARP
intensity_mat_hq <- intensity_mat[!grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(intensity_mat)),]
intensity_mat_lq <- intensity_mat[grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(intensity_mat)),]

non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)

intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]
intensity_mat_lq_wo_zeros <- intensity_mat_lq[,non_zeros]

#Apply changes to SPE object
spe <- spe[,non_zeros]
rowData(spe)$use_channel <- !grepl("mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))

##Matrix with morph. features
morph_mat_clipped <- (cbind("area"=spe$area_clipped, "solidity"=spe$solidity_clipped, "aspect_ratio"=spe$aspect_ratio_clipped))

tissue <- spe$tissue
cell_id <- colnames(spe)

mat <- intensity_mat_hq_wo_zeros

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_sample = setNames(sample(col_vector_433, length(unique(spe$sample))), unique(spe$sample))
```
We are excluding `r length(non_zeros[non_zeros==FALSE])` more cells here, as they are negative for all markers used in downstream analysis. 
To get a first grasp of the single-cell data and potential clusters, we will plot it in a single-cell heatmap.
spe_filtered <- spe_sub[, colData(spe_sub)$DNAIntensity >= low_dna_threshold]


## UMAP (for visualization only)
```{r umap}
set.seed(221228)

Sys.setenv(OMP_NUM_THREADS = 1)
Sys.setenv(MKL_NUM_THREADS = 1)

BPPARAM <- SnowParam(workers = 30, type = "SOCK", RNGseed = 221228)

mat_red <- calculateUMAP(
  mat,
  n_neighbors = 30,
  BPPARAM = BPPARAM
)

rownames(mat_red) <- colnames(spe)
reducedDim(spe, "UMAP") <- mat_red

p <- dittoDimPlot(spe, var = "fm_id", 
             reduction.use = "UMAP", size = 1,
             do.label = TRUE,
             legend.show=F) +
  theme(legend.title = element_blank()) +
  ggtitle("Sample ID on UMAP")


save_plot(p, "UMAP-vis-sampleID.pdf", width = 8, height = 6)

p

```



# UMAPs for each marker
```{r UMAP-all-markers, fig.width=24, fig.height=32}
excluded_markers <- "mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3"
max_cells_for_umap <- 20000

marker_vars <- rownames(spe)[
  !grepl(excluded_markers, rownames(spe))
]

set.seed(221228)
marker_umap_ncells <- min(max_cells_for_umap, ncol(spe))
marker_umap_idx <- sample(seq_len(ncol(spe)), marker_umap_ncells)
spe_umap_markers <- spe[, marker_umap_idx]

p_markers <- multi_dittoDimPlot(
  spe_umap_markers,
  vars = marker_vars,
  reduction.use = "UMAP",
  size = 0.5,
  ncol = 5,
  color.panel = viridis(100, direction = -1)
)

save_plot(p_markers, "UMAP-all-markers.pdf", width = 24, height = 32)
p_markers
```

## Bubble plot: Mean marker expression per cluster
Let's plot the mean marker expressions per cluster in a bubble plot.

```{r vis-data-bubbleplot}
# plot_cluster_bubble() and its colour helpers (color_seed_base,
# get_cluster_levels, generate_cluster_colors) live in code/bubbleplot.R so the
# AD and PT/BM pipelines share one implementation instead of each duplicating
# ~400 lines. The function auto-detects tissue: with a `tissue` column in
# colData(spe) it draws per-tissue (PT / BM) count barplots, otherwise a single
# count barplot (the AD, single-tissue case).
source(if (file.exists("code/bubbleplot.R")) "code/bubbleplot.R" else "../code/bubbleplot.R",
       local = TRUE)   # local=TRUE: define into the knit env so plot_cluster_bubble() sees spe/mat
```

```{r phenograph-umap-bubbleplot}
set.seed(221228)

k <- 30

# --- Reproducible phenograph clustering (see the reproducibility note) ---------
# Rphenoannoy uses Annoy (approximate NN, C++) + igraph::cluster_louvain, which
# are not bit-reproducible across igraph versions, compilers or CPUs (the same
# caveat as uwot/UMAP). We run a single k = 30 clustering, stored as `pg_clusters`.
# A frozen labelling is loaded from R_intermediary when present (keyed by unique
# cell id); otherwise it is computed here and saved for redistribution. Colour
# helpers come from code/bubbleplot.R (sourced above).
pg_name <- "pg_clusters_PTBM.rds"
inter_dirs <- c(Sys.getenv("MAPMET_INTERMEDIARY", ""),
                file.path(params$input,  "R_intermediary"),
                file.path(dirname(params$input), "R_intermediary"),
                file.path(params$output, "R_intermediary"))
inter_dirs <- inter_dirs[nzchar(inter_dirs)]
pg_file <- Find(file.exists, file.path(inter_dirs, pg_name))

if (!is.null(pg_file)) {
  pg <- readRDS(pg_file)
  if (!all(colnames(spe) %in% names(pg)))
    stop("Frozen clustering ", pg_file, " does not cover all cells; ",
         "regenerate it (code/generate_phenograph_intermediary.R).")
  lab <- as.integer(pg[colnames(spe)])
  spe$pg_clusters <- factor(lab, levels = sort(unique(lab)))
  message("Loaded frozen phenograph clustering: ", pg_file)
} else {
  Rphenograph_out <- Rphenoannoy(t(mat), k = k)
  spe$pg_clusters <- factor(membership(Rphenograph_out[[2]]))
  pg_out <- file.path(params$output, "R_intermediary")
  dir.create(pg_out, showWarnings = FALSE, recursive = TRUE)
  saveRDS(setNames(as.character(spe$pg_clusters), colnames(spe)),
          file.path(pg_out, pg_name))
  message("Saved frozen phenograph clustering: ", file.path(pg_out, pg_name))
}

# Cluster colours (get_cluster_levels / generate_cluster_colors / color_seed_base
# are defined in code/bubbleplot.R)
if (is.null(metadata(spe)$color_vectors)) {
  metadata(spe)$color_vectors <- list()
}
cluster_levels <- get_cluster_levels(spe$pg_clusters)
color_key <- "col_pg_clusters"
if (is.null(metadata(spe)$color_vectors[[color_key]]) ||
    !all(cluster_levels %in% names(metadata(spe)$color_vectors[[color_key]]))) {
  metadata(spe)$color_vectors[[color_key]] <- generate_cluster_colors(
    cluster_levels = cluster_levels
  )
}
col_clusters <- metadata(spe)$color_vectors[[color_key]][cluster_levels]

# UMAP coloured by pg_clusters
p <- dittoDimPlot(
  spe, var = "pg_clusters", reduction.use = "UMAP",
  size = 0.5, do.label = TRUE, legend.show = FALSE,
  color.panel = col_clusters
) +
  theme(legend.title = element_blank()) +
  ggtitle("pg_clusters (k = 30) on UMAP")

label_layer <- which(
  sapply(p$layers, function(x) class(x$geom)[1]) %in% c("GeomLabelRepel", "GeomLabel")
)
label_data <- layer_data(p, label_layer)
label_fills <- col_clusters[as.character(label_data$label)]
p$layers[[label_layer]]$aes_params$fill <- label_fills
p$layers[[label_layer]]$aes_params$colour <- "black"

save_plot(p, "umap_pg_clusters.pdf", width = 8, height = 6)
print(p)

# Bubble plot for pg_clusters
plot_cluster_bubble(
  cluster_col = "pg_clusters",
  save_name = "bubbleplot_pg_clusters.pdf",
  ncells_break_at = 100000
)
```


## Subclustering of mixed clusters using GMM

```{r gmm-subclustering}
library(mclust)

set.seed(221228)

# NOTE: subcluster_map / marker_map below are keyed to the pg_clusters
# (k=30) cluster IDs; re-check them against the frozen k=30 clustering.
spe$pg_clusters_beforesubclustering <- spe$pg_clusters


spe$pg_clusters <- spe$pg_clusters_beforesubclustering

# Ensure pg_clusters is character for easy reassignment
spe$pg_clusters <- as.character(spe$pg_clusters)

# Mapping: original cluster -> new subcluster label for the second component
subcluster_map <- list(
  "5"  = "35",
  "12" = "36",
  "17" = "37"
)

# define marker sets per cluster
marker_map <- list(
  "5"  = c("CD3_"),
  "12" = c("CD4_"),
  "17" = c("CD3_", "CD10")
)

for (cl in names(subcluster_map)) {
  new_cl <- subcluster_map[[cl]]
  
  idx <- which(spe$pg_clusters == cl)
  if (length(idx) < 4) next
  
  sub_mat <- t(mat[, idx])  # cells x markers
  
  # --- select cluster-specific markers ---
  if (cl %in% names(marker_map)) {
    markers <- marker_map[[cl]]
    keep_cols <- grep(paste0("^(", paste(markers, collapse="|"), ")"), colnames(sub_mat))
    sub_mat <- sub_mat[, keep_cols, drop = FALSE]
  }
  
  print(cl)
  print(head(sub_mat))
  
  # --- clean data ---
  sub_mat <- sub_mat[complete.cases(sub_mat), , drop = FALSE]
  sub_mat <- sub_mat[, apply(sub_mat, 2, var) > 0, drop = FALSE]
  print(cl)
  print(head(sub_mat))
  
  #if (nrow(sub_mat) < 4 || ncol(sub_mat) < 2) next
  
  sub_mat <- scale(sub_mat)
  
  # --- GMM ---
  gmm_fit <- tryCatch(
    Mclust(sub_mat, G = 2, verbose = FALSE),
    error = function(e) NULL
  )
  
  if (is.null(gmm_fit)) next
  
  comp2 <- which(gmm_fit$classification == 2)
  spe$pg_clusters[idx[comp2]] <- new_cl
}
```

## Single-cell heatmaps for subclustered clusters

```{r subcluster-heatmaps}
library(dittoSeq)
library(ggplot2)
library(viridis)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]

# Define the pairs: original cluster and its new sibling
subcluster_pairs <- list(
  list(orig = "5",  new = "35"),
  list(orig = "12", new = "36"),
  list(orig = "17", new = "37")
)


set.seed(221228)
for (pair in subcluster_pairs) {
  orig_cl <- pair$orig
  new_cl  <- pair$new

  # Cells belonging to either subcluster
  idx <- which(spe$pg_clusters %in% c(orig_cl, new_cl))

  if (length(idx) == 0) next

  # Sample up to 8000 cells
  n_cells   <- min(8000, length(idx))
  rand_idx  <- sample(idx, n_cells)

  spe_cl <- spe[, rand_idx]

  # Order cells by subcluster so the two groups appear side-by-side
  cell_order <- order(spe_cl$pg_clusters)

  p1 <- dittoHeatmap(
    spe_cl[, cell_order],
    genes          = genes_use,
    cluster_cols   = FALSE,          # preserve subcluster ordering
    scaled.to.max  = TRUE,
    heatmap.colors.max.scaled = viridis(100),
    annot.by       = "pg_clusters"   # colour bar showing subclusters
  )

  save_plot(
    p1,
    paste0("heatmap_subcluster_", orig_cl, "_", new_cl, ".pdf"),
    width  = 14,
    height = 6
  )
}
```

```{r pg-clusters-final-bubbleplot}

plot_cluster_bubble(
      cluster_col = "pg_clusters",
      ncells_break_at = 100000,
      save_name = "bubbleplot_pg_clusters.pdf")

```


## Define final clustering and plot clustered heatmaps for each cluster

```{r}
library(dittoSeq)
library(ggplot2)

# get cluster IDs
clusters <- unique(colData(spe)$pg_clusters)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]


for (cl in clusters) {
  
  # subset cells for this cluster
  spe_cl <- spe[, colData(spe)$pg_clusters == cl]
  
  # skip empty clusters (just in case)
  if (ncol(spe_cl) == 0) next
  
  # sample up to 8000 cells
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  # make heatmap
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  # save plot
  save_plot(
    p1,
    paste0("heatmap_after_subclustering_", cl, ".pdf"),
    width = 14,
    height = 6
  )
}

```


## Write cluster proportions for manual annotation (Excel)

The chunk below regenerates the cluster-annotation template that is filled in manually and published on Zenodo as a Supplementary Table (read back in the next section). The whole block is disabled with `eval=FALSE` so the pipeline does not depend on `writexl`; enable it only if you need to regenerate the template.

```{r cluster-annotation-excel, eval=FALSE}
library(writexl)

# Ensure cluster is character
spe$pg_clusters <- as.character(spe$pg_clusters)

# -----------------------------
# 1. Absolute counts
# -----------------------------
count_table <- as.data.frame(table(
  cluster = spe$pg_clusters,
  tissue  = spe$tissue
))

count_wide <- count_table %>%
  pivot_wider(
    names_from = tissue,
    values_from = Freq,
    values_fill = 0
  )

# Ensure all tissues exist
for(t in c("BM","PT")){
  if(!t %in% colnames(count_wide)){
    count_wide[[t]] <- 0
  }
}

count_wide <- count_wide %>%
  select(cluster, BM, PT)

# -----------------------------
# 2. Percentages WITHIN cluster
# -----------------------------
percent_within_cluster <- count_wide %>%
  rowwise() %>%
  mutate(
    total = BM + PT,
    BM_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * BM / total), "0.0%"),
    PT_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * PT / total), "0.0%")
  ) %>%
  ungroup() %>%
  select(-total)

# -----------------------------
# 3. Add annotation columns
# -----------------------------
annotation_template <- percent_within_cluster %>%
  mutate(
    celltype = "",
    metacluster = "",
    comments = ""
  ) %>%
  arrange(as.numeric(cluster))

# -----------------------------
# 4. Write Excel
# -----------------------------
write_xlsx(
  annotation_template,
  file.path(params$output, "cluster_annotation_template.xlsx")
)
```


## Annotation and merging of clusters

Next, we read assigned celltype and metacluster-labels for clusters from the Supplementary Table provided on Zenodo [(Supplementary_Tables/20260526_Suppl_Table3.xlsx)](https://doi.org/10.5281/zenodo.13220634). These have been assigned based on thorough examination of images and co-expressions and using literature. The sheet must contain the columns `cluster`, `celltype` and `metacluster`.
Let's start with celltype assignments...

```{r annotations-celltypes}
# Supplementary_Tables may live inside the cohort dir (container layout) or
# beside it (the Zenodo bundle keeps it next to MapMetIP_ProcessedDataset).
.suppl_cands <- file.path(c(params$input, dirname(params$input)), "Supplementary_Tables", "20260526_Suppl_Table3.xlsx")
file_path <- .suppl_cands[file.exists(.suppl_cands)][1]
if (is.na(file_path)) stop("Supplementary table not found in:\n  ", paste(.suppl_cands, collapse = "\n  "))
annotation <- read_excel(file_path, sheet = "Sheet1")

cluster_names <- setNames(annotation$celltype, annotation$cluster)
celltypes <- recode(spe$pg_clusters, !!!cluster_names)
spe$celltype <-  as.vector(celltypes)

metacluster_names <- setNames(annotation$metacluster, annotation$cluster)
metacelltypes <- recode(spe$pg_clusters, !!!metacluster_names)
spe$metacluster <-  as.vector(metacelltypes)

```

## Assign colors 
Finally, we assign colors to all celltypes and metaclusters.

```{r colors-celltypes raw_k30}
# Cell-type and metacluster colors are defined centrally in 00_MapMet_color_code.R
for (p in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(p)) { source(p); break }

col_celltype <- COL_CELLTYPE

metadata(spe)$color_vectors$col_celltype <- col_celltype
```

```{r colors-metaclusters}
col_metacluster <- COL_METACLUSTER
metadata(spe)$color_vectors$col_metacluster <- col_metacluster
```

## Visualizations
For visualization purpose, we will give the celltypes a meaningful order. This order can be downloaded from a csv file from Zenodo [(celltype_order.csv)](https://doi.org/10.5281/zenodo.13220634 )

```{r order-celltypes}
clusternames_ordered <- read.csv(file.path(params$input, "celltype_order.csv"), sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)
spe$celltype_ordered <- recode(spe$celltype, !!!codes)

metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
names(metadata(spe)$color_vectors$col_celltype_ordered) <- recode(names(metadata(spe)$color_vectors$col_celltype), !!!codes)
```

As a last step, we generate UMAP and bubble plots for the final celltypes/metaclusters.


```{r bubbleplot-celltypes, fig.width=16, fig.height=14}

plot_cluster_bubble(
  cluster_col = "celltype",
  save_name = "bubbleplot-celltypes.pdf",
  annotation_level = "metacluster",
  ncells_break_at = 65000,
  plot_width      = 13,
  plot_height     = 11.5
)

```

```{r umap-final-celltypes, fig.width=10}
# aspect.ratio = 1 fixes the UMAP panel to a square, independent of the legend
# width, so the celltype and metacluster UMAPs render identically. The celltype
# plot is only wider overall because its legend has more entries.
p1 <- dittoDimPlot(spe, var = "celltype_ordered",
             reduction.use = "UMAP", size = 0.3,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Celltypes on UMAP")

save_plot(p1, "umap-celltypes.pdf", width = 16, height = 8)

p1
```

We also plot the same celltype UMAP but colour only the primary tumor (PT)
cells, showing all other tissues as light-grey background dots.

```{r umap-celltype-PT-only, fig.width=10}
# cells.use restricts colouring to PT cells; show.others draws the remaining
# cells as light-grey background dots. Same square-panel format as above.
p1_pt <- dittoDimPlot(spe, var = "celltype_ordered",
             reduction.use = "UMAP", size = 0.3,
             cells.use = spe$tissue == "PT",
             show.others = TRUE,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Celltypes on UMAP (PT only)")

save_plot(p1_pt, "umap-celltype-PT-only.pdf", width = 16, height = 8)

p1_pt
```

```{r vis-metacluster}
p2 <- dittoDimPlot(spe, var = "metacluster",
             reduction.use = "UMAP", size = 0.3, labels.size=4,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_metacluster) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Metaclusters on UMAP")

save_plot(p2, "umap-metaclusters.pdf", width = 13, height = 8)

p2
```


Let's plot heatmap of clustered cells as a validation:


```{r}
library(dittoSeq)
library(ggplot2)
library(viridis)

# Ensure output directory exists
outdir <- file.path(plot_dir)
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

# Get cluster IDs
clusters <- unique(colData(spe)$celltype)

# Select genes to use (exclude certain markers)
genes_use <- rownames(spe)[
  !grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", 
         rownames(spe))
]

for (cl in clusters) {
  
  # --- 1. Make cluster name safe for filenames ---
  safe_name <- cl
  safe_name <- gsub("\\+", "plus", safe_name)
  safe_name <- gsub("-", "minus", safe_name)
  safe_name <- gsub("[^A-Za-z0-9_]", "_", safe_name)  # replaces /, spaces, special chars
  
  # --- 2. Subset cells for this cluster ---
  spe_cl <- spe[, colData(spe)$celltype == cl]
  
  # Skip empty clusters
  if (ncol(spe_cl) == 0) next
  
  # --- 3. Sample cells ---
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  # --- 4. Create heatmap ---
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  # --- 5. Save heatmap safely ---
  filename <- paste0("heatmap_celltype_", safe_name, ".pdf")
  full_path <- file.path(outdir, filename)
  
  # If save_plot() is your custom function
  save_plot(p1, filename = filename, width = 14, height = 6)
  
  # Or using ggsave directly:
  # ggsave(full_path, plot = p1, width = 14, height = 6, dpi = 300)
}
```
## Save object

```{r save-objects}

saveRDS(spe, file.path(params$output, "spe_04_phenotyping.rds"))

```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
