---
title: "Cell phenotyping"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# Plots directory
plot_dir <- file.path(params$output, "plots/04_phenotyping")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# Helper: save a plot (ggplot, ComplexHeatmap, or base R)
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300, subdir = NULL) {
  out_dir <- if (is.null(subdir)) plot_dir else file.path(plot_dir, subdir)
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  full_path <- file.path(out_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # base R / already-rendered plots
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  message("Saved: ", full_path)
}

# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Cell phenotyping {#phenotyping}
In this Rmd file, we will
1. run UMAP for visualization
2. cluster cells with phenograph
3. subcluster mixed clusters with a Gaussian mixture model
4. annotate and merge clusters into celltypes and metaclusters using the external annotation table
5. define color codes for all celltypes and metaclusters

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ComplexHeatmap)
library(RColorBrewer)
library(viridis)
library(scater)
library(BiocParallel)
library(Rphenoannoy)
library(tidyr)
library(tidyHeatmap)
library(bruceR)
library(dittoSeq)
library(dplyr)
library(mclust)
library(SingleR)
library(circlize)
library(readxl)
library(ggplot2)
library(patchwork)
library(harmony)
library(batchelor)
library(cluster)
```
</details>

## Read data
We will first read the data that passed the [second quality control step](#qc-2) and save the count data as a matrix. Low-quality markers (low SNR in a lot of samples) as well as all IF and nuclear IMC markers will be excluded for clustering. Moreover, mean intensity will be used to represent marker expression, to average out the signals arising from lateral spillover or artifacts. As explained in [the previous chapter](#qc2-save), we are using counts measured with respect to dilated masks in downstream analysis to capture membrane signal.

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_03_QC_2.rds"))

# Counts matrix; keep mean-intensity channels (drop the mean-80 variants)
intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- colnames(spe)
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]

# Keep only the high-quality clustering markers
intensity_mat_hq <- intensity_mat[!grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(intensity_mat)),]

# Drop cells negative for all clustering markers
non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)
intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]

spe <- spe[,non_zeros]
rowData(spe)$use_channel <- !grepl("mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))

mat <- intensity_mat_hq_wo_zeros
```
We are excluding `r length(non_zeros[non_zeros==FALSE])` more cells here, as they are negative for all markers used in downstream analysis.
To get a first look at the single-cell data, we compute a UMAP embedding (used for visualization only).


## UMAP (for visualization only)
```{r umap}
set.seed(221228)

Sys.setenv(OMP_NUM_THREADS = 1)
Sys.setenv(MKL_NUM_THREADS = 1)

BPPARAM <- SnowParam(workers = 30, type = "SOCK", RNGseed = 221228)

mat_red <- calculateUMAP(
  mat,
  n_neighbors = 30,
  BPPARAM = BPPARAM
)

rownames(mat_red) <- colnames(spe)
reducedDim(spe, "UMAP") <- mat_red

p <- dittoDimPlot(spe, var = "fm_id", 
             reduction.use = "UMAP", size = 1,
             do.label = TRUE,
             legend.show=F) +
  theme(legend.title = element_blank()) +
  ggtitle("Sample ID on UMAP")


save_plot(p, "UMAP-vis-sampleID.pdf", width = 8, height = 6)

p

```



# UMAPs for each marker
```{r UMAP-all-markers, fig.width=24, fig.height=32}
excluded_markers <- "mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3"
max_cells_for_umap <- 20000

marker_vars <- rownames(spe)[
  !grepl(excluded_markers, rownames(spe))
]

set.seed(221228)
marker_umap_ncells <- min(max_cells_for_umap, ncol(spe))
marker_umap_idx <- sample(seq_len(ncol(spe)), marker_umap_ncells)
spe_umap_markers <- spe[, marker_umap_idx]

p_markers <- multi_dittoDimPlot(
  spe_umap_markers,
  vars = marker_vars,
  reduction.use = "UMAP",
  size = 0.5,
  ncol = 5,
  color.panel = viridis(100, direction = -1)
)

save_plot(p_markers, "UMAP-all-markers.pdf", width = 24, height = 32)
p_markers
```

## Phenograph clustering and marker bubble plot
We source the shared bubble-plot helper, cluster the cells with phenograph (k = 30), and summarise each cluster in a bubble plot where dot colour encodes scaled marker expression and dot size the fraction of cells expressing the marker.

```{r vis-data-bubbleplot}
# Shared bubble-plot helper (also used by the AD pipeline).
source(if (file.exists("code/bubbleplot.R")) "code/bubbleplot.R" else "../code/bubbleplot.R",
       local = TRUE)   # local=TRUE so the helper sees spe/mat
```

```{r phenograph-umap-bubbleplot}
set.seed(221228)

k <- 30

# Phenograph (k = 30) is not bit-reproducible across machines, so load a frozen
# labelling from R_intermediary when present, else cluster here and save it.
pg_name <- "pg_clusters_PTBM.rds"
inter_dirs <- c(Sys.getenv("MAPMET_INTERMEDIARY", ""),
                file.path(params$input,  "R_intermediary"),
                file.path(dirname(params$input), "R_intermediary"),
                file.path(params$output, "R_intermediary"))
inter_dirs <- inter_dirs[nzchar(inter_dirs)]
pg_file <- Find(file.exists, file.path(inter_dirs, pg_name))

if (!is.null(pg_file)) {
  pg <- readRDS(pg_file)
  if (!all(colnames(spe) %in% names(pg)))
    stop("Frozen clustering ", pg_file, " does not cover all cells; ",
         "regenerate it (code/generate_phenograph_intermediary.R).")
  lab <- as.integer(pg[colnames(spe)])
  spe$pg_clusters <- factor(lab, levels = sort(unique(lab)))
  message("Loaded frozen phenograph clustering: ", pg_file)
} else {
  Rphenograph_out <- Rphenoannoy(t(mat), k = k)
  spe$pg_clusters <- factor(membership(Rphenograph_out[[2]]))
  pg_out <- file.path(params$output, "R_intermediary")
  dir.create(pg_out, showWarnings = FALSE, recursive = TRUE)
  saveRDS(setNames(as.character(spe$pg_clusters), colnames(spe)),
          file.path(pg_out, pg_name))
  message("Saved frozen phenograph clustering: ", file.path(pg_out, pg_name))
}

# Cluster colours (helpers from bubbleplot.R)
if (is.null(metadata(spe)$color_vectors)) {
  metadata(spe)$color_vectors <- list()
}
cluster_levels <- get_cluster_levels(spe$pg_clusters)
color_key <- "col_pg_clusters"
if (is.null(metadata(spe)$color_vectors[[color_key]]) ||
    !all(cluster_levels %in% names(metadata(spe)$color_vectors[[color_key]]))) {
  metadata(spe)$color_vectors[[color_key]] <- generate_cluster_colors(
    cluster_levels = cluster_levels
  )
}
col_clusters <- metadata(spe)$color_vectors[[color_key]][cluster_levels]

# UMAP coloured by pg_clusters
p <- dittoDimPlot(
  spe, var = "pg_clusters", reduction.use = "UMAP",
  size = 0.5, do.label = TRUE, legend.show = FALSE,
  color.panel = col_clusters
) +
  theme(legend.title = element_blank()) +
  ggtitle("pg_clusters (k = 30) on UMAP")

label_layer <- which(
  sapply(p$layers, function(x) class(x$geom)[1]) %in% c("GeomLabelRepel", "GeomLabel")
)
label_data <- layer_data(p, label_layer)
label_fills <- col_clusters[as.character(label_data$label)]
p$layers[[label_layer]]$aes_params$fill <- label_fills
p$layers[[label_layer]]$aes_params$colour <- "black"

save_plot(p, "umap_pg_clusters.pdf", width = 8, height = 6)
print(p)

# Bubble plot for pg_clusters
plot_cluster_bubble(
  cluster_col = "pg_clusters",
  save_name = "bubbleplot_pg_clusters.pdf",
  ncells_break_at = 100000
)
```


## Subclustering of mixed clusters using GMM

```{r gmm-subclustering}
library(mclust)

set.seed(221228)

# NOTE: subcluster_map / marker_map are keyed to the k=30 pg_clusters IDs.
spe$pg_clusters_beforesubclustering <- spe$pg_clusters


spe$pg_clusters <- spe$pg_clusters_beforesubclustering

# character, for reassignment below
spe$pg_clusters <- as.character(spe$pg_clusters)

# original cluster -> label for its second GMM component
subcluster_map <- list(
  "5"  = "35",
  "12" = "36",
  "17" = "37"
)

# define marker sets per cluster
marker_map <- list(
  "5"  = c("CD3_"),
  "12" = c("CD4_"),
  "17" = c("CD3_", "CD10")
)

for (cl in names(subcluster_map)) {
  new_cl <- subcluster_map[[cl]]
  
  idx <- which(spe$pg_clusters == cl)
  if (length(idx) < 4) next
  
  sub_mat <- t(mat[, idx])  # cells x markers
  
  # select cluster-specific markers
  if (cl %in% names(marker_map)) {
    markers <- marker_map[[cl]]
    keep_cols <- grep(paste0("^(", paste(markers, collapse="|"), ")"), colnames(sub_mat))
    sub_mat <- sub_mat[, keep_cols, drop = FALSE]
  }
  
  print(cl)
  print(head(sub_mat))
  
  # clean data
  sub_mat <- sub_mat[complete.cases(sub_mat), , drop = FALSE]
  sub_mat <- sub_mat[, apply(sub_mat, 2, var) > 0, drop = FALSE]
  print(cl)
  print(head(sub_mat))

  sub_mat <- scale(sub_mat)

  gmm_fit <- tryCatch(
    Mclust(sub_mat, G = 2, verbose = FALSE),
    error = function(e) NULL
  )
  
  if (is.null(gmm_fit)) next
  
  comp2 <- which(gmm_fit$classification == 2)
  spe$pg_clusters[idx[comp2]] <- new_cl
}
```

## Single-cell heatmaps for subclustered clusters

```{r subcluster-heatmaps}
library(dittoSeq)
library(ggplot2)
library(viridis)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]

# original/new subcluster pairs
subcluster_pairs <- list(
  list(orig = "5",  new = "35"),
  list(orig = "12", new = "36"),
  list(orig = "17", new = "37")
)


set.seed(221228)
for (pair in subcluster_pairs) {
  orig_cl <- pair$orig
  new_cl  <- pair$new

  # cells in either subcluster
  idx <- which(spe$pg_clusters %in% c(orig_cl, new_cl))

  if (length(idx) == 0) next

  n_cells   <- min(8000, length(idx))
  rand_idx  <- sample(idx, n_cells)

  spe_cl <- spe[, rand_idx]

  # order cells by subcluster (groups side-by-side)
  cell_order <- order(spe_cl$pg_clusters)

  p1 <- dittoHeatmap(
    spe_cl[, cell_order],
    genes          = genes_use,
    cluster_cols   = FALSE,          # keep our column order
    scaled.to.max  = TRUE,
    heatmap.colors.max.scaled = viridis(100),
    annot.by       = "pg_clusters"
  )

  save_plot(
    p1,
    paste0("heatmap_subcluster_", orig_cl, "_", new_cl, ".pdf"),
    width  = 14,
    height = 6
  )
}
```

```{r pg-clusters-final-bubbleplot}

# Bubble plot after subclustering (distinct filename).
plot_cluster_bubble(
      cluster_col = "pg_clusters",
      ncells_break_at = 100000,
      save_name = "bubbleplot_pg_clusters_after_subclustering.pdf")

# UMAP after subclustering; regenerate colours for the new levels and fix the
# factor level order so the colour panel lines up.
cluster_levels_after <- get_cluster_levels(spe$pg_clusters)
col_clusters_after   <- generate_cluster_colors(cluster_levels = cluster_levels_after)[cluster_levels_after]
spe$pg_clusters_after_plot <- factor(as.character(spe$pg_clusters), levels = cluster_levels_after)

p_after <- dittoDimPlot(
  spe, var = "pg_clusters_after_plot", reduction.use = "UMAP",
  size = 0.5, do.label = TRUE, legend.show = FALSE,
  color.panel = col_clusters_after
) +
  theme(legend.title = element_blank()) +
  ggtitle("pg_clusters after subclustering on UMAP")

label_layer <- which(
  sapply(p_after$layers, function(x) class(x$geom)[1]) %in% c("GeomLabelRepel", "GeomLabel")
)
label_data  <- layer_data(p_after, label_layer)
label_fills <- col_clusters_after[as.character(label_data$label)]
p_after$layers[[label_layer]]$aes_params$fill   <- label_fills
p_after$layers[[label_layer]]$aes_params$colour <- "black"

save_plot(p_after, "umap_pg_clusters_after_subclustering.pdf", width = 8, height = 6)
print(p_after)

```


## Clustered heatmaps for each cluster

```{r}
library(dittoSeq)
library(ggplot2)

clusters <- unique(colData(spe)$pg_clusters)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]


for (cl in clusters) {

  spe_cl <- spe[, colData(spe)$pg_clusters == cl]

  # skip empty clusters
  if (ncol(spe_cl) == 0) next

  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)

  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )

  save_plot(
    p1,
    paste0("heatmap_after_subclustering_", cl, ".pdf"),
    width = 14,
    height = 6,
    subdir = "heatmap_after_subclustering"
  )
}

```


## Write cluster proportions for manual annotation (Excel)

The chunk below regenerates the cluster-annotation template that is filled in manually and published on Zenodo as a Supplementary Table (read back in the next section). The whole block is disabled with `eval=FALSE` so the pipeline does not depend on `writexl`; enable it only if you need to regenerate the template.

```{r cluster-annotation-excel, eval=FALSE}
library(writexl)

spe$pg_clusters <- as.character(spe$pg_clusters)

# Absolute counts per tissue
count_table <- as.data.frame(table(
  cluster = spe$pg_clusters,
  tissue  = spe$tissue
))

count_wide <- count_table %>%
  pivot_wider(
    names_from = tissue,
    values_from = Freq,
    values_fill = 0
  )

# Ensure all tissues exist
for(t in c("BM","PT")){
  if(!t %in% colnames(count_wide)){
    count_wide[[t]] <- 0
  }
}

count_wide <- count_wide %>%
  select(cluster, BM, PT)

# Percentages within cluster
percent_within_cluster <- count_wide %>%
  rowwise() %>%
  mutate(
    total = BM + PT,
    BM_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * BM / total), "0.0%"),
    PT_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * PT / total), "0.0%")
  ) %>%
  ungroup() %>%
  select(-total)

# Empty annotation columns to fill in
annotation_template <- percent_within_cluster %>%
  mutate(
    celltype = "",
    metacluster = "",
    comments = ""
  ) %>%
  arrange(as.numeric(cluster))

# Write Excel
write_xlsx(
  annotation_template,
  file.path(params$output, "cluster_annotation_template.xlsx")
)
```


## Annotation and merging of clusters

Next, we read assigned celltype and metacluster-labels for clusters from the Supplementary Table provided on Zenodo [(Supplementary_Tables/20260526_Suppl_Table3.xlsx)](https://doi.org/10.5281/zenodo.13220634). These have been assigned based on thorough examination of images and co-expressions and using literature. The sheet must contain the columns `cluster`, `celltype` and `metacluster`.
Let's start with celltype assignments...

```{r annotations-celltypes}
# Supplementary table: inside the cohort dir or beside it.
.suppl_cands <- file.path(c(params$input, dirname(params$input)), "Supplementary_Tables", "20260526_Suppl_Table3.xlsx")
file_path <- .suppl_cands[file.exists(.suppl_cands)][1]
if (is.na(file_path)) stop("Supplementary table not found in:\n  ", paste(.suppl_cands, collapse = "\n  "))
annotation <- read_excel(file_path, sheet = "Sheet1")

cluster_names <- setNames(annotation$celltype, annotation$cluster)
celltypes <- recode(spe$pg_clusters, !!!cluster_names)
spe$celltype <-  as.vector(celltypes)

metacluster_names <- setNames(annotation$metacluster, annotation$cluster)
metacelltypes <- recode(spe$pg_clusters, !!!metacluster_names)
spe$metacluster <-  as.vector(metacelltypes)

```

## Assign colors 
Finally, we assign colors to all celltypes and metaclusters.

```{r colors-celltypes raw_k30}
# Colours are defined centrally in 00_MapMet_color_code.R
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }

col_celltype <- COL_CELLTYPE

metadata(spe)$color_vectors$col_celltype <- col_celltype
```

```{r colors-metaclusters}
col_metacluster <- COL_METACLUSTER
metadata(spe)$color_vectors$col_metacluster <- col_metacluster
```

## Visualizations
For visualization purpose, we will give the celltypes a meaningful order. This order can be downloaded from a csv file from Zenodo [(celltype_order.csv)](https://doi.org/10.5281/zenodo.13220634 )

```{r order-celltypes}
# celltype_order.csv ships in R_intermediary (falls back to params$input).
order_dirs <- c(inter_dirs, params$input)
order_file <- Find(file.exists, file.path(order_dirs, "celltype_order.csv"))
if (is.null(order_file))
  stop("celltype_order.csv not found in: ", paste(order_dirs, collapse = ", "))
clusternames_ordered <- read.csv(order_file, sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)

# The "NN_" prefix imposes display order via alphabetical sort; store
# celltype_ordered as a factor in that order but with the prefix stripped.
strip_order_prefix   <- function(x) sub("^[0-9]+_", "", x)
celltype_prefixed    <- recode(spe$celltype, !!!codes)            # e.g. "01_CD44+ Tu"
ordered_levels       <- strip_order_prefix(sort(unique(celltype_prefixed)))
spe$celltype_ordered <- factor(strip_order_prefix(celltype_prefixed), levels = ordered_levels)

# col_celltype is keyed by the same (unprefixed) names, so reuse it.
metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
```

As a last step, we generate UMAP and bubble plots for the final celltypes/metaclusters.


```{r bubbleplot-celltypes, fig.width=16, fig.height=14}

plot_cluster_bubble(
  cluster_col = "celltype",
  save_name = "bubbleplot-celltypes.pdf",
  annotation_level = "metacluster",
  ncells_break_at = 65000,
  plot_width      = 13,
  plot_height     = 11.5
)

```

```{r umap-final-celltypes, fig.width=10}
# aspect.ratio = 1 keeps the UMAP panel square regardless of legend width.
p1 <- dittoDimPlot(spe, var = "celltype_ordered",
             reduction.use = "UMAP", size = 0.3,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Celltypes on UMAP")

save_plot(p1, "umap-celltypes.pdf", width = 16, height = 8)

p1
```

We also plot the same celltype UMAP but colour only the primary tumor (PT)
cells, showing all other tissues as light-grey background dots.

```{r umap-celltype-PT-only, fig.width=10}
# colour PT cells only, others as grey background
p1_pt <- dittoDimPlot(spe, var = "celltype_ordered",
             reduction.use = "UMAP", size = 0.3,
             cells.use = spe$tissue == "PT",
             show.others = TRUE,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Celltypes on UMAP (PT only)")

save_plot(p1_pt, "umap-celltype-PT-only.pdf", width = 16, height = 8)

p1_pt
```

```{r vis-metacluster}
p2 <- dittoDimPlot(spe, var = "metacluster",
             reduction.use = "UMAP", size = 0.3, labels.size=4,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_metacluster) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Metaclusters on UMAP")

save_plot(p2, "umap-metaclusters.pdf", width = 13, height = 8)

p2
```


Finally, we plot a validation heatmap of the cells in each celltype.


```{r}
library(dittoSeq)
library(ggplot2)
library(viridis)

# celltypes to iterate over
clusters <- unique(colData(spe)$celltype)

genes_use <- rownames(spe)[
  !grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", 
         rownames(spe))
]

for (cl in clusters) {
  
  # safe filename from celltype
  safe_name <- cl
  safe_name <- gsub("\\+", "plus", safe_name)
  safe_name <- gsub("-", "minus", safe_name)
  safe_name <- gsub("[^A-Za-z0-9_]", "_", safe_name)
  
  spe_cl <- spe[, colData(spe)$celltype == cl]
  
  # skip empty celltypes
  if (ncol(spe_cl) == 0) next

  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  filename <- paste0("heatmap_celltype_", safe_name, ".pdf")
  save_plot(p1, filename = filename, width = 14, height = 6, subdir = "heatmap_celltype")
}
```
## Save object

```{r save-objects}

saveRDS(spe, file.path(params$output, "spe_04_phenotyping.rds"))

```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
