---
title: "Cell phenotyping"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/04_phenotyping")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# create phenograph heatmap folder
outdir <- paste0(plot_dir,"/pg_heatmap")
dir.create(outdir, showWarnings = FALSE)


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Cell phenotyping {#phenotyping}
In this Rmd file, we will perform 
1. dimensionality reduction
2. phenograph clustering
3. annotate and merge clusters & metaclusters externally 
4. retrieve tumor cells from GD2-MPO double positive cells (caused by spillover) 
5. re-classify granulocyte metaclusters into maturity states based on morphology
6. define color codes for all celltypes and metaclusters

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ComplexHeatmap)
library(RColorBrewer)
library(viridis)
library(scater)
library(BiocParallel)
library(Rphenoannoy)
library(tidyr)
library(tidyHeatmap)
library(bruceR)
library(dittoSeq)
library(dplyr)
library(mclust)
library(SingleR)
library(circlize)
library(readxl)
library(writexl)
library(ggplot2)
library(patchwork)
library(harmony)
library(batchelor)
library(cluster)
```
</details>

## Read data
We will first read the data that passed the [second quality control step](#qc-2) and save the count data as a matrix. Low-quality markers (low SNR in a lot of samples) as well as all IF and nuclear IMC markers will be excluded for clustering. Moreover, mean intensity will be used to represent marker expression, to average out the signals arising from lateral spillover or artifacts. As explained in [the previous chapter](#qc2-save), we are using counts measured with respect to dilated masks in downstream analysis do capture membrane signal. 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_qc2.rds"))

#computing matrix
intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- colnames(spe)

#use mean intensity
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]

#exclude low-quality markers and functional marker cl.PARP
intensity_mat_hq <- intensity_mat[!grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(intensity_mat)),]
intensity_mat_lq <- intensity_mat[grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(intensity_mat)),]

non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)

intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]
intensity_mat_lq_wo_zeros <- intensity_mat_lq[,non_zeros]

#Apply changes to SPE object
spe <- spe[,non_zeros]
rowData(spe)$use_channel <- !grepl("mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))

##Matrix with morph. features
morph_mat_clipped <- (cbind("area"=spe$area_clipped, "solidity"=spe$solidity_clipped, "aspect_ratio"=spe$aspect_ratio_clipped))

tissue <- spe$tissue
cell_id <- colnames(spe)

mat <- intensity_mat_hq_wo_zeros

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_sample = setNames(sample(col_vector_433, length(unique(spe$sample))), unique(spe$sample))
```
We are excluding `r length(non_zeros[non_zeros==FALSE])` more cells here, as they are negative for all markers used in downstream analysis. 
To get a first grasp of the single-cell data and potential clusters, we will plot it in a single-cell heatmap.
spe_filtered <- spe_sub[, colData(spe_sub)$DNAIntensity >= low_dna_threshold]


## UMAP (for visualization only)
```{r umap}
set.seed(221228)

Sys.setenv(OMP_NUM_THREADS = 1)
Sys.setenv(MKL_NUM_THREADS = 1)

BPPARAM <- SnowParam(workers = 30, type = "SOCK", RNGseed = 221228)

mat_red <- calculateUMAP(
  mat,
  n_neighbors = 30,
  BPPARAM = BPPARAM
)

rownames(mat_red) <- colnames(spe)
reducedDim(spe, "UMAP") <- mat_red

p <- dittoDimPlot(spe, var = "fm_id", 
             reduction.use = "UMAP", size = 1,
             do.label = TRUE,
             legend.show=F) +
  theme(legend.title = element_blank()) +
  ggtitle("Sample ID on UMAP")


save_plot(p2, "UMAP-vis-sampleID.png", width = 8, height = 6)

p

```



# UMAPs for each marker
```{r UMAP-all-markers, fig.width=24, fig.height=32}
excluded_markers <- "mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3"
max_cells_for_umap <- 20000

marker_vars <- rownames(spe)[
  !grepl(excluded_markers, rownames(spe))
]

set.seed(221228)
marker_umap_ncells <- min(max_cells_for_umap, ncol(spe))
marker_umap_idx <- sample(seq_len(ncol(spe)), marker_umap_ncells)
spe_umap_markers <- spe[, marker_umap_idx]

p_markers <- multi_dittoDimPlot(
  spe_umap_markers,
  vars = marker_vars,
  reduction.use = "UMAP",
  size = 0.5,
  ncol = 5,
  color.panel = viridis(100, direction = -1)
)

save_plot(p_markers, "UMAP-all-markers.png", width = 24, height = 32)
p_markers
```

# For later annotation, we generate one clustered heatmap per sample

```{r}
library(dittoSeq)
library(ggplot2)

# get cluster IDs
clusters <- unique(colData(spe)$pg_clusters)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]


for (cl in clusters) {
  
  # subset cells for this cluster
  spe_cl <- spe[, colData(spe)$pg_clusters == cl]
  
  # skip empty clusters (just in case)
  if (ncol(spe_cl) == 0) next
  
  # sample up to 8000 cells
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  # make heatmap
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  # save plot
  save_plot(
    p1,
    paste0("heatmap_phenocluster_", cl, ".png"),
    width = 14,
    height = 6
  )
}

```

## Bubble plot: Mean marker expression per cluster
Let's plot the mean marker expressions per cluster in a bubble plot.

```{r vis-data-bubbleplot, fig.height=14, fig.width=16}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(viridis)
library(cowplot)



# Suppress the extra print() re-draw in the for-loop: the function handles
# its own display so the returned marker object is a silent no-op when printed.
print.heatmap_drawn <- function(x, ...) invisible(NULL)

plot_cluster_bubble <- function(cluster_col, save_name,
                                annotation_level = c("cluster", "metacluster"),
                                col_annotation  = NULL,
                                metacluster_col = "metacluster",
                                ncells_break_at = 50000) {
  annotation_level <- match.arg(annotation_level)
  barplot_height <- unit(4, "cm")
  
  # ----------------------------
  # Cluster vector
  # ----------------------------
  cluster_vec <- as.character(colData(spe)[[cluster_col]])
  tissue_vec  <- colData(spe)$tissue

  # ----------------------------
  # Expression matrix
  # ----------------------------
  mat_df <- as.data.frame(t(mat))

  exp_mat_ <- aggregate(
    mat_df,
    by = list(cluster = cluster_vec),
    FUN = function(x) {
      pos_vals <- x[x > 0]
      if (length(pos_vals) == 0) return(0)
      median(log1p(pos_vals))
    }
  )

  cluster_names <- exp_mat_$cluster

  exp_mat <- exp_mat_[, -1, drop = FALSE]
  rownames(exp_mat) <- cluster_names
  exp_mat <- t(as.matrix(exp_mat))

  # ----------------------------
  # Row scaling
  # ----------------------------
  rmin <- apply(exp_mat, 1, min, na.rm = TRUE)
  rmax <- apply(exp_mat, 1, max, na.rm = TRUE)
  den  <- rmax - rmin

  exp_mat <- (exp_mat - rmin) / den

  exp_mat[is.nan(exp_mat)] <- 0
  exp_mat[is.infinite(exp_mat)] <- 0

  clusters <- colnames(exp_mat)

  # ----------------------------
  # Percent expressed
  # ----------------------------
  percent_mat <- sapply(seq_len(nrow(mat)), function(i) {

    row_vals <- mat[i, ]
    pos <- row_vals[row_vals > 0]

    if (length(pos) == 0) {

      out <- rep(0, length(clusters))
      names(out) <- clusters
      return(out)
    }

    thresh <- quantile(pos, 0.2, na.rm = TRUE)

    row_bin <- ifelse(row_vals > thresh, 1, 0)

    tapply(
      row_bin,
      factor(cluster_vec, levels = clusters),
      mean
    )
  })

  percent_mat <- t(percent_mat)

  percent_mat <- percent_mat[, clusters, drop = FALSE]

  percent_mat[is.na(percent_mat)] <- 0
  percent_mat[percent_mat < 0.1] <- 0.1

  # ----------------------------
  # Tissue counts
  # ----------------------------
  bm_count <- table(factor(cluster_vec[tissue_vec == "BM"], levels = clusters))
  pt_count <- table(factor(cluster_vec[tissue_vec == "PT"], levels = clusters))

  # ----------------------------
  # Barplot info helper (supports optional axis break)
  # ----------------------------
  make_barplot_info <- function(v) {
    v_num   <- as.numeric(v)
    max_val <- max(v_num, na.rm = TRUE)

    if (max_val <= ncells_break_at) {
      y_axis_max <- max(10000, ceiling(max_val / 10000) * 10000)
      at <- seq(0, y_axis_max, by = 10000)
      return(list(
        values      = v_num,
        upper       = y_axis_max,
        axis_param  = list(
          at     = at,
          labels = ifelse(at == 0, "0", paste0(at / 1000, "k")),
          gp     = gpar(fontsize = 12)
        ),
        needs_break = FALSE
      ))
    }

    # Break case: compress above-break values into a small top band
    # (top_frac of the display height is reserved for the above-break region)
    top_frac   <- 0.18
    break_frac <- 1 - top_frac
    top_ext    <- ncells_break_at * (top_frac / break_frac)
    ylim_max   <- ncells_break_at + top_ext

    # Above-break bars are drawn at 90% of the top band
    above_display <- ncells_break_at + top_ext * 0.9
    values_plot   <- ifelse(v_num > ncells_break_at, above_display, v_num)

    at_below   <- seq(0, ncells_break_at, by = 10000)
    at_all     <- c(at_below, above_display)
    labels_all <- c(
      ifelse(at_below == 0, "0", paste0(at_below / 1000, "k")),
      paste0(round(max_val / 1000), "k")
    )

    list(
      values           = values_plot,
      upper            = ylim_max,
      axis_param       = list(
        at     = at_all,
        labels = labels_all,
        gp     = gpar(fontsize = 12)
      ),
      needs_break      = TRUE,
      break_y_npc      = ncells_break_at / ylim_max,
      above_break_orig = which(v_num > ncells_break_at),
      n_total          = length(v_num)
    )
  }
  
  # ----------------------------
  # Cluster / metacluster colors
  # ----------------------------
  if (annotation_level == "metacluster") {
    # Keep legacy behavior from the old bubbleplot block:
    # aggregate(cluster -> metacluster, FUN = max)
    metaclust_by_cluster <- aggregate(
      colData(spe)[[metacluster_col]],
      list(cluster = cluster_vec),
      max
    )
    meta_by_cluster <- setNames(
      as.character(metaclust_by_cluster$x),
      as.character(metaclust_by_cluster$cluster)
    )
    annot_values <- meta_by_cluster[clusters]
    annot_name <- "metacluster"
    annot_colors <- metadata(spe)$color_vectors$col_metacluster
  } else {
    color_key <- paste0("col_", cluster_col)
    seed_offset <- match(cluster_col, cluster_vars)
    if (is.na(seed_offset)) {
      seed_offset <- sum(utf8ToInt(cluster_col))
    }

    existing_colors <- metadata(spe)$color_vectors[[color_key]]
    if (is.null(existing_colors) ||
        !all(clusters %in% names(existing_colors)) ||
        any(is.na(existing_colors[clusters]))) {
      existing_colors <- generate_cluster_colors(
        cluster_levels = clusters,
        seed = color_seed_base + seed_offset
      )
      metadata(spe)$color_vectors[[color_key]] <- existing_colors
    }

    annot_values <- clusters
    annot_name <- cluster_col
    annot_colors <- existing_colors[clusters]
  }

  if (!is.null(col_annotation)) {
    annot_colors <- col_annotation
  }

  levels_from_colors <- names(annot_colors)
  annot_levels <- if (!is.null(levels_from_colors)) {
    intersect(levels_from_colors, unique(annot_values))
  } else {
    unique(annot_values)
  }
  if (length(annot_levels) == 0) {
    annot_levels <- unique(annot_values)
  }

  col_for_annot <- setNames(list(annot_colors), annot_name)

  pt_info <- make_barplot_info(pt_count)
  bm_info <- make_barplot_info(bm_count)
  cluster_annotation_list <- setNames(
    list(factor(annot_values, levels = annot_levels)),
    annot_name
  )
  
  # ----------------------------
  # Heatmap colors
  # ----------------------------
  col_fun <- circlize::colorRamp2(
    c(0, 0.5, 1),
    viridis(100)[c(1, 50, 100)]
  )

  # ----------------------------
  # Column annotation
  # ----------------------------
  column_ha <- do.call(
    HeatmapAnnotation,
    c(
      cluster_annotation_list,
      list(
        ncells_PT = anno_barplot(
          pt_info$values,
          ylim = c(0, pt_info$upper),
          height = barplot_height,
          gp = gpar(fill = "brown"),
          axis_param = pt_info$axis_param
        ),
        ncells_BM = anno_barplot(
          bm_info$values,
          ylim = c(0, bm_info$upper),
          height = barplot_height,
          gp = gpar(fill = "aquamarine4"),
          axis_param = bm_info$axis_param
        ),
        col = col_for_annot,
        show_legend = TRUE
      )
    )
  )

  # ----------------------------
  # Bubble cell function
  # ----------------------------
  cell_fun <- function(j, i, x, y, w, h, fill) {

    grid.rect(
      x, y, w, h,
      gp = gpar(col = NA, fill = NA)
    )

    grid.circle(
      x = x,
      y = y,
      r = percent_mat[i, j] / 2 * min(unit.c(w, h)),
      gp = gpar(
        fill = col_fun(exp_mat[i, j]),
        col = NA
      )
    )
  }

  # ----------------------------
  # Bubble legend
  # ----------------------------
  lgd_list <- list(

    Legend(
      labels = c(0, 0.25, 0.5, 0.75, 1),
      title = "percentage expressed",

      graphics = lapply(
        c(0, 0.25, 0.5, 0.75, 1),
        function(r) {

          function(x, y, w, h)
            grid.circle(
              x = x,
              y = y,
              r = r * unit(2, "mm"),
              gp = gpar(fill = "black")
            )
        }
      )
    )
  )

  # ----------------------------
  # Heatmap
  # ----------------------------
  hp <- Heatmap(
    exp_mat,

    heatmap_legend_param = list(title = "expression"),

    column_title = paste0(
      "Bubble plot - ",
      cluster_col
    ),

    top_annotation = column_ha,

    col = col_fun,

    rect_gp = gpar(type = "none"),

    clustering_distance_columns = "pearson",

    clustering_method_columns = "average",

    row_labels = sapply(
      strsplit(rownames(exp_mat), "_"),
      "[[",
      1
    ),

    cell_fun = cell_fun,

    show_row_dend = FALSE,

    column_dend_height = unit(1.5, "cm"),

    row_names_gp = gpar(fontsize = 15),

    column_names_gp = gpar(fontsize = 15),

    column_names_rot = 45,

    border = "black"
  )

  # ----------------------------
  # Break-mark decoration helper
  # ----------------------------
  add_break_marks <- function(anno_name, info, drawn_hp) {
    if (!info$needs_break) return(invisible(NULL))

    col_ord <- column_order(drawn_hp)
    if (is.list(col_ord)) col_ord <- unlist(col_ord)
    n_tot <- info$n_total
    b_y   <- info$break_y_npc
    gap   <- 0.04   # height of the white erasure gap (npc units)

    decorate_annotation(anno_name, {
      pushViewport(viewport(clip = "off"))

      for (orig_idx in info$above_break_orig) {
        vis <- match(orig_idx, col_ord)
        if (is.na(vis)) next
        xc <- (vis - 0.5) / n_tot
        bw <- 0.8  / n_tot

        # White gap to visually break the bar
        grid.rect(
          x = xc, y = b_y - gap / 2,
          width = bw, height = gap,
          just  = c("center", "bottom"),
          default.units = "npc",
          gp = gpar(fill = "white", col = NA)
        )

        # Zigzag lines across the bar
        xs <- c(xc - bw/2, xc - bw/4, xc, xc + bw/4, xc + bw/2)
        ys <- b_y + c(-gap * 0.6, gap * 0.6, -gap * 0.6, gap * 0.6, -gap * 0.6)
        grid.polyline(
          x = xs, y = ys,
          default.units = "npc",
          gp = gpar(col = "black", lwd = 1.5)
        )
      }

      # Zigzag break mark on the y-axis (drawn to the left of the annotation)
      grid.polyline(
        x = unit(c(-2.5, -1.5, -2.5), "mm"),
        y = unit(c(b_y - gap * 0.6, b_y, b_y + gap * 0.6), "npc"),
        gp = gpar(col = "black", lwd = 1.5)
      )

      popViewport()
    })
  }

  # ----------------------------
  # Draw, decorate, and save
  # ----------------------------
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", save_name))
  pdf(full_path, width = 16, height = 14)
  hp_drawn <- draw(hp, annotation_legend_list = lgd_list)
  add_break_marks("ncells_PT", pt_info, hp_drawn)
  add_break_marks("ncells_BM", bm_info, hp_drawn)
  dev.off()
  message("Saved: ", full_path)

  # Draw once more for the notebook / interactive display (with decorations).
  # Return a silent marker so the caller's print() in the cluster_vars for-loop
  # (line 884) does not trigger a third, undecorated re-draw via the
  # ComplexHeatmap show() method.
  hp_drawn <- draw(hp, annotation_legend_list = lgd_list)
  add_break_marks("ncells_PT", pt_info, hp_drawn)
  add_break_marks("ncells_BM", bm_info, hp_drawn)

  invisible(structure(list(), class = "heatmap_drawn"))
}


```

```{r phenograph-umap-dotplot}
set.seed(221228)

k_values <- c(15, 30, 45, 60)

for (k in k_values) {
  Rphenograph_out <- Rphenoannoy(t(mat), k = k)
  clusters <- factor(membership(Rphenograph_out[[2]]))
  colData(spe)[[paste0("pg_clusters_k", k)]] <- clusters
}

cluster_vars <- c(
  "pg_clusters_k15",
  "pg_clusters_k30",
  "pg_clusters_k45",
  "pg_clusters_k60"
)

color_seed_base <- 221228

get_cluster_levels <- function(values) {
  cluster_levels <- unique(as.character(values))
  cluster_levels <- cluster_levels[!is.na(cluster_levels)]
  cluster_levels <- cluster_levels[nzchar(cluster_levels)]

  if (all(grepl("^\\d+$", cluster_levels))) {
    cluster_levels <- as.character(sort(as.integer(cluster_levels)))
  } else {
    cluster_levels <- sort(cluster_levels)
  }

  cluster_levels
}

generate_cluster_colors <- function(cluster_levels, seed = color_seed_base) {
  non_gray_colors <- colors()[
    grep("\\bgr[ae]y", grDevices::colors(), invert = TRUE)
  ]

  set.seed(seed)
  setNames(
    sample(
      non_gray_colors,
      length(cluster_levels),
      replace = length(cluster_levels) > length(non_gray_colors)
    ),
    cluster_levels
  )
}

if (is.null(metadata(spe)$color_vectors)) {
  metadata(spe)$color_vectors <- list()
}

for (var_name in cluster_vars) {
  cluster_levels <- get_cluster_levels(spe[[var_name]])
  color_key <- paste0("col_", var_name)

  if (is.null(metadata(spe)$color_vectors[[color_key]]) ||
      !all(cluster_levels %in% names(metadata(spe)$color_vectors[[color_key]]))) {
    metadata(spe)$color_vectors[[color_key]] <- generate_cluster_colors(
      cluster_levels = cluster_levels,
      seed = color_seed_base + match(var_name, cluster_vars)
    )
  }

  col_clusters <- metadata(spe)$color_vectors[[color_key]][cluster_levels]

  p <- dittoDimPlot(
    spe,
    var = var_name,
    reduction.use = "UMAP",
    size = 0.5,
    do.label = TRUE,
    legend.show = FALSE,
    color.panel = col_clusters
  ) +
    theme(legend.title = element_blank()) +
    ggtitle(paste(var_name, "on UMAP"))

  label_layer <- which(
    sapply(
      p$layers,
      function(x) class(x$geom)[1]
    ) %in% c("GeomLabelRepel", "GeomLabel")
  )

  label_data <- layer_data(p, label_layer)
  label_fills <- col_clusters[as.character(label_data$label)]
  p$layers[[label_layer]]$aes_params$fill <- label_fills
  p$layers[[label_layer]]$aes_params$colour <- "black"

  save_plot(
    p,
    paste0("umap_", var_name, ".png"),
    width = 8,
    height = 6
  )

  print(p)
}

for (cluster_col in cluster_vars) {
  message("Processing: ", cluster_col)
  print(
    plot_cluster_bubble(
      cluster_col = cluster_col,
      save_name = paste0("bubbleplot_", cluster_col, ".pdf"),
      ncells_break_at = 100000
    )
  )
}
```

## Define final clustering set
```{r pg-clusters}
spe$pg_clusters <- spe$pg_clusters_k30
```

## Subclustering of mixed clusters using GMM

```{r gmm-subclustering}
library(mclust)

set.seed(221228)

spe$pg_clusters_beforesubclustering <- spe$pg_clusters_30


spe$pg_clusters <- spe$pg_clusters_beforesubclustering

# Ensure pg_clusters is character for easy reassignment
spe$pg_clusters <- as.character(spe$pg_clusters)

# Mapping: original cluster -> new subcluster label for the second component
subcluster_map <- list(
  "7" = "36",
  "8" = "37",
  "14"  = "38",
  "17" = "39"
)

# define marker sets per cluster
marker_map <- list(
  "7"  = c("CD3_"),
  "8" = c("GD2_"),
  "14" = c("CD4_", "CD3_"),
  "17" = c("CD3_",  "CD10")
)

for (cl in names(subcluster_map)) {
  new_cl <- subcluster_map[[cl]]
  
  idx <- which(spe$pg_clusters == cl)
  if (length(idx) < 4) next
  
  sub_mat <- t(mat[, idx])  # cells x markers
  
  # --- select cluster-specific markers ---
  if (cl %in% names(marker_map)) {
    markers <- marker_map[[cl]]
    keep_cols <- grep(paste0("^(", paste(markers, collapse="|"), ")"), colnames(sub_mat))
    sub_mat <- sub_mat[, keep_cols, drop = FALSE]
  }
  
  print(cl)
  print(head(sub_mat))
  
  # --- clean data ---
  sub_mat <- sub_mat[complete.cases(sub_mat), , drop = FALSE]
  sub_mat <- sub_mat[, apply(sub_mat, 2, var) > 0, drop = FALSE]
  print(cl)
  print(head(sub_mat))
  
  #if (nrow(sub_mat) < 4 || ncol(sub_mat) < 2) next
  
  sub_mat <- scale(sub_mat)
  
  # --- GMM ---
  gmm_fit <- tryCatch(
    Mclust(sub_mat, G = 2, verbose = FALSE),
    error = function(e) NULL
  )
  
  if (is.null(gmm_fit)) next
  
  comp2 <- which(gmm_fit$classification == 2)
  spe$pg_clusters[idx[comp2]] <- new_cl
}
```

## Single-cell heatmaps for subclustered clusters

```{r subcluster-heatmaps}
library(dittoSeq)
library(ggplot2)
library(viridis)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]

# Define the pairs: original cluster and its new sibling
subcluster_pairs <- list(
  list(orig = "7",  new = "36"),
  list(orig = "8", new = "37"),
  list(orig = "14", new = "38"),
  list(orig = "17", new = "39")
)

set.seed(221228)
for (pair in subcluster_pairs) {
  orig_cl <- pair$orig
  new_cl  <- pair$new

  # Cells belonging to either subcluster
  idx <- which(spe$pg_clusters %in% c(orig_cl, new_cl))

  if (length(idx) == 0) next

  # Sample up to 8000 cells
  n_cells   <- min(8000, length(idx))
  rand_idx  <- sample(idx, n_cells)

  spe_cl <- spe[, rand_idx]

  # Order cells by subcluster so the two groups appear side-by-side
  cell_order <- order(spe_cl$pg_clusters)

  p1 <- dittoHeatmap(
    spe_cl[, cell_order],
    genes          = genes_use,
    cluster_cols   = FALSE,          # preserve subcluster ordering
    scaled.to.max  = TRUE,
    heatmap.colors.max.scaled = viridis(100),
    annot.by       = "pg_clusters"   # colour bar showing subclusters
  )

  save_plot(
    p1,
    paste0("heatmap_subcluster_", orig_cl, "_", new_cl, ".png"),
    width  = 14,
    height = 6
  )
}
```

```{r celltype-bubbleplot}

plot_cluster_bubble(
      cluster_col = "pg_clusters",
      ncells_break_at = 100000,
      save_name = "bubbleplot_pg_clusters.pdf")

```


## Define final clustering and plot clustered heatmaps for each cluster

```{r}
library(dittoSeq)
library(ggplot2)

# get cluster IDs
clusters <- unique(colData(spe)$pg_clusters)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]


for (cl in clusters) {
  
  # subset cells for this cluster
  spe_cl <- spe[, colData(spe)$pg_clusters == cl]
  
  # skip empty clusters (just in case)
  if (ncol(spe_cl) == 0) next
  
  # sample up to 8000 cells
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  # make heatmap
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  # save plot
  save_plot(
    p1,
    paste0("heatmap_after_subclustering_", cl, ".png"),
    width = 14,
    height = 6
  )
}

```


## Write cluster proportions for manual annotation (Excel)

```{r cluster-annotation-excel}

# Ensure cluster is character
spe$pg_clusters <- as.character(spe$pg_clusters)

# -----------------------------
# 1. Absolute counts
# -----------------------------
count_table <- as.data.frame(table(
  cluster = spe$pg_clusters,
  tissue  = spe$tissue
))

count_wide <- count_table %>%
  pivot_wider(
    names_from = tissue,
    values_from = Freq,
    values_fill = 0
  )

# Ensure all tissues exist
for(t in c("BM","PT")){
  if(!t %in% colnames(count_wide)){
    count_wide[[t]] <- 0
  }
}

count_wide <- count_wide %>%
  select(cluster, BM, PT)

# -----------------------------
# 2. Percentages WITHIN cluster
# -----------------------------
percent_within_cluster <- count_wide %>%
  rowwise() %>%
  mutate(
    total = BM + PT,
    BM_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * BM / total), "0.0%"),
    PT_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * PT / total), "0.0%")
  ) %>%
  ungroup() %>%
  select(-total)

# -----------------------------
# 3. Add annotation columns
# -----------------------------
annotation_template <- percent_within_cluster %>%
  mutate(
    celltype = "",
    metacluster = "",
    comments = ""
  ) %>%
  arrange(as.numeric(cluster))

# -----------------------------
# 4. Write Excel
# -----------------------------
write_xlsx(
  annotation_template,
  file.path(params$output, "cluster_annotation_template.xlsx")
)
```

```{r stop-here}
knitr::opts_chunk$set(eval = FALSE)
```

## Annotation and merging of clusters

Next, we read assigned celltype and metacluster-labels for clusters from the metadata file provided on Zenodo [(Supplementary_Tables/20240808_Suppl_Table3.xlsx)](https://doi.org/10.5281/zenodo.13220634 ). These have been assigned based on thorough examination of images and co-expressions and using literature.
Let's start with celltype assignments...

```{r annotations-celltypes}
file_path <- file.path('/home/rstudio/data/output', 'cluster_annotation_completed.xlsx')
annotation <- read_excel(file_path, sheet = "Sheet1")

cluster_names <- setNames(annotation$celltype, annotation$cluster)
celltypes <- recode(spe$pg_clusters, !!!cluster_names)
spe$celltype <-  as.vector(celltypes)

metacluster_names <- setNames(annotation$metacluster, annotation$cluster)
metacelltypes <- recode(spe$pg_clusters, !!!metacluster_names)
spe$metacluster <-  as.vector(metacelltypes)

```

## Assign colors 
Finally, we assign colors to all celltypes and metaclusters.

```{r colors-celltypes raw_k30}
col_celltype <- c(

  ## ---- Tumor ----
  "CXCR4+ Tu"     = rgb(120, 180,  40, maxColorValue = 255),
  "CD44+ Tu"      = rgb(65,  208,  57, maxColorValue = 255),
  "GD2lo Tu"      = rgb(40,  150,130, maxColorValue = 255),
  "Ki67+ Tu"      = rgb(173, 255,  47, maxColorValue = 255),
  "early NB Tu"   = rgb(150, 220, 170, maxColorValue = 255),
  "marker-lo Tu"  = rgb(40,  120,  70, maxColorValue = 255),
  "bridge-like Tu" = rgb(0, 160, 110, maxColorValue = 255),

  ## ---- Proliferation ----
  "Ki67+ cells" = rgb(255, 200,   0, maxColorValue = 255),
  "CXCR4+ cells" = rgb(255, 245, 160, maxColorValue = 255),

  ## ---- Progenitor ----
  "CD34+ HSPCs" = rgb(255,  95,   0, maxColorValue = 255),

  ## ---- T cells ----
  "CD8+ T cells"        = rgb(220,  20,  60, maxColorValue = 255),
  "CD4+ T cells"        = rgb(232,  80,  70, maxColorValue = 255),
  "CD3+ T cells"        = rgb(180,  30,  50, maxColorValue = 255),
  "S100B+ T cells"      = rgb(199,  21, 133, maxColorValue = 255),
  "GZMB+ T cells"       = rgb(220,  90,  40, maxColorValue = 255),
  "dense T cell region" = rgb(170,  30, 120, maxColorValue = 255),

  ## ---- Myeloid ----
  "granulocytes"       = rgb(8,   48, 107, maxColorValue = 255),
  "CD14+ MO"           = rgb(33, 102, 172, maxColorValue = 255),
  "monoblast-like"     = rgb(64, 180, 170, maxColorValue = 255),
  "GZMB+ DC/NK cells"  = rgb(107,174, 214, maxColorValue = 255),

  ## ---- B cells ----
  "CD20+ B cells" = rgb(136, 86, 167, maxColorValue = 255),

  ## ---- Stromal ----
  "S100B+ schwann cells" = rgb(166, 97,  26, maxColorValue = 255),
  "fibroblast/endothel"  = rgb(101, 67,  33, maxColorValue = 255),


  ## ---- Other ----
  "other" = rgb(120, 120, 120, maxColorValue = 255)
)

metadata(spe)$color_vectors$col_celltype <- col_celltype
```

```{r colors-metaclusters}
col_metacluster <- c(
  "tumor" = "#238B45",
  "T cells" = "#E34A33",
  "granulocytes" = "#08306B",   # dark blue (matches rgb(8,48,107))
  "MO/DC/NK" = "#41B6C4",
  "B cells" = "#88419D",
  "mesenchymal" = "#8C510A",
  "proliferating cells" = "#FFD700",
  "progenitors" = "#F59E0B",   
  "other" = "#969696"
)
metadata(spe)$color_vectors$col_metacluster <- col_metacluster
```

## Visualizations
For visualization purpose, we will give the celltypes a meaningful order. This order can be downloaded from a csv file from Zenodo [(R_intermediary/celltype_order.csv)](https://doi.org/10.5281/zenodo.13220634 )

```{r order-celltypes}
clusternames_ordered <- read.csv(file.path(params$input, "celltype_order.csv"), sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)
spe$celltype_ordered <- recode(spe$celltype, !!!codes)

metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
names(metadata(spe)$color_vectors$col_celltype_ordered) <- recode(names(metadata(spe)$color_vectors$col_celltype), !!!codes)
```

As a last step, we generate UMAP and bubble plots for the final celltypes/metaclusters.

```{r umap-final-celltypes, fig.width=10}
p1 <- dittoDimPlot(spe, var = "celltype_ordered", 
             reduction.use = "UMAP", size = 0.3, 
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(legend.title = element_blank()) +
  ggtitle("Celltypes on UMAP")

p2 <- dittoDimPlot(spe, var = "metacluster", 
             reduction.use = "UMAP", size = 0.3, labels.size=4,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_metacluster) +
  theme(legend.title = element_blank()) +
  ggtitle("Metaclusters on UMAP")

# Save both
save_plot(p1, "umap-celltypes.png", width = 16, height = 8)

p1
```

```{r vis-metacluster}
save_plot(p2, "umap-metaclusters.png", width = 13, height = 8)
p2
```


```{r}

plot_cluster_bubble(
  cluster_col = "celltype",
  save_name = "bubbleplot-celltypes.pdf",
  annotation_level = "metacluster"
)

```


Let's plot heatmap of clustered cells as a validation:


```{r}
library(dittoSeq)
library(ggplot2)
library(viridis)

# Ensure output directory exists
outdir <- file.path(plot_dir)
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

# Get cluster IDs
clusters <- unique(colData(spe)$celltype)

# Select genes to use (exclude certain markers)
genes_use <- rownames(spe)[
  !grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", 
         rownames(spe))
]

for (cl in clusters) {
  
  # --- 1. Make cluster name safe for filenames ---
  safe_name <- cl
  safe_name <- gsub("\\+", "plus", safe_name)
  safe_name <- gsub("-", "minus", safe_name)
  safe_name <- gsub("[^A-Za-z0-9_]", "_", safe_name)  # replaces /, spaces, special chars
  
  # --- 2. Subset cells for this cluster ---
  spe_cl <- spe[, colData(spe)$celltype == cl]
  
  # Skip empty clusters
  if (ncol(spe_cl) == 0) next
  
  # --- 3. Sample cells ---
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  # --- 4. Create heatmap ---
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  # --- 5. Save heatmap safely ---
  filename <- paste0("heatmap_celltype_", safe_name, ".png")
  full_path <- file.path(outdir, filename)
  
  # If save_plot() is your custom function
  save_plot(p1, filename = filename, width = 14, height = 6)
  
  # Or using ggsave directly:
  # ggsave(full_path, plot = p1, width = 14, height = 6, dpi = 300)
}
```


## Visual inspection of dense T cell region
```{r}

library(cytomapper)
library(cowplot)

## --- 1. Find ROI with most cells annoated as dense T cell region ---
roi_counts <- table(colData(spe)$sample_id[colData(spe)$celltype == "dense T cell region"])
sorted_rois <- sort(roi_counts, decreasing = TRUE)
top_roi <- names(sorted_rois[1])

## --- 2. Subset object ---
tmp <- spe[, spe$sample_id == top_roi]

## --- 3. Load images ---
masks <- cytomapper::loadImages(
  file.path(params$input, "masks", paste0(top_roi, ".tif")),
  as.is = TRUE
)

img <- cytomapper::loadImages(
  file.path(params$input, "img", paste0(top_roi, ".tif"))
)

mcols(masks) <- DataFrame(sample_id = names(masks))
mcols(img) <- DataFrame(sample_id = names(img))

channelNames(img) <- rownames(spe)[1:44]

## --- 4. Create plots ---
# GD2
p_gd2 <- plotPixels(
  img,
  colour_by = "GD2_Gd155_mean",
  bcg = list(GD2_Gd155_mean = c(0, 1, 1)),
  legend = NULL,
  return_plot = TRUE
)

save_plot(p_gd2, "ROI_dense_tcell_overview_GD2.png", width = 12, height = 10)

# Vim
p_vim <- plotPixels(
  img,
  colour_by = "Vimentin_Pt196_mean",
  bcg = list(Vimentin_Pt196_mean = c(0, 1, 1)),
  legend = NULL,
  return_plot = TRUE
)
save_plot(p_vim, "ROI_dense_tcell_overview_Vim.png", width = 12, height = 10)

# CD3
p_CD3 <- plotPixels(
  img,
  colour_by = "CD3_Sm152_mean",
  bcg = list(CD3_Sm152_mean = c(0, 1, 1)),
  legend = NULL,
  return_plot = TRUE
)
save_plot(p_CD3, "ROI_dense_tcell_overview_CD3.png", width = 12, height = 10)


# Celltype
p_celltype <- plotCells(
  masks,
  object = tmp,
  cell_id = "ObjectNumber",
  img_id = "sample_id",
  colour_by = "celltype",
  colour = list(celltype = metadata(spe)$color_vectors$col_celltype),
  missing_colour = "white",
  return_plot = TRUE
)
save_plot(p_celltype, "ROI_dense_tcell_overview_celltype.png", width = 12, height = 10)

```


## Save object

```{r save-objects}

saveRDS(spe, file.path(params$output, "spe_final_clustering.rds"))

```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
