---
title: "Cell phenotyping"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/04_phenotyping")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# create phenograph heatmap folder
outdir <- paste0(plot_dir,"/pg_heatmap")
dir.create(outdir, showWarnings = FALSE)


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Cell phenotyping {#phenotyping}
In this Rmd file, we will perform 
1. dimensionality reduction
2. phenograph clustering
3. annotate and merge clusters & metaclusters externally 
4. retrieve tumor cells from GD2-MPO double positive cells (caused by spillover) 
5. re-classify granulocyte metaclusters into maturity states based on morphology
6. define color codes for all celltypes and metaclusters

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(ComplexHeatmap)
library(RColorBrewer)
library(viridis)
library(scater)
library(BiocParallel)
library(Rphenoannoy)
library(tidyr)
library(tidyHeatmap)
library(bruceR)
library(dittoSeq)
library(dplyr)
library(mclust)
library(SingleR)
library(circlize)
library(readxl)
library(writexl)
library(ggplot2)
library(patchwork)
library(harmony)
library(batchelor)
library(cluster)
```
</details>

## Read data
We will first read the data that passed the [second quality control step](#qc-2) and save the count data as a matrix. Low-quality markers (low SNR in a lot of samples) as well as all IF and nuclear IMC markers will be excluded for clustering. Moreover, mean intensity will be used to represent marker expression, to average out the signals arising from lateral spillover or artifacts. As explained in [the previous chapter](#qc2-save), we are using counts measured with respect to dilated masks in downstream analysis do capture membrane signal. 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_qc2.rds"))

#computing matrix
intensity_mat <- assay(spe, "counts")
colnames(intensity_mat) <- colnames(spe)

#use mean intensity
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]

#exclude low-quality markers and functional marker cl.PARP
intensity_mat_hq <- intensity_mat[!grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(intensity_mat)),]
intensity_mat_lq <- intensity_mat[grepl("PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(intensity_mat)),]

non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)

intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]
intensity_mat_lq_wo_zeros <- intensity_mat_lq[,non_zeros]

#Apply changes to SPE object
spe <- spe[,non_zeros]
rowData(spe)$use_channel <- !grepl("mean-80|PARP|IDO|DAPI|IF2_GD2|IF3_CD45_CD56|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))

##Matrix with morph. features
morph_mat_clipped <- (cbind("area"=spe$area_clipped, "solidity"=spe$solidity_clipped, "aspect_ratio"=spe$aspect_ratio_clipped))

tissue <- spe$tissue
cell_id <- colnames(spe)

mat <- intensity_mat_hq_wo_zeros

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_sample = setNames(sample(col_vector_433, length(unique(spe$sample))), unique(spe$sample))
```
We are excluding `r length(non_zeros[non_zeros==FALSE])` more cells here, as they are negative for all markers used in downstream analysis. 
To get a first grasp of the single-cell data and potential clusters, we will plot it in a single-cell heatmap.
spe_filtered <- spe_sub[, colData(spe_sub)$DNAIntensity >= low_dna_threshold]


## UMAP (for visualization only)
```{r umap}
set.seed(221228)

Sys.setenv(OMP_NUM_THREADS = 1)
Sys.setenv(MKL_NUM_THREADS = 1)

BPPARAM <- SnowParam(workers = 30, type = "SOCK", RNGseed = 221228)

mat_red <- calculateUMAP(
  mat,
  n_neighbors = 30,
  BPPARAM = BPPARAM
)

rownames(mat_red) <- colnames(spe)
reducedDim(spe, "UMAP") <- mat_red

p <- dittoDimPlot(spe, var = "fm_id", 
             reduction.use = "UMAP", size = 1,
             do.label = TRUE,
             legend.show=F) +
  theme(legend.title = element_blank()) +
  ggtitle("Sample ID on UMAP")


save_plot(p2, "UMAP-vis-sampleID.png", width = 8, height = 6)

p

```



## Phenograph clustering on normalized data

```{r phenograph-clustering}
set.seed(221228)

k_values <- c(15, 30, 45, 60)

for (k in k_values) {

  # Run Phenograph
  Rphenograph_out <- Rphenoannoy(t(mat), k = k)

  # Extract clusters
  clusters <- factor(membership(Rphenograph_out[[2]]))

  # Add to SPE object
  colData(spe)[[paste0("pg_clusters_", k)]] <- clusters
}
```

## UMAP: Phenograph clusters
Lets visualize phenograph clusters and tissue types in UMAP space. Here `dittoDimPlot` can be used directly on the SPE object.

```{r UMAP-vis-clusters}
cluster_vars <- c(
  "pg_clusters_15",
  "pg_clusters_30",
  "pg_clusters_45",
  "pg_clusters_60"
)

color_seed_base <- 221228

get_cluster_levels <- function(values) {
  cluster_levels <- unique(as.character(values))
  cluster_levels <- cluster_levels[!is.na(cluster_levels)]
  
  if (all(grepl("^\\d+$", cluster_levels))) {
    cluster_levels <- as.character(sort(as.integer(cluster_levels)))
  } else {
    cluster_levels <- sort(cluster_levels)
  }
  
  cluster_levels
}

generate_cluster_colors <- function(cluster_levels, seed = color_seed_base) {
  col_vector_433 <- colors()[
    grep("gr(a|e)y", grDevices::colors(), invert = TRUE)
  ]
  
  set.seed(seed)
  setNames(sample(col_vector_433, length(cluster_levels)), cluster_levels)
}

if (is.null(metadata(spe)$color_vectors)) {
  metadata(spe)$color_vectors <- list()
}

for (var_name in cluster_vars) {
  cluster_levels <- get_cluster_levels(spe[[var_name]])
  color_key <- paste0("col_", var_name)
  
  if (is.null(metadata(spe)$color_vectors[[color_key]]) ||
      !all(cluster_levels %in% names(metadata(spe)$color_vectors[[color_key]]))) {
    metadata(spe)$color_vectors[[color_key]] <- generate_cluster_colors(
      cluster_levels = cluster_levels,
      seed = color_seed_base + match(var_name, cluster_vars)
    )
  }
  
  col_clusters <- metadata(spe)$color_vectors[[color_key]][cluster_levels]

  # Create plot
  p <- dittoDimPlot(
    spe,
    var = var_name,
    reduction.use = "UMAP",
    size = 0.5,
    do.label = TRUE,
    legend.show = FALSE,
    color.panel = col_clusters
  ) +
    theme(
      legend.title = element_blank()
    ) +
    ggtitle(paste(var_name, "on UMAP"))
    
  # Find label layer
  label_layer <- which(
    sapply(
      p$layers,
      function(x) class(x$geom)[1]
    ) %in% c("GeomLabelRepel", "GeomLabel")
  )
  
  # Get label data (one row per cluster)
  label_data <- layer_data(p, label_layer)
  
  # Match colors to labels
  label_fills <- col_clusters[
    as.character(label_data$label)
  ]
  
  # Set label appearance
  p$layers[[label_layer]]$aes_params$fill <- label_fills
  p$layers[[label_layer]]$aes_params$colour <- "black"

  
  p <- p +
    theme(
      legend.title = element_blank()
    ) +
    ggtitle(paste(var_name, "on UMAP"))

  # Save plot
  save_plot(
    p,
    paste0("umap_", var_name, ".png"),
    width = 8,
    height = 6
  )

  print(p)
}
```

## Subclustering of clusters 6, 18, and 31 using GMM

```{r gmm-subclustering}
library(mclust)

set.seed(221228)

spe$pg_clusters_beforesubclustering <- spe$pg_clusters_30


spe$pg_clusters <- spe$pg_clusters_beforesubclustering

# Ensure pg_clusters is character for easy reassignment
spe$pg_clusters <- as.character(spe$pg_clusters)

# Mapping: original cluster -> new subcluster label for the second component
subcluster_map <- list(
  "7" = "36",
  "8" = "37",
  "14"  = "38",
  "17" = "39"
)

# define marker sets per cluster
marker_map <- list(
  "7"  = c("CD3_"),
  "8" = c("GD2"),
  "14" = c("CD4_", "CD3_"),
  "17" = c("CD3_",  "CD10")
)

for (cl in names(subcluster_map)) {
  new_cl <- subcluster_map[[cl]]
  
  idx <- which(spe$pg_clusters == cl)
  if (length(idx) < 4) next
  
  sub_mat <- t(mat[, idx])  # cells x markers
  
  # --- select cluster-specific markers ---
  if (cl %in% names(marker_map)) {
    markers <- marker_map[[cl]]
    keep_cols <- grep(paste0("^(", paste(markers, collapse="|"), ")"), colnames(sub_mat))
    sub_mat <- sub_mat[, keep_cols, drop = FALSE]
  }
  
  print(cl)
  print(head(sub_mat))
  
  # --- clean data ---
  sub_mat <- sub_mat[complete.cases(sub_mat), , drop = FALSE]
  sub_mat <- sub_mat[, apply(sub_mat, 2, var) > 0, drop = FALSE]
  print(cl)
  print(head(sub_mat))
  
  #if (nrow(sub_mat) < 4 || ncol(sub_mat) < 2) next
  
  sub_mat <- scale(sub_mat)
  
  # --- GMM ---
  gmm_fit <- tryCatch(
    Mclust(sub_mat, G = 2, verbose = FALSE),
    error = function(e) NULL
  )
  
  if (is.null(gmm_fit)) next
  
  comp2 <- which(gmm_fit$classification == 2)
  spe$pg_clusters[idx[comp2]] <- new_cl
}
```

## Single-cell heatmaps for subclustered clusters (6/42, 18/43, 31/44)

```{r subcluster-heatmaps}
library(dittoSeq)
library(ggplot2)
library(viridis)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]

# Define the pairs: original cluster and its new sibling
subcluster_pairs <- list(
  list(orig = "7",  new = "36"),
  list(orig = "8", new = "37"),
  list(orig = "14", new = "38"),
  list(orig = "17", new = "39")
)

set.seed(221228)
for (pair in subcluster_pairs) {
  orig_cl <- pair$orig
  new_cl  <- pair$new

  # Cells belonging to either subcluster
  idx <- which(spe$pg_clusters %in% c(orig_cl, new_cl))

  if (length(idx) == 0) next

  # Sample up to 8000 cells
  n_cells   <- min(8000, length(idx))
  rand_idx  <- sample(idx, n_cells)

  spe_cl <- spe[, rand_idx]

  # Order cells by subcluster so the two groups appear side-by-side
  cell_order <- order(spe_cl$pg_clusters)

  p1 <- dittoHeatmap(
    spe_cl[, cell_order],
    genes          = genes_use,
    cluster_cols   = FALSE,          # preserve subcluster ordering
    scaled.to.max  = TRUE,
    heatmap.colors.max.scaled = viridis(100),
    annot.by       = "pg_clusters"   # colour bar showing subclusters
  )

  save_plot(
    p1,
    paste0("heatmap_subcluster_", orig_cl, "_", new_cl, ".png"),
    width  = 14,
    height = 6
  )
}
```

# For later annotation, we generate one clustered heatmap per sample

```{r}
library(dittoSeq)
library(ggplot2)

# get cluster IDs
clusters <- unique(colData(spe)$pg_clusters)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]


for (cl in clusters) {
  
  # subset cells for this cluster
  spe_cl <- spe[, colData(spe)$pg_clusters == cl]
  
  # skip empty clusters (just in case)
  if (ncol(spe_cl) == 0) next
  
  # sample up to 8000 cells
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  # make heatmap
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  # save plot
  save_plot(
    p1,
    paste0("heatmap_phenocluster_", cl, ".png"),
    width = 14,
    height = 6
  )
}

```

## Bubble plot: Mean marker expression per cluster
Let's plot the mean marker expressions per cluster in a bubble plot.

```{r vis-data-bubbleplot, fig.height=14, fig.width=16}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(viridis)
library(cowplot)

cluster_vars <- c(
  "pg_clusters_15",
  "pg_clusters_30",
  "pg_clusters_45",
  "pg_clusters_60"
)

plot_cluster_bubble <- function(cluster_col, save_name) {
  barplot_height <- unit(1.4, "cm")
  
  # ----------------------------
  # Cluster vector
  # ----------------------------
  cluster_vec <- as.character(colData(spe)[[cluster_col]])
  tissue_vec  <- colData(spe)$tissue

  # ----------------------------
  # Expression matrix
  # ----------------------------
  mat_df <- as.data.frame(t(mat))

  exp_mat_ <- aggregate(
    mat_df,
    by = list(cluster = cluster_vec),
    FUN = function(x) {
      if (all(x == 0)) return(0)
      median(log1p(x[x > 0]))
    }
  )

  cluster_names <- exp_mat_$cluster

  exp_mat <- exp_mat_[, -1, drop = FALSE]
  rownames(exp_mat) <- cluster_names
  exp_mat <- t(as.matrix(exp_mat))

  # ----------------------------
  # Row scaling
  # ----------------------------
  rmin <- apply(exp_mat, 1, min, na.rm = TRUE)
  rmax <- apply(exp_mat, 1, max, na.rm = TRUE)
  den  <- rmax - rmin

  exp_mat <- (exp_mat - rmin) / den

  exp_mat[is.nan(exp_mat)] <- 0
  exp_mat[is.infinite(exp_mat)] <- 0

  clusters <- colnames(exp_mat)

  # ----------------------------
  # Percent expressed
  # ----------------------------
  percent_mat <- sapply(seq_len(nrow(mat)), function(i) {

    row_vals <- mat[i, ]
    pos <- row_vals[row_vals > 0]

    if (length(pos) == 0) {

      out <- rep(0, length(clusters))
      names(out) <- clusters
      return(out)
    }

    thresh <- quantile(pos, 0.2, na.rm = TRUE)

    row_bin <- ifelse(row_vals > thresh, 1, 0)

    tapply(
      row_bin,
      factor(cluster_vec, levels = clusters),
      mean
    )
  })

  percent_mat <- t(percent_mat)

  percent_mat <- percent_mat[, clusters, drop = FALSE]

  percent_mat[is.na(percent_mat)] <- 0
  percent_mat[percent_mat < 0.1] <- 0.1

  # ----------------------------
  # Tissue counts
  # ----------------------------
  bm_count <- table(factor(cluster_vec[tissue_vec == "BM"], levels = clusters))
  pt_count <- table(factor(cluster_vec[tissue_vec == "PT"], levels = clusters))

  # ----------------------------
  # Axis helper
  # ----------------------------
  axis_10k_ticks <- function(v) {
    vmax <- max(as.numeric(v), na.rm = TRUE)
    upper <- max(10000, ceiling(vmax / 10000) * 10000)
    at <- seq(10000, upper, by = 10000)

    list(
      upper = upper,
      axis_param = list(
        at = at,
        labels = paste0(at / 1000, "k"),
        gp = gpar(fontsize = 12)
      )
    )
  }
  
  # ----------------------------
  # Cluster colors
  # ----------------------------
  color_key <- paste0("col_", cluster_col)
  
  if (is.null(metadata(spe)$color_vectors)) {
    metadata(spe)$color_vectors <- list()
  }
  
  if (is.null(metadata(spe)$color_vectors[[color_key]]) ||
      !all(clusters %in% names(metadata(spe)$color_vectors[[color_key]]))) {
    seed_offset <- match(cluster_col, cluster_vars)
    if (is.na(seed_offset)) seed_offset <- 0
    
    metadata(spe)$color_vectors[[color_key]] <- generate_cluster_colors(
      cluster_levels = clusters,
      seed = color_seed_base + seed_offset
    )
  }
  
  col_clusters <- metadata(spe)$color_vectors[[color_key]][clusters]
  pt_axis <- axis_10k_ticks(pt_count)
  bm_axis <- axis_10k_ticks(bm_count)
  cluster_anno <- setNames(list(factor(clusters, levels = clusters)), cluster_col)
  
  # ----------------------------
  # Heatmap colors
  # ----------------------------
  col_fun <- circlize::colorRamp2(
    c(0, 0.5, 1),
    viridis(100)[c(1, 50, 100)]
  )

  # ----------------------------
  # Column annotation
  # ----------------------------
  column_ha <- do.call(
    HeatmapAnnotation,
    c(
      cluster_anno,
      list(
        ncells_PT = anno_barplot(
          as.numeric(pt_count),
          ylim = c(0, pt_axis$upper),
          height = barplot_height,
          gp = gpar(fill = "brown"),
          axis_param = pt_axis$axis_param
        ),
        ncells_BM = anno_barplot(
          as.numeric(bm_count),
          ylim = c(0, bm_axis$upper),
          height = barplot_height,
          gp = gpar(fill = "aquamarine4"),
          axis_param = bm_axis$axis_param
        ),
        col = setNames(list(col_clusters), cluster_col),
        show_legend = TRUE
      )
    )
  )

  # ----------------------------
  # Bubble cell function
  # ----------------------------
  cell_fun <- function(j, i, x, y, w, h, fill) {

    grid.rect(
      x, y, w, h,
      gp = gpar(col = NA, fill = NA)
    )

    grid.circle(
      x = x,
      y = y,
      r = percent_mat[i, j] / 2 * min(unit.c(w, h)),
      gp = gpar(
        fill = col_fun(exp_mat[i, j]),
        col = NA
      )
    )
  }

  # ----------------------------
  # Bubble legend
  # ----------------------------
  lgd_list <- list(

    Legend(
      labels = c(0, 0.25, 0.5, 0.75, 1),
      title = "percentage expressed",

      graphics = lapply(
        c(0, 0.25, 0.5, 0.75, 1),
        function(r) {

          function(x, y, w, h)
            grid.circle(
              x = x,
              y = y,
              r = r * unit(2, "mm"),
              gp = gpar(fill = "black")
            )
        }
      )
    )
  )

  # ----------------------------
  # Heatmap
  # ----------------------------
  hp <- Heatmap(
    exp_mat,

    heatmap_legend_param = list(title = "expression"),

    column_title = paste0(
      "Bubble plot - ",
      cluster_col
    ),

    top_annotation = column_ha,

    col = col_fun,

    rect_gp = gpar(type = "none"),

    clustering_distance_columns = "pearson",

    clustering_method_columns = "average",

    row_labels = sapply(
      strsplit(rownames(exp_mat), "_"),
      "[[",
      1
    ),

    cell_fun = cell_fun,

    show_row_dend = FALSE,

    column_dend_height = unit(1.5, "cm"),

    row_names_gp = gpar(fontsize = 15),

    column_names_gp = gpar(fontsize = 15),

    column_names_rot = 45,

    border = "black"
  )

  hp_drawn <- draw(
    hp,
    annotation_legend_list = lgd_list
  )
  
  save_plot(
    hp_drawn, 
    save_name,
    width = 16, 
    height = 14
  )

  hp_drawn
}

for (cluster_col in cluster_vars) {
  message("Processing: ", cluster_col)
  print(
    plot_cluster_bubble(
      cluster_col = cluster_col,
      save_name = paste0("bubbleplot_", cluster_col, ".pdf")
    )
  )
}
```

## Define final clustering and plot clustered heatmaps for each cluster

```{r}
library(dittoSeq)
library(ggplot2)


# Define final clustering set
spe$pg_clusters <- spe$pg_clusters_30

# get cluster IDs
clusters <- unique(colData(spe)$pg_clusters)

genes_use <- rownames(spe)[!grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", rownames(spe))]


for (cl in clusters) {
  
  # subset cells for this cluster
  spe_cl <- spe[, colData(spe)$pg_clusters == cl]
  
  # skip empty clusters (just in case)
  if (ncol(spe_cl) == 0) next
  
  # sample up to 8000 cells
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  # make heatmap
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  # save plot
  save_plot(
    p1,
    paste0("heatmap_phenocluster_", cl, ".png"),
    width = 14,
    height = 6
  )
}

```


## Write cluster proportions for manual annotation (Excel)

```{r cluster-annotation-excel}

# Ensure cluster is character
spe$pg_clusters <- as.character(spe$pg_clusters)

# -----------------------------
# 1. Absolute counts
# -----------------------------
count_table <- as.data.frame(table(
  cluster = spe$pg_clusters,
  tissue  = spe$tissue
))

count_wide <- count_table %>%
  pivot_wider(
    names_from = tissue,
    values_from = Freq,
    values_fill = 0
  )

# Ensure all tissues exist
for(t in c("BM","PT")){
  if(!t %in% colnames(count_wide)){
    count_wide[[t]] <- 0
  }
}

count_wide <- count_wide %>%
  select(cluster, BM, PT)

# -----------------------------
# 2. Percentages WITHIN cluster
# -----------------------------
percent_within_cluster <- count_wide %>%
  rowwise() %>%
  mutate(
    total = BM + PT,
    BM_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * BM / total), "0.0%"),
    PT_percent = ifelse(total > 0, sprintf("%.1f%%", 100 * PT / total), "0.0%")
  ) %>%
  ungroup() %>%
  select(-total)

# -----------------------------
# 3. Add annotation columns
# -----------------------------
annotation_template <- percent_within_cluster %>%
  mutate(
    celltype = "",
    metacluster = "",
    comments = ""
  ) %>%
  arrange(as.numeric(cluster))

# -----------------------------
# 4. Write Excel
# -----------------------------
write_xlsx(
  annotation_template,
  file.path(params$output, "cluster_annotation_template.xlsx")
)
```

```{r stop-here}
knitr::opts_chunk$set(eval = FALSE)
```

## Annotation and merging of clusters

Next, we read assigned celltype and metacluster-labels for clusters from the metadata file provided on Zenodo [(Supplementary_Tables/20240808_Suppl_Table3.xlsx)](https://doi.org/10.5281/zenodo.13220634 ). These have been assigned based on thorough examination of images and co-expressions and using literature.
Let's start with celltype assignments...

```{r annotations-celltypes}
file_path <- file.path('/home/rstudio/data/output', 'cluster_annotation_raw_k30_subclustering.xlsx')
annotation <- read_excel(file_path, sheet = "Sheet1")

cluster_names <- setNames(annotation$celltype, annotation$cluster)
celltypes <- recode(spe$pg_clusters, !!!cluster_names)
spe$celltype <-  as.vector(celltypes)

metacluster_names <- setNames(annotation$metacluster, annotation$cluster)
metacelltypes <- recode(spe$pg_clusters, !!!metacluster_names)
spe$metacluster <-  as.vector(metacelltypes)

```

## Assign colors 
Finally, we assign colors to all celltypes and metaclusters.

```{r colors-celltypes raw_k30}
col_celltype <- c(

  ## ---- Tumor (unchanged) ----
  "GATA3hi Tu"    = rgb(0,   90,  50, maxColorValue = 255),
  "CXCR4hi Tu"    = rgb(120, 180, 40, maxColorValue = 255),
  "CD44+ Tu"      = rgb(65,  208, 57, maxColorValue = 255),
  "GD2lo Tu"      = rgb(40,  150,130, maxColorValue = 255),
  "Ki67hi Tu"     = rgb(173, 255, 47, maxColorValue = 255),
  "early NB Tu"   = rgb(150, 220,170, maxColorValue = 255),
  "marker-lo Tu"  = rgb(40,  120, 70, maxColorValue = 255),

  ## ---- Proliferation ----
  "Ki67+ cells"        = rgb(255, 200,   0, maxColorValue = 255),
  "Ki67+ CXCR4+ cells" = rgb(255, 245, 160, maxColorValue = 255),

  ## ---- Progenitor ----
  "MSCs"          = rgb(255, 165,  60, maxColorValue = 255),
  "CD34+ HSPCs"   = rgb(255,  95,   0, maxColorValue = 255),

  ## ---- T cells ----
  ## ---- T cells (rebalanced distinct reds) ----
  
  "CD8+ naïve T cells"   = rgb(220, 20, 60,  maxColorValue = 255),  # crimson (clean bright red)
  
  "CD4+ naïve T cells"   = rgb(232, 80, 70, maxColorValue = 255),  # firebrick (deep structured red)
  
  "CD8+ S100B+ T cells"   = rgb(199, 21, 133, maxColorValue = 255),  # medium violet-red (pinkish red, distinct hue shift)
    
  "CD4+ PD1+ T cells" = rgb(123, 0, 60,  maxColorValue = 255),
  
  "GZMB+ T cells" = rgb(220, 90, 40, maxColorValue = 255),  # true deep red (kept warm red identity)
  
  "dense T cell region"  = rgb(170, 30, 120, maxColorValue = 255),   # dark crimson (slightly different tone than GZMB)
  ## ---- Myeloid (UPDATED: clearly distinct blues) ----
  "granulocytes"      = rgb(8, 48,107, maxColorValue = 255),    # deep navy
  "CD14+ PDL1- MO"    = rgb(33,102,172, maxColorValue = 255),   # strong blue
  "CD14+ PDL1+ MO"    = rgb(67,147,195, maxColorValue = 255),   # lighter sky blue
  "monoblast-like"    = rgb(64, 180, 170, maxColorValue = 255),  # very light blue
  "GZMB+ DC/NK"       = rgb(107,174,214, maxColorValue = 255),  # mid-light blue
  "pDC"               = rgb(198,219,239, maxColorValue = 255),  # pale blue

  ## ---- B cells ----
  "B cells" = rgb(136,86,167, maxColorValue = 255),

  ## ---- Stromal (renamed) ----
  "schwann cells"        = rgb(166,97,26, maxColorValue = 255),
  "fibroblast/endothel" = rgb(101, 67, 33, maxColorValue = 255),  # dark brown,

  ## ---- Other ----
  "other" = rgb(120,120,120, maxColorValue = 255)
)

metadata(spe)$color_vectors$col_celltype <- col_celltype
```

```{r colors-metaclusters}
col_metacluster <- c(
  "tumor" = "#238B45",
  "T cells" = "#E34A33",
  "granulocytes" = "#08306B",   # dark blue (matches rgb(8,48,107))
  "MO/DC/NK" = "#41B6C4",
  "B cells" = "#88419D",
  "mesenchymal" = "#8C510A",
  "proliferating cells" = "#FFD700",
  "progenitors" = "#F59E0B",   
  "other" = "#969696"
)
metadata(spe)$color_vectors$col_metacluster <- col_metacluster
```

## Visualizations
For visualization purpose, we will give the celltypes a meaningful order. This order can be downloaded from a csv file from Zenodo [(R_intermediary/celltype_order.csv)](https://doi.org/10.5281/zenodo.13220634 )

```{r order-celltypes}
clusternames_ordered <- read.csv(file.path(params$input, "celltype_order.csv"), sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)
spe$celltype_ordered <- recode(spe$celltype, !!!codes)

metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
names(metadata(spe)$color_vectors$col_celltype_ordered) <- recode(names(metadata(spe)$color_vectors$col_celltype), !!!codes)
```

As a last step, we generate UMAP and bubble plots for the final celltypes/metaclusters.

```{r umap-final-celltypes, fig.width=10}
p1 <- dittoDimPlot(spe, var = "celltype_ordered", 
             reduction.use = "UMAP", size = 0.3, 
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(legend.title = element_blank()) +
  ggtitle("Celltypes on UMAP")

p2 <- dittoDimPlot(spe, var = "metacluster", 
             reduction.use = "UMAP", size = 0.3, labels.size=4,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(spe)$color_vectors$col_metacluster) +
  theme(legend.title = element_blank()) +
  ggtitle("Metaclusters on UMAP")

# Save both
save_plot(p1, "umap-celltypes.png", width = 16, height = 8)
save_plot(p2, "umap-metaclusters.png", width = 13, height = 8)

p1
```

```{r vis-metacluster}
p2
```


```{r bubbleplot-final-celltypes, fig.height=10, fig.width=12}
# -----------------------------
# Expression matrix
# -----------------------------

mat_df <- as.data.frame(t(mat))

exp_mat_ <- aggregate(
  mat_df,
  by = list(cluster = spe$celltype),
  FUN = function(x) {
    if (all(x == 0)) return(0)
    median(log1p(x[x > 0]))
  }
)

cluster_names <- exp_mat_$cluster
exp_mat <- exp_mat_[, -1, drop = FALSE]
rownames(exp_mat) <- cluster_names
exp_mat <- t(as.matrix(exp_mat))

# Row-wise scaling (per marker)
rmin <- apply(exp_mat, 1, min, na.rm = TRUE)
rmax <- apply(exp_mat, 1, max, na.rm = TRUE)
den  <- rmax - rmin

exp_mat <- (exp_mat - rmin) / den
exp_mat[is.nan(exp_mat)] <- 0
exp_mat[is.infinite(exp_mat)] <- 0


# -----------------------------
# Percentage expressed
# -----------------------------

clusters <- colnames(exp_mat)

percent_mat <- sapply(seq_len(nrow(mat)), function(i) {
  row_vals <- mat[i, ]
  pos <- row_vals[row_vals > 0]

  if (length(pos) == 0) {
    out <- rep(0, length(clusters))
    names(out) <- clusters
    return(out)
  }

  thresh <- quantile(pos, 0.2, na.rm = TRUE)
  row_bin <- ifelse(row_vals > thresh, 1, 0)

  tapply(row_bin, factor(spe$celltype, levels = clusters), mean)
})

percent_mat <- t(percent_mat)
percent_mat <- percent_mat[, clusters, drop = FALSE]
percent_mat[is.na(percent_mat)] <- 0

# Optional visualization floor
percent_mat[percent_mat < 0.1] <- 0.1

# Annotation bars
bm_count <- table(spe[, spe$tissue == "BM"]$celltype)
bm_count[setdiff(colnames(exp_mat), names(bm_count))] <- 0
bm_count <- bm_count[colnames(exp_mat)]

pt_count <- table(spe[, spe$tissue == "PT"]$celltype)
pt_count[setdiff(colnames(exp_mat), names(pt_count))] <- 0
pt_count <- pt_count[colnames(exp_mat)]

mean_metaclust <- aggregate(spe$metacluster, list(spe$celltype), max)

# Barplot axis: show only the max tick (no 0 tick)
axis_maxtick <- function(v) {
  vmax <- max(as.numeric(v), na.rm = TRUE)
  list(
    at = vmax,
    labels = as.character(vmax),
    gp = gpar(fontsize = 12)
  )
}

column_ha = HeatmapAnnotation(
  metacluster = mean_metaclust$x,
  
  ncells_PT = anno_barplot(
    unname(pt_count),
    add_numbers = FALSE,
    height = unit(0.7, "cm"),
    gp = gpar(fill = "brown"),
    axis_param = axis_maxtick(pt_count)
  ),
  
  ncells_BM = anno_barplot(
    unname(bm_count),
    add_numbers = FALSE,
    height = unit(0.7, "cm"),
    gp = gpar(fill = "aquamarine4"),
    axis_param = axis_maxtick(bm_count)
  ),
  
  col = list(
    metacluster = metadata(spe)$color_vectors$col_metacluster
  ),
  
  show_legend = TRUE
)

#Function to set dot size based on percentage expressed
cell_fun = function(j, i, x, y, w, h, fill){
          grid.rect(x = x, y = y, width = w, height = h, 
                    gp = gpar(col = NA, fill = NA))
          grid.circle(x=x,y=y,r= percent_mat[i, j]/2 * min(unit.c(w, h)),
                      gp = gpar(fill = col_fun(exp_mat[i, j]), col = NA))}

#Function to set dot color based on mean expression
col_fun = circlize::colorRamp2(c(min(exp_mat),(max(exp_mat)-min(exp_mat))/2, max(exp_mat)), viridis(100)[c(1,50,100)])

#Function to set legend size according to percentage expressed
layer_fun = function(j, i, x, y, w, h, fill){
          grid.rect(x = x, y = y, width = w, height = h, 
                    gp = gpar(col = NA, fill = NA))
          grid.circle(x=x,y=y,r= pindex(percent_mat, i, j)/1 * unit(1.5, "mm"),
                      gp = gpar(fill = col_fun(pindex(exp_mat, i, j)), col = NA))}

#Generate legend for percentage expressed
lgd_list = list(
    Legend( labels = c(0,0.25,0.5,0.75,1), title = "percentage expressed",
            graphics = list(
              function(x, y, w, h) grid.circle(x = x, y = y, r = 0 * unit(2, "mm"),
                                               gp = gpar(fill = "black")),
              function(x, y, w, h) grid.circle(x = x, y = y, r = 0.25 * unit(2, "mm"),
                                               gp = gpar(fill = "black")),
              function(x, y, w, h) grid.circle(x = x, y = y, r = 0.5 * unit(2, "mm"),
                                               gp = gpar(fill = "black")),
              function(x, y, w, h) grid.circle(x = x, y = y, r = 0.75 * unit(2, "mm"),
                                               gp = gpar(fill = "black")),
              function(x, y, w, h) grid.circle(x = x, y = y, r = 1 * unit(2, "mm"),
                                               gp = gpar(fill = "black")))
            ))

#Heatmap
hp <- Heatmap(exp_mat,
        heatmap_legend_param=list(title="expression"),
        column_title = "Bubble plot - normalized mean expression and %expressed", 
        top_annotation = column_ha, 
        col=col_fun,
        rect_gp = gpar(type = "none"),
        clustering_distance_columns="pearson", 
        clustering_method_columns="average", 
        layer_fun = layer_fun,
        row_labels=sapply(strsplit(rownames(exp_mat), "_"), "[[", 1),
        cell_fun = cell_fun,
        show_row_dend = F,
        column_dend_height = unit(1.5, "cm"),
        row_names_gp = gpar(fontsize = 10),
        column_names_gp = gpar(fontsize = 10),
        column_names_rot = 45,
        border = "black")

# Save and draw
hp_drawn <- draw(hp, annotation_legend_list = lgd_list)
save_plot(hp_drawn, "bubbleplot-celltypes_raw_k30.pdf", width = 12, height = 10)
hp_drawn
```

Let's plot IF GD2 by cluster as a validation:


```{r}
library(dittoSeq)
library(ggplot2)
library(viridis)

# Ensure output directory exists
outdir <- file.path(plot_dir)
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

# Get cluster IDs
clusters <- unique(colData(spe)$celltype)

# Select genes to use (exclude certain markers)
genes_use <- rownames(spe)[
  !grepl("-80|PARP|IDO|IF2_GD2|IF3_CD45_CD56|DAPI|DNA|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro|FOXP3", 
         rownames(spe))
]

for (cl in clusters) {
  
  # --- 1. Make cluster name safe for filenames ---
  safe_name <- cl
  safe_name <- gsub("\\+", "plus", safe_name)
  safe_name <- gsub("-", "minus", safe_name)
  safe_name <- gsub("[^A-Za-z0-9_]", "_", safe_name)  # replaces /, spaces, special chars
  
  # --- 2. Subset cells for this cluster ---
  spe_cl <- spe[, colData(spe)$celltype == cl]
  
  # Skip empty clusters
  if (ncol(spe_cl) == 0) next
  
  # --- 3. Sample cells ---
  n_cells <- min(8000, ncol(spe_cl))
  rand_cells <- sample(seq_len(ncol(spe_cl)), n_cells)
  
  # --- 4. Create heatmap ---
  p1 <- dittoHeatmap(
    spe_cl[, rand_cells],
    genes = genes_use,
    cluster_cols = TRUE,
    scaled.to.max = TRUE,
    heatmap.colors.max.scaled = viridis(100)
  )
  
  # --- 5. Save heatmap safely ---
  filename <- paste0("heatmap_celltype_", safe_name, ".png")
  full_path <- file.path(outdir, filename)
  
  # If save_plot() is your custom function
  save_plot(p1, filename = filename, width = 14, height = 6)
  
  # Or using ggsave directly:
  # ggsave(full_path, plot = p1, width = 14, height = 6, dpi = 300)
}
```


## Save object

```{r save-objects}

saveRDS(spe, file.path(params$output, "spe_final_clustering.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
