---
title: "Identifying cellular communities in diagnostic primary tumors (all cell types)"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/05.1_PT_DE_cellularcomm")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {

    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)

  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {

    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()

  } else {

    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }

  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Identifying cellular communities in diagnostic primary tumors (all cell types) {#pt-dx-allct}
Here we identify cellular communities by clustering diagnostic PT samples on the composition of **all cell types** (tumor and microenvironment together). The combined figure shows the all-cell-type abundance on top, the tumor-only abundance in place of the metacluster panel, and the immune abundance below.

## Load libraries
<details>
   <summary>Load libraries</summary>

```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(gridExtra)
library(ggpubr)
install.packages("ggdendro")
library(ggdendro)
library(patchwork)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping).

```{r read-data}
# Start from the phenotyped clustering object produced by 04_phenotyping.Rmd.
spe <- readRDS(file.path(params$output, "spe_04_phenotyping.rds"))
```

We will then transform the SPE object into a catalyst-compatible object.

```{r catalyst-comp-object}
keep_rows <- rowData(spe)$use_channel
keep_cols <- spe$celltype != "other" & spe$tissue=="PT" & (spe$timepoint=="DE")

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id, "_", sce$tissue))

#recode metadata on 1p loss
sce$X1p_loss <- factor(spe[keep_rows, keep_cols]$X1p_loss)
X1p_loss <- recode(sce$X1p_loss, `1` = 'yes', `2` = 'het', `0` = 'no', `9` = 'unknown')
sce$X1p_loss <- as.factor(X1p_loss)

sce$age_diag <- spe[keep_rows, keep_cols]$age_diag

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)

sce$condition <- 'na'

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))
```

## Cluster cellular communities

Next, we cluster proportions of **all cell types** (tumor and microenvironment, including the `other` category) across all diagnostic PT samples (instead of conditions). The composition is built directly from raw cell counts (`n / sum(n)` per sample). Because cell-type proportions are compositional data (they sum to 1 per sample), we apply a centered log-ratio (CLR) transformation before clustering the samples with Euclidean distance and Ward.D2 linkage, and plot the result in barplots:

```{r cohort-overview}
set.seed(20231103)

distance <- "euclidean"
linkage <- "ward.D2"
n_communities <- 2

# Pseudocount (in "cells") added before the CLR transform to handle zero
# proportions. It is scaled per sample by the number of cells, so it always
# represents the same fractional cell regardless of sample depth.
CLR_PSEUDO <- 0.5

# Tumor subset is kept for the tumor-only abundance panel and heterogeneity below
sce_tumor <- sce[,sce$metacluster=="tumor"]

# ---- Build the all-cell-type composition DIRECTLY from cell counts ----
# The clustering input is built from raw cell counts (NOT plotAbundances()$Freq,
# which CATALYST returns as a percentage 0-100) over ALL cell types INCLUDING
# "other", per sample. Proportions are n / sum(n); a per-sample pseudocount
# (CLR_PSEUDO cells / total cells in the sample) is then added on the same 0-1
# scale before the CLR transform. This mirrors the reference cluster_patients()
# recipe and keeps the pseudocount on the correct (proportion) scale.
md_clust <- as.data.frame(colData(spe)) %>%
  filter(tissue == "PT", timepoint == "DE") %>%             # keep "other" cells
  mutate(
    progress  = recode(factor(progression), `0` = "NoProg", `1` = "Prog"),
    mna       = recode(factor(MYCN_amp), `1` = "MNA", `2` = "het", `0` = "nMNA", `9` = "unknown"),
    sample_id = paste0(progress, "_", mna, "_", patient_id, "_", fm_id, "_", tissue),
    celltype  = as.character(celltype)
  )

pt_comp <- md_clust %>%
  count(sample_id, celltype, name = "n") %>%
  group_by(sample_id) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  pivot_wider(id_cols = sample_id, names_from = celltype, values_from = prop, values_fill = 0)

cmat <- as.matrix(pt_comp[, setdiff(names(pt_comp), "sample_id")])
rownames(cmat) <- pt_comp$sample_id

# Per-sample pseudocount (CLR_PSEUDO cells / total cells in the sample)
n_total    <- md_clust %>% count(sample_id, name = "n_total")
pseudo_vec <- CLR_PSEUDO / n_total$n_total[match(rownames(cmat), n_total$sample_id)]
cmat_adj   <- sweep(cmat, 1, pseudo_vec, "+")
clr_df     <- t(apply(cmat_adj, 1, function(x) log(x) - mean(log(x))))

# Cluster samples (euclidean distance, ward.D2 linkage)
dist_matrix <- dist(clr_df, method = distance)
hc <- hclust(dist_matrix, method = linkage)

# Order samples by dendrogram leaf order so the barplot follows the CLR clustering
clustered_order <- rownames(clr_df)[hc$order]
sce$sample_id <- factor(as.character(sce$sample_id), levels = clustered_order)

# All-cell-type abundance barplot in clustered order
p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust = FALSE,
                    k_pal = metadata(spe)$color_vectors$col_celltype_ordered)

p

save_plot(p, "Fig3aBis1_PT_abundance_patients_clustered_allcelltypes.pdf", width = 8, height = 6)
```

along with the dendogram:

```{r dendogram, fig.height=8, fig.width=7}
# Dendogram of the CLR-based clustering computed above
plot(hc)

save_plot(plot(hc), "Fig3aBis2_PT_abundance_patients_clustered_dendrogram.pdf", width = 8, height = 6)
```
We then save the order of clustered samples along with cluster labels.

```{r}
clustered_samples <- sapply(strsplit(levels(p$data$sample_id), "_"), function(x) paste(x[5], x[6], sep = "_"))

#Extract clusters of cellular communities
cellular_communities <- as.data.frame(cutree(hc, k = n_communities)[levels(p$data$sample_id)])
colnames(cellular_communities) <- c("community_cluster")
# Renumber cutree labels to consecutive IDs (1..n_communities) for plotting.
cluster_values <- sort(unique(cellular_communities$community_cluster))
cluster_names <- setNames(seq_along(cluster_values), cluster_values)
cellular_communities$community_cluster <- recode(cellular_communities$community_cluster, !!!cluster_names)
community_levels <- sort(unique(cellular_communities$community_cluster))

# Rename rownames to fm_id_tissue format for downstream matching
rownames(cellular_communities) <- sapply(strsplit(as.vector(rownames(cellular_communities)), "_"), function(x) paste(x[length(x) - 1], x[length(x)], sep = "_"))

# Add the all-cell-type-clustering community column to spe. This is the community
# assignment obtained by clustering samples on the composition of all cell types
# (tumor and microenvironment, including "other").
spe_key <- paste0(spe$fm_id, "_", spe$tissue)
spe$allcelltype_community <- cellular_communities[spe_key, "community_cluster"]
```

## Plot tumor cellular profiles

We plot the tumor cell-type profiles for these samples in the clustered order (this panel replaces the metacluster panel in the combined figure below).

```{r plot-tumor, fig.width=12, fig.height=5}
#order tumor subset based on above clustering
sce_tumor$sample_id <- factor(as.character(sce_tumor$sample_id), levels = clustered_order)

p_tumor <- plotAbundances(sce_tumor, k = "celltype", by = "sample", col_clust = FALSE,
                          k_pal = metadata(spe)$color_vectors$col_celltype_ordered)

p_tumor

save_plot(p_tumor, "Fig3aBis_PT_abundance_patients_clustered_tumor.pdf", width = 12, height = 5)
```

## Plot TME cellular profiles

We will then plot the cellular profiles of the TME for these samples (kept as is).

```{r plot-TME, fig.width=12, fig.height=5}
#order samples based on above clustering
sce$sample_id <- factor(paste0(spe[keep_rows, keep_cols]$fm_id, "_", spe[keep_rows, keep_cols]$tissue), levels=clustered_samples)

sce_ME <- sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]

p_ME <- plotAbundances(sce_ME, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_celltype_ordered)

p_ME

save_plot(p_ME, "Fig3aBis_PT_abundance_patients_clustered_immune.pdf", width = 12, height = 5)
```

## Compute tumor and TME heterogeneity

We compute tumor and TME intra-sample heterogeneity.

```{r heterogeneity}
proportions <- plotAbundances(sce, k = "celltype", by = "sample")
prop <- proportions$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)
prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)

me_clusters <- unique(sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]$cluster_id)
me_prop <- prop[rownames(prop)%in%me_clusters,]
me_het <- apply(me_prop,2, entropy)

tumor_clusters <- unique(sce[,sce$metacluster=="tumor"]$cluster_id)
tumor_prop <- prop[rownames(prop)%in%tumor_clusters,]
tumor_het <- apply(tumor_prop,2, entropy)
```

## Plot metadata

Then we plot intra-sample heterogeneity together with sample-specific metadata for the clustered sammples.

```{r plot-metadata}
#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna),
            X1p_loss = first(X1p_loss),
            progress=first(progress),
            age_diag = first(age_diag)
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


#assign color codes
# Sample-annotation colors are defined centrally in 00_MapMet_color_code.R
for (p in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(p)) { source(p); break }
col_code <- COL_CODE

col_age_diag = colorRamp2(c(min(sample_summary$age_diag, na.rm = TRUE), (quantile(sample_summary$age_diag, 0.95, na.rm = TRUE)-min(sample_summary$age_diag, na.rm = TRUE))/2, quantile(sample_summary$age_diag,0.95, na.rm = TRUE)), viridis(3))
col_me_het = colorRamp2(c(min(me_het, na.rm = TRUE), (max(me_het, na.rm = TRUE)-min(me_het, na.rm = TRUE))/2, max(me_het, na.rm = TRUE)), viridis(3))
col_tumor_het = colorRamp2(c(min(tumor_het, na.rm = TRUE), (max(tumor_het, na.rm = TRUE)-min(tumor_het, na.rm = TRUE))/2, max(tumor_het, na.rm = TRUE)), viridis(3))

# Plot metadata
p_anno <- columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(X1p_loss=sample_summary$X1p_loss, col=list(X1p_loss=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(age=sample_summary$age_diag, col=list(age=col_age_diag)) %v%
  columnAnnotation(TUM_het=tumor_het, col=list(TUM_het=col_tumor_het)) %v%
  columnAnnotation(ME_het=me_het, col=list(ME_het=col_me_het))

save_plot(p_anno, "Fig3aBis4_PT_abundance_patients_clustered_metadata.pdf", width = 8, height = 6)

```

## Plot continous and discrete summary statistics

We then plot summary statistics of continuous and discrete metadata in boxplots and piecharts, respectively.

```{r piecharts}
sample_summary[rownames(cellular_communities),"community_clusters"] <- cellular_communities$community_cluster

discrete_metadata <- c("mna", "X1p_loss", "progress")

discrete_metadata_colors <- list(col_code, col_code, col_code)

plot_list <- list()
for (d in seq_along(discrete_metadata)) {
  sample_meta <- sample_summary[,c(discrete_metadata[d], "community_clusters")]
  sample_meta <- sample_meta[rowSums(sample_meta == "unknown") == 0, ]
  colnames(sample_meta) <- c("dis_var", "condition")

  result <- sample_meta %>%
    group_by(condition, dis_var) %>%
    summarise(Freq = n(), .groups = 'drop')

  result[,"dis_var"]<- as.factor(unlist(result[,"dis_var"]))

plot_list[[d]] <- ggplot(result, aes(x="", y=Freq, fill=dis_var, group=dis_var)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = discrete_metadata_colors[[d]], name=discrete_metadata[d]) +
    coord_polar("y", start=0)  +
    facet_grid(.~ condition) +
    theme_void()
}

p_pie <- do.call("grid.arrange", c(plot_list, ncol=1))

save_plot(p_pie, "Fig3bBis_PT_abundance_piecharts_metadata.pdf", width = 6, height = 8)
```
```{r boxplots}
het_m <- as.vector(unname(me_het[clustered_samples]))
het_t <- as.vector(unname(tumor_het[clustered_samples]))

age <- sample_summary$age_diag

cont_df <- data.frame(cbind(het_m, het_t, age, cellular_communities),row.names=clustered_samples)
cont_df$community_cluster <- factor(cont_df$community_cluster, levels = community_levels)
# Keep a stable base palette so the first community colors stay consistent
# even if n_communities changes between analyses.
# Eight provides enough distinct baseline colors for this analysis while keeping
# the first assigned colors stable when fewer communities are used.
base_palette_size <- 8
community_palette <- viridis(max(base_palette_size, length(community_levels)))[seq_along(community_levels)]

plot_list <- list()

for (d in colnames(cont_df)[1:ncol(cont_df)-1]) {

  df <- cont_df[,c(d, "community_cluster")]

plot_list[[d]] <- ggboxplot(df, x = "community_cluster", y = d,
                color = "community_cluster", palette = community_palette,
                add = "jitter")+
                stat_compare_means() +
                theme(aspect.ratio = 1.5, legend.position="none")
}

p_box <- do.call("grid.arrange", c(plot_list, ncol=3))

save_plot(p_box, "Fig3cBis_PT_abundance_het_boxplot_metadata.pdf", width = 6, height = 8)
```

## Plot metaclusters

For reference, we also plot the proportions of metaclusters in the clustered samples (kept as a standalone figure; it is not part of the combined figure below).

```{r plot-metaclusters}
sce_mc <- sce
sce_mc$cluster_id <- sce_mc$metacluster
metadata(sce_mc)$cluster_codes <- data.frame(celltype = factor(sce_mc$metacluster))
p_all <- plotAbundances(sce_mc, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_metacluster)

p_all

save_plot(p_all, "Fig3aBis_PT_abundance_patients_clustered_metacluster.pdf", width = 12, height = 5)
```

## Combined Figure 3a-bis (stacked panels)

All panels share the same sample order (dendrogram order). The dendrogram on top
is a single connected tree: its root spans the two communities. To keep the
communities visually separated, a blank spacer column is inserted at each
community boundary in every panel (an empty column with no sample and no
dendrogram leaf), and the connecting root segment simply crosses that gap.
Because the dendrogram is drawn on a continuous x-axis matched to the discrete
barplot/tile axes (same `[0.5, M+0.5]` range and expansion, where `M` includes
the spacer columns), each dendrogram leaf is centered on its column in every
panel.

```{r fig3a-combined, fig.width=8, fig.height=15}
# Shared discrete x-axis expansion — set on every panel so the continuous
# dendrogram axis and the discrete barplot/tile axes share the same physical
# range [0.5, M + 0.5] and therefore align column-by-column.
X_EXPAND <- expansion(add = 0.5)

# ---- Community layout: blank spacer column between communities ----
# cellular_communities is contiguous in clustered order (each cutree cluster is a
# subtree of the dendrogram), so communities change at single boundaries. We add
# one empty column at each boundary: it carries no sample (and no dendrogram
# leaf), producing a visible gap that the single connected dendrogram spans.
comm_seq     <- cellular_communities[clustered_samples, "community_cluster"]
N            <- length(clustered_samples)
boundary_idx <- which(diff(as.integer(comm_seq)) != 0)  # last index of each community (except the last)
n_spacers    <- length(boundary_idx)
M            <- N + n_spacers

# Build a factor-level vector with a unique blank spacer inserted after each boundary
insert_spacers <- function(ids) {
  if (n_spacers == 0) return(ids)
  out <- character(0); prev <- 0L
  for (b in seq_along(boundary_idx)) {
    out  <- c(out, ids[(prev + 1L):boundary_idx[b]], paste0("__gap__", b))
    prev <- boundary_idx[b]
  }
  c(out, ids[(prev + 1L):length(ids)])
}

# Map an original x (1..N, no spacer) onto the spaced axis (1..M)
remap_x  <- function(v) v + vapply(v, function(z) sum(boundary_idx < z), numeric(1))
real_pos <- remap_x(seq_len(N))   # spaced positions of the N real columns

# Spaced level vectors for the two sample-id formats used by the panels
lev_allct <- insert_spacers(clustered_order)    # progress_..._tissue ids (all-cell-type & tumor panels)
lev_fm    <- insert_spacers(clustered_samples)  # fm_id_tissue ids (immune panel & metadata strips)

# ---- Helper: drop faceting, apply spaced levels, remove x-axis (shared axis) ----
# drop = FALSE keeps the empty spacer levels so they render as blank gap columns.
prep_bar <- function(pl, lev) {
  pl <- pl + facet_null()
  pl$data$sample_id <- factor(as.character(pl$data$sample_id), levels = lev)
  pl + theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    plot.title   = element_blank(),
    panel.border = element_blank()
  ) +
  scale_x_discrete(drop = FALSE, expand = X_EXPAND)
}

# ---- 1. Single connected dendrogram (no faceting), leaves on real columns ----
ddata     <- dendro_data(hc, type = "rectangle")
segs      <- segment(ddata)
segs$x    <- remap_x(segs$x)      # shift coordinates onto the spaced axis
segs$xend <- remap_x(segs$xend)
h_root    <- max(c(segs$y, segs$yend))

# Community labels centered over each community's (spaced) leaf range
comm_label_df <- data.frame(
  x     = vapply(community_levels, function(cc) mean(real_pos[comm_seq == cc]), numeric(1)),
  label = as.character(community_levels),
  y     = h_root * 1.08
)

p_dendro <- ggplot(segs) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = comm_label_df, aes(x = x, y = y, label = label),
            fontface = "bold", size = 4, inherit.aes = FALSE) +
  scale_x_continuous(limits = c(0.5, M + 0.5), expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Patient group") +
  theme_minimal() +
  theme(
    axis.text       = element_blank(),
    axis.ticks      = element_blank(),
    axis.title      = element_blank(),
    plot.title      = element_text(face = "bold", size = 11, hjust = 0.5),
    plot.title.position = "plot",
    panel.grid      = element_blank(),
    plot.margin     = margin(2, 2, 0, 2)
  )

# ---- 2. Metadata strips as geom_tile ggplots ----
meta_df <- sample_summary[clustered_samples, c("mna", "X1p_loss", "progress", "age_diag")]
meta_df$TUM_het   <- tumor_het[clustered_samples]
meta_df$ME_het    <- me_het[clustered_samples]
meta_df$sample_id <- factor(clustered_samples, levels = lev_fm)

# Colorbar guide for continuous variables
cbar_guide <- guide_colorbar(
  barwidth       = unit(0.5,  "cm"),
  barheight      = unit(2.5,  "cm"),
  title.position = "top",
  ticks          = FALSE
)

# Shared ggplot layers for each metadata strip (drop = FALSE keeps spacer columns)
meta_layers <- list(
  geom_tile(),
  scale_x_discrete(drop = FALSE, expand = X_EXPAND),
  theme_void(),
  theme(
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text     = element_text(size = 9),
    legend.title    = element_text(size = 10, face = "bold"),
    axis.title.y    = element_text(angle = 0, vjust = 0.5, size = 7),
    plot.margin     = margin(0, 2, 0, 2)
  )
)

p_mna_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = as.character(mna))) +
  meta_layers +
  scale_fill_manual(values = col_code, name = "MYCN", na.value = "white") +
  labs(y = "MYCN")

p_1p_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = as.character(X1p_loss))) +
  meta_layers +
  scale_fill_manual(values = col_code, name = "1p_loss", na.value = "white") +
  labs(y = "1p_loss")

p_prog_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = as.character(progress))) +
  meta_layers +
  scale_fill_manual(values = col_code, name = "Prog", na.value = "white") +
  labs(y = "Prog")

# Use scale_fill_viridis_c for all continuous metadata (avoids fragile attr() calls
# on colorRamp2 functions; viridis palette matches the colorRamp2 definitions above)
p_age_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = age_diag)) +
  meta_layers +
  scale_fill_viridis_c(name = "Age", na.value = "white", guide = cbar_guide) +
  labs(y = "Age")

p_tumhet_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = TUM_het)) +
  meta_layers +
  scale_fill_viridis_c(name = "TUM_het", na.value = "white", guide = cbar_guide) +
  labs(y = "TUM_het")

p_mehet_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = ME_het)) +
  meta_layers +
  scale_fill_viridis_c(name = "ME_het", na.value = "white", guide = cbar_guide) +
  labs(y = "ME_het")

# ---- 3. Clean barplots (drop facet, apply spacer levels, remove x-axis) ----
# Panel order: all cell types (was tumor) -> metadata -> tumor only (was
# metacluster) -> immune (kept as is).
p_allct_c  <- prep_bar(p,       lev_allct)
p_tumor_c  <- prep_bar(p_tumor, lev_allct)
p_immune_c <- prep_bar(p_ME,    lev_fm)

# ---- 4. Stack all panels ----
# guides = "collect" consolidates all legends into one area, eliminating overlap
p_fig3a <- (p_dendro /
  p_allct_c /
  p_mna_m / p_1p_m / p_prog_m / p_age_m / p_tumhet_m / p_mehet_m /
  p_tumor_c /
  p_immune_c) +
  plot_layout(ncol = 1, heights = c(2, 4, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 4, 4),
              guides = "collect")

p_fig3a

save_plot(p_fig3a, "Fig3aBis_combined_stacked.pdf", width = 8, height = 15)
```

## Save SPE

```{r save-spe}
# Persist the object carrying the all-cell-type community assignment
# (spe$allcelltype_community) for the downstream DA analysis (05.2_PT_DE_DA.Rmd).
saveRDS(spe, file.path(params$output, "spe_05.1_PT_DE_cellularcomm.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>

```{r, echo = FALSE}
sessionInfo()
```
</details>
