---
title: "Differential abundance between genetic subtypes in diagnostic primary tumors"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# Create plots directory
plot_dir <- file.path(params$output, "plots/05.2_PT_DE_DA")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# Helper to save plots
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
    
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {
    
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
    
  } else {
    
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  
  message("Saved: ", full_path)
}

# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Differential abundance in diagnostic primary tumors {#pt-dx}
Here we test for differential cell-type abundance between genetic subtypes (MYCN status) and between cellular communities, across tumor and TME compartments.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(SingleCellExperiment)
library(CATALYST)
library(diffcyt)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)
library(grid)
library(patchwork)
```
</details>

## Read data
First, we load the object produced by the previous step (05.1), which carries the `tumor_community` assignment on top of the phenotyped single-cell data.

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_05.1_PT_DE_cellularcomm.rds"))
```

We will then build a base catalyst-compatible object from all diagnostic PT cells, and run differential abundance analyses across multiple compartments and comparisons.

```{r build-base-sce}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$tissue == "PT" & spe$timepoint == "DE"

sce <- SingleCellExperiment(assays = list(
  counts = assay(spe, "counts")[keep_rows, keep_cols],
  exprs  = assay(spe, "counts")[keep_rows, keep_cols]
))

# Recode progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
sce$progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')

# Recode MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
sce$mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')

sce$patient        <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id          <- spe[keep_rows, keep_cols]$fm_id
sce$tissue         <- factor(spe[keep_rows, keep_cols]$tissue)
sce$metacluster    <- factor(spe[keep_rows, keep_cols]$metacluster)
sce$celltype_ordered <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$condition      <- 'na'

# tumor_community (from 05.1): samples clustered on tumor cell-type composition.
# The community DA below runs only when this column is present.
if ("tumor_community" %in% colnames(colData(spe))) {
  sce$tumor_community <- spe[keep_rows, keep_cols]$tumor_community
}

# Define cell type and state markers
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", "other"))

# Build marker_info once for use in all DA analyses
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 2)
marker_name  <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 1)
marker_class_vec <- ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                           ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", "other"))
marker_info <- data.frame(cbind(channel_name, marker_name), row.names = marker_name)
marker_info$marker_class <- as.factor(marker_class_vec)
```

## DA analysis function

```{r define-da-function}
# run_da_analysis: run DA for one compartment/comparison and draw one combined
# boxplot. comparison is "mna" (MNA vs nMNA) or a community column
# ("tumor_community"), comparing the two `selected_communities`.
# min_samples_frac scales the diffcyt min_samples threshold (NULL = 1).

run_da_analysis <- function(sce_base, cell_filter, cluster_level, comparison,
                            ord, selected_communities = NULL,
                            min_samples_frac = NULL, fig_suffix) {

  # --- 1. Subset cells ---
  sce_sub <- sce_base[, cell_filter]

  # --- 2. Apply comparison filter and set condition ---
  if (comparison == "mna") {
    sce_sub  <- sce_sub[, sce_sub$mna %in% c("nMNA", "MNA")]
    sce_sub$condition <- droplevels(sce_sub$mna)
    g1 <- "nMNA"; g2 <- "MNA"
  } else {
    if (is.null(colData(sce_sub)[[comparison]])) {
      stop("Community column not found in SCE: ", comparison,
           ". Run the corresponding cellular-community script first.")
    }
    if (is.null(selected_communities) || length(selected_communities) != 2) {
      stop("Please provide exactly two communities in `selected_communities`.")
    }
    selected_communities <- as.character(selected_communities)
    sce_sub <- sce_sub[, !is.na(colData(sce_sub)[[comparison]])]
    community_values <- as.character(colData(sce_sub)[[comparison]])
    sce_sub <- sce_sub[, community_values %in% selected_communities]
    community_values <- as.character(colData(sce_sub)[[comparison]])
    missing_communities <- setdiff(selected_communities, unique(community_values))
    if (length(missing_communities) > 0) {
      stop(
        "Selected communities not present after subsetting (", comparison, "): ",
        paste(missing_communities, collapse = ", ")
      )
    }
    sce_sub$condition <- factor(community_values, levels = selected_communities)
    g1 <- selected_communities[1]; g2 <- selected_communities[2]
  }

  # --- 3. Set cluster_id and colour vector ---
  if (cluster_level == "celltype") {
    sce_sub$cluster_id <- droplevels(sce_sub$celltype_ordered)
    metadata(sce_sub)$cluster_codes <- data.frame(celltype = droplevels(sce_sub$celltype_ordered))
    col_vec <- metadata(spe)$color_vectors$col_celltype_ordered
  } else {
    sce_sub$cluster_id <- droplevels(sce_sub$metacluster)
    metadata(sce_sub)$cluster_codes <- data.frame(celltype = droplevels(sce_sub$metacluster))
    col_vec <- metadata(spe)$color_vectors$col_metacluster
  }

  # --- 4. Set sample_id and experiment_info ---
  sce_sub$sample_id <- factor(paste0(sce_sub$progress, "_", sce_sub$mna, "_",
                                     sce_sub$patient, "_", sce_sub$fm_id, "_", sce_sub$tissue))

  metadata(sce_sub)$experiment_info <- as.data.frame(colData(sce_sub)) %>%
    group_by(sample_id, patient, mna, condition, progress) %>%
    summarise(n_cells = n(), .groups = 'drop') %>%
    as.data.frame()

  # --- 5. Per-sample cell-type abundances (for the boxplot, not rendered here) ---
  # plotAbundances() is used only to obtain per-sample cluster frequencies.
  abund_plot <- plotAbundances(sce_sub, k = "celltype", by = "sample")
  abund_data <- abund_plot$data %>%
    mutate(
      cluster_id = factor(as.character(cluster_id), levels = levels(sce_sub$cluster_id)),
      condition  = factor(as.character(condition),  levels = ord)
    ) %>%
    filter(!is.na(cluster_id))
  abund_data$cluster_id <- droplevels(abund_data$cluster_id)

  # --- 6. Prepare data for diffcyt ---
  df_list <- lapply(unique(sce_sub$sample_id), function(x) {
    as.data.frame(t(counts(sce_sub[, sce_sub$sample_id == x])))
  })
  names(df_list) <- unique(sce_sub$sample_id)

  experiment_info <- metadata(sce_sub)$experiment_info
  experiment_info <- experiment_info[match(names(df_list), experiment_info$sample_id), ]

  d_se <- prepareData(df_list, experiment_info, marker_info)
  rowData(d_se)$cluster_id <- sce_sub[, rownames(d_se)]$cluster_id
  metadata(d_se)$cluster_codes <- metadata(sce_sub)$cluster_codes
  d_se      <- transformData(d_se)
  d_counts  <- calcCounts(d_se)

  # --- 7. DA test ---
  design   <- createDesignMatrix(experiment_info, cols_design = "condition")
  # contrast c(0,1) = second vs first level; logFC is negated below so it reads
  # log2(g2 / g1) in our group ordering.
  contrast <- createContrast(c(0, 1))
  min_samples <- if (!is.null(min_samples_frac)) round(ncol(d_counts) * min_samples_frac) else 1

  res_DA <- testDA_voom(d_counts, design, contrast, min_cells = 3, min_samples = min_samples)
  FDR_cutoff <- 0.1
  print(rowData(res_DA)[which(rowData(res_DA)$p_adj < FDR_cutoff), ])
  res <- data.frame(rowData(res_DA))

  # --- 8. Combined grouped boxplot (one group per cluster) ---
  # For each cell type the two conditions are drawn side by side, with the DA
  # adjusted p-value bracket over the pair.
  cluster_present <- levels(abund_data$cluster_id)

  stats_df <- do.call(rbind, lapply(cluster_present, function(cl) {
    if (cl %in% rownames(res)) {
      p_adj_val <- round(res[cl, "p_adj"], 3)
      logFC_val <- -1 * round(res[cl, "logFC"], 2)
    } else {
      p_adj_val <- NA_real_; logFC_val <- NA_real_
    }
    signif <- if (!is.na(p_adj_val)) stars.pval(p_adj_val) else "ns"
    if (signif == " ") signif <- "ns"
    label <- if (is.na(p_adj_val)) "ns" else paste0("p=", p_adj_val, " ", signif)
    cl_freq <- abund_data$Freq[abund_data$cluster_id == cl]
    ymax <- if (length(cl_freq)) max(cl_freq, na.rm = TRUE) else 0
    data.frame(cluster_id = cl, group1 = g1, group2 = g2,
               p_adj = p_adj_val, logFC = logFC_val, label = label,
               y.position = ymax * 1.05, stringsAsFactors = FALSE)
  }))
  stats_df$cluster_id <- factor(stats_df$cluster_id, levels = cluster_present)

  # Community comparisons label the groups "Group 1"/"Group 2" (MNA keeps
  # nMNA/MNA); relabel the condition factor and bracket groups together.
  if (comparison != "mna") {
    levels(abund_data$condition) <- paste("Group", levels(abund_data$condition))
    stats_df$group1 <- paste("Group", stats_df$group1)
    stats_df$group2 <- paste("Group", stats_df$group2)
  }

  # Boxes are coloured by cell type / metacluster; missing colours fall back to grey.
  cluster_colors <- col_vec[cluster_present]
  cluster_colors[is.na(cluster_colors)] <- "gray50"
  names(cluster_colors) <- cluster_present

  long_labels <- any(nchar(cluster_present) > 6)

  # --- Combined grouped boxplot on a NUMERIC x-axis --------------------------
  # Cell type i sits at integer i, its two conditions at i +/- 0.2. Cell-type
  # names are drawn as a separate geom_text strip below (via patchwork) so the
  # 45 deg labels right-anchor under each cell type.
  n_clusters <- length(cluster_present)
  ct_idx     <- setNames(seq_along(cluster_present), cluster_present)
  cond_lev   <- levels(abund_data$condition)          # two levels: g1 then g2
  cond_off   <- setNames(c(-0.2, 0.2), cond_lev)

  abund_data$xpos <- ct_idx[as.character(abund_data$cluster_id)] +
                     cond_off[as.character(abund_data$condition)]
  abund_data$box  <- interaction(abund_data$cluster_id, abund_data$condition,
                                 drop = TRUE)

  # One row per box, in x order, for the condition (inner) axis tick labels.
  box_df <- data.frame(
    x    = as.vector(ct_idx[rep(cluster_present, each = 2)] +
                     rep(cond_off, times = n_clusters)),
    cond = rep(cond_lev, times = n_clusters),
    stringsAsFactors = FALSE
  )
  box_df <- box_df[order(box_df$x), ]

  # Bracket geometry, drawn manually at the numeric pair positions.
  stats_df$xc   <- ct_idx[as.character(stats_df$cluster_id)]
  stats_df$xmin <- stats_df$xc - 0.2
  stats_df$xmax <- stats_df$xc + 0.2
  y_tip <- 0.02 * max(stats_df$y.position, na.rm = TRUE)
  x_lims <- c(0.5, n_clusters + 0.5)

  # size (not linewidth): linewidth on segments needs ggplot2 >= 3.4.0.
  p_main <- ggplot(abund_data, aes(x = xpos, y = Freq, group = box)) +
    geom_boxplot(aes(fill = cluster_id, colour = cluster_id),
                 position = "identity", width = 0.32,
                 outlier.shape = NA, alpha = 0.45, show.legend = FALSE) +
    geom_jitter(aes(colour = cluster_id), width = 0.07, size = 1, alpha = 0.8,
                show.legend = FALSE) +
    geom_segment(data = stats_df, inherit.aes = FALSE, size = 0.3,
                 aes(x = xmin, xend = xmax, y = y.position, yend = y.position)) +
    geom_segment(data = stats_df, inherit.aes = FALSE, size = 0.3,
                 aes(x = xmin, xend = xmin, y = y.position - y_tip, yend = y.position)) +
    geom_segment(data = stats_df, inherit.aes = FALSE, size = 0.3,
                 aes(x = xmax, xend = xmax, y = y.position - y_tip, yend = y.position)) +
    geom_text(data = stats_df, inherit.aes = FALSE, size = 2.6, vjust = -0.3,
              aes(x = xc, y = y.position, label = label)) +
    scale_fill_manual(values = cluster_colors, drop = FALSE) +
    scale_colour_manual(values = cluster_colors, drop = FALSE) +
    scale_x_continuous(breaks = box_df$x, labels = box_df$cond,
                       limits = x_lims, expand = c(0, 0)) +
    scale_y_continuous(expand = expansion(mult = c(0.03, 0.18))) +
    labs(x = NULL, y = "% abundance") +
    theme_pubr(base_size = 11) +
    theme(
      # Group / MNA condition labels stay fully vertical, under each box.
      axis.text.x     = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
      plot.margin     = margin(t = 5, r = 5, b = 2, l = 5),
      legend.position = "none"
    )

  # Thin strip of cell-type names, sharing the main panel's x range so patchwork
  # aligns them column-by-column.
  ct_lab_df <- data.frame(x = seq_along(cluster_present), label = cluster_present)
  p_ctlab <- ggplot(ct_lab_df, aes(x = x, y = 1, label = label)) +
    geom_text(angle = 45, hjust = 1, vjust = 1, size = 2.8) +
    scale_x_continuous(limits = x_lims, expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    coord_cartesian(clip = "off") +
    theme_void() +
    theme(plot.margin = margin(t = 2, r = 5, b = 34, l = 5))

  p_boxplot <- p_main / p_ctlab + plot_layout(heights = c(1, 0.42))

  print(p_boxplot)
  save_plot(p_boxplot, paste0("boxplot_", fig_suffix, ".pdf"),
            width  = max(8, n_clusters * 1.1 + 1.5) / 2,
            height = (if (long_labels) 5 else 4))

  invisible(list(res = res, abund_data = abund_data,
                 stats = stats_df, p_boxplot = p_boxplot))
}
```

## Run differential abundance analyses

We perform DA at four cluster levels — metacluster, all cell types (including
"other"), tumor cell types, and immune cell types — and, for each level, across
the available comparisons: MYCN status (MNA vs nMNA) and, when the community
column is present in the object, the cellular community (`tumor_community`,
samples clustered on their tumor cell-type composition). Each analysis produces
one combined boxplot figure.

```{r run-all-analyses, fig.width=16, fig.height=9}
immune_metaclusters <- c("granulocytes", "T cells", "B cells", "myeloid/NK")

# Two communities to compare for every community-based comparison.
community_comparison <- c("1", "2")

# --- Cluster levels: metacluster, all cell types (incl. "other"), tumor, immune ---
cluster_level_configs <- list(
  list(key = "metacluster",     cluster_level = "metacluster",
       cell_filter = rep(TRUE, ncol(sce)),                     min_samples_frac = NULL),
  list(key = "celltype_all",    cluster_level = "celltype",
       cell_filter = rep(TRUE, ncol(sce)),                     min_samples_frac = NULL),
  list(key = "celltype_tumor",  cluster_level = "celltype",
       cell_filter = sce$metacluster == "tumor",               min_samples_frac = 0.2),
  list(key = "celltype_immune", cluster_level = "celltype",
       cell_filter = sce$metacluster %in% immune_metaclusters, min_samples_frac = NULL)
)

# --- Comparisons: MNA always; community comparison only if its column is present ---
comparison_configs <- list(
  list(key = "mna", comparison = "mna", ord = c("nMNA", "MNA"), selected_communities = NULL)
)
community_col_labels <- c(tumor_community = "tumorcomm")
for (cc in names(community_col_labels)) {
  if (cc %in% colnames(colData(sce))) {
    comparison_configs[[length(comparison_configs) + 1]] <- list(
      key = community_col_labels[[cc]], comparison = cc,
      ord = community_comparison, selected_communities = community_comparison
    )
  } else {
    message("Community column '", cc, "' not found - skipping its DA analyses. ",
            "Run the corresponding cellular-community script first.")
  }
}

# --- Full set of analyses: every cluster level x every available comparison ---
analysis_configs <- list()
for (cl in cluster_level_configs) {
  for (cm in comparison_configs) {
    suffix <- if (cm$comparison == "mna") {
      paste0(cl$key, "_mna")
    } else {
      paste0(cl$key, "_", cm$key, "_", paste(cm$selected_communities, collapse = "_vs_"))
    }
    # Short figure name: <level>_<mna|grp>, dropping the "celltype_" prefix.
    fig_name <- paste0(sub("^celltype_", "", cl$key),
                       if (cm$comparison == "mna") "_mna" else "_grp")
    analysis_configs[[suffix]] <- list(
      name             = suffix,
      cell_filter      = cl$cell_filter,
      cluster_level    = cl$cluster_level,
      comparison       = cm$comparison,
      ord              = cm$ord,
      selected_communities = cm$selected_communities,
      min_samples_frac = cl$min_samples_frac,
      fig_suffix       = fig_name
    )
  }
}

da_results <- list()
for (cfg in analysis_configs) {
  message("\n### Running DA analysis: ", cfg$name, " ###\n")
  da_results[[cfg$name]] <- run_da_analysis(
    sce_base         = sce,
    cell_filter      = cfg$cell_filter,
    cluster_level    = cfg$cluster_level,
    comparison       = cfg$comparison,
    ord              = cfg$ord,
    selected_communities = cfg$selected_communities,
    min_samples_frac = cfg$min_samples_frac,
    fig_suffix       = cfg$fig_suffix
  )
}
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
