---
title: "Validation of cell phenotypes using the fetal adrenal gland (AD) reference"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  public: !r Sys.getenv("MAPMET_PUBLIC", "/mnt/data/public_datasets")
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# Plot output directory
plot_dir <- file.path(params$output, "plots/06.1_correlation_AD")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot to plot_dir, dispatching on its object type
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  message("Saved: ", full_path)
}

knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Validation using the fetal adrenal gland (AD) reference {#ad}
To evaluate the cell phenotype annotations, we compare the PT/BM cell expression profiles to the fetal adrenal gland (AD) reference cohort, which is processed by the parallel IMC pipeline in `analysis_AD/`. Using SingleR we transfer the AD fetal cell-type labels onto the PT/BM cells (at cluster and single-cell level) and visualise how strongly each neuroblastoma cluster resembles the fetal adrenal cell types.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(SingleR)
library(Seurat)
library(bruceR)
library(ComplexHeatmap)
library(viridis)
library(RColorBrewer)
library(BiocParallel)
library(dittoSeq)
```
</details>

```{r}
# Load packages and the AD reference (spe_AD) + PT/BM query (spe_PTBM)
library(SpatialExperiment)
library(SingleR)
library(scater)
library(dplyr)
library(ggplot2)
library(pheatmap)

spe_AD <- readRDS(file.path(params$output,"spe_04_phenotyping_AD.rds"))
spe_PTBM <- readRDS(file.path(params$output,"spe_04_phenotyping.rds"))

# Colors from the AD reference, reused by the predicted-fetal plots
col_celltype_AD <- metadata(spe_AD)$color_vectors$col_celltype

# AD celltype order (celltype_order_AD.csv), used to order plot axes/legends
ad_order_dirs <- c(Sys.getenv("MAPMET_INTERMEDIARY", ""),
                   file.path(params$input,  "R_intermediary"),
                   file.path(dirname(params$input), "R_intermediary"),
                   file.path(params$output, "R_intermediary"),
                   params$input)
ad_order_dirs <- ad_order_dirs[nzchar(ad_order_dirs)]
ad_order_file <- Find(file.exists, file.path(ad_order_dirs, "celltype_order_AD.csv"))
if (is.null(ad_order_file))
  stop("celltype_order_AD.csv not found in: ", paste(ad_order_dirs, collapse = ", "))
ad_order <- read.csv(ad_order_file, sep = ";", colClasses = c("order" = "character"))
ordered_AD_celltypes <- ad_order$celltype[order(ad_order$order)]

# Order predicted-fetal labels by the AD celltype order (unknown labels appended)
order_fetal <- function(x) {
  x <- as.character(x)
  factor(x, levels = c(ordered_AD_celltypes,
                       setdiff(unique(x), ordered_AD_celltypes)))
}
```

```{r}

# Reference (AD) and query (PT/BM) count matrices
ref_counts <- assay(spe_AD, "counts")
ref_labels <- spe_AD$celltype

query_counts <- assay(spe_PTBM, "counts")

# SingleR label transfer, one label per PT/BM cluster
singleR_results <- SingleR(
  test = query_counts,
  ref = ref_counts,
  labels = ref_labels,
  method = "cluster",
  clusters = spe_PTBM$celltype
)

# Attach the predicted fetal label to each PT/BM cell
cluster_to_label <- setNames(singleR_results$labels, rownames(singleR_results))
spe_PTBM$predicted_fetal <- order_fetal(cluster_to_label[as.character(spe_PTBM$celltype)])

```

```{r}

# Assemble a per-cell table of metadata + marker means (used to build the heatmap below)
marker_df <- as.data.frame(t(assay(spe_PTBM, "counts")))  # cells x markers
colnames(marker_df) <- rownames(spe_PTBM)
meta_df <- as.data.frame(colData(spe_PTBM))
df <- cbind(meta_df, marker_df)
```

```{r}
library(tidyr)

# Mean expression per cluster x predicted-fetal label, reshaped to a clusters x labels matrix
mean_expr <- df %>%
  group_by(pg_clusters, predicted_fetal) %>%
  summarize(across(ends_with("_mean"), mean), .groups = "drop")

heatmap_df <- mean_expr %>%
  pivot_wider(names_from = predicted_fetal, values_from = ends_with("_mean"),
              values_fn = mean)

library(tibble)

heatmap_matrix <- heatmap_df %>%
  column_to_rownames("pg_clusters") %>%
  as.matrix()

# Correlate NB clusters (rows) against fetal types (columns) and plot
sim_matrix <- cor(t(heatmap_matrix), use = "pairwise.complete.obs")
sim_matrix[is.na(sim_matrix)] <- 0

ph <- pheatmap(sim_matrix,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         main = "Neuroblastoma clusters vs Fetal progenitors")
save_plot(ph, "correlation-clusters-vs-fetal.pdf", width = 8, height = 6)

# UMAP coloured by cluster-level predicted fetal label
p_umap_fetal <- plotUMAP(spe_PTBM, colour_by = "predicted_fetal") +
  scale_color_manual(values = col_celltype_AD) +
  # Closed panel border (all UMAPs)
  theme(aspect.ratio = 1,
        axis.line    = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5))
save_plot(p_umap_fetal, "umap-predicted-fetal.pdf", width = 16, height = 8)
p_umap_fetal

# Spatial map coloured by predicted fetal label
coords <- as.data.frame(spatialCoords(spe_PTBM))
coords$predicted_fetal <- spe_PTBM$predicted_fetal

p_spatial_fetal <- ggplot(coords, aes(x = Pos_X, y = Pos_Y, color = predicted_fetal)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = col_celltype_AD) +
  labs(title = "Spatial mapping of predicted fetal progenitors") +
  theme_minimal()
save_plot(p_spatial_fetal, "spatial-predicted-fetal.pdf", width = 8, height = 6)
p_spatial_fetal

```
```{r}
# SingleR score heatmap: PT/BM clusters (rows) x AD fetal celltypes (columns)
library(ggplot2)
library(tidyr)
library(dplyr)
library(forcats)

# Ordered cluster names (celltype_order.csv, from R_intermediary or input)
order_dirs <- c(Sys.getenv("MAPMET_INTERMEDIARY", ""),
                file.path(params$input,  "R_intermediary"),
                file.path(dirname(params$input), "R_intermediary"),
                file.path(params$output, "R_intermediary"),
                params$input)
order_dirs <- order_dirs[nzchar(order_dirs)]
order_file <- Find(file.exists, file.path(order_dirs, "celltype_order.csv"))
if (is.null(order_file))
  stop("celltype_order.csv not found in: ", paste(order_dirs, collapse = ", "))
clusternames_ordered <- read.csv(order_file,
                                sep = ";", colClasses = c("order" = "character"))
ordered_clusters <- clusternames_ordered$celltype

# SingleR scores matrix, rows relabelled with cluster IDs and reordered
singleR_clusters <- rownames(singleR_results)
score_matrix <- singleR_results$scores
rownames(score_matrix) <- singleR_clusters
score_matrix <- score_matrix[match(ordered_clusters, singleR_clusters), , drop = FALSE]

# Long format for ggplot, with clusters ordered on the y-axis
score_df <- as.data.frame(score_matrix) %>%
  rownames_to_column("Neuroblastoma_cluster") %>%
  pivot_longer(-Neuroblastoma_cluster, names_to = "Fetal_AD_celltype", values_to = "score")
score_df$Neuroblastoma_cluster <- factor(score_df$Neuroblastoma_cluster, levels = ordered_clusters)

p_singler_scores <- ggplot(score_df, aes(x = Fetal_AD_celltype, y = Neuroblastoma_cluster, fill = score)) +
  geom_tile() +
  scale_fill_viridis_c() +
  scale_x_discrete(limits = intersect(ordered_AD_celltypes,
                                      unique(as.character(score_df$Fetal_AD_celltype)))) +
  scale_y_discrete(limits = ordered_clusters) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Fetal AD celltype", y = "Neuroblastoma cluster", fill = "SingleR score")
save_plot(p_singler_scores, "singler-scores-heatmap.pdf", width = 8, height = 6)
p_singler_scores
```

```{r singler-scores-tumor-metacluster}
# Same score heatmap, but on a hybrid grouping: tumor -> fine celltype, else -> metacluster
hybrid_group <- ifelse(as.character(spe_PTBM$metacluster) == "tumor",
                       as.character(spe_PTBM$celltype),
                       as.character(spe_PTBM$metacluster))

singleR_hybrid <- SingleR(
  test     = query_counts,
  ref      = ref_counts,
  labels   = ref_labels,
  method   = "cluster",
  clusters = hybrid_group
)

# Row order: tumor celltypes first (celltype_order.csv), then other groups (alphabetical)
tumor_cts <- ordered_clusters[ordered_clusters %in%
             unique(as.character(spe_PTBM$celltype)[spe_PTBM$metacluster == "tumor"])]
score_matrix_h <- singleR_hybrid$scores
rownames(score_matrix_h) <- rownames(singleR_hybrid)
present_rows    <- rownames(score_matrix_h)
row_order_hybrid <- c(intersect(tumor_cts, present_rows),
                      sort(setdiff(present_rows, tumor_cts)))
score_matrix_h <- score_matrix_h[row_order_hybrid, , drop = FALSE]

score_df_h <- as.data.frame(score_matrix_h) %>%
  rownames_to_column("Neuroblastoma_group") %>%
  pivot_longer(-Neuroblastoma_group, names_to = "Fetal_AD_celltype", values_to = "score")
score_df_h$Neuroblastoma_group <- factor(score_df_h$Neuroblastoma_group, levels = row_order_hybrid)

p_singler_scores_hybrid <- ggplot(score_df_h,
    aes(x = Fetal_AD_celltype, y = Neuroblastoma_group, fill = score)) +
  geom_tile() +
  scale_fill_viridis_c() +
  scale_x_discrete(limits = intersect(ordered_AD_celltypes,
                                      unique(as.character(score_df_h$Fetal_AD_celltype)))) +
  scale_y_discrete(limits = row_order_hybrid) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Fetal AD celltype", y = "PT/BM tumor celltype / metacluster",
       fill = "SingleR score")
save_plot(p_singler_scores_hybrid,
          "singler-scores-heatmap-tumor-celltypes-metaclusters.pdf",
          width = 8, height = 6)
p_singler_scores_hybrid
```
```{r}

ref_counts <- assay(spe_AD, "counts")
ref_labels <- spe_AD$celltype

query_counts <- assay(spe_PTBM, "counts")

# SingleR label transfer at single-cell level (one label per PT/BM cell)
singleR_cell <- SingleR(
  test = query_counts,
  ref = ref_counts,
  labels = ref_labels
)

spe_PTBM$predicted_fetal_cell <- order_fetal(singleR_cell$labels)
```

```{r}
p_umap_fetal_cell <- plotUMAP(spe_PTBM, colour_by = "predicted_fetal_cell", point_size = 0.05) +
  scale_color_manual(values = col_celltype_AD,
                     name = "Predicted foetal adrenal gland celltype") +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  # Closed panel border (all UMAPs); size, not linewidth, for the older ggplot2
  theme(aspect.ratio = 1,
        axis.line    = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5))
save_plot(p_umap_fetal_cell, "umap-predicted-fetal-cell.pdf", width = 16, height = 8)
p_umap_fetal_cell
```

```{r singler-cell-composition-heatmap}
# Cell-level composition heatmap: % of cells of each PT/BM celltype (row) assigned
# to each AD celltype (column), from the per-cell transfer. Rows sum to 100%.
comp_tab <- table(
  Neuroblastoma_cluster = factor(as.character(spe_PTBM$celltype), levels = ordered_clusters),
  Fetal_AD_celltype     = factor(as.character(singleR_cell$labels))
)
comp_prop <- prop.table(comp_tab, margin = 1) * 100   # row-normalised percentages
comp_df   <- as.data.frame(comp_prop, responseName = "percent")

p_singler_composition <- ggplot(comp_df,
    aes(x = Fetal_AD_celltype, y = Neuroblastoma_cluster, fill = percent)) +
  geom_tile() +
  scale_fill_viridis_c() +
  scale_x_discrete(limits = intersect(ordered_AD_celltypes,
                                      unique(as.character(comp_df$Fetal_AD_celltype)))) +
  scale_y_discrete(limits = ordered_clusters) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Fetal AD celltype", y = "Neuroblastoma cluster",
       fill = "% of cells")
save_plot(p_singler_composition, "singler-celltype-composition-heatmap.pdf",
          width = 8, height = 6)
p_singler_composition
```

```{r singler-composition-tumor-metacluster}
# Same composition heatmap on the hybrid grouping (tumor celltype / metacluster),
# reusing singleR_cell, hybrid_group and row_order_hybrid from above. Rows sum to 100%.
comp_tab_h <- table(
  Neuroblastoma_group = factor(hybrid_group, levels = row_order_hybrid),
  Fetal_AD_celltype   = factor(as.character(singleR_cell$labels))
)
comp_prop_h <- prop.table(comp_tab_h, margin = 1) * 100   # row-normalised percentages
comp_df_h   <- as.data.frame(comp_prop_h, responseName = "percent")

p_singler_composition_hybrid <- ggplot(comp_df_h,
    aes(x = Fetal_AD_celltype, y = Neuroblastoma_group, fill = percent)) +
  geom_tile() +
  scale_fill_viridis_c() +
  scale_x_discrete(limits = intersect(ordered_AD_celltypes,
                                      unique(as.character(comp_df_h$Fetal_AD_celltype)))) +
  scale_y_discrete(limits = row_order_hybrid) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Fetal AD celltype", y = "PT/BM tumor celltype / metacluster",
       fill = "% of cells")
save_plot(p_singler_composition_hybrid,
          "singler-celltype-composition-heatmap-tumor-celltypes-metaclusters.pdf",
          width = 8, height = 6)
p_singler_composition_hybrid
```
