---
title: "Validation of metaclusters using scRNA-seq data by Fetahu et al."
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
  public: !r Sys.getenv("MAPMET_PUBLIC", "/mnt/data/public_datasets")
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/06.2_correlation_Fetahu")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300, ...) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF (extra args forwarded to draw(), e.g. annotation_legend_list)
    pdf(full_path, width = width, height = height)
    draw(plot_obj, ...)
    dev.off()
  } else if (inherits(plot_obj, "pheatmap")) {
    # pheatmap objects (e.g. SingleR::plotScoreHeatmap) - save as PDF
    pdf(full_path, width = width, height = height)
    grid::grid.newpage()
    grid::grid.draw(plot_obj$gtable)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Validation using scRNA-seq data by Fetahu et al. {#fetahu}
To evaluate the validity of metacluster annnotations, we compare the cell expression profiles to our previously generated scRNA-seq dataset of NB bone marrow samples ([Fetahu et al.](https://doi.org/10.1038/s41467-023-39210-0) downloaded from [Zenodo](https://doi.org/10.5281/zenodo.7707614). 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(SingleR)
library(Seurat)
library(SingleCellExperiment)
library(bruceR)
library(ComplexHeatmap)
library(viridis)
library(RColorBrewer)
library(BiocParallel)
library(dittoSeq)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping) and the published Seurat object containing the scRNA-seq data. Please make sure to replace `params$public`with the path to the scRNA-seq data.

```{r read-data}
#read IMC data
protein <- readRDS(file.path(params$output,"spe_04_phenotyping.rds"))
protein <- protein[rowData(protein)$use_channel,]
col_celltype <- metadata(protein)$color_vectors$col_celltype

#read scRNA-seq data
monocle_cell_dataset <- readRDS(file.path(params$public,"Fetahu/rna_integrated_monocle.rds"))
# The published Fetahu object is a monocle3 `cell_data_set`, which is just a
# SingleCellExperiment subclass with a few extra slots. We only read its
# assays / UMAP / metadata here (no monocle3 analysis is run), so rather than
# depend on the heavy, hard-to-build monocle3 package purely to register the S4
# class, demote the object to its SingleCellExperiment parent when monocle3 is
# not installed. as.Seurat() then resolves to the very same SingleCellExperiment
# method it would use with monocle3 loaded, producing an identical Seurat object.
if (!requireNamespace("monocle3", quietly = TRUE)) {
  attr(monocle_cell_dataset, "class") <-
    structure("SingleCellExperiment", package = "SingleCellExperiment")
}
rna <- as.Seurat(monocle_cell_dataset, data=NULL)
rna <- RenameAssays(object = rna, originalexp = 'RNA')
celltype_details <- readRDS(file.path(params$public,"Fetahu/metadata.rds"))
rna@active.ident <- setNames(pull(celltype_details[match(celltype_details$cell,colnames(rna)),"cellont_cluster"], "cellont_cluster"),
                             pull(celltype_details[match(celltype_details$cell,colnames(rna)), "cell"], "cell"))
DefaultAssay(rna) <- "RNA"
```

## Set common features 
Next, we will set common features between the reference and query dataset. Therefore, we have to rename the features in the imaging dataset to the names of the corresponding genes. A csv files with protein-gene mappings can be obtained from Zenodo [(protein2gene.csv)](https://doi.org/10.5281/zenodo.13220634)

```{r set-common-features}
protein2gene <- read.csv(file.path(params$public,"protein2gene.csv"), sep=";")

code <- setNames(protein2gene$genes, protein2gene$protein)
rownames(protein) <- recode(rownames(protein), !!!code)
rowData(protein)$name <- rownames(protein)
rowData(protein)$channel <- rownames(protein)

#reduce metadata
metadata <- c("sample_id", "ObjectNumber", "sample", "tissue", "celltype", "metacluster")
colData(protein)[, metadata]
colData(protein) <- colData(protein)[, metadata]

common_features <- intersect(row.names(rna), row.names(protein))
```

## Visualize public data

Will will first visualize the scRNA-seq data. Therefore we first define a color code with similar colors for similar celltypes between IMC and scRNA-seq.

```{r define-colors}
# Fetahu reference cluster colors are defined centrally in 00_MapMet_color_code.R
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }
col_clusters <- COL_FETAHU_CLUSTERS
```


```{r mean-per-cluster-rna, fig.width=8, fig.height=8}
rna_clusters <- rna@active.ident
rna_data <- GetAssayData(rna, assay = "RNA", slot = "data")[common_features, ]
rna_mean <- aggregate(t(rna_data), list(rna_clusters), mean)

rna_mean_t <- t(rna_mean[,2:ncol(rna_mean)])
colnames(rna_mean_t) <- rna_mean[,"Group.1"]
rna_mean_scaled <- t(apply(rna_mean_t, 1, RESCALE, to=0:1))

column_ha = HeatmapAnnotation(ncells=anno_barplot(as.data.frame(table(rna_clusters))[,"Freq"], height = unit(1, "cm")),
                              show_legend=F)

h_rna <- Heatmap(rna_mean_scaled,
        column_title = "Mean per cluster (scRNA-seq data)",
        top_annotation = column_ha,
        cluster_columns=T,
        show_column_dend = T,
        show_column_names=T,
        show_row_names=T,
        col=viridis(100),
        column_names_gp = gpar(fontsize = 10),
        column_names_max_height = max_text_width(colnames(rna_mean_scaled), gp = gpar(fontsize = 10))
        )

save_plot(h_rna, "Fetahu_rna_mean_per_cluster_heatmap.pdf", width = 8, height = 8)

h_rna
```

```{r umap-rna}
p_umap_rna <- DimPlot(rna, reduction="UMAP", cols=col_clusters)

save_plot(p_umap_rna, "Fetahu_rna_umap.pdf", width = 8, height = 6)

p_umap_rna
```

## Transfer labels with singleR
Finally, we will investigate correlation between the datasets using label transfer by singleR. 

```{r single-r}
rna_mat <- GetAssayData(rna, assay = "RNA", slot = "data")[common_features, ]
rna_clusters <- rna@active.ident
protein_mat <- assay(protein, "counts")[common_features, ]

predicted_labels <- SingleR(test = protein_mat, ref = rna_mat, labels = rna_clusters, BPPARAM = MulticoreParam(workers=10, RNGseed=20231030))

```

Plotting
```{r}

p<- plotScoreHeatmap(predicted_labels[sample(seq_len(nrow(predicted_labels)), 2000),],annotation_colors=list(Labels = col_clusters), annotation_legend=F, treeheight_row=0,
                 rows.order=names(col_clusters)) 

protein$predicted_celltype <- predicted_labels$labels

save_plot(p, "SupplFig3a_Fetahu_labeltransfer.pdf", width = 8, height = 6)
```

For visualization purposes, we rename the predicted celltypes to their metacluster and assign colors.

```{r metacluster-assignment}

protein$predicted_metacluster <- sub(" \\(.*$", "", protein$predicted_celltype)

col_metacluster <- COL_FETAHU_METACLUSTER
                
metadata(protein)$color_vectors$col_predicted_meta <- col_metacluster
```

We will plot the predicted metalabels in a UMAP.

```{r predicted-umap}
p1 <- dittoDimPlot(protein, var = "predicted_metacluster", 
             reduction.use = "UMAP", size = 0.3, labels.size=2,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(protein)$color_vectors$col_predicted_meta) +
  theme(legend.title = element_blank()) +
  ggtitle("Predicted celltype on UMAP")

save_plot(p1, "Fig2b_Fetahu_labeltransfer.pdf", width = 8, height = 6)

p1
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
