---
title: "Validation of metaclusters using scRNA-seq data by Fetahu et al."
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
  public: !r Sys.getenv("MAPMET_PUBLIC", "/mnt/data/public_datasets")
---

```{r setup, include=FALSE}
# Plot output directory
plot_dir <- file.path(params$output, "plots/06.2_correlation_Fetahu")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot to plot_dir, dispatching on its object type
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300, ...) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj, ...)
    dev.off()
  } else if (inherits(plot_obj, "pheatmap")) {
    pdf(full_path, width = width, height = height)
    grid::grid.newpage()
    grid::grid.draw(plot_obj$gtable)
    dev.off()
  } else {
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Validation using scRNA-seq data by Fetahu et al. {#fetahu}
To evaluate the validity of the metacluster annotations, we compare the cell expression profiles to our previously generated scRNA-seq dataset of NB bone marrow samples ([Fetahu et al.](https://doi.org/10.1038/s41467-023-39210-0), downloaded from [Zenodo](https://doi.org/10.5281/zenodo.7707614)).

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(SingleR)
library(Seurat)
library(SingleCellExperiment)
library(bruceR)
library(ComplexHeatmap)
library(viridis)
library(RColorBrewer)
library(BiocParallel)
library(dittoSeq)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping) and the published Seurat object containing the scRNA-seq data. Please make sure to replace `params$public`with the path to the scRNA-seq data.

```{r read-data}
#read IMC data
protein <- readRDS(file.path(params$output,"spe_04_phenotyping.rds"))
protein <- protein[rowData(protein)$use_channel,]

#read scRNA-seq data
monocle_cell_dataset <- readRDS(file.path(params$public,"Fetahu/rna_integrated_monocle.rds"))
# The Fetahu object is a monocle3 cell_data_set (a SingleCellExperiment subclass).
# If monocle3 isn't installed, demote it to its SingleCellExperiment parent so
# as.Seurat() still works - we only read its assays/UMAP/metadata, not run monocle3.
if (!requireNamespace("monocle3", quietly = TRUE)) {
  attr(monocle_cell_dataset, "class") <-
    structure("SingleCellExperiment", package = "SingleCellExperiment")
}
rna <- as.Seurat(monocle_cell_dataset, data=NULL)
rna <- RenameAssays(object = rna, originalexp = 'RNA')
celltype_details <- readRDS(file.path(params$public,"Fetahu/metadata.rds"))
rna@active.ident <- setNames(pull(celltype_details[match(celltype_details$cell,colnames(rna)),"cellont_cluster"], "cellont_cluster"),
                             pull(celltype_details[match(celltype_details$cell,colnames(rna)), "cell"], "cell"))
DefaultAssay(rna) <- "RNA"
```

## Set common features 
Next, we will set common features between the reference and query dataset. Therefore, we have to rename the features in the imaging dataset to the names of the corresponding genes. A csv files with protein-gene mappings can be obtained from Zenodo [(protein2gene.csv)](https://doi.org/10.5281/zenodo.13220634)

```{r set-common-features}
protein2gene <- read.csv(file.path(params$public,"protein2gene.csv"), sep=";")

code <- setNames(protein2gene$genes, protein2gene$protein)
rownames(protein) <- recode(rownames(protein), !!!code)
rowData(protein)$name <- rownames(protein)
rowData(protein)$channel <- rownames(protein)

#reduce metadata
metadata <- c("sample_id", "ObjectNumber", "sample", "tissue", "celltype", "metacluster")
colData(protein) <- colData(protein)[, metadata]

common_features <- intersect(row.names(rna), row.names(protein))
```

## Visualize public data

We will first visualize the scRNA-seq data. Therefore we first define a color code with similar colors for similar celltypes between IMC and scRNA-seq.

```{r define-colors}
# Fetahu reference cluster colors are defined centrally in 00_MapMet_color_code.R
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }
col_clusters <- COL_FETAHU_CLUSTERS
```


```{r mean-per-cluster-rna, fig.width=8, fig.height=8}
rna_clusters <- rna@active.ident
rna_data <- GetAssayData(rna, assay = "RNA", slot = "data")[common_features, ]
rna_mean <- aggregate(t(rna_data), list(rna_clusters), mean)

rna_mean_t <- t(rna_mean[,2:ncol(rna_mean)])
colnames(rna_mean_t) <- rna_mean[,"Group.1"]
rna_mean_scaled <- t(apply(rna_mean_t, 1, RESCALE, to=0:1))

column_ha = HeatmapAnnotation(ncells=anno_barplot(as.data.frame(table(rna_clusters))[,"Freq"], height = unit(1, "cm")),
                              show_legend=F)

h_rna <- Heatmap(rna_mean_scaled,
        column_title = "Mean per cluster (scRNA-seq data)",
        top_annotation = column_ha,
        cluster_columns=T,
        show_column_dend = T,
        show_column_names=T,
        show_row_names=T,
        col=viridis(100),
        column_names_gp = gpar(fontsize = 10),
        column_names_max_height = max_text_width(colnames(rna_mean_scaled), gp = gpar(fontsize = 10))
        )

save_plot(h_rna, "Fetahu_rna_mean_per_cluster_heatmap.pdf", width = 8, height = 8)

h_rna
```

```{r umap-rna}
p_umap_rna <- DimPlot(rna, reduction="UMAP", cols=col_clusters) +
  # Harmonised square panel border (all UMAPs).
  theme(axis.line    = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5))

save_plot(p_umap_rna, "Fetahu_rna_umap.pdf", width = 8, height = 6)

p_umap_rna
```

## Transfer labels with singleR
Finally, we will investigate correlation between the datasets using label transfer by singleR. 

```{r single-r}
rna_mat <- GetAssayData(rna, assay = "RNA", slot = "data")[common_features, ]
rna_clusters <- rna@active.ident
protein_mat <- assay(protein, "counts")[common_features, ]

predicted_labels <- SingleR(test = protein_mat, ref = rna_mat, labels = rna_clusters, BPPARAM = MulticoreParam(workers=10, RNGseed=20231030))

```

Plotting
```{r}

p<- plotScoreHeatmap(predicted_labels[sample(seq_len(nrow(predicted_labels)), 2000),],annotation_colors=list(Labels = col_clusters), annotation_legend=F, treeheight_row=0,
                 rows.order=names(col_clusters)) 

protein$predicted_celltype <- predicted_labels$labels

save_plot(p, "SupplFig3a_Fetahu_labeltransfer.pdf", width = 8, height = 6)
```

For visualization purposes, we rename the predicted celltypes to their metacluster and assign colors.

```{r metacluster-assignment}

protein$predicted_metacluster <- sub(" \\(.*$", "", protein$predicted_celltype)

col_metacluster <- COL_FETAHU_METACLUSTER
                
metadata(protein)$color_vectors$col_predicted_meta <- col_metacluster
```

We will plot the predicted metaclusters in a UMAP.

```{r predicted-umap}
p1 <- dittoDimPlot(protein, var = "predicted_metacluster", 
             reduction.use = "UMAP", size = 0.3, labels.size=2,
             do.label = F,
             order = "randomize",
             legend.show=T) +
  scale_color_manual(values = metadata(protein)$color_vectors$col_predicted_meta) +
  theme(legend.title = element_blank(), aspect.ratio = 1) +
  ggtitle("Predicted celltype on UMAP")

# Standard UMAP proportions (16x8)
save_plot(p1, "Fig2b_Fetahu_labeltransfer.pdf", width = 16, height = 8)

p1
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
