---
title: "Validation of TME celltypes using scRNA-seq data by Lee et al."
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
  public: !r Sys.getenv("MAPMET_PUBLIC", "/mnt/data/public_datasets")
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/06.4_correlation_Lee")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300, ...) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF (extra args forwarded to draw(), e.g. annotation_legend_list)
    pdf(full_path, width = width, height = height)
    draw(plot_obj, ...)
    dev.off()
  } else if (inherits(plot_obj, "pheatmap")) {
    # pheatmap objects (e.g. SingleR::plotScoreHeatmap) - save as PDF
    pdf(full_path, width = width, height = height)
    grid::grid.newpage()
    grid::grid.draw(plot_obj$gtable)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Validation using scRNA-seq data by Lee et al. {#lee}
To evaluate the validity of tumor microenviroment (TME) cell phenotype annnotations, we compare the cell expression profiles to a previously generated scRNA-seq dataset of healthy human bone marrow of ages ranging from 2-84 years ([Lee et al.](https://doi.org/10.3389/fimmu.2023.1127879)) downloaded from [Zenodo]](https://doi.org/10.5281/zenodo.10275924). 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(Seurat)
library(dplyr)
library(bruceR)
library(ComplexHeatmap)
library(viridis)
library(BiocParallel)
library(SingleR)
library(dittoSeq)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping) and the published Seurat object containing the scRNA-seq data. Please make sure to replace `params$public`with the path to the scRNA-seq data.

```{r read-data}
# IMC data
protein <- readRDS(file.path(params$output,"spe_04_phenotyping.rds"))
protein <- protein[rowData(protein)$use_channel,]
col_celltype <- metadata(protein)$color_vectors$col_celltype
protein <- protein[,protein$metacluster!="tumor" & protein$metacluster!="other"]

#scRNA-seq data
rna <- readRDS(file.path(params$public,"Lee/disco_bone_marrow_v2.0.rds"))
DefaultAssay(rna) <- "RNA"
rna@assays$RNA@data <- rna@assays$RNA@counts
```

## Set common tumor markers 
Next, we will set common markers between the reference and query dataset. Therefore, we have to rename the features in the imaging dataset to the names of the corresponding genes. 
A csv files with protein-gene mappings can be obtained from Zenodo [(protein2gene.csv)](https://doi.org/10.5281/zenodo.13220634).

```{r set-common-tumormarkers}
protein2gene <- read.csv(file.path(params$public,"protein2gene.csv"), sep=";")

code <- setNames(protein2gene$genes, protein2gene$protein)
rownames(protein) <- recode(rownames(protein), !!!code)
rowData(protein)$name <- rownames(protein)
rowData(protein)$channel <- rownames(protein)

#reduce metadata
metadata <- c("sample_id", "ObjectNumber", "sample", "tissue", "celltype", "metacluster")
colData(protein) <- colData(protein)[, metadata]

common_features <- intersect(rownames(rna@assays$RNA@data), row.names(protein))
```
## Visualize public data

Will will first visualize the scRNA-seq data. Therefore we first define a color code with similar colors for similar celltypes between IMC and scRNA-seq. 

```{r define-colors}
# Extra label colors are defined centrally in 00_MapMet_color_code.R
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }
col_clusters <- read.csv(file.path(params$public, "Lee/celltype_metacluster.csv"), sep=";")
col_clust <- setNames(col_clusters$color, col_clusters$celltype)
col_clust <- c(col_clust, COL_LEE_EXTRA)
```


```{r mean-per-cluster-rna, fig.width=12, fig.height=10}
rna_clusters <- rna@active.ident
rna_data <- GetAssayData(rna, assay = "RNA", slot = "data")[common_features, ]
rna_mean <- aggregate(t(rna_data), list(rna_clusters), mean)

rna_mean_t <- t(rna_mean[,2:ncol(rna_mean)])
colnames(rna_mean_t) <- rna_mean[,"Group.1"]
rna_mean_scaled <- t(apply(rna_mean_t, 1, RESCALE, to=0:1))

column_ha = HeatmapAnnotation(ncells=anno_barplot(as.data.frame(table(rna_clusters))[,"Freq"], height = unit(1, "cm")),
                              show_legend=F)

h_rna <- Heatmap(rna_mean_scaled,
        column_title = "Mean per cluster (scRNA-seq data)",
        top_annotation = column_ha,
        cluster_columns=T,
        show_column_dend = T,
        show_column_names=T,
        show_row_names=T,
        col=viridis(100),
        column_names_gp = gpar(fontsize = 10),
        column_names_max_height = max_text_width(colnames(rna_mean_scaled), gp = gpar(fontsize = 10))
        )

save_plot(h_rna, "Lee_rna_mean_per_cluster_heatmap.pdf", width = 12, height = 10)

h_rna
```

```{r umap-rna, fig.width=15, fig.height=10}
p_umap_rna <- DimPlot(rna, reduction="umap", cols=col_clust)

save_plot(p_umap_rna, "Lee_rna_umap.pdf", width = 15, height = 10)

p_umap_rna
```

## Transfer labels with singleR
Finally, we will investigate correlation between the datasets using label transfer by singleR. 

```{r single-r, fig.width=13, fig.height=9}
rna_mat <- GetAssayData(rna, assay = "RNA", slot = "data")[common_features, ]
rna_clusters <- rna@active.ident
protein_mat <- assay(protein, "counts")[common_features, ]

#remove Seurat and SPE object prior to parallel computation
rm(rna)
rm(protein)

predicted_labels <- SingleR(test = protein_mat, ref = rna_mat, labels = rna_clusters, BPPARAM = MulticoreParam(workers=16, RNGseed=20231030))

# plotScoreHeatmap draws a `Labels` annotation whose factor levels are the SingleR
# reference labels; pheatmap aborts ("Factor levels on variable Labels do not
# match with annotation_colors") if annotation_colors does not cover every level.
# Guarantee coverage by giving any label absent from col_clust (a Lee reference
# celltype not listed in celltype_metacluster.csv / COL_LEE_EXTRA) a neutral grey
# - newer, stricter pheatmap versions error on the gap that older ones tolerated.
score_labels <- union(colnames(predicted_labels$scores),
                      as.character(predicted_labels$labels))
missing_labs <- setdiff(score_labels, names(col_clust))
if (length(missing_labs)) {
  col_clust[missing_labs] <- "grey80"
  message("plotScoreHeatmap: filled ", length(missing_labs),
          " uncolored label(s) with grey: ", paste(missing_labs, collapse = ", "))
}

p_score <- plotScoreHeatmap(predicted_labels[sample(seq_len(nrow(predicted_labels)), 2000),],annotation_colors=list(Labels = col_clust), annotation_legend=F, treeheight_row=0) #define colors first

save_plot(p_score, "Lee_singleR_labeltransfer_scoreheatmap.pdf", width = 13, height = 9)
```

We will plot the predicted metalabels in a UMAP.

```{r predicted-umap}
#save predicted labels in separate slot
protein <- readRDS(file.path(params$output,"spe_04_phenotyping.rds"))
protein$celltype_predicted <- "tumor"
protein[,protein$metacluster=="other"]$celltype_predicted <- "other"
protein[,rownames(predicted_labels)]$celltype_predicted <- predicted_labels$labels
metadata(protein)$color_vectors$col_predicted_label <- col_clust

# Full UMAP (all cells, original colours), no on-plot labels, with TWO titled
# legends: the Lee-predicted TME celltypes ("Lee predicted celltype", colour
# aesthetic) and the PT/BM background categories tumor/other ("PT/BM", fill
# aesthetic). Two aesthetics give two separate legends without extra packages;
# tumor/other keep their original col_clust colours.
col_pred  <- metadata(protein)$color_vectors$col_predicted_label
bg_types  <- intersect(c("tumor", "other"), as.character(protein$celltype_predicted))
lee_types <- setdiff(sort(unique(as.character(protein$celltype_predicted))), bg_types)

umap_df <- data.frame(
  UMAP_1 = SingleCellExperiment::reducedDim(protein, "UMAP")[, 1],
  UMAP_2 = SingleCellExperiment::reducedDim(protein, "UMAP")[, 2],
  label  = as.character(protein$celltype_predicted)
)
umap_df$lee <- factor(ifelse(umap_df$label %in% lee_types, umap_df$label, NA), levels = lee_types)
umap_df$bg  <- factor(ifelse(umap_df$label %in% bg_types,  umap_df$label, NA), levels = bg_types)

# tumor/other background first (bottom), Lee-predicted celltypes on top.
p1 <- ggplot(umap_df, aes(UMAP_1, UMAP_2)) +
  geom_point(data = subset(umap_df, !is.na(bg)),  aes(fill   = bg),  shape = 21, size = 0.3, stroke = 0) +
  geom_point(data = subset(umap_df, !is.na(lee)), aes(colour = lee), size = 0.3) +
  scale_fill_manual(name = "PT/BM",                    values = col_pred[bg_types]) +
  scale_colour_manual(name = "Lee predicted celltype", values = col_pred[lee_types]) +
  guides(colour = guide_legend(order = 1, override.aes = list(size = 3)),
         fill   = guide_legend(order = 2, override.aes = list(size = 3))) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  ggtitle("Predicted celltype on UMAP")

# Match the standard UMAP proportions used elsewhere (aspect.ratio = 1 + 16x8),
# e.g. umap-celltypes.pdf in analysis/04_phenotyping.Rmd.
save_plot(p1, "Lee_predicted_celltype_umap.pdf", width = 16, height = 8)

p1
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
