---
title: "Identifying cellular communities in diagnostic primary tumors"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/06_PT_DE_cellularcom")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
    
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {
    
    # save as PNG instead of PDF if you want consistency
    png(full_path, width = width * dpi, height = height * dpi, res = dpi)
    draw(plot_obj)
    dev.off()
    
  } else {
    
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Identifying cellular communities in diagnostic primary tumors {#pt-dx}
Following the path of tumor progression, we start with an analysis of the primary site at diagnosis. Therefore, we perform hierarchical clustering of tumor cell composition profiles in primary tumors.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(gridExtra)
library(ggpubr)
#install.packages("ggdendro")
library(ggdendro)
library(patchwork)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final_clustering.rds"))
```

We will then transform the SPE object into a catalyst-compatible object. 

```{r catalyst-comp-object}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$celltype != "other" & spe$tissue=="PT" & (spe$timepoint=="DE")

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id, "_", sce$tissue))

#recode metadata on 1p loss
sce$X1p_loss <- factor(spe[keep_rows, keep_cols]$X1p_loss)
X1p_loss <- recode(sce$X1p_loss, `1` = 'yes', `2` = 'het', `0` = 'no', `9` = 'unknown')
sce$X1p_loss <- as.factor(X1p_loss)

sce$age_diag <- spe[keep_rows, keep_cols]$age_diag

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)

sce$condition <- 'na' 

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))
```

## Cluster cellular communities

Next, we cluster proportions of tumor cell types across all diagnostic PT samples (instead of conditions) and plot the result in barplots:

```{r cohort-overview}
set.seed(20231103)
sce_tumor <- sce[,sce$metacluster=="tumor"]

distance <- "manhattan"
linkage <- "complete"
n_communities <- 3

p <- plotAbundances(sce_tumor, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                    linkage = linkage, distance= distance) 

p

save_plot(p, "Fig3a1_PT_abundance_patients_clustered_cancer.png", width = 8, height = 6)
```

along with the dendogram:

```{r dendogram, fig.height=8, fig.width=7}
#Dendogram
set.seed(20231103)
wide_df <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

wide_df <- data.frame(wide_df)
rownames(wide_df) <- wide_df$sample_id
wide_df <- wide_df %>% select(-sample_id)

dist_matrix <- dist(wide_df, method = distance)
hc <- hclust(dist_matrix, method = linkage)

plot(hc)

save_plot(plot(hc), "Fig3a2_PT_abundance_patients_clustered_dendrogram.png", width = 8, height = 6)
```
We then save the order of clustered samples along with cluster labels.

```{r}
clustered_samples <- sapply(strsplit(levels(p$data$sample_id), "_"), function(x) paste(x[5], x[6], sep = "_"))

#Extract clusters of cellular communities
cellular_communities <- as.data.frame(cutree(hc, k = n_communities)[levels(p$data$sample_id)])
colnames(cellular_communities) <- c("community_cluster")
# Renumber cutree labels to consecutive IDs (1..n_communities) for plotting.
cluster_values <- sort(unique(cellular_communities$community_cluster))
cluster_names <- setNames(seq_along(cluster_values), cluster_values)
cellular_communities$community_cluster <- recode(cellular_communities$community_cluster, !!!cluster_names)
community_levels <- sort(unique(cellular_communities$community_cluster))

# Rename rownames to fm_id_tissue format for downstream matching
rownames(cellular_communities) <- sapply(strsplit(as.vector(rownames(cellular_communities)), "_"), function(x) paste(x[length(x) - 1], x[length(x)], sep = "_"))

# Add tumor_community metacolumn to spe
spe_key <- paste0(spe$fm_id, "_", spe$tissue)
spe$tumor_community <- cellular_communities[spe_key, "community_cluster"]
```

We now display the tumor abundance barplot split by community for clearer group separation.

```{r fig3a1-split, fig.width=12, fig.height=5}
p$data$comm_key <- sapply(strsplit(as.character(p$data$sample_id), "_"), function(x) paste(x[length(x) - 1], x[length(x)], sep = "_"))
p$data$community <- factor(cellular_communities[p$data$comm_key, "community_cluster"], levels = community_levels)
p_split <- p + facet_grid(. ~ community, scales = "free_x", space = "free_x")
p_split

save_plot(p_split, "Fig3a1_PT_abundance_patients_clustered_cancer_split.png", width = 12, height = 5)
```

## Plot TME cellular profiles 

We will then plot the cellular profiles of the TME for these samples. 

```{r plot-TME, fig.width=12, fig.height=5}
#order samples based on above clustering
sce$sample_id <- factor(paste0(spe[keep_rows, keep_cols]$fm_id, "_", spe[keep_rows, keep_cols]$tissue), levels=clustered_samples)

sce_ME <- sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]

p_ME <- plotAbundances(sce_ME, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_celltype_ordered)

# Split barplot by community for clearer group separation
p_ME$data$community <- factor(cellular_communities[as.character(p_ME$data$sample_id), "community_cluster"], levels = community_levels)
p_ME <- p_ME + facet_grid(. ~ community, scales = "free_x", space = "free_x")

p_ME

save_plot(p_ME, "Fig3a3_PT_abundance_patients_clustered_immune.png", width = 12, height = 5)
```

## Compute tumor and TME heterogeneity

We compute tumor and TME intra-sample heterogeneity.

```{r heterogeneity}
proportions <- plotAbundances(sce, k = "celltype", by = "sample")
prop <- proportions$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)
prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)

me_clusters <- unique(sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]$cluster_id)
me_prop <- prop[rownames(prop)%in%me_clusters,]
me_het <- apply(me_prop,2, entropy)
  
tumor_clusters <- unique(sce[,sce$metacluster=="tumor"]$cluster_id)
tumor_prop <- prop[rownames(prop)%in%tumor_clusters,]
tumor_het <- apply(tumor_prop,2, entropy)
```

## Plot metadata 

Then we plot intra-sample heterogeneity together with sample-specific metadata for the clustered sammples.

```{r plot-metadata}
#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            X1p_loss = first(X1p_loss),
            progress=first(progress),
            age_diag = first(age_diag)
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


#assign color codes
col_code <- c("MNA"=rgb(202,0,32, maxColorValue = 255),
              "het"=rgb(146,197,222, maxColorValue = 255),
              "nMNA"=rgb(5,113,176, maxColorValue = 255),
              "unknown"="white",
              "yes"=rgb(202,0,32, maxColorValue = 255),
              "no"=rgb(5,113,176, maxColorValue = 255),
              "Prog"=rgb(202,0,32, maxColorValue = 255),
              "NoProg"=rgb(5,113,176, maxColorValue = 255)
              )

col_age_diag = colorRamp2(c(min(sample_summary$age_diag, na.rm = TRUE), (quantile(sample_summary$age_diag, 0.95, na.rm = TRUE)-min(sample_summary$age_diag, na.rm = TRUE))/2, quantile(sample_summary$age_diag,0.95, na.rm = TRUE)), viridis(3))
col_me_het = colorRamp2(c(min(me_het, na.rm = TRUE), (max(me_het, na.rm = TRUE)-min(me_het, na.rm = TRUE))/2, max(me_het, na.rm = TRUE)), viridis(3))
col_tumor_het = colorRamp2(c(min(tumor_het, na.rm = TRUE), (max(tumor_het, na.rm = TRUE)-min(tumor_het, na.rm = TRUE))/2, max(tumor_het, na.rm = TRUE)), viridis(3))

# Plot metadata
p <- columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(X1p_loss=sample_summary$X1p_loss, col=list(X1p_loss=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(age=sample_summary$age_diag, col=list(age=col_age_diag)) %v%
  columnAnnotation(TUM_het=tumor_het, col=list(TUM_het=col_tumor_het)) %v%
  columnAnnotation(ME_het=me_het, col=list(ME_het=col_me_het))

save_plot(p, "Fig3a4_PT_abundance_patients_clustered_metadata.png", width = 8, height = 6)
  
```

## Plot continous and discrete summary statistics

We then plot summary statistics of continuous and discrete metadata in boxplots and piecharts, respectively.

```{r piecharts}
sample_summary[rownames(cellular_communities),"community_clusters"] <- cellular_communities$community_cluster

discrete_metadata <- c("mna", "X1p_loss", "progress")

discrete_metadata_colors <- list(col_code, col_code, col_code)

plot_list <- list()
for (d in seq_along(discrete_metadata)) {
  sample_meta <- sample_summary[,c(discrete_metadata[d], "community_clusters")]
  sample_meta <- sample_meta[rowSums(sample_meta == "unknown") == 0, ] 
  colnames(sample_meta) <- c("dis_var", "condition")
  
  result <- sample_meta %>%
    group_by(condition, dis_var) %>%
    summarise(Freq = n(), .groups = 'drop')
  
  result[,"dis_var"]<- as.factor(unlist(result[,"dis_var"]))
  
plot_list[[d]] <- ggplot(result, aes(x="", y=Freq, fill=dis_var, group=dis_var)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = discrete_metadata_colors[[d]], name=discrete_metadata[d]) +
    coord_polar("y", start=0)  +  
    facet_grid(.~ condition) +
    theme_void()
}

p<-do.call("grid.arrange", c(plot_list, ncol=1))

save_plot(p, "Fig3b_PT_abundance_piecharts_metadata.png", width = 6, height = 8)
```
```{r boxplots}
het_m <- as.vector(unname(me_het[clustered_samples]))
het_t <- as.vector(unname(tumor_het[clustered_samples]))

age <- sample_summary$age_diag

cont_df <- data.frame(cbind(het_m, het_t, age, cellular_communities),row.names=clustered_samples)
cont_df$community_cluster <- factor(cont_df$community_cluster, levels = community_levels)
community_palette <- viridis(max(8, length(community_levels)))[seq_along(community_levels)]

plot_list <- list()

for (d in colnames(cont_df)[1:ncol(cont_df)-1]) {

  df <- cont_df[,c(d, "community_cluster")] 

plot_list[[d]] <- ggboxplot(df, x = "community_cluster", y = d,
                color = "community_cluster", palette = community_palette,
                add = "jitter")+ 
                stat_compare_means() +
                theme(aspect.ratio = 1.5, legend.position="none") 
}

p<-do.call("grid.arrange", c(plot_list, ncol=3))

save_plot(p, "Fig3c_PT_abundance_het_boxplot_metadata.png", width = 6, height = 8)
```

## Plot metaclusters

Finally, we plot the proportions of metaclusters in the clustered samples.

```{r plot-metaclusters}
sce$cluster_id <- sce$metacluster
metadata(sce)$cluster_codes <- data.frame(celltype = factor(sce$metacluster))
p_all <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_metacluster)

# Split barplot by community for clearer group separation
p_all$data$community <- factor(cellular_communities[as.character(p_all$data$sample_id), "community_cluster"], levels = community_levels)
p_all <- p_all + facet_grid(. ~ community, scales = "free_x", space = "free_x")

p_all

save_plot(p_all, "Fig3a3_PT_abundance_patients_clustered_metacluster.png", width = 12, height = 5)
```

## Combined Figure 3a (stacked panels)

All panels share the same sample order (dendrogram order) and community split.

```{r fig3a-combined, fig.width=8, fig.height=15}
# Shared inter-community gap in physical units — applied to every panel so all
# gaps are identical regardless of the panel's base theme.
COMM_GAP <- unit(4, "pt")

# Shared discrete x-axis expansion — set explicitly on every panel so the
# proportional widths (space = "free_x") are identical for barplots and tiles.
X_EXPAND <- expansion(add = 0.5)

# ---- Helper: remove x-axis elements (shared axis) ----
clean_x <- function(p) {
  p + theme(
    axis.text.x       = element_blank(),
    axis.ticks.x      = element_blank(),
    axis.title.x      = element_blank(),
    plot.title        = element_blank(),
    strip.text.x      = element_blank(),
    strip.background  = element_blank(),
    panel.border      = element_blank(),
    panel.spacing.x   = COMM_GAP
  ) +
  scale_x_discrete(expand = X_EXPAND)
}

# Community strip labeller used by the dendrogram panel
comm_labels <- setNames(paste("Community", community_levels), as.character(community_levels))
comm_labeller <- labeller(community = comm_labels)

# ---- 1. Dendrogram as ggplot, split by community ----
ddata <- dendro_data(hc, type = "rectangle")
segs  <- segment(ddata)

# Map leaf x-positions to community
leaf_x   <- as.integer(ddata$labels$x)
leaf_lab <- as.character(ddata$labels$label)
leaf_key <- sapply(strsplit(leaf_lab, "_"), function(x) paste(x[length(x)-1], x[length(x)], sep="_"))
leaf_comm <- cellular_communities[leaf_key, "community_cluster"]

community_ranges <- lapply(community_levels, function(comm) {
  comm_x <- leaf_x[leaf_comm == comm]
  c(min = min(comm_x), max = max(comm_x), mean = mean(comm_x))
})
names(community_ranges) <- as.character(community_levels)
# Order communities left-to-right in the dendrogram for root-arm placement.
community_order <- as.integer(names(sort(sapply(community_ranges, `[[`, "mean"))))

# 0.5 is the tolerance used to assign dendrogram segments to communities.
# Leaf positions are integers (1..n), so 0.5 ensures each segment whose
# x-range falls entirely within one community's leaf positions is assigned
# to that community; cross-community (root) segments are excluded.
segs$community <- factor(mapply(function(x1, x2) {
  lo <- min(x1, x2); hi <- max(x1, x2)
  in_comm <- vapply(community_levels, function(comm) {
    comm_range <- community_ranges[[as.character(comm)]]
    lo >= comm_range["min"] - 0.5 && hi <= comm_range["max"] + 0.5
  }, logical(1))
  if (sum(in_comm) == 1) return(community_levels[in_comm])
  NA_integer_
}, segs$x, segs$xend), levels = community_levels)

segs_c <- segs[!is.na(segs$community), ]

# ---- Root arms: connect communities with a horizontal bar ----
# segs_c already contains, for each community, the vertical "root-drop" segment
# that goes from h_root down to the community's highest merge node.  These pass
# the community filter because their x coordinate lies within one community's
# leaf range.  We must NOT add another vertical stub (that would duplicate the
# existing line, creating a "second root").  Instead we only add a short
# horizontal arm that extends from the root-drop's x position outward to the
# far panel edge.  When the two panels are placed side by side the two arms
# read as a single connected horizontal bar at h_root.
h_root     <- max(c(segs$y, segs$yend))

# Return the x coordinate of the vertical segment that reaches h_root in comm.
get_root_drop_x <- function(comm) {
  s_comm <- segs_c[segs_c$community == comm, ]
  # Vertical segments: x == xend
  s_v <- s_comm[abs(s_comm$x - s_comm$xend) < 1e-6, ]
  if (nrow(s_v) == 0) return(NULL)
  # The root-drop has its top endpoint at h_root
  s_v_root <- s_v[abs(pmax(s_v$y, s_v$yend) - h_root) < 1e-6, ]
  if (nrow(s_v_root) == 0) s_v_root <- s_v[which.max(pmax(s_v$y, s_v$yend)), ]
  s_v_root$x[1]
}

make_root_arm <- function(comm) {
  x_pos   <- get_root_drop_x(comm)
  if (is.null(x_pos)) return(NULL)
  comm_x <- leaf_x[leaf_comm == comm]
  if (length(comm_x) == 0) return(NULL)
  if (comm == community_order[1]) {
    # Panel on the left → arm extends rightward to right panel edge
    data.frame(x = x_pos, y = h_root, xend = max(comm_x) + 0.5, yend = h_root,
               community = factor(comm, levels = community_levels))
  } else if (comm == tail(community_order, 1)) {
    # Panel on the right → arm extends leftward to left panel edge
    data.frame(x = min(comm_x) - 0.5, y = h_root, xend = x_pos, yend = h_root,
               community = factor(comm, levels = community_levels))
  } else {
    # Panels between leftmost/rightmost extend fully across to bridge both sides
    data.frame(x = min(comm_x) - 0.5, y = h_root, xend = max(comm_x) + 0.5, yend = h_root,
               community = factor(comm, levels = community_levels))
  }
}

root_arms <- do.call(rbind, lapply(community_levels, make_root_arm))
segs_with_root <- if (is.null(root_arms)) segs_c else rbind(segs_c, root_arms)

p_dendro <- ggplot(segs_with_root) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  scale_x_continuous(expand = c(0, 0.5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  facet_grid(. ~ community, scales = "free_x", space = "free_x",
             labeller = comm_labeller) +
  theme_minimal() +
  theme(
    axis.text       = element_blank(),
    axis.ticks      = element_blank(),
    axis.title      = element_blank(),
    plot.title      = element_blank(),
    strip.text.x    = element_text(face = "bold", size = 11),
    strip.background = element_blank(),
    panel.grid      = element_blank(),
    panel.spacing.x = COMM_GAP,
    plot.margin     = margin(2, 2, 0, 2)
  )

# ---- 2. Metadata strips as geom_tile ggplots ----
meta_df <- sample_summary[clustered_samples, c("mna", "X1p_loss", "progress", "age_diag")]
meta_df$TUM_het   <- tumor_het[clustered_samples]
meta_df$ME_het    <- me_het[clustered_samples]
meta_df$sample_id <- factor(clustered_samples, levels = clustered_samples)
meta_df$community <- factor(cellular_communities[clustered_samples, "community_cluster"], levels = community_levels)

# Colorbar guide for continuous variables
cbar_guide <- guide_colorbar(
  barwidth       = unit(0.5,  "cm"),
  barheight      = unit(2.5,  "cm"),
  title.position = "top",
  ticks          = FALSE
)

# Shared ggplot layers for each metadata strip
meta_layers <- list(
  geom_tile(),
  facet_grid(. ~ community, scales = "free_x", space = "free_x"),
  scale_x_discrete(expand = X_EXPAND),
  theme_void(),
  theme(
    strip.text.x    = element_blank(),
    strip.background = element_blank(),
    panel.spacing.x = COMM_GAP,
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text     = element_text(size = 9),
    legend.title    = element_text(size = 10, face = "bold"),
    axis.title.y    = element_text(angle = 0, vjust = 0.5, size = 7),
    plot.margin     = margin(0, 2, 0, 2)
  )
)

p_mna_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = as.character(mna))) +
  meta_layers +
  scale_fill_manual(values = col_code, name = "MYCN", na.value = "white") +
  labs(y = "MYCN")

p_1p_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = as.character(X1p_loss))) +
  meta_layers +
  scale_fill_manual(values = col_code, name = "1p_loss", na.value = "white") +
  labs(y = "1p_loss")

p_prog_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = as.character(progress))) +
  meta_layers +
  scale_fill_manual(values = col_code, name = "Prog", na.value = "white") +
  labs(y = "Prog")

# Use scale_fill_viridis_c for all continuous metadata (avoids fragile attr() calls
# on colorRamp2 functions; viridis palette matches the colorRamp2 definitions above)
p_age_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = age_diag)) +
  meta_layers +
  scale_fill_viridis_c(name = "Age", na.value = "white", guide = cbar_guide) +
  labs(y = "Age")

p_tumhet_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = TUM_het)) +
  meta_layers +
  scale_fill_viridis_c(name = "TUM_het", na.value = "white", guide = cbar_guide) +
  labs(y = "TUM_het")

p_mehet_m <- ggplot(meta_df, aes(x = sample_id, y = 1, fill = ME_het)) +
  meta_layers +
  scale_fill_viridis_c(name = "ME_het", na.value = "white", guide = cbar_guide) +
  labs(y = "ME_het")

# ---- 3. Clean barplots (remove x-axis, title, and facet strips) ----
p_tumor_c  <- clean_x(p_split)
p_meta_c   <- clean_x(p_all)
p_immune_c <- clean_x(p_ME)

# ---- 4. Stack all panels ----
# guides = "collect" consolidates all legends into one area, eliminating overlap
p_fig3a <- (p_dendro /
  p_tumor_c /
  p_mna_m / p_1p_m / p_prog_m / p_age_m / p_tumhet_m / p_mehet_m /
  p_meta_c /
  p_immune_c) +
  plot_layout(ncol = 1, heights = c(1, 4, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 4, 4),
              guides = "collect")

p_fig3a

save_plot(p_fig3a, "Fig3a_combined_stacked.png", width = 8, height = 15)
```

## Save SPE

```{r save-spe}
#saveRDS(spe, file.path(params$output, "spe_final_communities.rds"))
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
