---
title: "Differential abundance between genetic subtypes in diagnostic primary tumors"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/06_PT_DE_DA")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
    
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {
    
    # save as PNG instead of PDF if you want consistency
    png(full_path, width = width * dpi, height = height * dpi, res = dpi)
    draw(plot_obj)
    dev.off()
    
  } else {
    
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Identifying cellular communities in diagnostic primary tumors {#pt-dx}
Herein, we explore differential cell type abundances in tumor and TME.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(SingleCellExperiment)
library(CATALYST)
library(diffcyt)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final_communities.rds"))
```

We will then build a base catalyst-compatible object from all diagnostic PT cells, and run differential abundance analyses across multiple compartments and comparisons.

```{r build-base-sce}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$tissue == "PT" & spe$timepoint == "DE"

sce <- SingleCellExperiment(assays = list(
  counts = assay(spe, "counts")[keep_rows, keep_cols],
  exprs  = assay(spe, "counts")[keep_rows, keep_cols]
))

# Recode progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
sce$progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')

# Recode MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
sce$mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')

sce$patient        <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id          <- spe[keep_rows, keep_cols]$fm_id
sce$tissue         <- factor(spe[keep_rows, keep_cols]$tissue)
sce$metacluster    <- factor(spe[keep_rows, keep_cols]$metacluster)
sce$celltype_ordered <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$tumor_community  <- spe[keep_rows, keep_cols]$tumor_community
sce$condition      <- 'na'

# Define cell type and state markers
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", "other"))

# Build marker_info once for use in all DA analyses
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 2)
marker_name  <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 1)
marker_class_vec <- ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                           ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", "other"))
marker_info <- data.frame(cbind(channel_name, marker_name), row.names = marker_name)
marker_info$marker_class <- as.factor(marker_class_vec)
```

## DA analysis function

```{r define-da-function}
# run_da_analysis: performs DA for one compartment/comparison combination and produces plots.
# sce_base       : base SCE with all PT DE cells and full metadata
# cell_filter    : logical vector over ncol(sce_base) selecting the cells of interest
# cluster_level  : "celltype" or "metacluster"
# comparison     : "mna" (MNA vs nMNA) or "tumor_community" (community 1 vs 2)
# ord            : display order of groups on boxplot x-axis
# min_samples_frac: fraction of samples to use as min_samples threshold (NULL = 1)
# fig_suffix     : string appended to all saved filenames for this analysis

run_da_analysis <- function(sce_base, cell_filter, cluster_level, comparison,
                            ord, min_samples_frac = NULL, fig_suffix) {

  # --- 1. Subset cells ---
  sce_sub <- sce_base[, cell_filter]

  # --- 2. Apply comparison filter and set condition ---
  if (comparison == "mna") {
    sce_sub  <- sce_sub[, sce_sub$mna %in% c("nMNA", "MNA")]
    sce_sub$condition <- droplevels(sce_sub$mna)
    g1 <- "nMNA"; g2 <- "MNA"
  } else {
    sce_sub  <- sce_sub[, !is.na(sce_sub$tumor_community)]
    sce_sub$condition <- factor(as.character(sce_sub$tumor_community))
    g1 <- "1"; g2 <- "2"
  }

  # --- 3. Set cluster_id and colour vector ---
  if (cluster_level == "celltype") {
    sce_sub$cluster_id <- droplevels(sce_sub$celltype_ordered)
    metadata(sce_sub)$cluster_codes <- data.frame(celltype = droplevels(sce_sub$celltype_ordered))
    col_vec <- metadata(spe)$color_vectors$col_celltype_ordered
  } else {
    sce_sub$cluster_id <- droplevels(sce_sub$metacluster)
    metadata(sce_sub)$cluster_codes <- data.frame(celltype = droplevels(sce_sub$metacluster))
    col_vec <- metadata(spe)$color_vectors$col_metacluster
  }

  # --- 4. Set sample_id and experiment_info ---
  sce_sub$sample_id <- factor(paste0(sce_sub$progress, "_", sce_sub$mna, "_",
                                     sce_sub$patient, "_", sce_sub$fm_id, "_", sce_sub$tissue))

  metadata(sce_sub)$experiment_info <- as.data.frame(colData(sce_sub)) %>%
    group_by(sample_id, patient, mna, condition, progress) %>%
    summarise(n_cells = n(), .groups = 'drop') %>%
    as.data.frame()

  # --- 5. Abundance barplot per sample ---
  p_abund <- plotAbundances(sce_sub, k = "celltype", by = "sample", col_clust = TRUE,
                            k_pal = col_vec, linkage = "ward.D", distance = "manhattan")
  print(p_abund)
  save_plot(p_abund, paste0("Fig_abundance_", fig_suffix, ".png"), width = 10, height = 6)

  # --- 6. Celltype abundance summaries (for celltype-level analyses) ---
  p_abund_bar <- NULL
  p_abund_box <- NULL
  if (cluster_level == "celltype") {
    abundance_data <- p_abund$data %>%
      mutate(
        cluster_id = factor(cluster_id, levels = levels(sce_sub$cluster_id)),
        freq_value = Freq
      )

    abundance_summary <- abundance_data %>%
      group_by(condition, cluster_id) %>%
      summarise(
        mean_freq = mean(freq_value, na.rm = TRUE),
        sd_freq = sd(freq_value, na.rm = TRUE),
        n = n(),
        se_freq = if_else(n > 1 & !is.na(sd_freq), sd_freq / sqrt(n), NA_real_),
        .groups = "drop"
      )

    p_abund_bar <- ggplot(abundance_summary, aes(x = cluster_id, y = mean_freq, fill = cluster_id)) +
      geom_col(width = 0.75, alpha = 0.45, show.legend = FALSE) +
      geom_errorbar(
        aes(ymin = mean_freq - se_freq, ymax = mean_freq + se_freq),
        width = 0.2,
        na.rm = TRUE
      ) +
      geom_jitter(
        data = abundance_data,
        aes(x = cluster_id, y = freq_value, color = cluster_id),
        inherit.aes = FALSE,
        width = 0.15,
        alpha = 0.8,
        size = 1.5,
        show.legend = FALSE
      ) +
      facet_wrap(~condition, nrow = 1) +
      scale_fill_manual(values = col_vec) +
      scale_color_manual(values = col_vec) +
      ylab("% abundance") +
      xlab("celltype") +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold")
      )
    print(p_abund_bar)
    save_plot(p_abund_bar, paste0("Fig_abundance_celltype_barplot_", fig_suffix, ".png"), width = 12, height = 6)

    p_abund_box <- ggplot(abundance_data, aes(x = cluster_id, y = freq_value, fill = cluster_id, color = cluster_id)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.45, width = 0.7, show.legend = FALSE) +
      geom_jitter(width = 0.15, alpha = 0.8, size = 1.5, show.legend = FALSE) +
      facet_wrap(~condition, nrow = 1) +
      scale_fill_manual(values = col_vec) +
      scale_color_manual(values = col_vec) +
      ylab("% abundance") +
      xlab("celltype") +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold")
      )
    print(p_abund_box)
    save_plot(p_abund_box, paste0("Fig_abundance_celltype_boxplot_", fig_suffix, ".png"), width = 12, height = 6)
  }

  # --- 7. Prepare data for diffcyt ---
  df_list <- lapply(unique(sce_sub$sample_id), function(x) {
    as.data.frame(t(counts(sce_sub[, sce_sub$sample_id == x])))
  })
  names(df_list) <- unique(sce_sub$sample_id)

  experiment_info <- metadata(sce_sub)$experiment_info
  experiment_info <- experiment_info[match(names(df_list), experiment_info$sample_id), ]

  d_se <- prepareData(df_list, experiment_info, marker_info)
  rowData(d_se)$cluster_id <- sce_sub[, rownames(d_se)]$cluster_id
  metadata(d_se)$cluster_codes <- metadata(sce_sub)$cluster_codes
  d_se      <- transformData(d_se)
  d_counts  <- calcCounts(d_se)

  # --- 8. DA test ---
  design   <- createDesignMatrix(experiment_info, cols_design = "condition")
  # contrast c(0,1) tests second condition level vs first (baseline).
  # logFC from testDA_voom is therefore log2(second/first); we negate it
  # throughout so reported values represent log2(g2/g1) in our ordering.
  contrast <- createContrast(c(0, 1))
  min_samples <- if (!is.null(min_samples_frac)) round(ncol(d_counts) * min_samples_frac) else 1

  res_DA <- testDA_voom(d_counts, design, contrast, min_cells = 3, min_samples = min_samples)
  FDR_cutoff <- 0.1
  print(rowData(res_DA)[rowData(res_DA)$p_adj < FDR_cutoff, ])
  res <- data.frame(rowData(res_DA))

  # --- 9. Boxplots per cluster ---
  plot_list_fc <- list()
  for (i in seq_along(unique(p_abund$data$cluster_id))) {
    cluster_name <- unique(p_abund$data$cluster_id)[i]
    proportions  <- data.frame(p_abund$data[p_abund$data$cluster_id == cluster_name,
                                            c("sample_id", "condition", "Freq")])

    if (as.character(cluster_name) %in% rownames(res)) {
      p_adj_val <- round(res[as.character(cluster_name), "p_adj"], 3)
      # logFC is negated to show fold-change in the direction of the second group (g2 vs g1)
      logFC_val <- -1 * round(res[as.character(cluster_name), "logFC"], 2)
    } else {
      p_adj_val <- NA; logFC_val <- NA
    }

    stats_df <- data.frame(p_adj = p_adj_val, group1 = g1, group2 = g2, logFC = logFC_val)
    stats_df$p_adj_signif <- ifelse(!is.na(p_adj_val), stars.pval(p_adj_val), "ns")
    stats_df$p_adj_signif[stats_df$p_adj_signif == " "] <- "ns"
    stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ") (", stats_df$p_adj_signif, ")")

    color <- unname(col_vec[as.character(cluster_name)])
    if (is.na(color) || is.null(color)) {
      warning("No color found for cluster: ", as.character(cluster_name), ". Using gray50.")
      color <- "gray50"
    }

    plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order = ord,
                color = "condition", palette = c(color, color),
                add.params = list(size = 1.5),
                add = "jitter",
                title = as.character(cluster_name), show.legend = FALSE) +
      stat_pvalue_manual(stats_df,
                         y.position = max(proportions$Freq, na.rm = TRUE),
                         step.increase = 0.4,
                         label = "p_fc") +
      theme(aspect.ratio = 1.5, legend.position = "none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
  }

  p_boxplot <- grid.arrange(grobs = plot_list_fc, ncol = 6)
  save_plot(p_boxplot, paste0("Fig_boxplot_", fig_suffix, ".png"), width = 12, height = 9)

  invisible(list(res = res, p_abund = p_abund, p_abund_bar = p_abund_bar,
                 p_abund_box = p_abund_box, p_boxplot = p_boxplot))
}
```

## Run differential abundance analyses

We perform DA at three levels (metacluster, tumor celltype, immune celltype) and two comparisons (MYCN status, tumor community).

```{r run-all-analyses, fig.width=12, fig.height=9}
immune_metaclusters <- c("granulocytes", "T cells", "B cells", "MO/DC/NK")

analysis_configs <- list(
  list(
    name            = "metacluster_mna",
    cell_filter     = rep(TRUE, ncol(sce)),
    cluster_level   = "metacluster",
    comparison      = "mna",
    ord             = c("nMNA", "MNA"),
    min_samples_frac = NULL,
    fig_suffix      = "metacluster_mna"
  ),
  list(
    name            = "metacluster_community",
    cell_filter     = rep(TRUE, ncol(sce)),
    cluster_level   = "metacluster",
    comparison      = "tumor_community",
    ord             = c("1", "2"),
    min_samples_frac = NULL,
    fig_suffix      = "metacluster_community"
  ),
  list(
    name            = "celltype_tumor_mna",
    cell_filter     = sce$metacluster == "tumor",
    cluster_level   = "celltype",
    comparison      = "mna",
    ord             = c("nMNA", "MNA"),
    min_samples_frac = 0.2,
    fig_suffix      = "celltype_tumor_mna"
  ),
  list(
    name            = "celltype_tumor_community",
    cell_filter     = sce$metacluster == "tumor",
    cluster_level   = "celltype",
    comparison      = "tumor_community",
    ord             = c("1", "2"),
    min_samples_frac = 0.2,
    fig_suffix      = "celltype_tumor_community"
  ),
  list(
    name            = "celltype_immune_mna",
    cell_filter     = sce$metacluster %in% immune_metaclusters,
    cluster_level   = "celltype",
    comparison      = "mna",
    ord             = c("nMNA", "MNA"),
    min_samples_frac = NULL,
    fig_suffix      = "celltype_immune_mna"
  ),
  list(
    name            = "celltype_immune_community",
    cell_filter     = sce$metacluster %in% immune_metaclusters,
    cluster_level   = "celltype",
    comparison      = "tumor_community",
    ord             = c("1", "2"),
    min_samples_frac = NULL,
    fig_suffix      = "celltype_immune_community"
  )
)

da_results <- list()
for (cfg in analysis_configs) {
  message("\n### Running DA analysis: ", cfg$name, " ###\n")
  da_results[[cfg$name]] <- run_da_analysis(
    sce_base         = sce,
    cell_filter      = cfg$cell_filter,
    cluster_level    = cfg$cluster_level,
    comparison       = cfg$comparison,
    ord              = cfg$ord,
    min_samples_frac = cfg$min_samples_frac,
    fig_suffix       = cfg$fig_suffix
  )
}
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
