---
title: "Spatial analysis in primary tumor samples"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# Plots directory
plot_dir <- file.path(params$output, "plots/07_spatial_analysis")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot (ggplot / ComplexHeatmap / base) into a figure subfolder
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  # Route figures into subfolders by filename. The compartment check runs before
  # patch so Fig4g3 panels (named "..._patch_...") land under compartments/.
  subdir <- if (grepl("_spatial_neighborhood|Fig4d_combined_stacked|Fig4d1|Fig4d2|Fig4d3", filename)) "neighborhood" else
            if (grepl("compartment|stackedbar|bordersize|Fig4g3|Fig4m", filename))                    "compartments" else
            if (grepl("patch|border_distance|Fig4h_DA|Fig4i_DA|Fig4j", filename))                      "patch"        else ""
  out_dir   <- if (nzchar(subdir)) file.path(plot_dir, subdir) else plot_dir
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  full_path <- file.path(out_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {
    
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
    
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {
    
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
    
  } else {
    
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  
  message("Saved: ", full_path)
}

# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Spatial analysis in PT {#sp-pt}

Herein, we explore spatial interactions in the primary tumor. 
Due to a [bug](https://github.com/BodenmillerGroup/imcRtools) in the `testInteractions` function in imcRtools versions < 1.5.5, we provide another docker image with a newer imcRtools version for spatial analysis. Please pull *lazdaria/mapmetsc_spatial:v1.0* from [docker hub]() and for the present RMD file.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(imcRtools)
library(scales)
library(BiocParallel)
library(RColorBrewer)
library(circlize)
library(patchwork)
library(ggplot2)
library(lisaClust)
library(tidyr)
library(diffcyt)
library(dplyr)
library(CATALYST)
library(ComplexHeatmap)
library(gtools)
library(ggpubr)
library(gridExtra)
library(ggridges)
library(forcats)
library(pheatmap)
library(tidyverse)
library(ggdendro)
library(lme4)
```
</details>

## Read data
First, we load the phenotyped single-cell data ([chapter 4](#phenotyping)) from the 05.1 checkpoint, which also carries the cellular-community (`tumor_community`) annotation. We perform spatial analysis only in the primary tumor samples, since the bone marrow samples are aspirates, where the spatial context is lost. Therefore, we first subset the SPE object to the PT samples and then load cell neighbors that were computed in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP). Therein, we have extracted neighbors by expanding individual high-resolution cell masks by 15 pixels. We load the cell neighbors for the primary-tumour samples directly with `read_steinbock`, using its `pattern` argument to select the `TU` samples from the shared `intensities-0px` folder (no file duplication needed).

```{r temp-folder}
# No temp folder / file duplication needed: read_steinbock() below selects the
# TU samples directly via its `pattern` argument.
```

We can now read in the neighbors.
For the spatial interaction analysis, we will use the metaclusters for non-tumor cells, since these represent the minority in the PT samples, and the celltype level annotations for the tumor cells (majority).

```{r read-data}
# Load the phenotyped object from 05.1 (carries the tumor_community column used below)
spe <- readRDS(file.path(params$output,"spe_05.1_PT_DE_cellularcomm.rds"))
spe <- spe[,spe$tissue=="PT"]

spe_neighbors <- read_steinbock(path=file.path(params$input), image_file=NULL, panel_file = NULL, intensities_folder = "intensities-0px",regionprops_folder = NULL, graphs_folder="neighbors", pattern = "TU")
colnames(spe_neighbors) <- paste0(spe_neighbors$sample_id, "_", spe_neighbors$ObjectNumber)
colPair(spe, "neighborhood") <- colPair(spe_neighbors[,colnames(spe)], "neighborhood")

#celltype 2
spe$celltype_2 <- spe$metacluster
spe[,spe$metacluster=="tumor"]$celltype_2 <- spe[,spe$metacluster=="tumor"]$celltype

# generate new color vector
tumor_names <- unique(spe[,spe$metacluster=="tumor"]$celltype)
metadata(spe)$color_vectors$col_celltype_2 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_2 <- c(metadata(spe)$color_vectors$col_celltype_2, metadata(spe)$color_vectors$col_celltype[tumor_names])

#celltype 3
spe$celltype_3 <- spe$metacluster
spe[,spe$metacluster=="T cells" | spe$metacluster=="B cells" | spe$metacluster=="granulocytes" | spe$metacluster=="myeloid/NK"]$celltype_3 <- "immune"

# generate new color vector
metadata(spe)$color_vectors$col_celltype_3 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_3 <- c(metadata(spe)$color_vectors$col_celltype_3, "immune" = "red")
```

## Interaction analysis

Next, we determine the interactions between cells on the metacluster level. For each image, the `testInteractions` function computes the averaged cell type interactions and compares this number against an empirical null distribution generated by permuting all cell labels, while maintaining tissue structure.

```{r interaction-analysis}
out <- testInteractions(spe,
                        group_by = "sample_id",
                        label = "metacluster",
                        colPairName = "neighborhood",
                        BPPARAM = SerialParam(RNGseed = 221029))
```

These interactions can be shown in a heatmap.

```{r interaction-heatmap, fig.width=5, fig.height=5}

data <- out %>% as_tibble() %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

data_wide <- data %>%
  pivot_wider(
    names_from = to_label,
    values_from = sum_sigval
  )

data_wide <- tibble::column_to_rownames(data_wide, var = "from_label")

# Fixed symmetric color range (+/-50), so breaks are centred on 0
max_range <- max(abs(50), abs(50))
breaks <- seq(-max_range, max_range, length.out = 101)

# blue-white-red scale
colors <- colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1)

p_heatmap <- pheatmap(data_wide,
         color = colors,
         breaks = breaks,
         cellwidth = 20,
         cellheight = 20,
         fontsize = 10,
         display_numbers = TRUE,
         number_format = "%.0f")

save_plot(p_heatmap, "Fig4b_cell-cell_interactions.pdf", width = 6, height = 6)
```

## Cellular neighborhoods

In a next step, we detect cellular neighborhoods (CNs). For each cell, we compute the fraction of each cell type (celltype_2) among its neighbors, and then aggregate cells into a predefined number of neighborhoods (here 6) by k-means.

```{r c-neighborhoods cn-analysis, fig.width=20, fig.height=50}
spe <- aggregateNeighbors(spe, 
                          colPairName = "neighborhood", 
                          aggregate_by = "metadata", 
                          count_by = "celltype_2")

set.seed(220705)

n_neighborhoods <- 6

cn_1 <- kmeans(spe$aggregatedNeighbors, centers = n_neighborhoods)
spe$cn_celltypes_2 <- as.factor(cn_1$cluster)

p_spatial <- plotSpatial(spe, 
            node_color_by = "cn_celltypes_2", 
            img_id = "sample_id", 
            ncols=6,
            node_size_fix = 0.5) +
    scale_color_brewer(palette = "Set3")

save_plot(p_spatial, "Fig4f_spatial_neighborhood_images.pdf", width = 20, height = 49)

p_spatial
```

Lastly, we can visualize the enrichment of cell types within cellular neighborhoods
using the `regionMap` function of the `lisaClust` package.

```{r regionmap, fig.height=6, fig.width=5}
library(lisaClust)
p_map <- regionMap(spe,
          cellType = "celltype_2",
          region = "cn_celltypes_2")

p_map <- p_map + ggplot2::coord_flip() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_map, "Fig4c_spatial_neighborhood_map.pdf", width = 8, height = 4)

p_map
```
         
### Differential abundance of CNs

Finally, we will generate a barplot showing the proportions of CNs per sample. To do so, we first transform the SPE object into a CATALYST-compatible SingleCellExperiment.

```{r transform-data, fig.height=6, fig.width=10}
sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts"), exprs=assay(spe, "counts")))

#recode metadata on progression
sce$progress <- factor(spe$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$tumor_community <- factor(spe$tumor_community)

sce$patient <- factor(spe$patient_id)
sce$fm_id <- spe$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe$cn_celltypes_2)
sce$metacluster <- factor(spe$metacluster)

sce$condition <- 'na' 

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe$cn_celltypes_2))
```

Next, we will generate a barplot showing the proportion of cells assigned to each of the 6 CNs per sample.

```{r cohort-overview}
set.seed(20231103)

distance <- "manhattan"
linkage <- "complete"

p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=brewer.pal(n_neighborhoods, "Set3"), 
                    linkage = linkage, distance= distance) 

save_plot(p, "Fig4d1_spatial_neighborhood_clustering.pdf", width = 6, height = 4)

p
```

We generate a dendogram for the hierarchical clustering of barplots.

```{r dendrogram, fig.height=8, fig.width=7}
set.seed(20231103)
wide_df <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

wide_df <- data.frame(wide_df)
rownames(wide_df) <- wide_df$sample_id
wide_df <- wide_df %>% select(-sample_id)

dist_matrix <- dist(wide_df, method = distance)
hc <- hclust(dist_matrix, method = linkage)

p2 <- plot(hc)

save_plot(p2, "Fig4d2_spatial_neighborhood_clustering_dendogram.pdf", width = 6, height = 6)
p2
```

In addition, we plot the metadata of these samples in the respective order.

```{r plot-metadata, fig.height=3}
clustered_samples <- levels(p$data$sample_id)
sce$sample_id <- factor(sce$sample_id, levels=clustered_samples)

#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
            tumor_community=first(tumor_community),
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


# Sample-annotation colors from 00_MapMet_color_code.R
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }
col_code <- COL_CODE

# Explicit patient-group palette (col_code's 1/2/3 = blue/red/green is used
# elsewhere); shared with the combined Fig4d strip below.
community_cols <- c("1" = "#E6C700", "2" = "#C71585", "3" = "#4DAF4A")


# Plot metadata
p_anno<- columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(patient_group=as.character(sample_summary$tumor_community), col=list(patient_group=community_cols))

save_plot(p_anno, "Fig4d3_spatial_neighborhood_clustering_metadata.pdf", width = 8, height = 4)

p_anno
```

Finally, we assemble the dendrogram, the CN-abundance barplot and the three metadata strips into a single stacked figure (Fig4d).

```{r fig4d-combined-stacked, fig.height=8.4, fig.width=8}

ddata <- dendro_data(hc, type = "rectangle")
segs <- segment(ddata)

p_dendro <- ggplot(segs) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  scale_x_continuous(expand = expansion(add = 0.5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    plot.margin = margin(2, 2, 0, 2)
  ) +
  labs(title = "Sample clustering based on cellular neighborhood composition")

meta_df <- sample_summary[clustered_samples, c("mna", "progress", "tumor_community")]
meta_df$sample_id <- factor(rownames(meta_df), levels = clustered_samples)

# One annotation strip per track, each with its own fill scale so the three
# legends render separately. patient_group uses the explicit community palette.
make_meta_strip <- function(df, ann_col, legend_title, cols) {
  ggplot(df, aes(x = sample_id, y = 1, fill = as.character(.data[[ann_col]]))) +
    # explicit height (y is constant)
    geom_tile(height = 1) +
    scale_fill_manual(values = cols, na.value = "white", name = legend_title) +
    scale_x_discrete(expand = expansion(add = 0.5)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(y = legend_title) +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_text(angle = 0, hjust = 1, vjust = 0.5, size = 9),
      legend.position = "right",
      legend.key.size = unit(0.4, "cm"),
      panel.grid = element_blank(),
      plot.margin = margin(0, 2, 0, 2)
    )
}

p_meta_mna   <- make_meta_strip(meta_df, "mna",                   "MYCN_amp",      col_code)
p_meta_prog  <- make_meta_strip(meta_df, "progress",              "progression",   col_code)
p_meta_group <- make_meta_strip(meta_df, "tumor_community", "patient_group", community_cols)
# extra bottom margin so the last strip is not clipped
p_meta_group <- p_meta_group + theme(plot.margin = margin(0, 2, 2, 2))

# Abundance barplot: rename the CN legend and drop the placeholder "na" facet
p_abund <- p +
  scale_fill_manual(values = brewer.pal(n_neighborhoods, "Set3"),
                    name = "Cellular neighborhood") +
  facet_null() +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    plot.title = element_blank(),
    strip.text = element_blank(),
    strip.background = element_blank(),
    plot.margin = margin(0, 2, 0, 2)
  )

# Relative panel heights: dendrogram / abundance / three metadata strips
panel_heights <- c(1.2, 4, 0.25, 0.25, 0.25)
p_fig4d <- (p_dendro / p_abund / p_meta_mna / p_meta_prog / p_meta_group) +
  plot_layout(ncol = 1, heights = panel_heights, guides = "collect")

p_fig4d

save_plot(p_fig4d, "Fig4d_combined_stacked.pdf", width = 8, height = 8.4)
```

We then compare the abundance of cellular neighborhoods (CNs) between nMNA and MNA samples.
Therefore, we first prepare data for DA analysis using the `diffcyt` package.

```{r transform-data2, fig.height=6, fig.width=10}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$MYCN_amp==0 | spe$MYCN_amp==1
sce <- SingleCellExperiment(assays=list(counts=assay(spe[keep_rows,keep_cols], "counts"), exprs=assay(spe[keep_rows,keep_cols], "counts")))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows,keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows,keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows,keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows,keep_cols]$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe[keep_rows,keep_cols]$cn_celltypes_2)
sce$metacluster <- factor(spe[keep_rows,keep_cols]$metacluster)

sce$condition <- sce$mna 

# Add experiment_info
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows,keep_cols]$cn_celltypes_2))

# Define cell type and state markers
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")
```

```{r prepare-data}
#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe[keep_rows,]) %in% type_markers, "type",
                                    ifelse(rownames(spe[keep_rows,]) %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

d_se <- transformData(d_se)
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing CN proportions between the two genetic subgroups (nMNA vs MNA).

```{r da-analysis}
# Create design matrix
design <- createDesignMatrix(experiment_info, cols_design = "condition")

# Create contrast matrix
contrast <- createContrast(c(0, 1)) 
nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_voom(d_counts,design, contrast
                       )

FDR_cutoff = 0.1
rowData(res_DA)[which(rowData(res_DA)$p_adj<FDR_cutoff),]
res<- data.frame(rowData(res_DA))
```

Finally, we visualize DA results in boxplots and barplots showing the logFC.

```{r visualize-da2, fig.width=12, fig.height=3}
p <- plotAbundances(sce, k = "celltype", by = "sample") 
comparisons <- c("nMNAvsMNA")

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  color <- brewer.pal(n_neighborhoods, "Set3")[cluster_name]
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("nMNA", "MNA"),
              color = "condition", palette =c(color, color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

p_grid <- grid.arrange(grobs = plot_list_fc, ncol = 6)

save_plot(p_grid, "Fig4e2_spatial_neighborhood_DA_mycn_boxplot.pdf", width = 12, height = 3)

p_grid


```

and in dodge barplots.

```{r barplot-logFC2}
df <- res[, c("logFC", "p_adj")]
df$p_adj_signif <- stars.pval(df$p_adj)
df$p_adj_signif[is.na(df$p_adj_signif) | df$p_adj_signif==" "] <- "ns"
df$cluster <- rownames(df)

p <- ggplot(df, aes(fill=cluster, y=logFC, x=cluster)) + 
  ggtitle("DA - CNs") +
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = F) + 
  scale_fill_manual(values = brewer.pal(6, "Set3")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_text(aes(label = p_adj_signif), vjust = 0)

save_plot(p, "Fig4e1_spatial_neighborhood_DA_mycn_barplot.pdf", width = 8, height = 4)

p
```

## Patch detection

Subsequently, we detect tumor patches. We first detect connected sets of tumor cells with a predefined min. number of cells making up a patch and then we expand the concave hull around the patch by 1 pixel.

```{r patchDetection-1, fig.height=20, fig.width=20}
spe <- patchDetection(spe, 
                      patch_cells = spe$metacluster == "tumor",
                      img_id = "sample_id",
                      min_patch_size=20,
                      expand_by=1,
                      colPairName = "neighborhood")
```

We can now plot the tumor patch in the neighborhood graph of individual samples.

```{r plot-patch}
p_s1<- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "GD2_Gd155_mean", 
            assay_type = "counts",
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "neighborhood", 
            nodes_first = FALSE, 
            node_size_by = "area", 
            directed = FALSE,
            edge_color_fix = "grey") + 
    scale_size_continuous(range = c(0.1, 2)) +
    ggtitle("GD2")


save_plot(p_s1, "Fig4g1_tumor_patch.pdf", width = 6, height = 6)

p_s2 <- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "patch_id", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    theme(legend.position = "none")

save_plot(p_s2, "Fig4g2_tumor_patch.pdf", width = 6, height = 6)

(p_s1 | p_s2)
```


Then we determine, per sample, the frequency of cells within tumor patches belonging to each metacluster, and compare these frequencies with respect to MYCN amplification status.

```{r patch-freq-mycn, fig.width=14, fig.height=7}
variable <- "MYCN_amp"
plot_list <- list()
celltypes <- unique(spe$metacluster)

for(i in 1:length(unique(celltypes))){
  
  c <- celltypes[i]
  
  cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
      group_by(patch_id, sample) %>%
      summarize(cell_count = sum(metacluster == c),
                patch_size = n(),
                cell_freq = cell_count / patch_size) %>%
      filter(!is.na(patch_id)) 
    
  mean_cell_freq_per_sample <- cell_freq_per_patch %>%
    group_by(sample) %>%
    summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
    arrange(desc(mean_cell_freq))
  
  df <- data.frame(
    sample = spe$sample,
    condition = colData(spe)[,variable]
  )
  
  progression_per_sample <- df %>%
    group_by(sample) %>%
    summarise(condition = first(condition))
  
  combined_data <- mean_cell_freq_per_sample %>%
    left_join(progression_per_sample, by = "sample")
  
  # log2FC of mean frequency between the two condition levels
  freq_by_cond <- combined_data %>%
    group_by(condition) %>%
    summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE))
  grp <- sort(unique(as.character(freq_by_cond$condition)))
  log2fc_df <- data.frame(log2FC = log2(
    freq_by_cond$mean_freq[match(grp[2], as.character(freq_by_cond$condition))] /
    freq_by_cond$mean_freq[match(grp[1], as.character(freq_by_cond$condition))]))
  
  stats_df <- compare_means(mean_cell_freq ~ condition,  
                            method = "wilcox.test",  
                            data = combined_data) 
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                            " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_metacluster[as.character(c)])
  
  plot_list[[c]] <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                  color = "condition", palette = rep(color, length(unique(combined_data$condition))),
                  add.params = list(size = 3),
                  add = "jitter",
                  title=c) + 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                  theme(aspect.ratio = 1.6, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
  
}

p_box <- grid.arrange(grobs = plot_list, ncol = 6)

save_plot(p_box, paste0("Fig4i_DA_",variable,".pdf"), width = 16, height = 12)

p_box

```

We repeat the same comparison across patient groups (`tumor_community`).

```{r patch-freq-comm, fig.width=14, fig.height=7}
variable <- "tumor_community"
plot_list <- list()
celltypes <- unique(spe$metacluster)

for(i in 1:length(unique(celltypes))){
  
  c <- celltypes[i]
  
  cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
      group_by(patch_id, sample) %>%
      summarize(cell_count = sum(metacluster == c),
                patch_size = n(),
                cell_freq = cell_count / patch_size) %>%
      filter(!is.na(patch_id)) 
    
  mean_cell_freq_per_sample <- cell_freq_per_patch %>%
    group_by(sample) %>%
    summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
    arrange(desc(mean_cell_freq))
  
  df <- data.frame(
    sample = spe$sample,
    condition = colData(spe)[,variable]
  )
  
  progression_per_sample <- df %>%
    group_by(sample) %>%
    summarise(condition = first(condition))
  
  combined_data <- mean_cell_freq_per_sample %>%
    left_join(progression_per_sample, by = "sample")
  
  # log2FC of mean frequency between the two condition levels
  freq_by_cond <- combined_data %>%
    group_by(condition) %>%
    summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE))
  grp <- sort(unique(as.character(freq_by_cond$condition)))
  log2fc_df <- data.frame(log2FC = log2(
    freq_by_cond$mean_freq[match(grp[2], as.character(freq_by_cond$condition))] /
    freq_by_cond$mean_freq[match(grp[1], as.character(freq_by_cond$condition))]))
  
  stats_df <- compare_means(mean_cell_freq ~ condition,  
                            method = "wilcox.test",  
                            data = combined_data) 
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                            " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_metacluster[as.character(c)])
  
  plot_list[[c]] <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                  color = "condition", palette = rep(color, length(unique(combined_data$condition))),
                  add.params = list(size = 3),
                  add = "jitter",
                  title=c) + 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                  theme(aspect.ratio = 1.6, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
  
}

p_box <- grid.arrange(grobs = plot_list, ncol = 6)

save_plot(p_box, paste0("Fig4h_DA_",variable,".pdf"), width = 16, height = 12)

p_box

```

We repeat the same for immune cells overall, here comparing their patch frequency with respect to progression status.

```{r patch-freq-immune}
variable <- "progression"
  
c <- "immune"

cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
    group_by(patch_id, sample) %>%
    summarize(cell_count = sum(celltype_3 == c),
              patch_size = n(),
              cell_freq = cell_count / patch_size) %>%
    filter(!is.na(patch_id)) 
  
mean_cell_freq_per_sample <- cell_freq_per_patch %>%
  group_by(sample) %>%
  summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
  arrange(desc(mean_cell_freq))

df <- data.frame(
  sample = spe$sample,
  condition = colData(spe)[,variable]
)

progression_per_sample <- df %>%
  group_by(sample) %>%
  summarise(condition = first(condition))

combined_data <- mean_cell_freq_per_sample %>%
  left_join(progression_per_sample, by = "sample")

# log2FC of mean frequency between the two condition levels
freq_by_cond <- combined_data %>%
  group_by(condition) %>%
  summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE))
grp <- sort(unique(as.character(freq_by_cond$condition)))
log2fc_df <- data.frame(log2FC = log2(
  freq_by_cond$mean_freq[match(grp[2], as.character(freq_by_cond$condition))] /
  freq_by_cond$mean_freq[match(grp[1], as.character(freq_by_cond$condition))]))

stats_df <- compare_means(mean_cell_freq ~ condition,  
                          method = "wilcox.test",  
                          data = combined_data) 
  
stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                          " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")

color <- unname(metadata(spe)$color_vectors$col_celltype_3[as.character(c)])

p <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                color = "condition", palette =c(color, color),
                add.params = list(size = 3),
                add = "jitter",
                title=c) + 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                theme(aspect.ratio = 1.6, legend.position="none") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
p
```

### Distance to tumor-stroma border

Based on detected patches, the distance to the tumor stroma border can be computed. Therefore the [minDistToCells](https://bodenmillergroup.github.io/imcRtools/reference/minDistToCells.html) can be used. Negative values indicate cells within patches and positive values cells outside of patches.

```{r minDistCells}
spe <- minDistToCells(spe, 
                      x_cells = !is.na(spe$patch_id), 
                      img_id = "sample_id")

plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "distToCells", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    scale_color_gradient2(low = "dark blue", mid = "white", high = "dark red")
```

The distance distribution per (fine-grained) celltype can be shown in ridge
plots. The core/border/stroma compartments (defined below using a 200 px
threshold on this same distance) are marked in two ways: once as shaded
background regions, and once as a green/red/blue compartment bar beneath a
plain-background ridge plot.

```{r ridges-distance, fig.height=5,fig.width=8}
# Border threshold shown here matches `canonical_border` set in the
# spatial-compartments section below (kept as a literal since that section
# runs after this chunk).
border_threshold <- 200

dist_data <- as.data.frame(colData(spe)) %>%
  mutate(celltype = fct_reorder(celltype, distToCells, .fun = median, .desc = TRUE))

x_range <- range(dist_data$distToCells, na.rm = TRUE)

# Shared ridge layers for both variants below.
ridge_layers <- list(
  stat_density_ridges(aes(distToCells, celltype, fill = celltype),
                      quantiles = c(0.5), quantile_lines = TRUE, scale = 3,
                      # plain filled legend keys (no median line)
                      key_glyph = "rect"),
  geom_vline(xintercept = 0, color = "dark red", linewidth = 0.5),
  geom_vline(xintercept = c(-border_threshold, border_threshold),
             color = "black", linewidth = 0.5, linetype = "dashed"),
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype, name = "Celltypes"),
  labs(y = "Celltype"),
  theme(legend.position = "none")
)

# --- Variant 1: compartments shown as shaded background regions ---
p_dist_shaded <- ggplot(dist_data) +
  annotate("rect", xmin = -Inf, xmax = -border_threshold, ymin = -Inf, ymax = Inf,
           fill = COL_COMPARTMENT[["core"]], alpha = 0.15) +
  annotate("rect", xmin = -border_threshold, xmax = border_threshold, ymin = -Inf, ymax = Inf,
           fill = COL_COMPARTMENT[["border"]], alpha = 0.15) +
  annotate("rect", xmin = border_threshold, xmax = Inf, ymin = -Inf, ymax = Inf,
           fill = COL_COMPARTMENT[["stroma"]], alpha = 0.15) +
  ridge_layers +
  coord_cartesian(xlim = x_range, expand = FALSE) +
  theme_ridges()

save_plot(p_dist_shaded, "Fig4j_border_distance_shaded.pdf", width = 8, height = 5)

p_dist_shaded

# --- Variant 2: plain white background, compartments shown as a bar below ---
p_dist_plain <- ggplot(dist_data) +
  ridge_layers +
  coord_cartesian(xlim = x_range, expand = FALSE) +
  theme_ridges() +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

compartment_bar <- ggplot() +
  annotate("rect", xmin = -Inf, xmax = -border_threshold, ymin = 0, ymax = 1,
           fill = COL_COMPARTMENT[["core"]]) +
  annotate("rect", xmin = -border_threshold, xmax = border_threshold, ymin = 0, ymax = 1,
           fill = COL_COMPARTMENT[["border"]]) +
  annotate("rect", xmin = border_threshold, xmax = Inf, ymin = 0, ymax = 1,
           fill = COL_COMPARTMENT[["stroma"]]) +
  geom_vline(xintercept = c(-border_threshold, border_threshold),
             color = "black", linewidth = 0.5, linetype = "dashed") +
  coord_cartesian(xlim = x_range, ylim = c(0, 1), expand = FALSE) +
  labs(x = "distToCells") +
  theme_minimal() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), panel.grid = element_blank(),
        panel.border = element_blank())

p_dist_bar <- p_dist_plain / compartment_bar + plot_layout(heights = c(20, 1))

save_plot(p_dist_bar, "Fig4j_border_distance_bar.pdf", width = 8, height = 5)

p_dist_bar
```

Based on the calculated distance for each cell, three compartments - tumor core, tumor/stroma border and stroma - can be defined using a distance threshold. We sweep border sizes from 100 to 300 pixels, visualizing the compartments and counting immune and tumor cells in each, and then adopt a border of 200 pixels for all downstream analyses. Image `20221018_20220906_03-018_TU_001` lacks tumor patches, therefore all cells are defined as `stroma` for this image.

```{r spatial-compartments}
thresholds <- seq(100, 300, by = 50)
panel_sample_id_1 <- "20220926_20220809_16-006_TU_003"
# sanitize sample IDs for file-safe output tags
sanitize_sample_tag <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
panel_sample_tag_1 <- sanitize_sample_tag(panel_sample_id_1)
y_axis_upper_expand <- 0.1

plot_list_img1 <- list()
results_list <- list()

for (th in thresholds) {

  # reset compartment
  spe$compartment <- "stroma"

  # border
  spe[
    ,
    abs(spe$distToCells) <= th &
      spe$sample_id != "20221018_20220906_03-018_TU_001"
  ]$compartment <- "border"

  # core
  spe[
    ,
    spe$compartment != "border" &
      spe$distToCells <= th &
      spe$sample_id != "20221018_20220906_03-018_TU_001"
  ]$compartment <- "core"

  # per-threshold compartment map for the panel sample
  p_s2 <- plotSpatial(
    spe[, spe$sample_id == panel_sample_id_1],
    node_color_by = "compartment",
    img_id = "sample_id",
    node_size_fix = 0.5
  ) +
    scale_color_manual(values = COL_COMPARTMENT) +
    ggtitle(paste("Border size =", th))

  save_plot(
    p_s2,
    paste0("Fig4g3_tumor_patch_", th, ".pdf"),
    width = 6,
    height = 6
  )

  plot_list_img1[[as.character(th)]] <- p_s2

  # immune counts per compartment
  immune_df <- as.data.frame(
    table(spe$compartment[spe$celltype_3 == "immune"])
  )

  colnames(immune_df) <- c("compartment", "count")

  immune_df$threshold <- th
  immune_df$cell_class <- "immune"

  # tumor counts per compartment
  tumor_df <- as.data.frame(
    table(spe$compartment[spe$metacluster == "tumor"])
  )

  colnames(tumor_df) <- c("compartment", "count")

  tumor_df$threshold <- th
  tumor_df$cell_class <- "tumor"

  # store
  results_list[[paste0("immune_", th)]] <- immune_df
  results_list[[paste0("tumor_", th)]] <- tumor_df
}

# Canonical compartment definition (border = 200) used by all downstream
# analyses; the loop above leaves spe$compartment at its last threshold (300).
canonical_border <- 200
spe$compartment <- "stroma"
spe[
  ,
  abs(spe$distToCells) <= canonical_border &
    spe$sample_id != "20221018_20220906_03-018_TU_001"
]$compartment <- "border"
spe[
  ,
  spe$compartment != "border" &
    spe$distToCells <= canonical_border &
    spe$sample_id != "20221018_20220906_03-018_TU_001"
]$compartment <- "core"

# combine all
results_df <- bind_rows(results_list)

panel_img1 <- wrap_plots(plot_list_img1, ncol = length(thresholds)) +
  plot_annotation(title = panel_sample_id_1)

save_plot(
  panel_img1,
  paste0("Fig4g3_tumor_patch_panel_all_thresholds_", panel_sample_tag_1, ".pdf"),
  width = 30,
  height = 6
)

panel_img1

# order compartments
results_df$compartment <- factor(
  results_df$compartment,
  levels = c("core", "border", "stroma")
)

# immune counts by compartment, per border size
immune_plot <- ggplot(
  results_df %>% filter(cell_class == "immune"),
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  labs(
    title = "Immune cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    immune_plot,
    paste0("Fig4g3_tumor_bordersize_immune_barplot.pdf"),
    width = 6,
    height = 6
  )

immune_plot

# tumor counts by compartment, per border size
tumor_plot <- ggplot(
  results_df %>% filter(cell_class == "tumor"),
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  labs(
    title = "Tumor cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    tumor_plot,
    paste0("Fig4g3_tumor_bordersize_tumor_barplot.pdf"),
    width = 6,
    height = 6
  )

tumor_plot

# combined immune + tumor barplot (one facet each)
results_df$cell_class <- factor(results_df$cell_class, levels = c("immune", "tumor"))

immune_tumor_plot <- ggplot(
  results_df,
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  facet_wrap(~cell_class, ncol = 1, scales = "free_y") +
  labs(
    title = "Immune and tumor cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    immune_tumor_plot,
    paste0("Fig4g3_tumor_bordersize_immune_tumor_barplot.pdf"),
    width = 6,
    height = 10
  )

immune_tumor_plot
```

Analogously to the Fig4f cellular-neighborhood spatial map, we can also show all
primary-tumour images at once, this time coloring each cell by its tumor
compartment (core / border / stroma) rather than by its cellular neighborhood.

```{r spatial-compartment-images, fig.width=20, fig.height=50}
# All PT images colored by tumor compartment (spe$compartment at border 200)
p_spatial_compartment <- plotSpatial(spe,
            node_color_by = "compartment",
            img_id = "sample_id",
            ncols = 6,
            node_size_fix = 0.5) +
    scale_color_manual(values = COL_COMPARTMENT)

save_plot(p_spatial_compartment, "Fig4f_spatial_compartment_images.pdf", width = 20, height = 49)

p_spatial_compartment
```

Similarly, we show all primary-tumour images at once colored by cell type
(`celltype_2`: tumor celltypes + non-tumor metaclusters).

```{r spatial-celltype-images, fig.width=20, fig.height=50}
# All PT images colored by cell type (celltype_2)
p_spatial_celltype <- plotSpatial(spe,
            node_color_by = "celltype_2",
            img_id = "sample_id",
            ncols = 6,
            node_size_fix = 0.5) +
    scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_2)

save_plot(p_spatial_celltype, "Fig4f_spatial_celltype_images.pdf", width = 20, height = 49)

p_spatial_celltype
```


### DA analysis in spatial compartments

Next, we will plot the differential abundance of cell types with respect to spatial compartment on a sample level. We analyse both tumor celltypes and immune metaclusters, looping over all three compartment comparison pairs (`core-border`, `core-stroma`, `border-stroma`). Differential abundance is tested with `diffcyt` (`testDA_edgeR`) controlling for sample-level pairing.

```{r define-markers}

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

```

```{r cluster-order-colors}
# Cluster display orders used throughout the compartment analyses (immune_order
# are metacluster names; tumor_order are current spe$celltype_2 names).
immune_order <- c("T cells", "B cells", "myeloid/NK", "granulocytes")

tumor_order <- c("adrenergic marker-low Tu", "GD2neg Tu", "CD44+ Tu", "Ki67+ Tu",
                 "neuroblast-like Tu", "early neuroblast-like Tu")
# keep only the celltypes actually present in the data (preserves order)
tumor_order <- intersect(tumor_order,
                         unique(as.character(spe$celltype_2[spe$metacluster == "tumor"])))

# Color lookup, falling back to grey if a celltype has no (or NA) color
cluster_color <- function(name) {
  col <- unname(metadata(spe)$color_vectors$col_celltype_2[as.character(name)])
  if (length(col) == 0 || is.na(col)) "grey50" else col
}
```






```{r compartment-boxplots-all-comparisons}
# For each cell class: run diffcyt (testDA_edgeR) for every compartment pair,
# then draw boxplots labelled with log2FC + adjusted p-value and stars.

cells_list       <- c("tumor", "immune")
comp_levels      <- c("core", "border", "stroma")
comparison_pairs <- list(c("core","border"), c("core","stroma"), c("border","stroma"))

# Freq is a percentage: panels use a fixed 0-100 y-axis with the pairwise
# brackets drawn above it (clip = "off").
y_axis_max        <- 100
bracket_base      <- 106
bracket_gap       <- 14
title_gap         <- 85
bracket_positions <- bracket_base + (seq_along(comparison_pairs) - 1) * bracket_gap
max_plot_columns  <- 5

keep_rows <- rowData(spe)$use_channel

# helper to build marker_info from the global marker vectors
.make_marker_info <- function(spe, keep_rows) {
  channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 2)
  marker_name  <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 1)
  marker_class <- ifelse(rownames(rowData(spe)[keep_rows, ]) %in% type_markers, "type",
                  ifelse(rownames(rowData(spe)[keep_rows, ]) %in% state_markers, "state", "other"))
  data.frame(channel_name, marker_name,
             marker_class = factor(marker_class),
             row.names = marker_name)
}

marker_info_da <- .make_marker_info(spe, keep_rows)

for (cells in cells_list) {

  order     <- if (cells == "immune") immune_order else tumor_order
  keep_meta <- if (cells == "immune") spe$metacluster %in% immune_order else spe$metacluster == "tumor"

  # 1. diffcyt DA results for every compartment pair
  diffcyt_results <- list()

  for (pair in comparison_pairs) {
    comps     <- pair
    key       <- paste(comps, collapse = "-")
    keep_cols <- keep_meta & spe$compartment %in% comps

    if (sum(keep_cols) < 10) { diffcyt_results[[key]] <- NULL; next }

    sce_da <- SingleCellExperiment(
      assays = list(
        counts = assay(spe, "counts")[keep_rows, keep_cols],
        exprs  = assay(spe, "counts")[keep_rows, keep_cols]
      )
    )
    sce_da$fm_id       <- spe[keep_rows, keep_cols]$fm_id
    sce_da$sample      <- factor(spe[keep_rows, keep_cols]$sample)
    sce_da$compartment <- factor(spe[keep_rows, keep_cols]$compartment, levels = comps)
    sce_da$sample_id   <- factor(paste0(sce_da$fm_id, "_", sce_da$compartment))
    sce_da$condition   <- factor(sce_da$compartment, levels = comps)
    sce_da$cluster_id  <- factor(spe[keep_rows, keep_cols]$celltype_2, levels = order)

    metadata(sce_da)$cluster_codes   <- data.frame(
      celltype = factor(spe[keep_rows, keep_cols]$celltype_2, levels = order)
    )
    metadata(sce_da)$experiment_info <- as.data.frame(colData(sce_da)) %>%
      dplyr::group_by(sample_id, condition, sample) %>%
      dplyr::summarise(n_cells = dplyr::n(), .groups = "drop") %>%
      as.data.frame()

    exp_info           <- metadata(sce_da)$experiment_info
    exp_info           <- exp_info[match(unique(sce_da$sample_id), exp_info$sample_id), ]
    exp_info$condition <- factor(exp_info$condition, levels = comps)

    df_list_da <- lapply(unique(sce_da$sample_id), function(x)
      as.data.frame(t(counts(sce_da[, sce_da$sample_id == x]))))
    names(df_list_da) <- unique(sce_da$sample_id)

    d_se_da <- prepareData(df_list_da, exp_info, marker_info_da)
    rowData(d_se_da)$cluster_id     <- sce_da[, rownames(d_se_da)]$cluster_id
    metadata(d_se_da)$cluster_codes <- metadata(sce_da)$cluster_codes
    d_se_da     <- transformData(d_se_da)
    d_counts_da <- calcCounts(d_se_da)

    design_da   <- createDesignMatrix(exp_info, cols_design = c("condition", "sample"))
    contrast_da <- createContrast(c(0, 1, rep(0, length(unique(exp_info$sample)) - 1)))

    res_DA_da <- tryCatch(
      testDA_edgeR(d_counts_da, design_da, contrast_da,
                   min_cells = 3,
                   min_samples = round(ncol(d_counts_da) * 0.2)),
      error = function(e) { message("diffcyt failed for ", key, ": ", e$message); NULL }
    )
    diffcyt_results[[key]] <- if (!is.null(res_DA_da)) data.frame(rowData(res_DA_da)) else NULL
  }

  # 2. Sample-level proportions (% within compartment) for plotting
  comp_df <- data.frame(
    fm_id       = spe$fm_id,
    compartment = spe$compartment,
    celltype_2  = spe$celltype_2,
    keep        = keep_meta
  )
  comp_df <- comp_df[comp_df$keep & comp_df$compartment %in% comp_levels, ]
  comp_df$compartment <- factor(comp_df$compartment, levels = comp_levels)

  sample_totals <- comp_df %>%
    dplyr::group_by(fm_id, compartment) %>%
    dplyr::summarise(total = dplyr::n(), .groups = "drop")

  plot_data_list <- list()
  for (cluster_name in order) {
    cluster_counts <- comp_df %>%
      dplyr::filter(celltype_2 == cluster_name) %>%
      dplyr::group_by(fm_id, compartment) %>%
      dplyr::summarise(n = dplyr::n(), .groups = "drop")

    plot_data_list[[cluster_name]] <- dplyr::left_join(
      sample_totals, cluster_counts, by = c("fm_id", "compartment")
    ) %>%
      dplyr::mutate(
        n         = dplyr::if_else(is.na(n), 0L, n),
        Freq      = 100 * n / total,
        condition = factor(compartment, levels = comp_levels),
        sample_id = fm_id
      )
  }

  # 3. Collate diffcyt stats for all pairs into one table
  all_stats_list <- list()
  for (cluster_name in order) {
    for (pair in comparison_pairs) {
      key      <- paste(pair, collapse = "-")
      res_pair <- diffcyt_results[[key]]
      if (!is.null(res_pair) && cluster_name %in% rownames(res_pair)) {
        p_adj <- res_pair[cluster_name, "p_adj"]
        logFC <- res_pair[cluster_name, "logFC"]
      } else {
        p_adj <- NA_real_; logFC <- NA_real_
      }
      p_signif <- if (!is.na(p_adj)) gtools::stars.pval(p_adj) else "NA"
      if (p_signif == " ") p_signif <- "ns"
      all_stats_list <- c(all_stats_list, list(data.frame(
        cluster_name = cluster_name,
        group1       = pair[1],
        group2       = pair[2],
        p_adj        = p_adj,
        logFC        = round(logFC, 2),
        label        = paste0(round(logFC, 2), " (",
                              if (!is.na(p_adj)) sprintf("%.3f", p_adj) else "NA",
                              " ", p_signif, ")")
      )))
    }
  }
  all_stats <- dplyr::bind_rows(all_stats_list)

  # 4. Boxplots with diffcyt annotation
  plot_list_comp <- list()
  for (cluster_name in order) {
    proportions <- plot_data_list[[cluster_name]]
    stats_df    <- all_stats[all_stats$cluster_name == cluster_name, ]

    # fixed bracket heights, identical for every panel
    stats_df$y.position <- bracket_positions[seq_len(nrow(stats_df))]

    color_cluster <- cluster_color(cluster_name)

    plot_list_comp[[cluster_name]] <- ggboxplot(
      proportions, x = "condition", y = "Freq", order = comp_levels,
      color = "condition", palette = rep(color_cluster, 3),
      add = "jitter"
    ) +
      stat_pvalue_manual(stats_df, y.position = "y.position", label = "label") +
      # wrap long celltype names so they are not cropped in narrow panels
      ggtitle(str_wrap(cluster_name, width = 18)) +
      scale_y_continuous(breaks = seq(0, y_axis_max, 25),
                         expand = expansion(mult = c(0.02, 0))) +
      # lock axis to 0-100; brackets draw outside it
      coord_cartesian(ylim = c(0, y_axis_max), clip = "off") +
      theme(aspect.ratio = 1.0, legend.position = "none",
            plot.title  = element_text(hjust = 0.5, size = 14,
                                       margin = margin(b = title_gap)),
            plot.margin = margin(t = 10, r = 8, b = 8, l = 8))
  }

  p_comp <- gridExtra::grid.arrange(
    grobs = plot_list_comp,
    ncol  = min(length(plot_list_comp), max_plot_columns)
  )

  save_plot(p_comp,
            paste0("Fig4m_boxplot_compartments_", cells, ".pdf"),
            width = 12, height = 8)
}
```

Finally, we summarize the cell-type composition of each compartment as stacked bars, drawn for every border size and shown for immune metaclusters, tumor celltypes, and the two pooled together.

```{r composition-stacked-bars}
# spe$compartment (the canonical-border assignment) is left untouched;
# each border's compartment is computed on the fly.

comp_levels <- c("core", "border", "stroma")

# Compartment (core / border / stroma) for a given border size, as a plain
# vector (same rule as the threshold loop; does not modify spe).
compartment_at <- function(spe, border) {
  comp <- rep("stroma", ncol(spe))
  ok   <- spe$sample_id != "20221018_20220906_03-018_TU_001"
  comp[which(abs(spe$distToCells) <= border & ok)] <- "border"
  comp[which(comp != "border" & spe$distToCells <= border & ok)] <- "core"
  comp
}

# One stacked-bar figure for a compartment vector and a cell class: "immune"
# (metaclusters), "tumor" (celltypes) or "tumor_immune" (the two pooled).
make_compartment_stackedbar <- function(comp_vec, cells, border) {
  if (cells == "immune") {
    keep <- spe$metacluster %in% immune_order;                       order <- immune_order
  } else if (cells == "tumor") {
    keep <- spe$metacluster == "tumor";                              order <- tumor_order
  } else {                                                           # tumor_immune (pooled)
    keep <- spe$metacluster == "tumor" | spe$metacluster %in% immune_order
    order <- c(tumor_order, immune_order)
  }

  df_bar <- data.frame(compartment = comp_vec, cluster_id = spe$celltype_2, keep = keep)
  df_bar <- df_bar[df_bar$keep & df_bar$compartment %in% comp_levels, ]

  totals <- df_bar %>% dplyr::group_by(compartment) %>%
    dplyr::summarise(N = dplyr::n(), .groups = "drop")

  df_prop <- df_bar %>%
    dplyr::group_by(compartment, cluster_id) %>% dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
    dplyr::group_by(compartment) %>% dplyr::mutate(pct = 100 * n / sum(n)) %>% dplyr::ungroup() %>%
    dplyr::mutate(compartment = factor(compartment, levels = comp_levels),
                  cluster_id  = factor(cluster_id,  levels = order))

  color_vec <- metadata(spe)$color_vectors$col_celltype_2[order]
  lab       <- if (cells == "tumor_immune") "tumor + immune" else cells

  p_stack <- ggplot(df_prop, aes(x = compartment, y = pct, fill = cluster_id)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(
      data = totals %>% dplyr::mutate(compartment = factor(compartment, levels = comp_levels)),
      aes(x = compartment, y = 103, label = paste0("n=", N)),
      inherit.aes = FALSE, vjust = 0, size = 3.5) +
    scale_fill_manual(values = color_vec, drop = FALSE, name = "Cell type") +
    scale_y_continuous(limits = c(0, 115), breaks = seq(0, 100, 25),
                       expand = expansion(mult = c(0, 0))) +
    labs(title = paste0("Compartment composition (", lab, ") - border ", border),
         x = "Compartment", y = "% cells") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.key.size = unit(0.4, "cm"))

  save_plot(p_stack,
            paste0("Fig4_suppl_stackedbar_compartments_", cells, "_bordersize_", border, ".pdf"),
            width = if (cells == "tumor_immune") 7 else 6, height = 5)
  print(p_stack)
}

for (border in thresholds) {
  comp_vec <- compartment_at(spe, border)
  for (cells in c("immune", "tumor", "tumor_immune"))
    make_compartment_stackedbar(comp_vec, cells, border)
}
```



## Save R object

Finally, we save the R object.

```{r save-rds}
saveRDS(spe, file.path(params$output,"spe_07_spatial_analysis.rds"))
```


## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
