---
title: "Spatial analysis in primary tumor samples"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# Plots directory
plot_dir <- file.path(params$output, "plots/07_spatial_analysis")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# Save a plot (ggplot / ComplexHeatmap / base) into a figure subfolder
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  # Route every figure into exactly one of three subfolders by filename:
  # cell-cell_interaction (Fig4b), neighborhood_level (anything CN/neighborhood-
  # related, incl. the combined Fig4d stack), celltype_level (everything else -
  # tumor/immune celltype breakdowns, spatial compartments, border distance by
  # celltype).
  subdir <- if (grepl("cell-cell_interactions", filename))               "cell-cell_interaction" else
            if (grepl("neighborhood|Fig4d_combined_stacked", filename))  "neighborhood_level"    else
                                                                          "celltype_level"
  out_dir   <- file.path(plot_dir, subdir)
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  full_path <- file.path(out_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {
    
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
    
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {
    
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
    
  } else {
    
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  
  message("Saved: ", full_path)
}

# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Spatial analysis in PT {#sp-pt}

Herein, we explore spatial interactions in the primary tumor. 
Due to a [bug](https://github.com/BodenmillerGroup/imcRtools) in the `testInteractions` function in imcRtools versions < 1.5.5, we provide another docker image with a newer imcRtools version for spatial analysis. Please pull *lazdaria/mapmetsc_spatial:v1.0* from [docker hub]() and for the present RMD file.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(imcRtools)
library(scales)
library(BiocParallel)
library(RColorBrewer)
library(circlize)
library(patchwork)
library(ggplot2)
library(lisaClust)
library(tidyr)
library(diffcyt)
library(dplyr)
library(CATALYST)
library(ComplexHeatmap)
library(gtools)
library(ggpubr)
library(gridExtra)
library(ggridges)
library(forcats)
library(pheatmap)
library(tidyverse)
library(ggdendro)
library(lme4)
```
</details>

## Read data
First, we load the phenotyped single-cell data ([chapter 4](#phenotyping)) from the 05.1 checkpoint, which also carries the cellular-community (`tumor_community`) annotation. We perform spatial analysis only in the primary tumor samples, since the bone marrow samples are aspirates, where the spatial context is lost. Therefore, we first subset the SPE object to the PT samples and then load cell neighbors that were computed in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP). Therein, we have extracted neighbors by expanding individual high-resolution cell masks by 15 pixels. We load the cell neighbors for the primary-tumour samples directly with `read_steinbock`, using its `pattern` argument to select the `TU` samples from the shared `intensities-0px` folder (no file duplication needed).

```{r temp-folder}
# No temp folder / file duplication needed: read_steinbock() below selects the
# TU samples directly via its `pattern` argument.
```

We can now read in the neighbors.
For the spatial interaction analysis, we will use the metaclusters for non-tumor cells, since these represent the minority in the PT samples, and the celltype level annotations for the tumor cells (majority).

```{r read-data}
# Sample-annotation and structural colors (COL_CODE, COL_COMPARTMENT, COL_CN)
# from 00_MapMet_color_code.R - sourced early so they are available as soon as
# the corresponding columns are derived below / further down this file.
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }

# Load the phenotyped object from 05.1 (carries the tumor_community column used below)
spe <- readRDS(file.path(params$output,"spe_05.1_PT_DE_cellularcomm.rds"))
spe <- spe[,spe$tissue=="PT"]

spe_neighbors <- read_steinbock(path=file.path(params$input), image_file=NULL, panel_file = NULL, intensities_folder = "intensities-0px",regionprops_folder = NULL, graphs_folder="neighbors", pattern = "TU")
colnames(spe_neighbors) <- paste0(spe_neighbors$sample_id, "_", spe_neighbors$ObjectNumber)
colPair(spe, "neighborhood") <- colPair(spe_neighbors[,colnames(spe)], "neighborhood")

#celltype 2
spe$celltype_2 <- spe$metacluster
spe[,spe$metacluster=="tumor"]$celltype_2 <- spe[,spe$metacluster=="tumor"]$celltype

# generate new color vector
tumor_names <- unique(spe[,spe$metacluster=="tumor"]$celltype)
metadata(spe)$color_vectors$col_celltype_2 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_2 <- c(metadata(spe)$color_vectors$col_celltype_2, metadata(spe)$color_vectors$col_celltype[tumor_names])

#celltype 3 (used below to count "immune" cells per compartment)
spe$celltype_3 <- spe$metacluster
spe[,spe$metacluster=="T cells" | spe$metacluster=="B cells" | spe$metacluster=="granulocytes" | spe$metacluster=="myeloid/NK"]$celltype_3 <- "immune"
```

## Interaction analysis

Next, we determine the interactions between cells on the metacluster level. For each image, the `testInteractions` function computes the averaged cell type interactions and compares this number against an empirical null distribution generated by permuting all cell labels, while maintaining tissue structure.

```{r interaction-analysis}
out <- testInteractions(spe,
                        group_by = "sample_id",
                        label = "metacluster",
                        colPairName = "neighborhood",
                        BPPARAM = SerialParam(RNGseed = 221029))
```

These interactions can be shown in a heatmap.

```{r interaction-heatmap, fig.width=5, fig.height=5}

data <- out %>% as_tibble() %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

data_wide <- data %>%
  pivot_wider(
    names_from = to_label,
    values_from = sum_sigval
  )

data_wide <- tibble::column_to_rownames(data_wide, var = "from_label")

# Fixed symmetric color range (+/-50), so breaks are centred on 0
max_range <- max(abs(50), abs(50))
breaks <- seq(-max_range, max_range, length.out = 101)

# blue-white-red scale
colors <- colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1)

p_heatmap <- pheatmap(data_wide,
         color = colors,
         breaks = breaks,
         cellwidth = 20,
         cellheight = 20,
         fontsize = 10,
         display_numbers = TRUE,
         number_format = "%.0f")

save_plot(p_heatmap, "Fig4b_cell-cell_interactions.pdf", width = 6, height = 6)
```

We repeat the same analysis, this time at the resolution of all (fine-grained)
celltypes rather than the celltype/metacluster mixture used above.

```{r interaction-analysis-celltype}
out_celltype <- testInteractions(spe,
                        group_by = "sample_id",
                        label = "celltype",
                        colPairName = "neighborhood",
                        BPPARAM = SerialParam(RNGseed = 221029))
```

```{r interaction-heatmap-celltype, fig.width=12, fig.height=12}

data_celltype <- out_celltype %>% as_tibble() %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

data_wide_celltype <- data_celltype %>%
  pivot_wider(
    names_from = to_label,
    values_from = sum_sigval
  )

data_wide_celltype <- tibble::column_to_rownames(data_wide_celltype, var = "from_label")

# Fixed symmetric color range (+/-50), so breaks are centred on 0
max_range_celltype <- max(abs(50), abs(50))
breaks_celltype <- seq(-max_range_celltype, max_range_celltype, length.out = 101)

# blue-white-red scale
colors_celltype <- colorRampPalette(c("blue", "white", "red"))(length(breaks_celltype) - 1)

p_heatmap_celltype <- pheatmap(data_wide_celltype,
         color = colors_celltype,
         breaks = breaks_celltype,
         cellwidth = 12,
         cellheight = 12,
         fontsize = 7,
         display_numbers = TRUE,
         number_format = "%.0f")

save_plot(p_heatmap_celltype, "Fig4b_cell-cell_interactions_all_celltypes.pdf", width = 14, height = 14)
```

## Cellular neighborhoods

In a next step, we detect cellular neighborhoods (CNs). For each cell, we compute the fraction of each cell type (celltype_2) among its neighbors, and then aggregate cells into a predefined number of neighborhoods (here 6) by k-means.

```{r c-neighborhoods cn-analysis, fig.width=20, fig.height=50}
spe <- aggregateNeighbors(spe, 
                          colPairName = "neighborhood", 
                          aggregate_by = "metadata", 
                          count_by = "celltype_2")

set.seed(220705)

n_neighborhoods <- 6

cn_1 <- kmeans(spe$aggregatedNeighbors, centers = n_neighborhoods)
spe$cn_celltypes_2 <- as.factor(cn_1$cluster)

# Centralized neighborhood palette (COL_CN, from 00_MapMet_color_code.R) so the
# same CN -> color mapping is reused everywhere below (and in 08_generate_images).
metadata(spe)$color_vectors$col_cn <- COL_CN[levels(spe$cn_celltypes_2)]

p_spatial <- plotSpatial(spe,
            node_color_by = "cn_celltypes_2",
            img_id = "sample_id",
            ncols=6,
            node_size_fix = 0.5) +
    scale_color_manual(values = metadata(spe)$color_vectors$col_cn)

save_plot(p_spatial, "Fig4f_spatial_neighborhood_images.pdf", width = 20, height = 49)

p_spatial
```

Lastly, we can visualize the enrichment of cell types within cellular neighborhoods
using the `regionMap` function of the `lisaClust` package.

```{r regionmap, fig.height=6, fig.width=5}
library(lisaClust)
p_map <- regionMap(spe,
          cellType = "celltype_2",
          region = "cn_celltypes_2")

p_map <- p_map + ggplot2::coord_flip() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_map, "Fig4c_spatial_neighborhood_map.pdf", width = 8, height = 4)

p_map
```
         
### Differential abundance of CNs

Finally, we will generate a barplot showing the proportions of CNs per sample. To do so, we first transform the SPE object into a CATALYST-compatible SingleCellExperiment.

```{r transform-data, fig.height=6, fig.width=10}
sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts"), exprs=assay(spe, "counts")))

#recode metadata on progression
sce$progress <- factor(spe$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$tumor_community <- factor(spe$tumor_community)

sce$patient <- factor(spe$patient_id)
sce$fm_id <- spe$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe$cn_celltypes_2)
sce$metacluster <- factor(spe$metacluster)

sce$condition <- 'na' 

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe$cn_celltypes_2))
```

Next, we will generate a barplot showing the proportion of cells assigned to each of the 6 CNs per sample.

```{r cohort-overview}
set.seed(20231103)

distance <- "manhattan"
linkage <- "complete"

p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_cn,
                    linkage = linkage, distance= distance)

save_plot(p, "Fig4d1_spatial_neighborhood_clustering.pdf", width = 6, height = 4)

p
```

We generate a dendogram for the hierarchical clustering of barplots.

```{r dendrogram, fig.height=8, fig.width=7}
set.seed(20231103)
wide_df <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

wide_df <- data.frame(wide_df)
rownames(wide_df) <- wide_df$sample_id
wide_df <- wide_df %>% select(-sample_id)

dist_matrix <- dist(wide_df, method = distance)
hc <- hclust(dist_matrix, method = linkage)

p2 <- plot(hc)

save_plot(p2, "Fig4d2_spatial_neighborhood_clustering_dendogram.pdf", width = 6, height = 6)
p2
```

In addition, we plot the metadata of these samples in the respective order.

```{r plot-metadata, fig.height=3}
clustered_samples <- levels(p$data$sample_id)
sce$sample_id <- factor(sce$sample_id, levels=clustered_samples)

#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
            tumor_community=first(tumor_community),
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


# Sample-annotation colors (COL_CODE sourced from 00_MapMet_color_code.R in the
# read-data chunk above)
col_code <- COL_CODE

# Explicit patient-group palette (col_code's 1/2/3 = blue/red/green is used
# elsewhere); shared with the combined Fig4d strip below.
community_cols <- c("1" = "#E6C700", "2" = "#C71585", "3" = "#4DAF4A")


# Plot metadata
p_anno<- columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(patient_group=as.character(sample_summary$tumor_community), col=list(patient_group=community_cols))

save_plot(p_anno, "Fig4d3_spatial_neighborhood_clustering_metadata.pdf", width = 8, height = 4)

p_anno
```

Finally, we assemble the dendrogram, the CN-abundance barplot and the three metadata strips into a single stacked figure (Fig4d).

```{r fig4d-combined-stacked, fig.height=8.4, fig.width=8}

ddata <- dendro_data(hc, type = "rectangle")
segs <- segment(ddata)

p_dendro <- ggplot(segs) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  scale_x_continuous(expand = expansion(add = 0.5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    plot.margin = margin(2, 2, 0, 2)
  ) +
  labs(title = "Sample clustering based on cellular neighborhood composition")

meta_df <- sample_summary[clustered_samples, c("mna", "progress", "tumor_community")]
meta_df$sample_id <- factor(rownames(meta_df), levels = clustered_samples)

# One annotation strip per track, each with its own fill scale so the three
# legends render separately. patient_group uses the explicit community palette.
make_meta_strip <- function(df, ann_col, legend_title, cols) {
  ggplot(df, aes(x = sample_id, y = 1, fill = as.character(.data[[ann_col]]))) +
    # explicit height (y is constant)
    geom_tile(height = 1) +
    scale_fill_manual(values = cols, na.value = "white", name = legend_title) +
    scale_x_discrete(expand = expansion(add = 0.5)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(y = legend_title) +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_text(angle = 0, hjust = 1, vjust = 0.5, size = 9),
      legend.position = "right",
      legend.key.size = unit(0.4, "cm"),
      panel.grid = element_blank(),
      plot.margin = margin(0, 2, 0, 2)
    )
}

p_meta_mna   <- make_meta_strip(meta_df, "mna",                   "MYCN_amp",      col_code)
p_meta_prog  <- make_meta_strip(meta_df, "progress",              "progression",   col_code)
p_meta_group <- make_meta_strip(meta_df, "tumor_community", "patient_group", community_cols)
# extra bottom margin so the last strip is not clipped
p_meta_group <- p_meta_group + theme(plot.margin = margin(0, 2, 2, 2))

# Abundance barplot: rename the CN legend and drop the placeholder "na" facet
p_abund <- p +
  scale_fill_manual(values = metadata(spe)$color_vectors$col_cn,
                    name = "Cellular neighborhood") +
  facet_null() +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    plot.title = element_blank(),
    strip.text = element_blank(),
    strip.background = element_blank(),
    plot.margin = margin(0, 2, 0, 2)
  )

# Relative panel heights: dendrogram / abundance / three metadata strips
panel_heights <- c(1.2, 4, 0.25, 0.25, 0.25)
p_fig4d <- (p_dendro / p_abund / p_meta_mna / p_meta_prog / p_meta_group) +
  plot_layout(ncol = 1, heights = panel_heights, guides = "collect")

p_fig4d

save_plot(p_fig4d, "Fig4d_combined_stacked.pdf", width = 8, height = 8.4)
```

We then compare the abundance of cellular neighborhoods (CNs) between nMNA and MNA samples.
Therefore, we first prepare data for DA analysis using the `diffcyt` package.

```{r transform-data2, fig.height=6, fig.width=10}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$MYCN_amp==0 | spe$MYCN_amp==1
sce <- SingleCellExperiment(assays=list(counts=assay(spe[keep_rows,keep_cols], "counts"), exprs=assay(spe[keep_rows,keep_cols], "counts")))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows,keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows,keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows,keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows,keep_cols]$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe[keep_rows,keep_cols]$cn_celltypes_2)
sce$metacluster <- factor(spe[keep_rows,keep_cols]$metacluster)

sce$condition <- sce$mna 

# Add experiment_info
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows,keep_cols]$cn_celltypes_2))

# Define cell type and state markers
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")
```

```{r prepare-data}
#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe[keep_rows,]) %in% type_markers, "type",
                                    ifelse(rownames(spe[keep_rows,]) %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

d_se <- transformData(d_se)
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing CN proportions between the two genetic subgroups (nMNA vs MNA).

```{r da-analysis}
# Create design matrix
design <- createDesignMatrix(experiment_info, cols_design = "condition")

# Create contrast matrix
contrast <- createContrast(c(0, 1)) 
nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_voom(d_counts,design, contrast
                       )

FDR_cutoff = 0.1
rowData(res_DA)[which(rowData(res_DA)$p_adj<FDR_cutoff),]
res<- data.frame(rowData(res_DA))
```

Finally, we visualize DA results in boxplots and barplots showing the logFC.

```{r visualize-da2, fig.width=12, fig.height=3}
p <- plotAbundances(sce, k = "celltype", by = "sample") 
comparisons <- c("nMNAvsMNA")

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("nMNA", "MNA"),
              fill = "condition", palette = COL_CODE[c("nMNA", "MNA")], color = "black",
              add.params = list(size = 1.5, color = "black"),
              add = "jitter",
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

p_grid <- grid.arrange(grobs = plot_list_fc, ncol = 6)

save_plot(p_grid, "Fig4e2_spatial_neighborhood_DA_mycn_boxplot.pdf", width = 12, height = 3)

p_grid


```

and in dodge barplots.

```{r barplot-logFC2}
df <- res[, c("logFC", "p_adj")]
df$p_adj_signif <- stars.pval(df$p_adj)
df$p_adj_signif[is.na(df$p_adj_signif) | df$p_adj_signif==" "] <- "ns"
df$cluster <- rownames(df)

p <- ggplot(df, aes(fill=cluster, y=logFC, x=cluster)) + 
  ggtitle("DA - CNs") +
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = F) + 
  scale_fill_manual(values = metadata(spe)$color_vectors$col_cn) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_text(aes(label = p_adj_signif), vjust = 0)

save_plot(p, "Fig4e1_spatial_neighborhood_DA_mycn_barplot.pdf", width = 8, height = 4)

p
```

## Patch detection

Subsequently, we detect tumor patches. We first detect connected sets of tumor cells with a predefined min. number of cells making up a patch and then we expand the concave hull around the patch by 1 pixel.

```{r patchDetection-1, fig.height=20, fig.width=20}
spe <- patchDetection(spe, 
                      patch_cells = spe$metacluster == "tumor",
                      img_id = "sample_id",
                      min_patch_size=20,
                      expand_by=1,
                      colPairName = "neighborhood")
```

### Distance to tumor-stroma border

Based on detected patches, the distance to the tumor stroma border can be computed. Therefore the [minDistToCells](https://bodenmillergroup.github.io/imcRtools/reference/minDistToCells.html) can be used. Negative values indicate cells within patches and positive values cells outside of patches.

```{r minDistCells}
spe <- minDistToCells(spe, 
                      x_cells = !is.na(spe$patch_id), 
                      img_id = "sample_id")

plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "distToCells", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    scale_color_gradient2(low = "dark blue", mid = "white", high = "dark red")
```

The distance distribution per celltype can be shown in ridge plots, with no
legend (the y-axis labels already name every group) so the ridges themselves
get the full plot width. We show this for both `celltype_2` (tumor celltypes +
non-tumor metaclusters) and the fine-grained `celltype`, and for each we mark
the core/border/stroma compartments (defined below using a 200 px threshold on
this same distance) in two ways: as translucent shaded background regions, or
as a green/red/blue compartment bar beneath a plain-background ridge plot. The
+/-200 px cutoff is also marked as dashed lines in every version.

```{r ridges-distance, fig.height=5,fig.width=8}
# Border threshold shown here matches `canonical_border` set in the
# spatial-compartments section below (kept as a literal since that section
# runs after this chunk).
border_threshold <- 200
x_range <- range(spe$distToCells, na.rm = TRUE)

# Compartment bar (green/red/blue core/border/stroma), shared by every "bar"
# distance-ridge figure below - only depends on border_threshold/x_range, not
# on the grouping variable.
compartment_bar <- ggplot() +
  annotate("rect", xmin = -Inf, xmax = -border_threshold, ymin = 0, ymax = 1,
           fill = COL_COMPARTMENT[["core"]]) +
  annotate("rect", xmin = -border_threshold, xmax = border_threshold, ymin = 0, ymax = 1,
           fill = COL_COMPARTMENT[["border"]]) +
  annotate("rect", xmin = border_threshold, xmax = Inf, ymin = 0, ymax = 1,
           fill = COL_COMPARTMENT[["stroma"]]) +
  geom_vline(xintercept = c(-border_threshold, border_threshold),
             color = "black", linewidth = 0.5, linetype = "dashed") +
  coord_cartesian(xlim = x_range, ylim = c(0, 1), expand = FALSE) +
  labs(x = "distToCells") +
  theme_minimal() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), panel.grid = element_blank(),
        panel.border = element_blank())

# Ridge plot (no legend - group names are the y-axis labels) for one grouping
# variable, in one of two compartment-display variants:
#  - "bar":    plain background, stacked on the shared compartment_bar below
#  - "shaded": core/border/stroma shown as translucent background regions
# Reused below for celltype_2, celltype, and cellular neighborhoods.
plot_dist_ridge <- function(group_var, y_label, fill_values, filename,
                            variant = c("bar", "shaded")) {
  variant <- match.arg(variant)

  dist_data <- as.data.frame(colData(spe)) %>%
    mutate(!!group_var := fct_reorder(.data[[group_var]], distToCells, .fun = median, .desc = TRUE))

  shaded_bg <- if (variant == "shaded") list(
    annotate("rect", xmin = -Inf, xmax = -border_threshold, ymin = -Inf, ymax = Inf,
             fill = COL_COMPARTMENT[["core"]], alpha = 0.15),
    annotate("rect", xmin = -border_threshold, xmax = border_threshold, ymin = -Inf, ymax = Inf,
             fill = COL_COMPARTMENT[["border"]], alpha = 0.15),
    annotate("rect", xmin = border_threshold, xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = COL_COMPARTMENT[["stroma"]], alpha = 0.15)
  ) else NULL

  p_ridge <- ggplot(dist_data) +
    shaded_bg +
    stat_density_ridges(aes(distToCells, .data[[group_var]], fill = .data[[group_var]]),
                        quantiles = c(0.5), quantile_lines = TRUE, scale = 3,
                        key_glyph = "rect") +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 0.5) +
    geom_vline(xintercept = c(-border_threshold, border_threshold),
               color = "black", linewidth = 0.5, linetype = "dashed") +
    scale_fill_manual(values = fill_values) +
    labs(y = y_label) +
    coord_cartesian(xlim = x_range, expand = FALSE) +
    theme_ridges() +
    theme(legend.position = "none")

  if (variant == "bar") {
    p_ridge <- p_ridge +
      theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
            axis.ticks.x = element_blank())
    p <- p_ridge / compartment_bar + plot_layout(heights = c(20, 1))
  } else {
    p <- p_ridge
  }

  save_plot(p, filename, width = 8, height = 5)
  p
}

# 1-2. celltype_2 (tumor celltypes + non-tumor metaclusters), bar and shaded
p_dist_celltype2_bar <- plot_dist_ridge(
  "celltype_2", "Celltype", metadata(spe)$color_vectors$col_celltype_2,
  "Fig4j_border_distance_celltype2_bar.pdf", variant = "bar")
p_dist_celltype2_bar

p_dist_celltype2_shaded <- plot_dist_ridge(
  "celltype_2", "Celltype", metadata(spe)$color_vectors$col_celltype_2,
  "Fig4j_border_distance_celltype2_shaded.pdf", variant = "shaded")
p_dist_celltype2_shaded

# 3-4. fine-grained celltype (all celltypes), bar and shaded
p_dist_celltype_bar <- plot_dist_ridge(
  "celltype", "Celltype (fine-grained)", metadata(spe)$color_vectors$col_celltype,
  "Fig4j_border_distance_celltype_bar.pdf", variant = "bar")
p_dist_celltype_bar

p_dist_celltype_shaded <- plot_dist_ridge(
  "celltype", "Celltype (fine-grained)", metadata(spe)$color_vectors$col_celltype,
  "Fig4j_border_distance_celltype_shaded.pdf", variant = "shaded")
p_dist_celltype_shaded
```

The same distance distribution, now grouped by cellular neighborhood (CN,
`cn_celltypes_2`, computed further below) instead of celltype.

```{r ridges-distance-neighborhood, fig.height=5, fig.width=8}
p_dist_cn <- plot_dist_ridge("cn_celltypes_2", "Cellular neighborhood",
                             metadata(spe)$color_vectors$col_cn,
                             "Fig4j_border_distance_neighborhood.pdf",
                             variant = "bar")

p_dist_cn
```

Based on the calculated distance for each cell, three compartments - tumor core, tumor/stroma border and stroma - can be defined using a distance threshold. We sweep border sizes from 100 to 300 pixels, visualizing the compartments and counting immune and tumor cells in each, and then adopt a border of 200 pixels for all downstream analyses. Image `20221018_20220906_03-018_TU_001` lacks tumor patches, therefore all cells are defined as `stroma` for this image.

```{r spatial-compartments}
thresholds <- seq(100, 300, by = 50)
panel_sample_id_1 <- "20220926_20220809_16-006_TU_003"
# sanitize sample IDs for file-safe output tags
sanitize_sample_tag <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
panel_sample_tag_1 <- sanitize_sample_tag(panel_sample_id_1)
y_axis_upper_expand <- 0.1

plot_list_img1 <- list()
results_list <- list()

for (th in thresholds) {

  # reset compartment
  spe$compartment <- "stroma"

  # border
  spe[
    ,
    abs(spe$distToCells) <= th &
      spe$sample_id != "20221018_20220906_03-018_TU_001"
  ]$compartment <- "border"

  # core
  spe[
    ,
    spe$compartment != "border" &
      spe$distToCells <= th &
      spe$sample_id != "20221018_20220906_03-018_TU_001"
  ]$compartment <- "core"

  # per-threshold compartment map for the panel sample
  p_s2 <- plotSpatial(
    spe[, spe$sample_id == panel_sample_id_1],
    node_color_by = "compartment",
    img_id = "sample_id",
    node_size_fix = 0.5
  ) +
    scale_color_manual(values = COL_COMPARTMENT) +
    ggtitle(paste("Border size =", th))

  save_plot(
    p_s2,
    paste0("Fig4g3_tumor_patch_", th, ".pdf"),
    width = 6,
    height = 6
  )

  plot_list_img1[[as.character(th)]] <- p_s2

  # immune counts per compartment
  immune_df <- as.data.frame(
    table(spe$compartment[spe$celltype_3 == "immune"])
  )

  colnames(immune_df) <- c("compartment", "count")

  immune_df$threshold <- th
  immune_df$cell_class <- "immune"

  # tumor counts per compartment
  tumor_df <- as.data.frame(
    table(spe$compartment[spe$metacluster == "tumor"])
  )

  colnames(tumor_df) <- c("compartment", "count")

  tumor_df$threshold <- th
  tumor_df$cell_class <- "tumor"

  # store
  results_list[[paste0("immune_", th)]] <- immune_df
  results_list[[paste0("tumor_", th)]] <- tumor_df
}

# Canonical compartment definition (border = 200) used by all downstream
# analyses; the loop above leaves spe$compartment at its last threshold (300).
canonical_border <- 200
spe$compartment <- "stroma"
spe[
  ,
  abs(spe$distToCells) <= canonical_border &
    spe$sample_id != "20221018_20220906_03-018_TU_001"
]$compartment <- "border"
spe[
  ,
  spe$compartment != "border" &
    spe$distToCells <= canonical_border &
    spe$sample_id != "20221018_20220906_03-018_TU_001"
]$compartment <- "core"

# combine all
results_df <- bind_rows(results_list)

panel_img1 <- wrap_plots(plot_list_img1, ncol = length(thresholds)) +
  plot_annotation(title = panel_sample_id_1)

save_plot(
  panel_img1,
  paste0("Fig4g3_tumor_patch_panel_all_thresholds_", panel_sample_tag_1, ".pdf"),
  width = 30,
  height = 6
)

panel_img1

# order compartments
results_df$compartment <- factor(
  results_df$compartment,
  levels = c("core", "border", "stroma")
)

# immune counts by compartment, per border size
immune_plot <- ggplot(
  results_df %>% filter(cell_class == "immune"),
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  labs(
    title = "Immune cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    immune_plot,
    paste0("Fig4g3_tumor_bordersize_immune_barplot.pdf"),
    width = 6,
    height = 6
  )

immune_plot

# tumor counts by compartment, per border size
tumor_plot <- ggplot(
  results_df %>% filter(cell_class == "tumor"),
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  labs(
    title = "Tumor cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    tumor_plot,
    paste0("Fig4g3_tumor_bordersize_tumor_barplot.pdf"),
    width = 6,
    height = 6
  )

tumor_plot

# combined immune + tumor barplot (one facet each)
results_df$cell_class <- factor(results_df$cell_class, levels = c("immune", "tumor"))

immune_tumor_plot <- ggplot(
  results_df,
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  facet_wrap(~cell_class, ncol = 1, scales = "free_y") +
  labs(
    title = "Immune and tumor cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    immune_tumor_plot,
    paste0("Fig4g3_tumor_bordersize_immune_tumor_barplot.pdf"),
    width = 6,
    height = 10
  )

immune_tumor_plot
```

Analogously to the Fig4f cellular-neighborhood spatial map, we can also show all
primary-tumour images at once, this time coloring each cell by its tumor
compartment (core / border / stroma) rather than by its cellular neighborhood.

```{r spatial-compartment-images, fig.width=20, fig.height=50}
# All PT images colored by tumor compartment (spe$compartment at border 200)
p_spatial_compartment <- plotSpatial(spe,
            node_color_by = "compartment",
            img_id = "sample_id",
            ncols = 6,
            node_size_fix = 0.5) +
    scale_color_manual(values = COL_COMPARTMENT)

save_plot(p_spatial_compartment, "Fig4f_spatial_compartment_images.pdf", width = 20, height = 49)

p_spatial_compartment
```

Similarly, we show all primary-tumour images at once colored by cell type
(`celltype_2`: tumor celltypes + non-tumor metaclusters).

```{r spatial-celltype-images, fig.width=20, fig.height=50}
# All PT images colored by cell type (celltype_2)
p_spatial_celltype <- plotSpatial(spe,
            node_color_by = "celltype_2",
            img_id = "sample_id",
            ncols = 6,
            node_size_fix = 0.5) +
    scale_color_manual(values = metadata(spe)$color_vectors$col_celltype_2)

save_plot(p_spatial_celltype, "Fig4f_spatial_celltype_images.pdf", width = 20, height = 49)

p_spatial_celltype
```


### DA analysis in spatial compartments

Next, we will plot the differential abundance of cell types with respect to spatial compartment on a sample level. We analyse tumor celltypes, immune metaclusters, and cellular neighborhoods, looping over all three compartment comparison pairs (`core-border`, `core-stroma`, `border-stroma`). Differential abundance is tested with `diffcyt` (`testDA_edgeR`) controlling for sample-level pairing, except for cellular neighborhoods, where the DA test (and its p-value annotation) is skipped and only the compartment boxplots are drawn.

```{r define-markers}

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

```

```{r cluster-order-colors}
# Cluster display orders used throughout the compartment analyses (immune_order
# are metacluster names; tumor_order are current spe$celltype_2 names;
# neighborhood_order are the cn_celltypes_2 cluster numbers, as characters).
immune_order <- c("T cells", "B cells", "myeloid/NK", "granulocytes")

tumor_order <- c("adrenergic marker-low Tu", "GD2neg Tu", "CD44+ Tu", "Ki67+ Tu",
                 "neuroblast-like Tu", "early neuroblast-like Tu")
# keep only the celltypes actually present in the data (preserves order)
tumor_order <- intersect(tumor_order,
                         unique(as.character(spe$celltype_2[spe$metacluster == "tumor"])))

neighborhood_order <- levels(spe$cn_celltypes_2)
```






```{r compartment-boxplots-all-comparisons}
# For each cell class: run diffcyt (testDA_edgeR) for every compartment pair,
# then draw boxplots. Boxes are filled by compartment (COL_COMPARTMENT), with
# black outlines/points; the diffcyt DA test and its log2FC/p-value annotation
# are computed and drawn for every class except "neighborhood".

cells_list       <- c("tumor", "immune", "neighborhood")
comp_levels      <- c("core", "border", "stroma")
comparison_pairs <- list(c("core","border"), c("core","stroma"), c("border","stroma"))

# Freq is a percentage: panels use a fixed 0-100 y-axis with the pairwise
# brackets drawn above it (clip = "off").
y_axis_max        <- 100
bracket_base      <- 106
bracket_gap       <- 14
title_gap         <- 85
bracket_positions <- bracket_base + (seq_along(comparison_pairs) - 1) * bracket_gap
max_plot_columns  <- 5

keep_rows <- rowData(spe)$use_channel

# helper to build marker_info from the global marker vectors
.make_marker_info <- function(spe, keep_rows) {
  channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 2)
  marker_name  <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 1)
  marker_class <- ifelse(rownames(rowData(spe)[keep_rows, ]) %in% type_markers, "type",
                  ifelse(rownames(rowData(spe)[keep_rows, ]) %in% state_markers, "state", "other"))
  data.frame(channel_name, marker_name,
             marker_class = factor(marker_class),
             row.names = marker_name)
}

marker_info_da <- .make_marker_info(spe, keep_rows)

for (cells in cells_list) {

  order     <- switch(cells,
                      immune       = immune_order,
                      tumor        = tumor_order,
                      neighborhood = neighborhood_order)
  keep_meta <- switch(cells,
                      immune       = spe$metacluster %in% immune_order,
                      tumor        = spe$metacluster == "tumor",
                      neighborhood = rep(TRUE, ncol(spe)))
  # celltype_2 for tumor/immune; cn_celltypes_2 for neighborhoods, so every
  # cell (no metacluster subsetting) counts.
  cluster_field <- if (cells == "neighborhood") "cn_celltypes_2" else "celltype_2"
  # Neighborhood boxplots skip the diffcyt DA test entirely (no p-value
  # annotation for cellular neighborhoods).
  compute_pvals <- cells != "neighborhood"

  # 1. diffcyt DA results for every compartment pair (skipped for neighborhoods)
  diffcyt_results <- list()

  if (compute_pvals) for (pair in comparison_pairs) {
    comps     <- pair
    key       <- paste(comps, collapse = "-")
    keep_cols <- keep_meta & spe$compartment %in% comps

    if (sum(keep_cols) < 10) { diffcyt_results[[key]] <- NULL; next }

    sce_da <- SingleCellExperiment(
      assays = list(
        counts = assay(spe, "counts")[keep_rows, keep_cols],
        exprs  = assay(spe, "counts")[keep_rows, keep_cols]
      )
    )
    sce_da$fm_id       <- spe[keep_rows, keep_cols]$fm_id
    sce_da$sample      <- factor(spe[keep_rows, keep_cols]$sample)
    sce_da$compartment <- factor(spe[keep_rows, keep_cols]$compartment, levels = comps)
    sce_da$sample_id   <- factor(paste0(sce_da$fm_id, "_", sce_da$compartment))
    sce_da$condition   <- factor(sce_da$compartment, levels = comps)
    sce_da$cluster_id  <- factor(spe[keep_rows, keep_cols][[cluster_field]], levels = order)

    metadata(sce_da)$cluster_codes   <- data.frame(
      celltype = factor(spe[keep_rows, keep_cols][[cluster_field]], levels = order)
    )
    metadata(sce_da)$experiment_info <- as.data.frame(colData(sce_da)) %>%
      dplyr::group_by(sample_id, condition, sample) %>%
      dplyr::summarise(n_cells = dplyr::n(), .groups = "drop") %>%
      as.data.frame()

    exp_info           <- metadata(sce_da)$experiment_info
    exp_info           <- exp_info[match(unique(sce_da$sample_id), exp_info$sample_id), ]
    exp_info$condition <- factor(exp_info$condition, levels = comps)

    df_list_da <- lapply(unique(sce_da$sample_id), function(x)
      as.data.frame(t(counts(sce_da[, sce_da$sample_id == x]))))
    names(df_list_da) <- unique(sce_da$sample_id)

    d_se_da <- prepareData(df_list_da, exp_info, marker_info_da)
    rowData(d_se_da)$cluster_id     <- sce_da[, rownames(d_se_da)]$cluster_id
    metadata(d_se_da)$cluster_codes <- metadata(sce_da)$cluster_codes
    d_se_da     <- transformData(d_se_da)
    d_counts_da <- calcCounts(d_se_da)

    design_da   <- createDesignMatrix(exp_info, cols_design = c("condition", "sample"))
    contrast_da <- createContrast(c(0, 1, rep(0, length(unique(exp_info$sample)) - 1)))

    res_DA_da <- tryCatch(
      testDA_edgeR(d_counts_da, design_da, contrast_da,
                   min_cells = 3,
                   min_samples = round(ncol(d_counts_da) * 0.2)),
      error = function(e) { message("diffcyt failed for ", key, ": ", e$message); NULL }
    )
    diffcyt_results[[key]] <- if (!is.null(res_DA_da)) data.frame(rowData(res_DA_da)) else NULL
  }

  # 2. Sample-level proportions (% within compartment) for plotting
  comp_df <- data.frame(
    fm_id       = spe$fm_id,
    compartment = spe$compartment,
    cluster_id  = spe[[cluster_field]],
    keep        = keep_meta
  )
  comp_df <- comp_df[comp_df$keep & comp_df$compartment %in% comp_levels, ]
  comp_df$compartment <- factor(comp_df$compartment, levels = comp_levels)

  sample_totals <- comp_df %>%
    dplyr::group_by(fm_id, compartment) %>%
    dplyr::summarise(total = dplyr::n(), .groups = "drop")

  plot_data_list <- list()
  for (cluster_name in order) {
    cluster_counts <- comp_df %>%
      dplyr::filter(cluster_id == cluster_name) %>%
      dplyr::group_by(fm_id, compartment) %>%
      dplyr::summarise(n = dplyr::n(), .groups = "drop")

    plot_data_list[[cluster_name]] <- dplyr::left_join(
      sample_totals, cluster_counts, by = c("fm_id", "compartment")
    ) %>%
      dplyr::mutate(
        n         = dplyr::if_else(is.na(n), 0L, n),
        Freq      = 100 * n / total,
        condition = factor(compartment, levels = comp_levels),
        sample_id = fm_id
      )
  }

  # 3. Collate diffcyt stats for all pairs into one table (skipped for
  # neighborhoods, alongside the diffcyt test itself in step 1)
  all_stats <- NULL
  if (compute_pvals) {
    all_stats_list <- list()
    for (cluster_name in order) {
      for (pair in comparison_pairs) {
        key      <- paste(pair, collapse = "-")
        res_pair <- diffcyt_results[[key]]
        if (!is.null(res_pair) && cluster_name %in% rownames(res_pair)) {
          p_adj <- res_pair[cluster_name, "p_adj"]
          logFC <- res_pair[cluster_name, "logFC"]
        } else {
          p_adj <- NA_real_; logFC <- NA_real_
        }
        p_signif <- if (!is.na(p_adj)) gtools::stars.pval(p_adj) else "NA"
        if (p_signif == " ") p_signif <- "ns"
        all_stats_list <- c(all_stats_list, list(data.frame(
          cluster_name = cluster_name,
          group1       = pair[1],
          group2       = pair[2],
          p_adj        = p_adj,
          logFC        = round(logFC, 2),
          label        = paste0(round(logFC, 2), " (",
                                if (!is.na(p_adj)) sprintf("%.3f", p_adj) else "NA",
                                " ", p_signif, ")")
        )))
      }
    }
    all_stats <- dplyr::bind_rows(all_stats_list)
  }

  # 4. Boxplots, filled by compartment (COL_COMPARTMENT) with black
  # outlines/points; diffcyt annotation added only when computed above.
  plot_list_comp <- list()
  for (cluster_name in order) {
    proportions <- plot_data_list[[cluster_name]]

    p_box <- ggboxplot(
      proportions, x = "condition", y = "Freq", order = comp_levels,
      fill = "condition", palette = COL_COMPARTMENT[comp_levels], color = "black",
      add = "jitter", add.params = list(color = "black")
    ) +
      # wrap long celltype names so they are not cropped in narrow panels
      ggtitle(str_wrap(cluster_name, width = 18)) +
      scale_y_continuous(breaks = seq(0, y_axis_max, 25),
                         expand = expansion(mult = c(0.02, 0))) +
      # lock axis to 0-100; brackets (when present) draw outside it
      coord_cartesian(ylim = c(0, y_axis_max), clip = "off") +
      theme(aspect.ratio = 1.0, legend.position = "none",
            plot.title  = element_text(hjust = 0.5, size = 14,
                                       margin = margin(b = title_gap)),
            plot.margin = margin(t = 10, r = 8, b = 8, l = 8))

    if (compute_pvals) {
      stats_df <- all_stats[all_stats$cluster_name == cluster_name, ]
      # fixed bracket heights, identical for every panel
      stats_df$y.position <- bracket_positions[seq_len(nrow(stats_df))]
      p_box <- p_box + stat_pvalue_manual(stats_df, y.position = "y.position", label = "label")
    }

    plot_list_comp[[cluster_name]] <- p_box
  }

  p_comp <- gridExtra::grid.arrange(
    grobs = plot_list_comp,
    ncol  = min(length(plot_list_comp), max_plot_columns)
  )

  save_plot(p_comp,
            paste0("Fig4m_boxplot_compartments_", cells, ".pdf"),
            width = 12, height = 8)
}
```

Finally, we summarize the cell-type composition of each compartment as stacked bars, drawn for every border size and shown for immune metaclusters, tumor celltypes, the two pooled together, and cellular neighborhoods.

```{r composition-stacked-bars}
# spe$compartment (the canonical-border assignment) is left untouched;
# each border's compartment is computed on the fly.

comp_levels <- c("core", "border", "stroma")

# Compartment (core / border / stroma) for a given border size, as a plain
# vector (same rule as the threshold loop; does not modify spe).
compartment_at <- function(spe, border) {
  comp <- rep("stroma", ncol(spe))
  ok   <- spe$sample_id != "20221018_20220906_03-018_TU_001"
  comp[which(abs(spe$distToCells) <= border & ok)] <- "border"
  comp[which(comp != "border" & spe$distToCells <= border & ok)] <- "core"
  comp
}

# One stacked-bar figure for a compartment vector and a cell class: "immune"
# (metaclusters), "tumor" (celltypes), "tumor_immune" (the two pooled), or
# "neighborhood" (cellular neighborhoods, all cells).
make_compartment_stackedbar <- function(comp_vec, cells, border) {
  if (cells == "immune") {
    keep <- spe$metacluster %in% immune_order;                       order <- immune_order
  } else if (cells == "tumor") {
    keep <- spe$metacluster == "tumor";                              order <- tumor_order
  } else if (cells == "neighborhood") {
    keep <- rep(TRUE, ncol(spe));                                    order <- neighborhood_order
  } else {                                                           # tumor_immune (pooled)
    keep <- spe$metacluster == "tumor" | spe$metacluster %in% immune_order
    order <- c(tumor_order, immune_order)
  }

  cluster_field <- if (cells == "neighborhood") "cn_celltypes_2" else "celltype_2"
  color_vec_all <- if (cells == "neighborhood") metadata(spe)$color_vectors$col_cn else
                                                 metadata(spe)$color_vectors$col_celltype_2

  df_bar <- data.frame(compartment = comp_vec, cluster_id = spe[[cluster_field]], keep = keep)
  df_bar <- df_bar[df_bar$keep & df_bar$compartment %in% comp_levels, ]

  totals <- df_bar %>% dplyr::group_by(compartment) %>%
    dplyr::summarise(N = dplyr::n(), .groups = "drop")

  df_prop <- df_bar %>%
    dplyr::group_by(compartment, cluster_id) %>% dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
    dplyr::group_by(compartment) %>% dplyr::mutate(pct = 100 * n / sum(n)) %>% dplyr::ungroup() %>%
    dplyr::mutate(compartment = factor(compartment, levels = comp_levels),
                  cluster_id  = factor(cluster_id,  levels = order))

  color_vec <- color_vec_all[order]
  lab       <- if (cells == "tumor_immune") "tumor + immune" else cells

  p_stack <- ggplot(df_prop, aes(x = compartment, y = pct, fill = cluster_id)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(
      data = totals %>% dplyr::mutate(compartment = factor(compartment, levels = comp_levels)),
      aes(x = compartment, y = 103, label = paste0("n=", N)),
      inherit.aes = FALSE, vjust = 0, size = 3.5) +
    scale_fill_manual(values = color_vec, drop = FALSE, name = "Cell type") +
    scale_y_continuous(limits = c(0, 115), breaks = seq(0, 100, 25),
                       expand = expansion(mult = c(0, 0))) +
    labs(title = paste0("Compartment composition (", lab, ") - border ", border),
         x = "Compartment", y = "% cells") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.key.size = unit(0.4, "cm"))

  save_plot(p_stack,
            paste0("Fig4_suppl_stackedbar_compartments_", cells, "_bordersize_", border, ".pdf"),
            width = if (cells == "tumor_immune") 7 else 6, height = 5)
  print(p_stack)
}

for (border in thresholds) {
  comp_vec <- compartment_at(spe, border)
  for (cells in c("immune", "tumor", "tumor_immune", "neighborhood"))
    make_compartment_stackedbar(comp_vec, cells, border)
}
```



## Save R object

Finally, we save the R object.

```{r save-rds}
saveRDS(spe, file.path(params$output,"spe_07_spatial_analysis.rds"))
```


## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
