---
title: "Spatial analysis in primary tumor samples"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input: !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/07_spatial_analysis")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  # Route figures into subfolders (there are many related plots) by filename:
  # neighborhood / compartments / patch, else the plot_dir root. Order matters -
  # the compartment check runs before the patch check so the Fig4g3 compartment
  # panels (named "..._patch_...") land under compartments/, not patch/.
  subdir <- if (grepl("_spatial_neighborhood|Fig4d_combined_stacked|Fig4d1|Fig4d2|Fig4d3", filename)) "neighborhood" else
            if (grepl("compartment|stackedbar|bordersize|Fig4g3|Fig4m", filename))                    "compartments" else
            if (grepl("patch|border_distance|Fig4h_DA|Fig4i_DA|Fig4j", filename))                      "patch"        else ""
  out_dir   <- if (nzchar(subdir)) file.path(plot_dir, subdir) else plot_dir
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  full_path <- file.path(out_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {
    
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
    
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {
    
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
    
  } else {
    
    dev.copy(pdf, full_path, width = width, height = height)
    dev.off()
  }
  
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Spatial analysis in PT {#sp-pt}

Herein, we explore spatial interactions in the primary tumor. 
Due to a [bug](https://github.com/BodenmillerGroup/imcRtools) in the `testInteractions` function in imcRtools versions < 1.5.5, we provide another docker image with a newer imcRtools version for spatial analysis. Please pull *lazdaria/mapmetsc_spatial:v1.0* from [docker hub]() and for the present RMD file.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(imcRtools)
library(scales)
library(BiocParallel)
library(RColorBrewer)
library(circlize)
library(patchwork)
library(ggplot2)
library(lisaClust)
library(tidyr)
library(diffcyt)
library(dplyr)
library(CATALYST)
library(ComplexHeatmap)
library(gtools)
library(ggpubr)
library(gridExtra)
library(ggridges)
library(forcats)
library(pheatmap)
library(tidyverse)
library(ggdendro)
library(lme4)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). We perform spatial analysis only in the primary tumor samples, since the bone marrow samples are aspirates, where the spatial context is lost. Therefore, we first subset the SPE object to the PT samples and then load cell neighbors that were computed in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP). Therein, we have extracted neighbors by expanding individual high-resolution cell masks by 15 pixels. We load the cell neighbors for the primary-tumour samples directly with `read_steinbock`, using its `pattern` argument to select the `TU` samples from the shared `intensities-0px` folder (no file duplication needed).

```{r temp-folder}
# No temporary folder / file duplication needed: the read_steinbock() call below
# selects the primary-tumour (TU) samples directly from the shared
# intensities-0px folder via its `pattern` argument.
```

We can now read in the neighbors.
For the spatial interaction analysis, we will use the metaclusters for non-tumor cells, since these represent the minority in the PT samples, and the celltype level annotations for the tumor cells (majority).

```{r read-data}
# Reads the cellular-community object from 05.1 and uses its
# `allcelltype_community` column for the community annotations below. The
# imcRtools::testInteractions fix (which once required a separate spatial image)
# is in the imcRtools 1.10.0 / Bioconductor 3.19 that this image pins.
spe <- readRDS(file.path(params$output,"spe_05.1_PT_DE_cellularcomm.rds"))
spe <- spe[,spe$tissue=="PT"]

spe_neighbors <- read_steinbock(path=file.path(params$input), image_file=NULL, panel_file = NULL, intensities_folder = "intensities-0px",regionprops_folder = NULL, graphs_folder="neighbors", pattern = "TU")
colnames(spe_neighbors) <- paste0(spe_neighbors$sample_id, "_", spe_neighbors$ObjectNumber)
colPair(spe, "neighborhood") <- colPair(spe_neighbors[,colnames(spe)], "neighborhood")

#celltype 2
spe$celltype_2 <- spe$metacluster
spe[,spe$metacluster=="tumor"]$celltype_2 <- spe[,spe$metacluster=="tumor"]$celltype

# generate new color vector
tumor_names <- unique(spe[,spe$metacluster=="tumor"]$celltype)
metadata(spe)$color_vectors$col_celltype_2 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_2 <- c(metadata(spe)$color_vectors$col_celltype_2, metadata(spe)$color_vectors$col_celltype[tumor_names])

#celltype 3
spe$celltype_3 <- spe$metacluster
spe[,spe$metacluster=="T cells" | spe$metacluster=="B cells" | spe$metacluster=="granulocytes" | spe$metacluster=="myeloid/NK"]$celltype_3 <- "immune"

# generate new color vector
metadata(spe)$color_vectors$col_celltype_3 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_3 <- c(metadata(spe)$color_vectors$col_celltype_3, "immune" = "red")
```

## Interaction analysis

Next, we determine the interactions between cells on the celltype_2 level. For each image, the `testInteractions` function computes the averaged cell type interactions and compares this number against an empirical null distribution generated by permuting all cell labels, while maintaining tissue structure. 

```{r interaction-analysis}
out <- testInteractions(spe,
                        group_by = "sample_id",
                        label = "metacluster",
                        colPairName = "neighborhood",
                        BPPARAM = SerialParam(RNGseed = 221029))
```

These interactions can be shown in a heatmap.

```{r interaction-heatmap, fig.width=5, fig.height=5}

data <- out %>% as_tibble() %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

# Reshaping the data to a wide format suitable for heatmap
data_wide <- data %>%
  pivot_wider(
    names_from = to_label,
    values_from = sum_sigval
  )

data_wide <- tibble::column_to_rownames(data_wide, var = "from_label")

# Define maximum range value for symmetric color scale
#max_range <- max(abs(max(data_wide)), abs(min(data_wide)))  
max_range <- max(abs(50), abs(50))

# Define breaks, making sure 0 is exactly in the middle
breaks <- seq(-max_range, max_range, length.out = 101)  # From -50 to 50

# Create a color palette from blue (negative) to white (zero) to red (positive)
colors <- colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1)

p_heatmap <- pheatmap(data_wide,
         color = colors,
         breaks = breaks,
         cellwidth = 20,
         cellheight = 20,
         fontsize = 10,
         display_numbers = TRUE,
         number_format = "%.0f")  # display the values in cells as round numbers

save_plot(p_heatmap, "Fig4b_cell-cell_interactions.pdf", width = 6, height = 6)
```

## Cellular neighborhoods

In a next step, we detect cellular neighborhoods (CNs). Cells are grouped into neighborhoods based on celltypes in their proximity. For each cell, the fraction of cells of a celltype among its neighbors, is detected. Based on these metrics, the cells can then be aggregated into neighborhoods. The number of neighborhoods has to be predefined. 

```{r c-neighborhoods cn-analysis, fig.width=20, fig.height=50}
spe <- aggregateNeighbors(spe, 
                          colPairName = "neighborhood", 
                          aggregate_by = "metadata", 
                          count_by = "celltype_2")

set.seed(220705)

n_neighborhoods <- 6

cn_1 <- kmeans(spe$aggregatedNeighbors, centers = n_neighborhoods)
spe$cn_celltypes_2 <- as.factor(cn_1$cluster)

p_spatial <- plotSpatial(spe, 
            node_color_by = "cn_celltypes_2", 
            img_id = "sample_id", 
            ncols=6,
            node_size_fix = 0.5) +
    scale_color_brewer(palette = "Set3")

save_plot(p_spatial, "Fig4f_spatial_neighborhood_images.pdf", width = 20, height = 49)

p_spatial
```

Lastly, we can visualize the enrichment of cell types within cellular neighborhoods
using the `regionMap` function of the `lisaClust` package.

```{r regionmap, fig.height=6, fig.width=5}
library(lisaClust)
p_map <- regionMap(spe,
          cellType = "celltype_2",
          region = "cn_celltypes_2")

p_map <- p_map + ggplot2::coord_flip() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_map, "Fig4c_spatial_neighborhood_map.pdf", width = 8, height = 4)

p_map
```
         
### Differential abundance of CNs

Finally, we will generate a barplot showing the proportions of CNs per sample. To be able to use the diffcyt package, we will first transform the SPE object into a catalyst-compatible object.

```{r transform-data, fig.height=6, fig.width=10}
sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts"), exprs=assay(spe, "counts")))

#recode metadata on progression
sce$progress <- factor(spe$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$allcelltype_community <- factor(spe$allcelltype_community)

sce$patient <- factor(spe$patient_id)
sce$fm_id <- spe$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe$cn_celltypes_2)
sce$metacluster <- factor(spe$metacluster)

sce$condition <- 'na' 

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe$cn_celltypes_2))
```

Next, we will generate a barplot showing the proportion of cells assigned to each of the 6 CNs per sample.

```{r cohort-overview}
set.seed(20231103)

distance <- "manhattan"
linkage <- "complete"

p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=brewer.pal(n_neighborhoods, "Set3"), 
                    linkage = linkage, distance= distance) 

save_plot(p, "Fig4d1_spatial_neighborhood_clustering.pdf", width = 6, height = 4)

p
```

We generate a dendogram for the hierarchical clustering of barplots.

```{r dendrogram, fig.height=8, fig.width=7}
#Dendrogram
set.seed(20231103)
wide_df <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

wide_df <- data.frame(wide_df)
rownames(wide_df) <- wide_df$sample_id
wide_df <- wide_df %>% select(-sample_id)

dist_matrix <- dist(wide_df, method = distance)
hc <- hclust(dist_matrix, method = linkage)

p2 <- plot(hc)

save_plot(p2, "Fig4d2_spatial_neighborhood_clustering_dendogram.pdf", width = 6, height = 6)
p2
```

In addition, we plot the metadata of these samples in the respective order.

```{r plot-metadata, fig.height=3}
clustered_samples <- levels(p$data$sample_id)
sce$sample_id <- factor(sce$sample_id, levels=clustered_samples)

#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
            allcelltype_community=first(allcelltype_community),
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


#assign color codes
# Sample-annotation colors are defined centrally in 00_MapMet_color_code.R
for (mapmet_cc_path in c("analysis/00_MapMet_color_code.R", "00_MapMet_color_code.R", "../analysis/00_MapMet_color_code.R"))
  if (file.exists(mapmet_cc_path)) { source(mapmet_cc_path); break }
col_code <- COL_CODE

# Patient-group (cellular-community) numbers get an explicit yellow/magenta
# palette. col_code maps 1/2/3 to blue/red/green (used elsewhere), which is not
# what we want for this annotation; defining the colors explicitly keeps the
# strip deterministic and lets Fig4d3 and the combined Fig4d strip below share
# one palette (group 1 = yellow, group 2 = magenta).
community_cols <- c("1" = "#E6C700", "2" = "#C71585", "3" = "#4DAF4A")


# Plot metadata
p_anno<- columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(patient_group=as.character(sample_summary$allcelltype_community), col=list(patient_group=community_cols))

save_plot(p_anno, "Fig4d3_spatial_neighborhood_clustering_metadata.pdf", width = 8, height = 4)

p_anno
```

```{r fig4d-combined-stacked, fig.height=8.4, fig.width=8}

ddata <- dendro_data(hc, type = "rectangle")
segs <- segment(ddata)

p_dendro <- ggplot(segs) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  scale_x_continuous(expand = expansion(add = 0.5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    plot.margin = margin(2, 2, 0, 2)
  ) +
  labs(title = "Sample clustering based on cellular neighborhood composition")

meta_df <- sample_summary[clustered_samples, c("mna", "progress", "allcelltype_community")]
meta_df$sample_id <- factor(rownames(meta_df), levels = clustered_samples)

# One horizontal annotation strip per track (MYCN_amp / progression /
# patient_group), each with its own fill scale so the three legends render
# separately, exactly as in Fig4d3. MYCN and progression reuse col_code;
# patient_group uses the explicit community palette (yellow/magenta) so the
# strip matches Fig4d3 rather than the col_code 1/2/3 = blue/red/green fallback.
make_meta_strip <- function(df, ann_col, legend_title, cols) {
  ggplot(df, aes(x = sample_id, y = 1, fill = as.character(.data[[ann_col]]))) +
    # explicit height: y is constant so geom_tile cannot infer it from resolution
    geom_tile(height = 1) +
    scale_fill_manual(values = cols, na.value = "white", name = legend_title) +
    scale_x_discrete(expand = expansion(add = 0.5)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(y = legend_title) +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_text(angle = 0, hjust = 1, vjust = 0.5, size = 9),
      legend.position = "right",
      legend.key.size = unit(0.4, "cm"),
      panel.grid = element_blank(),
      plot.margin = margin(0, 2, 0, 2)
    )
}

p_meta_mna   <- make_meta_strip(meta_df, "mna",                   "MYCN_amp",      col_code)
p_meta_prog  <- make_meta_strip(meta_df, "progress",              "progression",   col_code)
p_meta_group <- make_meta_strip(meta_df, "allcelltype_community", "patient_group", community_cols)
# extra bottom margin on the last strip so its lower edge is not clipped
p_meta_group <- p_meta_group + theme(plot.margin = margin(0, 2, 2, 2))

# Abundance barplot: rename the neighborhood legend to "Cellular neighborhood"
# (keeping the Set3 colors) and drop the "na" facet strip that plotAbundances
# adds from the placeholder condition = "na".
p_abund <- p +
  scale_fill_manual(values = brewer.pal(n_neighborhoods, "Set3"),
                    name = "Cellular neighborhood") +
  facet_null() +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    plot.title = element_blank(),
    strip.text = element_blank(),
    strip.background = element_blank(),
    plot.margin = margin(0, 2, 0, 2)
  )

# Empirically tuned panel proportions:
# - dendrogram = 1.2 keeps branch structure readable without dominating the figure
# - abundance = 4 prioritizes the main abundance signal (largest information density)
# - each metadata strip = 0.25 (halved relative size, so the annotation rows take
#   up less of the figure than the dendrogram/abundance panels)
panel_heights <- c(1.2, 4, 0.25, 0.25, 0.25)
p_fig4d <- (p_dendro / p_abund / p_meta_mna / p_meta_prog / p_meta_group) +
  plot_layout(ncol = 1, heights = panel_heights, guides = "collect")

p_fig4d

save_plot(p_fig4d, "Fig4d_combined_stacked.pdf", width = 8, height = 8.4)
```

We then compare the abundance of cellular communities between nMNA and MNA samples.
Therefore, we first prepare data for DA analysis using the `diffcyt` package. 

```{r transform-data2, fig.height=6, fig.width=10}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$MYCN_amp==0 | spe$MYCN_amp==1
sce <- SingleCellExperiment(assays=list(counts=assay(spe[keep_rows,keep_cols], "counts"), exprs=assay(spe[keep_rows,keep_cols], "counts")))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows,keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows,keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows,keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows,keep_cols]$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe[keep_rows,keep_cols]$cn_celltypes_2)
sce$metacluster <- factor(spe[keep_rows,keep_cols]$metacluster)

sce$condition <- sce$mna 

# Add experiment_info
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows,keep_cols]$cn_celltypes_2))

# Define cell type and state markers
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")
```

```{r prepare-data}
#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe[keep_rows,]) %in% type_markers, "type",
                                    ifelse(rownames(spe[keep_rows,]) %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between to genetic subgroups. 

```{r da-analysis}
# Create design matrix
design <- createDesignMatrix(experiment_info, cols_design = "condition")

# Create contrast matrix
contrast <- createContrast(c(0, 1)) 
nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_voom(d_counts,design, contrast
                       )

FDR_cutoff = 0.1
rowData(res_DA)[which(rowData(res_DA)$p_adj<FDR_cutoff),]
res<- data.frame(rowData(res_DA))
```

Finally, we visualize DA results in boxplots and barplots showing the logFC.

```{r visualize-da2, fig.width=12, fig.height=3}
p <- plotAbundances(sce, k = "celltype", by = "sample") 
comparisons <- c("nMNAvsMNA")

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  color <- brewer.pal(n_neighborhoods, "Set3")[cluster_name]
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("nMNA", "MNA"),
              color = "condition", palette =c(color, color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

p_grid <- grid.arrange(grobs = plot_list_fc, ncol = 6)

save_plot(p_anno, "Fig4e2_spatial_neighborhood_DA_mycn_boxplot.pdf", width = 8, height = 4)

p_grid


```

and in dodge barplots.

```{r barplot-logFC2}
df <- res[, c("logFC", "p_adj")]
df$p_adj_signif <- stars.pval(df$p_adj)
df$p_adj_signif[is.na(df$p_adj_signif) | df$p_adj_signif==" "] <- "ns"
df$cluster <- rownames(df)

p <- ggplot(df, aes(fill=cluster, y=logFC, x=cluster)) + 
  ggtitle("DA - CNs") +
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = F) + 
  scale_fill_manual(values = brewer.pal(6, "Set3")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_text(aes(label = p_adj_signif), vjust = 0)

save_plot(p, "Fig4e1_spatial_neighborhood_DA_mycn_barplot.pdf", width = 8, height = 4)

p
```

## Patch detection

Subsequently, we detect tumor patches. We first detect connected sets of tumor cells with a predefined min. number of cells making up a patch and then we expand the concave hull around the patch by 1 pixel.

```{r patchDetection-1, fig.height=20, fig.width=20}
spe <- patchDetection(spe, 
                      patch_cells = spe$metacluster == "tumor",
                      img_id = "sample_id",
                      min_patch_size=20,
                      expand_by=1,
                      colPairName = "neighborhood")
```

We can now plot the tumor patch in the neighborhood graph of individual samples.

```{r plot-patch}
p_s1<- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "GD2_Gd155_mean", 
            assay_type = "counts",
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "neighborhood", 
            nodes_first = FALSE, 
            node_size_by = "area", 
            directed = FALSE,
            edge_color_fix = "grey") + 
    scale_size_continuous(range = c(0.1, 2)) +
    ggtitle("GD2")


save_plot(p_s1, "Fig4g1_tumor_patch.pdf", width = 6, height = 6)

p_s2 <- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "patch_id", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    theme(legend.position = "none")

save_plot(p_s2, "Fig4g2_tumor_patch.pdf", width = 6, height = 6)

(p_s1 | p_s2)
```


Then we determine the frequency of cells within the patch belonging to a metacluster per sample with respect to progression status of the respective patient.

```{r patch-freq-mycn, fig.width=14, fig.height=7}
variable <- "MYCN_amp"
plot_list <- list()
celltypes <- unique(spe$metacluster)

for(i in 1:length(unique(celltypes))){
  
  c <- celltypes[i]
  
  cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
      group_by(patch_id, sample) %>%
      summarize(cell_count = sum(metacluster == c),
                patch_size = n(),
                cell_freq = cell_count / patch_size) %>%
      filter(!is.na(patch_id)) 
    
  mean_cell_freq_per_sample <- cell_freq_per_patch %>%
    group_by(sample) %>%
    summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
    arrange(desc(mean_cell_freq))
  
  df <- data.frame(
    sample = spe$sample,
    condition = colData(spe)[,variable]
  )
  
  progression_per_sample <- df %>%
    group_by(sample) %>%
    summarise(condition = first(condition))
  
  combined_data <- mean_cell_freq_per_sample %>%
    left_join(progression_per_sample, by = "sample")
  
  # log2FC = log2(second / first) over the two condition levels present, so it
  # works whether they are coded 0/1 (e.g. MYCN) or 1/2 (cellular communities).
  freq_by_cond <- combined_data %>%
    group_by(condition) %>%
    summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE))
  grp <- sort(unique(as.character(freq_by_cond$condition)))
  log2fc_df <- data.frame(log2FC = log2(
    freq_by_cond$mean_freq[match(grp[2], as.character(freq_by_cond$condition))] /
    freq_by_cond$mean_freq[match(grp[1], as.character(freq_by_cond$condition))]))
  
  stats_df <- compare_means(mean_cell_freq ~ condition,  
                            method = "wilcox.test",  
                            data = combined_data) 
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                            " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_metacluster[as.character(c)])
  
  plot_list[[c]] <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                  color = "condition", palette = rep(color, length(unique(combined_data$condition))),
                  add.params = list(size = 3),
                  add = "jitter",
                  title=c) + 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                  theme(aspect.ratio = 1.6, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
  
}

p_box <- grid.arrange(grobs = plot_list, ncol = 6)

save_plot(p_box, paste0("Fig4i_DA_",variable,".pdf"), width = 16, height = 12)

p_box

```

```{r patch-freq-comm, fig.width=14, fig.height=7}
variable <- "allcelltype_community"
plot_list <- list()
celltypes <- unique(spe$metacluster)

for(i in 1:length(unique(celltypes))){
  
  c <- celltypes[i]
  
  cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
      group_by(patch_id, sample) %>%
      summarize(cell_count = sum(metacluster == c),
                patch_size = n(),
                cell_freq = cell_count / patch_size) %>%
      filter(!is.na(patch_id)) 
    
  mean_cell_freq_per_sample <- cell_freq_per_patch %>%
    group_by(sample) %>%
    summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
    arrange(desc(mean_cell_freq))
  
  df <- data.frame(
    sample = spe$sample,
    condition = colData(spe)[,variable]
  )
  
  progression_per_sample <- df %>%
    group_by(sample) %>%
    summarise(condition = first(condition))
  
  combined_data <- mean_cell_freq_per_sample %>%
    left_join(progression_per_sample, by = "sample")
  
  # log2FC = log2(second / first) over the two condition levels present, so it
  # works whether they are coded 0/1 (e.g. MYCN) or 1/2 (cellular communities).
  freq_by_cond <- combined_data %>%
    group_by(condition) %>%
    summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE))
  grp <- sort(unique(as.character(freq_by_cond$condition)))
  log2fc_df <- data.frame(log2FC = log2(
    freq_by_cond$mean_freq[match(grp[2], as.character(freq_by_cond$condition))] /
    freq_by_cond$mean_freq[match(grp[1], as.character(freq_by_cond$condition))]))
  
  stats_df <- compare_means(mean_cell_freq ~ condition,  
                            method = "wilcox.test",  
                            data = combined_data) 
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                            " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_metacluster[as.character(c)])
  
  plot_list[[c]] <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                  color = "condition", palette = rep(color, length(unique(combined_data$condition))),
                  add.params = list(size = 3),
                  add = "jitter",
                  title=c) + 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                  theme(aspect.ratio = 1.6, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
  
}

p_box <- grid.arrange(grobs = plot_list, ncol = 6)

save_plot(p_box, paste0("Fig4h_DA_",variable,".pdf"), width = 16, height = 12)

p_box

```

We repeat the same for immune cells overall. 

```{r patch-freq-immune}
variable <- "progression"
  
c <- "immune"

cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
    group_by(patch_id, sample) %>%
    summarize(cell_count = sum(celltype_3 == c),
              patch_size = n(),
              cell_freq = cell_count / patch_size) %>%
    filter(!is.na(patch_id)) 
  
mean_cell_freq_per_sample <- cell_freq_per_patch %>%
  group_by(sample) %>%
  summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
  arrange(desc(mean_cell_freq))

df <- data.frame(
  sample = spe$sample,
  condition = colData(spe)[,variable]
)

progression_per_sample <- df %>%
  group_by(sample) %>%
  summarise(condition = first(condition))

combined_data <- mean_cell_freq_per_sample %>%
  left_join(progression_per_sample, by = "sample")

# log2FC = log2(second / first) over the two condition levels present.
freq_by_cond <- combined_data %>%
  group_by(condition) %>%
  summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE))
grp <- sort(unique(as.character(freq_by_cond$condition)))
log2fc_df <- data.frame(log2FC = log2(
  freq_by_cond$mean_freq[match(grp[2], as.character(freq_by_cond$condition))] /
  freq_by_cond$mean_freq[match(grp[1], as.character(freq_by_cond$condition))]))

stats_df <- compare_means(mean_cell_freq ~ condition,  
                          method = "wilcox.test",  
                          data = combined_data) 
  
stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
stats_df$p_adj_signif[is.na(stats_df$p_adj_signif) | stats_df$p_adj_signif==" "] <- "ns"
stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                          " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")

color <- unname(metadata(spe)$color_vectors$col_celltype_3[as.character(c)])

p <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                color = "condition", palette =c(color, color),
                add.params = list(size = 3),
                add = "jitter",
                title=c) + 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                theme(aspect.ratio = 1.6, legend.position="none") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
p
```

### Distance to tumor-stroma border

Based on detected patches, the distance to the tumor stroma border can be computed. Therefore the [minDistToCells](https://bodenmillergroup.github.io/imcRtools/reference/minDistToCells.html) can be used. Negative values indicate cells within patches and positive values cells outside of patches.

```{r minDistCells}
spe <- minDistToCells(spe, 
                      x_cells = !is.na(spe$patch_id), 
                      img_id = "sample_id")

plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "distToCells", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    scale_color_gradient2(low = "dark blue", mid = "white", high = "dark red")
```

The distance distribution per celltype can be shown in ridge plots.

```{r ridges-distance, fig.height=10,fig.width=16}
p_dist <- ggplot(as.data.frame(colData(spe)) %>%
         mutate(celltype_2 = fct_reorder(celltype_2, distToCells, .fun = median))) +
  stat_density_ridges(aes(distToCells, celltype_2, fill = celltype_2),
                      quantiles = c(0.5), quantile_lines = TRUE, scale = 3,
                      # plain filled rectangles in the legend keys (no median
                      # quantile line running through the color blocks)
                      key_glyph = "rect") +
  geom_vline(xintercept = 0, color = "dark red", linewidth = 0.5) +
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_2, name = "Celltypes") +
  theme_ridges()

save_plot(p_dist, "Fig4j_border_distance.pdf", width = 12, height = 10)

p_dist
```

Based on the calculated distance for each cell, three compartments - tumor core, tumor/stroma border and stroma - can be defined using a distance threshold. Image `20221018_20220906_03-018_TU_001` lacks tumor patches, therefore all cells are defined as `stroma` for this image.

```{r spatial-compartments}
thresholds <- seq(100, 300, by = 50)
panel_sample_id_1 <- "20220926_20220809_16-006_TU_003"
# sanitize sample IDs for file-safe output tags
sanitize_sample_tag <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
panel_sample_tag_1 <- sanitize_sample_tag(panel_sample_id_1)
y_axis_upper_expand <- 0.1

plot_list_img1 <- list()
results_list <- list()

for (th in thresholds) {

  # reset compartment
  spe$compartment <- "stroma"

  # border
  spe[
    ,
    abs(spe$distToCells) <= th &
      spe$sample_id != "20221018_20220906_03-018_TU_001"
  ]$compartment <- "border"

  # core
  spe[
    ,
    spe$compartment != "border" &
      spe$distToCells <= th &
      spe$sample_id != "20221018_20220906_03-018_TU_001"
  ]$compartment <- "core"

  # -----------------------------
  # PLOTS
  # -----------------------------

  p_s2 <- plotSpatial(
    spe[, spe$sample_id == panel_sample_id_1],
    node_color_by = "compartment",
    img_id = "sample_id",
    node_size_fix = 0.5
  ) +
    ggtitle(paste("Border size =", th))

  save_plot(
    p_s2,
    paste0("Fig4g3_tumor_patch_", th, ".pdf"),
    width = 6,
    height = 6
  )

  plot_list_img1[[as.character(th)]] <- p_s2

  # -----------------------------
  # IMMUNE COUNTS
  # -----------------------------
  
  immune_df <- as.data.frame(
    table(spe$compartment[spe$celltype_3 == "immune"])
  )

  colnames(immune_df) <- c("compartment", "count")

  immune_df$threshold <- th
  immune_df$cell_class <- "immune"

  # -----------------------------
  # TUMOR COUNTS
  # -----------------------------

  tumor_df <- as.data.frame(
    table(spe$compartment[spe$metacluster == "tumor"])
  )

  colnames(tumor_df) <- c("compartment", "count")

  tumor_df$threshold <- th
  tumor_df$cell_class <- "tumor"

  # store
  results_list[[paste0("immune_", th)]] <- immune_df
  results_list[[paste0("tumor_", th)]] <- tumor_df
}

# ---------------------------------------------------------------------------
# Canonical compartment definition for all downstream analyses.
# The loop above leaves spe$compartment set to its last threshold (300). Here we
# fix it to a border size of 200, which is the value used from this point on:
# the Fig4f compartment images, the Fig4m compartment boxplots and the
# Fig4_suppl compartment stacked bars all read this spe$compartment.
canonical_border <- 200
spe$compartment <- "stroma"
spe[
  ,
  abs(spe$distToCells) <= canonical_border &
    spe$sample_id != "20221018_20220906_03-018_TU_001"
]$compartment <- "border"
spe[
  ,
  spe$compartment != "border" &
    spe$distToCells <= canonical_border &
    spe$sample_id != "20221018_20220906_03-018_TU_001"
]$compartment <- "core"

# combine all
results_df <- bind_rows(results_list)

panel_img1 <- wrap_plots(plot_list_img1, ncol = length(thresholds)) +
  plot_annotation(title = panel_sample_id_1)

save_plot(
  panel_img1,
  paste0("Fig4g3_tumor_patch_panel_all_thresholds_", panel_sample_tag_1, ".pdf"),
  width = 30,
  height = 6
)

panel_img1

# order compartments
results_df$compartment <- factor(
  results_df$compartment,
  levels = c("core", "border", "stroma")
)

# -----------------------------
# IMMUNE PLOT
# -----------------------------

immune_plot <- ggplot(
  results_df %>% filter(cell_class == "immune"),
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  labs(
    title = "Immune cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    immune_plot,
    paste0("Fig4g3_tumor_bordersize_immune_barplot.pdf"),
    width = 6,
    height = 6
  )

immune_plot

# -----------------------------
# TUMOR PLOT
# -----------------------------

tumor_plot <- ggplot(
  results_df %>% filter(cell_class == "tumor"),
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  labs(
    title = "Tumor cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    tumor_plot,
    paste0("Fig4g3_tumor_bordersize_tumor_barplot.pdf"),
    width = 6,
    height = 6
  )

tumor_plot

# -----------------------------
# COMBINED IMMUNE + TUMOR PLOT
# -----------------------------
# Same border-size sweep as the two panels above, but immune and tumor cell
# counts shown together in a single figure (one facet each). The individual
# immune-only and tumor-only barplots above are kept unchanged.

results_df$cell_class <- factor(results_df$cell_class, levels = c("immune", "tumor"))

immune_tumor_plot <- ggplot(
  results_df,
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  facet_wrap(~cell_class, ncol = 1, scales = "free_y") +
  labs(
    title = "Immune and tumor cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    immune_tumor_plot,
    paste0("Fig4g3_tumor_bordersize_immune_tumor_barplot.pdf"),
    width = 6,
    height = 10
  )

immune_tumor_plot
```

Analogously to the Fig4f cellular-neighborhood spatial map, we can also show all
primary-tumour images at once, this time coloring each cell by its tumor
compartment (core / border / stroma) rather than by its cellular neighborhood.

```{r spatial-compartment-images, fig.width=20, fig.height=50}
# Equivalent of Fig4f_spatial_neighborhood_images, but node_color_by the tumor
# compartment instead of the cellular neighborhood. Uses spe$compartment at the
# canonical border size of 200 (set above, after the threshold loop) and the
# same default compartment color scheme as the per-image panel (p_s2),
# so core / border / stroma are colored consistently across the compartment
# figures.
p_spatial_compartment <- plotSpatial(spe,
            node_color_by = "compartment",
            img_id = "sample_id",
            ncols = 6,
            node_size_fix = 0.5)

save_plot(p_spatial_compartment, "Fig4f_spatial_compartment_images.pdf", width = 20, height = 49)

p_spatial_compartment
```


### DA analysis in spatial compartments

Next, we will plot the differential abundance of metaclusters with respect to spatial compartment on a sample level. We analyse both tumor celltypes and immune metaclusters, looping over all three compartment comparison pairs (`core-border`, `core-stroma`, `border-stroma`). Differential abundance is tested with `diffcyt` (`testDA_edgeR`) controlling for sample-level pairing.

```{r define-markers}

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

```

```{r cluster-order-colors}
# Canonical cluster orders used consistently throughout the compartment
# analyses. immune_order are metacluster names.
immune_order <- c("T cells", "B cells", "myeloid/NK", "granulocytes")

# Curated tumor celltype order. The previous labels were renamed during
# phenotyping and no longer match spe$celltype_2 / the col_celltype_2 color
# vector (a stale name maps to an NA color and makes ggpubr::ggboxplot() fail
# with "missing value where TRUE/FALSE needed"). Updated to the current names,
# keeping the same order:
#   marker-lo Tu   -> adrenergic marker-low Tu
#   GD2lo Tu       -> GD2neg Tu
#   bridge-like Tu -> neuroblast-like Tu
#   early NB Tu    -> early neuroblast-like Tu
#   CXCR4+ Tu      -> dropped (no longer exists)
tumor_order <- c("adrenergic marker-low Tu", "GD2neg Tu", "CD44+ Tu", "Ki67+ Tu",
                 "neuroblast-like Tu", "early neuroblast-like Tu")
# keep only the tumor celltypes actually present in the data (preserves order)
tumor_order <- intersect(tumor_order,
                         unique(as.character(spe$celltype_2[spe$metacluster == "tumor"])))

# Safe color lookup: fall back to grey if a celltype has no color (or an NA
# color), so ggboxplot() never receives an NA palette.
cluster_color <- function(name) {
  col <- unname(metadata(spe)$color_vectors$col_celltype_2[as.character(name)])
  if (length(col) == 0 || is.na(col)) "grey50" else col
}
```






```{r compartment-boxplots-all-comparisons}
# Runs diffcyt (testDA_edgeR) for every compartment pair × cell type,
# then draws combined boxplots labelled with the log2 fold-change + BH-adjusted
# p-value and significance stars, e.g. "0.37 (0.017 *)".

cells_list       <- c("tumor", "immune")
comp_levels      <- c("core", "border", "stroma")
comparison_pairs <- list(c("core","border"), c("core","stroma"), c("border","stroma"))

# Freq is a percentage, so every panel uses a fixed 0-100 y-axis and the pairwise
# brackets are drawn *above* it (outside the axis) via clip = "off".
#   y_axis_max   : top of the visible axis (data can never exceed 100)
#   bracket_base : height of the lowest bracket, just above the axis
#   bracket_gap  : vertical spacing between the stacked brackets (bigger = more room)
#   title_gap    : bottom margin (pt) reserved under the title so it clears the brackets
y_axis_max        <- 100
bracket_base      <- 106
bracket_gap       <- 14
title_gap         <- 85
bracket_positions <- bracket_base + (seq_along(comparison_pairs) - 1) * bracket_gap
max_plot_columns  <- 5

keep_rows <- rowData(spe)$use_channel

# helper to build marker_info from the global marker vectors
.make_marker_info <- function(spe, keep_rows) {
  channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 2)
  marker_name  <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 1)
  marker_class <- ifelse(rownames(rowData(spe)[keep_rows, ]) %in% type_markers, "type",
                  ifelse(rownames(rowData(spe)[keep_rows, ]) %in% state_markers, "state", "other"))
  data.frame(channel_name, marker_name,
             marker_class = factor(marker_class),
             row.names = marker_name)
}

marker_info_da <- .make_marker_info(spe, keep_rows)

for (cells in cells_list) {

  order     <- if (cells == "immune") immune_order else tumor_order
  keep_meta <- if (cells == "immune") spe$metacluster %in% immune_order else spe$metacluster == "tumor"

  # ------------------------------------------------------------------
  # 1. Precompute diffcyt DA results for every compartment pair
  # ------------------------------------------------------------------
  diffcyt_results <- list()

  for (pair in comparison_pairs) {
    comps     <- pair
    key       <- paste(comps, collapse = "-")
    keep_cols <- keep_meta & spe$compartment %in% comps

    if (sum(keep_cols) < 10) { diffcyt_results[[key]] <- NULL; next }

    sce_da <- SingleCellExperiment(
      assays = list(
        counts = assay(spe, "counts")[keep_rows, keep_cols],
        exprs  = assay(spe, "counts")[keep_rows, keep_cols]
      )
    )
    sce_da$fm_id       <- spe[keep_rows, keep_cols]$fm_id
    sce_da$sample      <- factor(spe[keep_rows, keep_cols]$sample)
    sce_da$compartment <- factor(spe[keep_rows, keep_cols]$compartment, levels = comps)
    sce_da$sample_id   <- factor(paste0(sce_da$fm_id, "_", sce_da$compartment))
    sce_da$condition   <- factor(sce_da$compartment, levels = comps)
    sce_da$cluster_id  <- factor(spe[keep_rows, keep_cols]$celltype_2, levels = order)

    metadata(sce_da)$cluster_codes   <- data.frame(
      celltype = factor(spe[keep_rows, keep_cols]$celltype_2, levels = order)
    )
    metadata(sce_da)$experiment_info <- as.data.frame(colData(sce_da)) %>%
      dplyr::group_by(sample_id, condition, sample) %>%
      dplyr::summarise(n_cells = dplyr::n(), .groups = "drop") %>%
      as.data.frame()

    exp_info           <- metadata(sce_da)$experiment_info
    exp_info           <- exp_info[match(unique(sce_da$sample_id), exp_info$sample_id), ]
    exp_info$condition <- factor(exp_info$condition, levels = comps)

    df_list_da <- lapply(unique(sce_da$sample_id), function(x)
      as.data.frame(t(counts(sce_da[, sce_da$sample_id == x]))))
    names(df_list_da) <- unique(sce_da$sample_id)

    d_se_da <- prepareData(df_list_da, exp_info, marker_info_da)
    rowData(d_se_da)$cluster_id     <- sce_da[, rownames(d_se_da)]$cluster_id
    metadata(d_se_da)$cluster_codes <- metadata(sce_da)$cluster_codes
    d_se_da     <- transformData(d_se_da)
    d_counts_da <- calcCounts(d_se_da)

    design_da   <- createDesignMatrix(exp_info, cols_design = c("condition", "sample"))
    contrast_da <- createContrast(c(0, 1, rep(0, length(unique(exp_info$sample)) - 1)))

    res_DA_da <- tryCatch(
      testDA_edgeR(d_counts_da, design_da, contrast_da,
                   min_cells = 3,
                   min_samples = round(ncol(d_counts_da) * 0.2)),
      error = function(e) { message("diffcyt failed for ", key, ": ", e$message); NULL }
    )
    diffcyt_results[[key]] <- if (!is.null(res_DA_da)) data.frame(rowData(res_DA_da)) else NULL
  }

  # ------------------------------------------------------------------
  # 2. Sample-level proportions (% within compartment) for plotting
  # ------------------------------------------------------------------
  comp_df <- data.frame(
    fm_id       = spe$fm_id,
    compartment = spe$compartment,
    celltype_2  = spe$celltype_2,
    keep        = keep_meta
  )
  comp_df <- comp_df[comp_df$keep & comp_df$compartment %in% comp_levels, ]
  comp_df$compartment <- factor(comp_df$compartment, levels = comp_levels)

  sample_totals <- comp_df %>%
    dplyr::group_by(fm_id, compartment) %>%
    dplyr::summarise(total = dplyr::n(), .groups = "drop")

  plot_data_list <- list()
  for (cluster_name in order) {
    cluster_counts <- comp_df %>%
      dplyr::filter(celltype_2 == cluster_name) %>%
      dplyr::group_by(fm_id, compartment) %>%
      dplyr::summarise(n = dplyr::n(), .groups = "drop")

    plot_data_list[[cluster_name]] <- dplyr::left_join(
      sample_totals, cluster_counts, by = c("fm_id", "compartment")
    ) %>%
      dplyr::mutate(
        n         = dplyr::if_else(is.na(n), 0L, n),
        Freq      = 100 * n / total,
        condition = factor(compartment, levels = comp_levels),
        sample_id = fm_id
      )
  }

  # ------------------------------------------------------------------
  # 3. Collate diffcyt stats for all pairs into one table
  # ------------------------------------------------------------------
  all_stats_list <- list()
  for (cluster_name in order) {
    for (pair in comparison_pairs) {
      key      <- paste(pair, collapse = "-")
      res_pair <- diffcyt_results[[key]]
      if (!is.null(res_pair) && cluster_name %in% rownames(res_pair)) {
        p_adj <- res_pair[cluster_name, "p_adj"]
        logFC <- res_pair[cluster_name, "logFC"]
      } else {
        p_adj <- NA_real_; logFC <- NA_real_
      }
      p_signif <- if (!is.na(p_adj)) gtools::stars.pval(p_adj) else "NA"
      if (p_signif == " ") p_signif <- "ns"
      all_stats_list <- c(all_stats_list, list(data.frame(
        cluster_name = cluster_name,
        group1       = pair[1],
        group2       = pair[2],
        p_adj        = p_adj,
        logFC        = round(logFC, 2),
        label        = paste0(round(logFC, 2), " (",
                              if (!is.na(p_adj)) sprintf("%.3f", p_adj) else "NA",
                              " ", p_signif, ")")
      )))
    }
  }
  all_stats <- dplyr::bind_rows(all_stats_list)

  # ------------------------------------------------------------------
  # 4. Boxplots with diffcyt annotation
  # ------------------------------------------------------------------
  plot_list_comp <- list()
  for (cluster_name in order) {
    proportions <- plot_data_list[[cluster_name]]
    stats_df    <- all_stats[all_stats$cluster_name == cluster_name, ]

    # brackets sit at fixed heights ABOVE the 0-100 axis, identical for every panel
    stats_df$y.position <- bracket_positions[seq_len(nrow(stats_df))]

    color_cluster <- cluster_color(cluster_name)

    plot_list_comp[[cluster_name]] <- ggboxplot(
      proportions, x = "condition", y = "Freq", order = comp_levels,
      color = "condition", palette = rep(color_cluster, 3),
      add = "jitter"
    ) +
      stat_pvalue_manual(stats_df, y.position = "y.position", label = "label") +
      # wrap + shrink the title so long celltype names (e.g. "adrenergic
      # marker-low Tu") are not cropped in the narrow panels
      ggtitle(str_wrap(cluster_name, width = 18)) +
      scale_y_continuous(breaks = seq(0, y_axis_max, 25),
                         expand = expansion(mult = c(0.02, 0))) +
      # lock the visible axis to 0-100 and let the brackets draw outside it
      coord_cartesian(ylim = c(0, y_axis_max), clip = "off") +
      theme(aspect.ratio = 1.2, legend.position = "none",
            # smaller, wrapped, centred title clears the brackets and fits the panel
            plot.title  = element_text(hjust = 0.5, size = 10,
                                       margin = margin(b = title_gap)),
            plot.margin = margin(t = 10, r = 8, b = 8, l = 8))
  }

  p_comp <- gridExtra::grid.arrange(
    grobs = plot_list_comp,
    ncol  = min(length(plot_list_comp), max_plot_columns)
  )

  save_plot(p_comp,
            paste0("Fig4m_boxplot_compartments_", cells, ".pdf"),
            width = 12, height = 8)
}
```

```{r composition-stacked-bars}
# Stacked-bar charts of cell-type composition per compartment, drawn for EVERY
# border size (not just the canonical one). For each border we produce three
# figures: immune metaclusters, tumor celltypes, and the two pooled together.
# spe$compartment (the canonical-border assignment used elsewhere) is left
# untouched - each border's compartment is computed on the fly.

comp_levels <- c("core", "border", "stroma")

# Compartment (core / border / stroma) for a given border size, returned as a
# plain vector without modifying spe. Same rule as the threshold loop above;
# which() drops any NA distances safely.
compartment_at <- function(spe, border) {
  comp <- rep("stroma", ncol(spe))
  ok   <- spe$sample_id != "20221018_20220906_03-018_TU_001"
  comp[which(abs(spe$distToCells) <= border & ok)] <- "border"
  comp[which(comp != "border" & spe$distToCells <= border & ok)] <- "core"
  comp
}

# One stacked-bar figure for a compartment vector and a cell class: "immune"
# (metaclusters), "tumor" (celltypes) or "tumor_immune" (the two pooled).
make_compartment_stackedbar <- function(comp_vec, cells, border) {
  if (cells == "immune") {
    keep <- spe$metacluster %in% immune_order;                       order <- immune_order
  } else if (cells == "tumor") {
    keep <- spe$metacluster == "tumor";                              order <- tumor_order
  } else {                                                           # tumor_immune (pooled)
    keep <- spe$metacluster == "tumor" | spe$metacluster %in% immune_order
    order <- c(tumor_order, immune_order)
  }

  df_bar <- data.frame(compartment = comp_vec, cluster_id = spe$celltype_2, keep = keep)
  df_bar <- df_bar[df_bar$keep & df_bar$compartment %in% comp_levels, ]

  totals <- df_bar %>% dplyr::group_by(compartment) %>%
    dplyr::summarise(N = dplyr::n(), .groups = "drop")

  df_prop <- df_bar %>%
    dplyr::group_by(compartment, cluster_id) %>% dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
    dplyr::group_by(compartment) %>% dplyr::mutate(pct = 100 * n / sum(n)) %>% dplyr::ungroup() %>%
    dplyr::mutate(compartment = factor(compartment, levels = comp_levels),
                  cluster_id  = factor(cluster_id,  levels = order))

  color_vec <- metadata(spe)$color_vectors$col_celltype_2[order]
  lab       <- if (cells == "tumor_immune") "tumor + immune" else cells

  p_stack <- ggplot(df_prop, aes(x = compartment, y = pct, fill = cluster_id)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(
      data = totals %>% dplyr::mutate(compartment = factor(compartment, levels = comp_levels)),
      aes(x = compartment, y = 103, label = paste0("n=", N)),
      inherit.aes = FALSE, vjust = 0, size = 3.5) +
    scale_fill_manual(values = color_vec, drop = FALSE, name = "Cell type") +
    scale_y_continuous(limits = c(0, 115), breaks = seq(0, 100, 25),
                       expand = expansion(mult = c(0, 0))) +
    labs(title = paste0("Compartment composition (", lab, ") - border ", border),
         x = "Compartment", y = "% cells") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.key.size = unit(0.4, "cm"))

  save_plot(p_stack,
            paste0("Fig4_suppl_stackedbar_compartments_", cells, "_bordersize_", border, ".pdf"),
            width = if (cells == "tumor_immune") 7 else 6, height = 5)
  print(p_stack)
}

for (border in thresholds) {
  comp_vec <- compartment_at(spe, border)
  for (cells in c("immune", "tumor", "tumor_immune"))
    make_compartment_stackedbar(comp_vec, cells, border)
}
```



## Save R object

Finally, we save the R object.

```{r save-rds}
saveRDS(spe, file.path(params$output,"spe_07_spatial_analysis.rds"))
```


## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
