---
title: "Differential abundance between clinical conditions in diagnostic bone marrow samples"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/08_BM_DE_DA")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# DA analysis between clinical conditions in BM samples {#bm-da}

Herein, we explore differential cell type abundances in tumor and TME.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(SingleCellExperiment)
library(CATALYST)
library(diffcyt)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)
library(tidyr)
library(BioQC)
```
</details>

## Read data

First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final_clustering.rds"))
```

We will then transform the SPE object into a catalyst-compatible object. Select whether you want to investigate differential abundance (DA) in the tumor (`tumor`) or TME (`TME`) compartment. We decide whether we want to compare between progression (`prog`) or MYCN (`mycn`) status. Reduce the dataset to only `nMNA` or `MNA` samples with `reduce_to`.

```{r catalyst-comp-object}
compartment <- "tumor"
status <- "prog"
reduce_to <- "none"

if (compartment == "tumor"){
  # Step 1: count tumor cells per fm_id at DE
  counts <- table(spe$fm_id[spe$metacluster == "tumor" & spe$timepoint == "DE" & spe$tissue=="BM"])
  
  # Step 2: keep fm_id with >100 tumor cells
  valid_fm <- names(counts[counts > 100])
  
  # Step 3: filter
  keep_cols <- spe$fm_id %in% valid_fm & 
               spe$metacluster == "tumor" & 
               spe$timepoint == "DE" &
              spe$tissue=="BM"
}else{
  keep_cols <- spe$celltype != "other" & spe$timepoint=="DE" & spe$metacluster!="tumor" & spe$metacluster!="mesenchymal"  & spe$tissue=="BM"
}

if (status == "mycn"){
  keep_cols <- keep_cols & (spe$MYCN_amp == 0 | spe$MYCN_amp == 1)}

if (reduce_to == "nMNA"){keep_cols <- keep_cols & spe$MYCN_amp == 0}
if (reduce_to == "MNA"){keep_cols <- keep_cols & spe$MYCN_amp == 1}

keep_rows <- rowData(spe)$use_channel 

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$sample_id <- factor(spe[keep_rows, keep_cols]$sample_id)
sce$age_diag <- spe[keep_rows, keep_cols]$age_diag

sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))
col_vector <- metadata(spe)$color_vectors$col_celltype_ordered

if (status == "prog"){
  sce$condition <- sce$progress
}else{
  sce$condition <- sce$mna
}

# Add experiment_info
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell type and state markers
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))
```

## Plot cellular proportions per sample per condition

We then visualize the cellular proportions per sample and condition in barplots.

```{r cohort-overview, fig.width=10, fig.height=5}
set.seed(20231103)

p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=col_vector, 
                    linkage="ward.D", distance="manhattan") 

save_plot(p1, paste0("BM_abundance_", status, "_stackedbarplot.png"),  width = 18, height = 10)

p
```
We save the order of the clustered samples to be able to plot corresponding metadata.

```{r store-order}
condition <- levels(sce$condition)
clustered_samples <- list()
for (i in seq_along(condition)){
  clustered_samples <- c(clustered_samples, levels(droplevels(p$data[p$data$condition==condition[i],]$sample_id)))
  }
clustered_samples <- as.character(clustered_samples)
```

## Compute heterogeneity

We compute intra-sample heterogeneity.

```{r heterogeneity}
proportions <- plotAbundances(sce, k = "celltype", by = "sample")
prop <- proportions$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)

prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)

rownames(prop) <- levels(sce$cluster_id)

prop <- prop[,clustered_samples]
het <- apply(prop,2, entropy)
```

## Plot metadata 

Then we plot intra-sample heterogeneity together with sample-specific metadata for the clustered samples.

```{r plot-metadata}
#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(age_diag = first(age_diag),
            condition = first(condition)
  )
sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)
sample_summary <- sample_summary[clustered_samples,]

age <- sample_summary$age_diag
condition <- sample_summary$condition

het_t <- as.vector(unname(het[clustered_samples]))

#calculate tumor infiltration per sample
total_cells_per_sample <- table(spe$sample_id)[clustered_samples]
comp_cells_per_sample <- table(sce$sample_id)[clustered_samples]
perc_tumor <- round((as.vector(comp_cells_per_sample) / as.vector(total_cells_per_sample)) * 100, 3)

# Plot metadata
cont_df <- data.frame(perc_tumor, het_t, age, condition,row.names=clustered_samples)

plot_list <- list()

for (d in colnames(cont_df)[1:ncol(cont_df)-1]) {

  df <- cont_df[,c(d, "condition")] 
  
  colnames(df) <- c("y", "condition")
  stats_df <- compare_means(y ~ condition,  
                        p.adjust.method = "BH", 
                        method='wilcox.test', 
                        data = df)
  
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0(stats_df$p.adj, " (", stats_df$p_adj_signif, ")")
  print(d)
  print(stats_df)

plot_list[[d]] <- ggboxplot(df, x = "condition", y = "y",
                color = "condition", palette = c("slategray1", "slategray"),
                add.params = list(size = 1.5),
                add = "jitter") + 
  stat_pvalue_manual(stats_df, label = "p_long", 
                     y.position = max(df$y, na.rm = TRUE) * 1.1, 
                     step.increase = 0.2) +
  theme(aspect.ratio = 2, legend.position = "none")
}

p_boxplot <- do.call("grid.arrange", c(plot_list, ncol=3))

save_plot(p_boxplot, paste0("Fig6e_BM_abundance_", status, "_boxplot_metadata.png"), width = 18, height = 10)
```

## Differential abundance analysis

We first prepare data for DA analysis using the `diffcyt` package. 

```{r prepare-data}

#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                                    ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between clinical conditions. 

```{r da-analysis}
# Create design matrix
design <- createDesignMatrix(experiment_info, cols_design = "condition")

# Create contrast matrix
contrast <- createContrast(c(0, 1)) 
nrow(contrast) == ncol(design)

if(compartment == "tumor"){
  min_sam = round(dim(d_counts)[2]*0.2)
}else{
  min_sam = round(dim(d_counts)[2]*0.05)
}

# Test for differential abundance (DA) of clusters
res_DA <- testDA_voom(d_counts,design, contrast, 
                       min_cells=3, 
                       min_samples=min_sam
                       )

FDR_cutoff = 0.1
rowData(res_DA)[rowData(res_DA)$p_adj<FDR_cutoff,]
res<- data.frame(rowData(res_DA))
```
## Visualization of DA results

Finally, we visualize DA results in boxplots and barplots showing the logFC.

```{r visualize-da, fig.width=12, fig.height=10}
if(status=="prog"){
  comparisons <- c("NoProgvsProg")
  ord <- c("NoProg", "Prog")
}else{
  comparisons <- c("nMNAvsMNA")
  ord <- c("nMNA", "MNA")
}

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  color <- unname(col_vector[as.character(cluster_name)])
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=ord,
              color = "condition", palette =c(color, color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

grid.arrange(grobs = plot_list_fc, ncol = 5)

save_plot(p_boxplot, paste0("Fig6g_BM_abundance_", status, "_boxplot_celltype.png"), width = 18, height = 10)
```
```{r barplot-logFC, fig.width=5, fig.height=8}
df <- res[, c("logFC", "p_adj")]
df$p_adj_signif <- stars.pval(df$p_adj)
df$p_adj_signif[df$p_adj_signif==" "] <- "ns"
df$cluster <- rownames(df)

p1 <- ggplot(df, aes(fill=cluster, y=logFC, x=cluster)) + 
  ggtitle(colnames(contrasts)[i]) +
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = F) + 
  scale_fill_manual(values = col_vector) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_text(aes(label = p_adj_signif), vjust = 0)

save_plot(p1, paste0("Fig6g2_BM_abundance_", status, "_barplot.png"),  width = 18, height = 10)
  
p1
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
