---
title: "Identifying cellular communities in diagnostic bone marrow samples"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/home/rstudio/data/input_all"
  output: "/scratch/research/taschnermandl/rstudio/mapmetsc"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/08_BM_DE_cellularcomm")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    # ggplot2 objects
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, "Heatmap") || inherits(plot_obj, "HeatmapList")) {
    # ComplexHeatmap objects - save as PDF
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    # Base R plots or other objects - assume already rendered, save via dev.copy
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Identifying cellular communities in diagnostic bone marrow samples {#bm-dx}

We next explore the contribution of celltypes to disease progression. Therefore, we perform hierarchical clustering of cell composition profiles in bone marrow samples.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(gtools)
library(edgeR)
```
</details>

## Read data

First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final_clustering.rds"))
```

We will then transform the SPE object into a catalyst-compatible object. 

```{r catalyst-comp-object}

keep_rows <- rowData(spe)$use_channel 
# Step 1: count tumor cells per fm_id at DE
counts <- table(spe$fm_id[spe$metacluster == "tumor" & spe$timepoint == "DE" & spe$tissue=="BM"])

# Step 2: keep fm_id with >100 tumor cells
valid_fm <- names(counts[counts > 100])
samples_with_tumor_cells <- valid_fm

# Step 3: filter
keep_cols <- spe$fm_id %in% valid_fm & 
             spe$celltype != "other" & 
             spe$timepoint == "DE" &
            spe$tissue=="BM"

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id, "_", sce$tissue))

sce$age_diag <- spe[keep_rows, keep_cols]$age_diag
sce$solidity_clipped <- spe[keep_rows, keep_cols]$solidity_clipped

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)

sce$condition <- 'na' 

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))
```

## Cluster cellular communities

Next, we cluster proportions of tumor cell types across all diagnostic BM samples (instead of conditions) and plot the result in barplots:

```{r cohort-overview, fig.width=11, figh.height=5}
set.seed(20231103)
sce_tumor <- sce[,sce$metacluster=="tumor"]

p <- plotAbundances(sce_tumor, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                    linkage="ward.D", distance="manhattan") 

save_plot(p, "Fig6a1_BM_abundance_clustering_tumor.png", width = 18, height = 10)

p
```

along with the dendogram:

```{r dendogram, fig.height=8, fig.width=11}
#Dendogram
set.seed(20231103)
wide_df <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

wide_df <- data.frame(wide_df)
rownames(wide_df) <- wide_df$sample_id
wide_df <- wide_df %>% select(-sample_id)

dist_matrix <- dist(wide_df, method = "manhattan")
hc <- hclust(dist_matrix, method = "ward.D")

save_plot(plot(hc), "Fig6a2_BM_abundance_clustering_dendrogram.png", width = 8, height = 6)

plot(hc)
```
We then save the order of clustered samples along with cluster labels.

```{r}
clustered_samples <- levels(p$data$sample_id)

#Extract clusters of cellular communities
cellular_communities <- as.data.frame(cutree(hc, k = 2)[levels(p$data$sample_id)])
colnames(cellular_communities) <- c("community_cluster")
cluster_names <- setNames(order(unique(cellular_communities$community_cluster)), c(1:max(cellular_communities$community_cluster)))
cellular_communities$community_cluster <- recode(cellular_communities$community_cluster, !!!cluster_names)
```

## Plot TME cellular profiles 

We will then plot the cellular profiles of the TME for these samples. 

```{r plot-TME, fig.width=13, fig.height=6}
#order samples based on above clustering
sce$sample_id <- factor(sce$sample_id, levels=clustered_samples)

sce_ME <- sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]

p_ME <- plotAbundances(sce_ME, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_celltype_ordered)

save_plot(plot(hc), "Fig6a3_BM_abundance_clustering_immune.png", width = 8, height = 6)

p_ME
```

## Compute tumor and TME heterogeneity

We compute tumor and TME intra-sample heterogeneity.

```{r heterogeneity}
proportions <- plotAbundances(sce, k = "celltype", by = "sample")
prop <- proportions$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)
prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)

me_clusters <- unique(sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]$cluster_id)
me_prop <- prop[rownames(prop)%in%me_clusters,]
me_het <- apply(me_prop,2, entropy)
  
tumor_clusters <- unique(sce[,sce$metacluster=="tumor"]$cluster_id)
tumor_prop <- prop[rownames(prop)%in%tumor_clusters,]
tumor_het <- apply(tumor_prop,2, entropy)
```

## Plot metadata 

Then we plot intra-sample heterogeneity together with sample-specific metadata for the clustered sammples.

```{r plot-metadata}
#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
            age_diag = first(age_diag)
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


#assign color codes
col_code <- c("MNA"=rgb(202,0,32, maxColorValue = 255),
              "het"=rgb(146,197,222, maxColorValue = 255),
              "nMNA"=rgb(5,113,176, maxColorValue = 255),
              "Prog"=rgb(202,0,32, maxColorValue = 255),
              "NoProg"=rgb(5,113,176, maxColorValue = 255)
              )

col_age_diag = colorRamp2(c(min(sample_summary$age_diag, na.rm = TRUE), (quantile(sample_summary$age_diag, 0.95, na.rm = TRUE)-min(sample_summary$age_diag, na.rm = TRUE))/2, quantile(sample_summary$age_diag,0.95, na.rm = TRUE)), viridis(3))
col_me_het = colorRamp2(c(min(me_het, na.rm = TRUE), (max(me_het, na.rm = TRUE)-min(me_het, na.rm = TRUE))/2, max(me_het, na.rm = TRUE)), viridis(3))
col_tumor_het = colorRamp2(c(min(tumor_het, na.rm = TRUE), (max(tumor_het, na.rm = TRUE)-min(tumor_het, na.rm = TRUE))/2, max(tumor_het, na.rm = TRUE)), viridis(3))

# Plot metadata
p_col <- columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(age=sample_summary$age_diag, col=list(age=col_age_diag)) %v%
  columnAnnotation(TUM_het=tumor_het, col=list(TUM_het=col_tumor_het)) %v%
  columnAnnotation(ME_het=me_het, col=list(ME_het=col_me_het))

save_plot(p_col, "Fig6a4_BM_abundance_clustering_metadata.png", width = 8, height = 6)
  
```

## Plot contiunous and discrete summary statistics

We then plot summary statistics of continuous and discrete metadata in boxplots and piecharts, respectively.

```{r piecharts, fig.height=4, fig.width=5}
sample_summary[rownames(cellular_communities),"community_clusters"] <- cellular_communities$community_cluster

discrete_metadata <- c("mna", "progress")

discrete_metadata_colors <- list(col_code, col_code)

plot_list <- list()
for (d in seq_along(discrete_metadata)) {
  sample_meta <- sample_summary[,c(discrete_metadata[d], "community_clusters")]
  sample_meta <- sample_meta[rowSums(sample_meta == "unknown") == 0, ] 
  colnames(sample_meta) <- c("dis_var", "condition")
  
  result <- sample_meta %>%
    group_by(condition, dis_var) %>%
    summarise(Freq = n(), .groups = 'drop')
  
  result[,"dis_var"]<- as.factor(unlist(result[,"dis_var"]))
  
plot_list[[d]] <- ggplot(result, aes(x="", y=Freq, fill=dis_var, group=dis_var)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = discrete_metadata_colors[[d]], name=discrete_metadata[d]) +
    coord_polar("y", start=0)  +  
    facet_grid(.~ condition) +
    theme_void()
}

p_pie <- do.call("grid.arrange", c(plot_list, ncol=1))

save_plot(p_pie, "Fig6b_BM_piecharts_metadata.png", width = 8, height = 6)

p_pie
```

```{r boxplots}
het_t <- as.vector(unname(tumor_het[clustered_samples]))
perc_tumor <- round((as.vector(unname(table(sce_tumor$sample_id)[clustered_samples]))/as.vector(unname(table(sce$sample_id)[clustered_samples])))*100,3)
age <- sample_summary$age_diag

cont_df <- data.frame(cbind(het_t, perc_tumor, age, cellular_communities),row.names=clustered_samples)
comp <- list( c("1", "2"), c("1", "3"), c("2", "3") )

plot_list <- list()

for (d in colnames(cont_df)[1:ncol(cont_df)-1]) {

  df <- cont_df[,c(d, "community_cluster")] 
  
  colnames(df) <- c("y", "community_cluster")
  stats_df <- compare_means(y ~ community_cluster, 
                        comparisons = comp, 
                        p.adjust.method = "BH", 
                        method='wilcox.test', 
                        data = df)
  
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0(stats_df$p.adj, " (", stats_df$p_adj_signif, ")")

plot_list[[d]] <- ggboxplot(df, x = "community_cluster", y = d,
                color = "community_cluster", palette =c("slategray1", "slategray", "slategray3"),
                add.params = list(size = 3),
                add = "jitter")+ 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(df$y)+0.5, step.increase = 0.2) +
                theme(aspect.ratio = 2, legend.position="none") 
}

p_boxplot <- do.call("grid.arrange", c(plot_list, ncol=3))

save_plot(p_boxplot, "Fig6c_BM_het_boxplot_metadata.png", width = 8, height = 6)
```

## Plot metaclusters

Finally, we plot the proportions of metaclusters in the clustered samples.

```{r plot-metaclusters, fig.width=11, figh.height=5}
sce_tmp <- sce
sce_tmp$cluster_id <- sce_tmp$metacluster
metadata(sce_tmp)$cluster_codes <- data.frame(celltype = factor(sce_tmp$metacluster))
p_all <- plotAbundances(sce_tmp, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_metacluster)
rm(sce_tmp)


save_plot(p_all, "Fig6a5_BM_abundance_clustering_metacluster.png", width = 8, height = 6)

p_all
```

# Differential abundance analysis between cell community clusters

In this second part, we perform differential abundance analysis between identified cellular community clusters. Therefore we set the community cluster as new condition. Select whether you want to plot DA for tumor (`tumor`) or TME (`TME`) celltypes. 

```{r set-condition}
compartment <- "tumor"

keep_rows <- rowData(spe)$use_channel 

if (compartment=="tumor"){
  keep_cols <- spe$fm_id %in% valid_fm & spe$celltype != "other" & spe$tissue=="BM" & spe$timepoint=="DE" & spe$metacluster == "tumor"
}else{
  keep_cols <- spe$fm_id %in% valid_fm & spe$celltype != "other" & spe$tissue=="BM" & spe$timepoint=="DE" & spe$metacluster != "tumor" & spe$metacluster != "mesenchymal"
}

sce_comp <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce_comp$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce_comp$progress, `0` = 'NoProg', `1` = 'Prog')
sce_comp$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce_comp$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce_comp$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce_comp$mna <- as.factor(mna)

sce_comp$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce_comp$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce_comp$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce_comp$sample_id <- factor(paste0(sce_comp$progress, "_", sce_comp$mna, "_", sce_comp$patient, "_", sce_comp$fm_id, "_", sce_comp$tissue), levels=clustered_samples)

sce_comp$age_diag <- spe[keep_rows, keep_cols]$age_diag
sce_comp$solidity_clipped <- spe[keep_rows, keep_cols]$solidity_clipped

sce_comp$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce_comp$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)

sce_comp$condition <- sce_comp$sample_id
sample_clusters <- setNames(paste0("c_", cellular_communities$community_cluster), rownames(cellular_communities))
sce_comp$condition <- as.factor(recode(sce_comp$condition, !!!sample_clusters))

# Add celltype information to metadata
metadata(sce_comp)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
metadata(sce_comp)$experiment_info <- as.data.frame(colData(sce_comp)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to sce
rowData(sce_comp)$marker_class <- ifelse(rownames(sce_comp) %in% type_markers, "type",
                                    ifelse(rownames(sce_comp) %in% state_markers, "state", 
                                    "other"))
```

## Prepare data

We first prepare data for DA analysis using the `diffcyt` package. 

```{r prepare-data}
# list of samples (one data.frame per sample)
df_list <- lapply(unique(sce_comp$sample_id), function(x) {
  as.data.frame(t(counts(sce_comp[, sce_comp$sample_id == x])))
})
names(df_list) <- unique(sce_comp$sample_id)

# experiment info
experiment_info <- metadata(sce_comp)$experiment_info
experiment_info <- experiment_info[match(names(df_list), experiment_info$sample_id), ]

# marker info
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 2)
marker_name  <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 1)

marker_class <- ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                 ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", "other"))

marker_info <- data.frame(channel_name, marker_name, row.names = marker_name)
marker_info$marker_class <- factor(marker_class)

# prepare data
d_se <- prepareData(df_list, experiment_info, marker_info)

# add cluster info
rowData(d_se)$cluster_id <- sce_comp[, rownames(d_se)]$cluster_id
metadata(d_se)$cluster_codes <- metadata(sce_comp)$cluster_codes

# transform + count
d_se     <- transformData(d_se)
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between community clusters.

```{r da-analysis}
#Create design matrix
design <- model.matrix(~0+condition, experiment_info) 

#comparing between cellular communities
colnames(design) <- c("c_1", "c_2")
contrasts <- makeContrasts(c_1vsc_2 = c_1-c_2,
                           levels = design)

if (compartment=="tumor"){
  min_sam <- round(dim(d_counts)[2]*0.2)
}else{
  min_sam <- 1
}

res <- list()

for (i in 1:ncol(contrasts)){

    res_DA <- testDA_voom(d_counts,design, contrasts[,i],
                       min_cells=3, 
                       min_samples = min_sam
                       )
    
    res <- append(res, list(data.frame(rowData(res_DA))))
}
```

## Visualization of DA results

Finally, we visualize DA results in boxplots and barplots.

```{r visualize-da, fig.width=12, fig.height=10}
p <- plotAbundances(sce_comp, k = "celltype", by = "cluster_id", shape_by = "sample_id")

comparisons <-colnames(contrasts)

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:ncol(contrasts)){
  stats <- res[[j]]
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_celltype_ordered[as.character(cluster_name)])
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("c_1", "c_2", "c_3"),
              color = "condition", palette=c(color, color, color), 
              add.params = list(size = 1.5),
              add = "jitter", title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 2, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
              
}

p_boxplot <- grid.arrange(grobs = plot_list_fc, ncol = 5)

save_plot(p_boxplot, "Fig6associated_BM_abundance_clustering_boxplot_celltype.png", width = 16, height = 12)
```
```{r barplot-logFC, fig.width=16, fig.height=8}
plot_list <- list()

for (i in 1: length(colnames(contrasts))){
  df <- res[[i]][, c("logFC", "p_adj")]
  df$p_adj_signif <- stars.pval(df$p_adj)
  df$p_adj_signif[df$p_adj_signif==" "] <- "ns"
  df$cluster <- rownames(df)
  
  plot_list[[i]] <- ggplot(df, aes(fill=cluster, y=logFC, x=cluster)) + 
    ggtitle(colnames(contrasts)[i]) +
    geom_bar(
      stat="identity",
      position="dodge",
      show.legend = F) + 
    scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    geom_text(aes(label = p_adj_signif), vjust = 0)
  
}

p_barplot <- grid.arrange(grobs = plot_list, ncol = 1)

save_plot(p_barplot, "Fig6d_BM_abundance_clustering_barplot_celltype.png", width = 10, height = 6)

p_barplot
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
