---
title: "Supplementary: region and marker images"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input:  !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
  # Image(s) to render. Give a vector of ids in `image_ids` to render several in
  # one pass; a single `image_id` is still accepted (back-compat). Override from
  # run_analysis.R (see the render_step call). If `image_ids` is set it wins.
  image_id:  "20220926_20220809_16-006_TU_003"
  image_ids: !r NULL
  # Markers rendered individually and as a colour merge. Either full channel
  # names (as in rownames(spe), e.g. "CD3_Sm152_mean") or the short marker name
  # (e.g. "CD3") - short names are resolved to the matching channel below.
  # (DNA1_Ir193_mean is the Ir193 DNA channel, i.e. DNA2.)
  markers: !r c("Vimentin_Pt196_mean", "CD3_Sm152_mean", "DNA1_Ir193_mean", "CD14_Lu175_mean", "GD2_Gd155_mean", "CD15_Bi209_mean")
  # Colour for each marker, used for BOTH its individual image and the merge.
  # Keyed by the short marker name (token before the first "_"); "DNA1" keys the
  # DNA channel. Markers without an entry fall back to a spare-colour palette.
  marker_colours: !r c(CD3 = "red", Vimentin = "purple", CD14 = "cyan", DNA1 = "blue", GD2 = "green", CD15 = "yellow")
  # Short marker names to draw individually but leave OUT of the colour merge.
  # None here: all six markers go into the merge (GD2 now takes CD56's green slot).
  merge_exclude: !r character(0)
  # Zoom window(s) c(xmin, xmax, ymin, ymax) in pixel coordinates of the image
  # array (x = first image dimension, y = second). Both the whole image AND any
  # zoom are rendered. `crops` is a list with one entry per `image_ids` id
  # (NULL = whole image only, the default for every image); `crop` is a single
  # window for the back-compat single-image path. NULL renders whole image only.
  crop:  !r NULL
  crops: !r NULL
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/08_generate_images")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList"))) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

knitr::opts_chunk$set(
  echo = TRUE, error = FALSE, message = FALSE, warning = FALSE,
  dev = "png", dpi = 150, fig.width = 8, fig.height = 8
)
```

# Region and marker images

This standalone supplementary report renders, for a single image — as **both the
whole image and a zoomed region**:

* each requested **marker** as a separate pixel image,
* a colour **merge** of all requested markers,
* the **cell types**,
* the **spatial compartment** (tumor core / border / stroma), and
* the **cellular neighborhood**.

The cell-level panels (cell types, compartment, neighborhood) are drawn on the
segmentation **masks** (`plotCells`), not as points. It is meant to be rendered
from `run_analysis.R` with a *customised* parameter list (image id, zoom crop,
marker list) rather than the shared `main_params` — see the
`render_step("analysis/08_generate_images.Rmd", ...)` call there. The
`compartment` and `cn_celltypes_2` annotations are read from
`spe_07_spatial_analysis.rds`, so the chosen image must be one of the
primary-tumour (PT) images processed in [chapter 07](07_spatial_analysis.html).

## Load libraries
```{r load-libraries}
library(SpatialExperiment)
library(cytomapper)
library(cowplot)
library(RColorBrewer)
library(ggplot2)
```

## Load data and images
```{r load-data}
spe <- readRDS(file.path(params$output, "spe_07_spatial_analysis.rds"))

# --- images to render + their optional zoom crops ---------------------------
# New interface: params$image_ids (a character vector) renders several images in
# one pass, with params$crops a list holding one zoom window per id (a NULL entry
# = whole image only). Back-compat: a single params$image_id (+ params$crop).
# Every crop defaults to NULL, i.e. whole image only.
image_ids <- if (!is.null(params$image_ids)) params$image_ids else params$image_id
if (is.null(image_ids) || !length(image_ids))
  stop("No image id provided: set params$image_ids (or params$image_id).")
crops <- if (!is.null(params$crops)) params$crops else
         if (!is.null(params$crop)) rep(list(params$crop), length(image_ids)) else
         vector("list", length(image_ids))          # all NULL -> whole image only
if (length(crops) != length(image_ids))
  stop("params$crops must have one entry per image id (NULL = whole image only). ",
       "Got ", length(crops), " crop(s) for ", length(image_ids), " id(s).")

missing_ids <- setdiff(image_ids, unique(as.character(spe$sample_id)))
if (length(missing_ids))
  stop("image id(s) not among the PT images in spe_07_spatial_analysis.rds: ",
       paste(missing_ids, collapse = ", "), ". Available images include: ",
       paste(head(unique(as.character(spe$sample_id))), collapse = ", "), " ...")

# Channel (_mean) marker names from the spe. Image channels are named from these
# (in order), and requested markers are resolved against them.
marker_names <- rownames(spe)[grepl("_mean$", rownames(spe))]

# Resolve requested markers: accept either a full channel name or a short marker
# name (the token before the first "_", e.g. "CD3" -> "CD3_Sm152_mean").
resolve_markers <- function(requested, channels) {
  token <- sub("_.*$", "", channels)
  vapply(requested, function(r) {
    if (r %in% channels) return(r)
    hit <- channels[token == r]
    if (length(hit)) hit[1] else NA_character_
  }, character(1))
}

markers_full <- resolve_markers(params$markers, marker_names)
if (anyNA(markers_full))
  warning("Markers not found among the spe channels (dropped): ",
          paste(params$markers[is.na(markers_full)], collapse = ", "))
markers <- unname(markers_full[!is.na(markers_full)])
if (!length(markers)) stop("None of the requested markers match the spe channels.")
message("Rendering markers: ", paste(markers, collapse = ", "))

# Load one image + its mask + the matching cells (tmp). Channels are named from
# marker_names using the image's own channel count (robust to raw images that
# still carry channels dropped from the intensity tables). Dimensions are logged
# so an out-of-range crop - the usual cause of a black zoom panel - is easy to spot.
load_image <- function(image_id) {
  tmp   <- spe[, spe$sample_id == image_id]
  img   <- loadImages(file.path(params$input, "img",   paste0(image_id, ".tif")))
  masks <- loadImages(file.path(params$input, "masks", paste0(image_id, ".tif")), as.is = TRUE)
  mcols(img)   <- DataFrame(sample_id = names(img))
  mcols(masks) <- DataFrame(sample_id = names(masks))
  message("Loaded '", image_id, "': image ", dim(img[[1]])[1], " x ", dim(img[[1]])[2],
          " px (", dim(img[[1]])[3], " channels); mask ",
          dim(masks[[1]])[1], " x ", dim(masks[[1]])[2], " px.")
  n_ch <- dim(img[[1]])[3]
  channelNames(img) <- if (n_ch <= length(marker_names)) {
    marker_names[seq_len(n_ch)]
  } else {
    c(marker_names, paste0("extra_channel_", seq_len(n_ch - length(marker_names))))
  }
  list(img = img, masks = masks, tmp = tmp)
}
```

## Colours and helpers
```{r helpers}
# Clamp/shift a requested [lo, hi] window so it fits inside [1, n]. If the window
# is at least as wide as the axis it falls back to the whole axis; otherwise it
# is shifted inward to stay in range. Returns integer c(lo, hi).
.clamp_axis <- function(lo, hi, n) {
  lo <- as.integer(round(lo)); hi <- as.integer(round(hi))
  if (lo > hi) { tmp <- lo; lo <- hi; hi <- tmp }
  w <- hi - lo
  if (w >= n - 1L) return(c(1L, n))
  lo <- max(1L, min(lo, n - w))
  c(lo, lo + w)
}

# Crop a CytoImageList to c(xmin, xmax, ymin, ymax). The window is clamped to the
# image bounds (an out-of-range crop is shifted in / falls back to the whole
# axis, with a message), and every cropped frame is rebuilt as a proper EBImage
# Image - plain `[` subsetting can drop the Image class and yield a blank
# CytoImageList that plotPixels/plotCells then render black.
crop_cil <- function(cil, crop) {
  d0 <- dim(cil[[1]])
  xr <- .clamp_axis(crop[1], crop[2], d0[1])
  yr <- .clamp_axis(crop[3], crop[4], d0[2])
  if (!identical(c(xr, yr), as.integer(c(crop[1], crop[2], crop[3], crop[4]))))
    message("Crop x[", crop[1], "-", crop[2], "] y[", crop[3], "-", crop[4],
            "] adjusted to x[", xr[1], "-", xr[2], "] y[", yr[1], "-", yr[2],
            "] to fit the ", d0[1], " x ", d0[2], " px image.")
  xs <- seq(xr[1], xr[2]); ys <- seq(yr[1], yr[2])
  imgs <- lapply(seq_along(cil), function(i) {
    im  <- cil[[i]]
    dat <- EBImage::imageData(im)
    sub <- if (length(dim(im)) == 3L) dat[xs, ys, , drop = FALSE]
           else                       dat[xs, ys,    drop = FALSE]
    EBImage::Image(sub, colormode = EBImage::colorMode(im))
  })
  out <- CytoImageList(imgs)
  names(out) <- names(cil)
  mcols(out) <- mcols(cil)
  if (!is.null(channelNames(cil))) channelNames(out) <- channelNames(cil)
  out
}

# Rotate a CytoImageList 90 degrees counter-clockwise (lossless: transpose then
# vertical flip, so integer mask labels are preserved). The raw cytomapper
# images come out 90 degrees clockwise relative to the plotSpatial figures in 07
# (which plot cell coordinates, not the image matrix); this brings the panels
# into the same orientation as 07_spatial_analysis.
rotate90_ccw <- function(cil) {
  out <- CytoImageList(lapply(seq_along(cil), function(i)
    EBImage::flip(EBImage::transpose(cil[[i]]))))
  names(out) <- names(cil)
  mcols(out) <- mcols(cil)
  if (!is.null(channelNames(cil))) channelNames(out) <- channelNames(cil)
  out
}

# core / border / stroma coloured with the ggplot2 default discrete palette
# (scales::hue_pal, alphabetical order: border, core, stroma), so the mask image
# matches the plotSpatial compartment figures in 07_spatial_analysis, which use
# those default colours.
compartment_cols <- setNames(scales::hue_pal()(3), c("border", "core", "stroma"))

# Same Set3 palette / neighborhood numbering as Fig4f in chapter 07.
cn_levels <- levels(factor(spe$cn_celltypes_2))
cn_cols   <- setNames(brewer.pal(max(3, length(cn_levels)), "Set3")[seq_along(cn_levels)],
                      cn_levels)

# Scale bar (IMC pixel size is 1 um, so a length in px is a length in um). It is
# drawn ONLY on the merge panel; cex sizes the "100 um" label - small enough to
# stay inside the frame (the default was cropped), large enough to read. The
# image name / title is removed from every panel (image_title = NULL).
scale_bar_spec <- list(length = 100, label = "100 µm", cex = 1.5)

# Evaluate a plotPixels/plotCells call lazily; on success save + show it and
# return TRUE, on any error report it, skip, and return FALSE - so one bad panel
# never aborts the whole render. A flat channel (a marker with no signal in this
# view) makes EBImage::normalize fail with "clipping range is 0"; such markers
# fail here and are simply left out (also of the merge). `plot_call` is a
# promise, so the plotting only runs inside the tryCatch below.
render_and_save <- function(plot_call, filename, width, height) {
  po <- tryCatch(plot_call,
                 error = function(e) {
                   message("Skipped ", filename, ": ", conditionMessage(e)); NULL
                 })
  if (is.null(po)) return(invisible(FALSE))
  p <- ggdraw(po$plot)
  save_plot(p, filename, width = width, height = height)
  print(p)
  invisible(TRUE)
}

# Colour for each marker (used for BOTH its individual image and the merge),
# keyed by the short marker name (token before the first "_"). Markers without an
# explicit colour fall back to spare hues so they stay distinct in a merge.
.marker_col_map <- local({
  fallback <- c("white", "orange", "yellow", "magenta", "pink", "brown")
  tok  <- sub("_.*$", "", markers)
  cols <- unname(params$marker_colours[tok])   # NA where the short name is unmapped
  miss <- which(is.na(cols))
  if (length(miss))
    cols[miss] <- fallback[((seq_along(miss) - 1) %% length(fallback)) + 1]
  setNames(cols, markers)
})

# Marker channels to draw individually but leave OUT of the colour merge,
# matched by short marker name via params$merge_exclude.
.merge_exclude <- markers[sub("_.*$", "", markers) %in% params$merge_exclude]

# Render every panel (individual markers, merge, celltype, compartment,
# neighborhood) for one view (whole image or a crop). tag goes into the filename.
# No legends are drawn - each panel is just the image.
plot_region <- function(img_v, masks_v, tmp, markers, tag) {

  # --- individual markers, each in its assigned merge colour; keep the ones
  #     that render (a marker absent from this view is a flat channel, skipped) ---
  usable <- character(0)
  for (m in markers) {
    ok <- render_and_save(
      plotPixels(img_v, colour_by = m,
                 colour = setNames(list(c("black", .marker_col_map[[m]])), m),
                 bcg = setNames(list(c(0, 1, 1)), m),
                 image_title = NULL, scale_bar = NULL,
                 legend = NULL, return_plot = TRUE),
      paste0(tag, "_marker_", m, ".png"), 8, 8)
    if (isTRUE(ok)) usable <- c(usable, m)
  }

  # --- merge of the markers that rendered, minus any flagged individual-only;
  #     same per-marker colours. The scale bar is drawn on this panel only.
  #     cytomapper::plotPixels composites at most 6 markers, so cap the merge at
  #     6 (dropping the extras, which still have their own individual panels). ---
  merge_markers <- setdiff(usable, .merge_exclude)
  if (length(merge_markers) > 6) {
    message("Merge capped at 6 markers (cytomapper::plotPixels max); leaving out: ",
            paste(sub("_.*$", "", merge_markers[-seq_len(6)]), collapse = ", "))
    merge_markers <- merge_markers[seq_len(6)]
  }
  if (length(merge_markers)) {
    colour_list <- setNames(lapply(merge_markers, function(mm) c("black", .marker_col_map[[mm]])), merge_markers)
    bcg_list    <- setNames(lapply(merge_markers, function(x) c(0, 1, 1)), merge_markers)
    render_and_save(
      plotPixels(img_v, colour_by = merge_markers, colour = colour_list,
                 bcg = bcg_list, image_title = NULL, scale_bar = scale_bar_spec,
                 legend = NULL, return_plot = TRUE),
      paste0(tag, "_markers_merge.png"), 8, 8)
  }

  # --- cell types (on the masks) ---
  render_and_save(
    plotCells(masks_v, object = tmp, cell_id = "ObjectNumber", img_id = "sample_id",
              colour_by = "celltype",
              colour = list(celltype = metadata(spe)$color_vectors$col_celltype),
              missing_colour = "white", image_title = NULL, scale_bar = NULL,
              legend = NULL, return_plot = TRUE),
    paste0(tag, "_celltype.png"), 10, 8)

  # --- spatial compartment (on the masks) ---
  render_and_save(
    plotCells(masks_v, object = tmp, cell_id = "ObjectNumber", img_id = "sample_id",
              colour_by = "compartment",
              colour = list(compartment = compartment_cols),
              missing_colour = "white", image_title = NULL, scale_bar = NULL,
              legend = NULL, return_plot = TRUE),
    paste0(tag, "_compartment.png"), 10, 8)

  # --- cellular neighborhood (on the masks) ---
  render_and_save(
    plotCells(masks_v, object = tmp, cell_id = "ObjectNumber", img_id = "sample_id",
              colour_by = "cn_celltypes_2",
              colour = list(cn_celltypes_2 = cn_cols),
              missing_colour = "white", image_title = NULL, scale_bar = NULL,
              legend = NULL, return_plot = TRUE),
    paste0(tag, "_cellular_neighborhood.png"), 10, 8)
}
```

## Whole image and zoom
Every image in `image_ids` is rendered as a whole image; when that image has a
zoom window in `crops`, the zoomed region is rendered too. Every output filename
carries the image id and `_whole` / `_zoom`.

```{r generate}
for (i in seq_along(image_ids)) {
  image_id <- image_ids[[i]]
  crop     <- crops[[i]]
  message("=== image: ", image_id, " (", i, "/", length(image_ids), ") ===")

  loaded <- load_image(image_id)
  img <- loaded$img; masks <- loaded$masks; tmp <- loaded$tmp

  # Always the whole image; add the zoom when this image has crop coordinates.
  views <- list(whole = NULL)
  if (!is.null(crop)) views[["zoom"]] <- crop

  for (vn in names(views)) {
    cr      <- views[[vn]]
    # Crop in the native image orientation (so crop coordinates stay in raw-image
    # pixel space), then rotate the view to match the 07 plotSpatial orientation.
    img_v   <- if (is.null(cr)) img   else crop_cil(img,   cr)
    masks_v <- if (is.null(cr)) masks else crop_cil(masks, cr)
    img_v   <- rotate90_ccw(img_v)
    masks_v <- rotate90_ccw(masks_v)
    message("=== view: ", vn, " (", paste(dim(img_v[[1]])[1:2], collapse = " x "), " px) ===")
    plot_region(img_v, masks_v, tmp, markers, paste0(image_id, "_", vn))
  }
}
```
