---
title: "Supplementary: region and marker images"
author: "LazDaria"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params:
  input:  !r Sys.getenv("MAPMET_INPUT", "/mnt/data/input_all")
  output: !r Sys.getenv("MAPMET_OUTPUT", "/mnt/data/output")
  # Image to render. Override from run_analysis.R (see the render_step call).
  image_id: "20220926_20220809_16-006_TU_003"
  # Markers rendered individually and as a colour merge. Either full channel
  # names (as in rownames(spe), e.g. "CD3_Sm152_mean") or the short marker name
  # (e.g. "CD3") - short names are resolved to the matching channel below.
  # (DNA1_Ir193_mean is the Ir193 DNA channel, i.e. DNA2.)
  markers: !r c("CD20_Nd148_mean", "CD3_Sm152_mean", "DNA1_Ir193_mean", "CD14_Lu175_mean", "CD56_Er170_mean")
  # Colour for each marker, used for BOTH its individual image and the merge.
  # Keyed by the short marker name (token before the first "_"); "DNA1" keys the
  # DNA channel. Markers without an entry fall back to a spare-colour palette.
  marker_colours: !r c(CD3 = "red", CD20 = "purple", CD56 = "green", CD14 = "cyan", DNA1 = "blue")
  # Zoom window c(xmin, xmax, ymin, ymax) in pixel coordinates of the image
  # array (x = first image dimension, y = second). Both the whole image AND this
  # zoom are rendered. Set to NULL to render the whole image only.
  crop: !r c(2000, 3000, 2000, 3000)
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/08_generate_images")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))

  if (inherits(plot_obj, "ggplot")) {
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi, bg = "white")
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList"))) {
    pdf(full_path, width = width, height = height)
    draw(plot_obj)
    dev.off()
  } else {
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  message("Saved: ", full_path)
}

knitr::opts_chunk$set(
  echo = TRUE, error = FALSE, message = FALSE, warning = FALSE,
  dev = "png", dpi = 150, fig.width = 8, fig.height = 8
)
```

# Region and marker images

This standalone supplementary report renders, for a single image — as **both the
whole image and a zoomed region**:

* each requested **marker** as a separate pixel image,
* a colour **merge** of all requested markers,
* the **cell types**,
* the **spatial compartment** (tumor core / border / stroma), and
* the **cellular neighborhood**.

The cell-level panels (cell types, compartment, neighborhood) are drawn on the
segmentation **masks** (`plotCells`), not as points. It is meant to be rendered
from `run_analysis.R` with a *customised* parameter list (image id, zoom crop,
marker list) rather than the shared `main_params` — see the
`render_step("analysis/08_generate_images.Rmd", ...)` call there. The
`compartment` and `cn_celltypes_2` annotations are read from
`spe_07_spatial_analysis.rds`, so the chosen image must be one of the
primary-tumour (PT) images processed in [chapter 07](07_spatial_analysis.html).

## Load libraries
```{r load-libraries}
library(SpatialExperiment)
library(cytomapper)
library(cowplot)
library(RColorBrewer)
library(ggplot2)
```

## Load data and images
```{r load-data}
spe <- readRDS(file.path(params$output, "spe_07_spatial_analysis.rds"))

image_id <- params$image_id
if (!image_id %in% spe$sample_id)
  stop("image_id '", image_id, "' is not among the PT images in ",
       "spe_07_spatial_analysis.rds. Available images include: ",
       paste(head(unique(as.character(spe$sample_id))), collapse = ", "), " ...")

tmp <- spe[, spe$sample_id == image_id]

# Multichannel image (pixels) and segmentation mask (cells)
img   <- loadImages(file.path(params$input, "img",   paste0(image_id, ".tif")))
masks <- loadImages(file.path(params$input, "masks", paste0(image_id, ".tif")), as.is = TRUE)

mcols(img)   <- DataFrame(sample_id = names(img))
mcols(masks) <- DataFrame(sample_id = names(masks))

# Name the image channels from the marker (_mean) names, using the image's
# actual channel count (same robust rule as analysis_AD/03_QC_2.Rmd, in case the
# raw image still carries channels dropped from the intensity tables).
marker_names <- rownames(spe)[grepl("_mean$", rownames(spe))]
n_ch <- dim(img[[1]])[3]
channelNames(img) <- if (n_ch <= length(marker_names)) {
  marker_names[seq_len(n_ch)]
} else {
  c(marker_names, paste0("extra_channel_", seq_len(n_ch - length(marker_names))))
}

# Resolve requested markers: accept either a full channel name or a short marker
# name (the token before the first "_", e.g. "CD3" -> "CD3_Sm152_mean").
resolve_markers <- function(requested, channels) {
  token <- sub("_.*$", "", channels)
  vapply(requested, function(r) {
    if (r %in% channels) return(r)
    hit <- channels[token == r]
    if (length(hit)) hit[1] else NA_character_
  }, character(1))
}

markers_full <- resolve_markers(params$markers, channelNames(img))
if (anyNA(markers_full))
  warning("Markers not found as image channels (dropped): ",
          paste(params$markers[is.na(markers_full)], collapse = ", "))
markers <- unname(markers_full[!is.na(markers_full)])
if (!length(markers)) stop("None of the requested markers match the image channels.")
message("Rendering markers: ", paste(markers, collapse = ", "))
```

## Colours and helpers
```{r helpers}
# Cropping a CytoImageList to c(xmin, xmax, ymin, ymax). Handles the
# multichannel image (X, Y, C) and the single-channel mask (X, Y), and clamps
# the window to the image bounds so an out-of-range crop can never error.
crop_cil <- function(cil, crop) {
  imgs <- lapply(seq_along(cil), function(i) {
    im <- cil[[i]]
    d  <- dim(im)
    x1 <- max(1L, min(crop[1], d[1])); x2 <- max(1L, min(crop[2], d[1]))
    y1 <- max(1L, min(crop[3], d[2])); y2 <- max(1L, min(crop[4], d[2]))
    xs <- seq(min(x1, x2), max(x1, x2)); ys <- seq(min(y1, y2), max(y1, y2))
    if (length(d) == 3L) im[xs, ys, , drop = FALSE] else im[xs, ys, drop = FALSE]
  })
  out <- CytoImageList(imgs)
  names(out) <- names(cil)
  mcols(out) <- mcols(cil)
  if (!is.null(channelNames(cil))) channelNames(out) <- channelNames(cil)
  out
}

# core / border / stroma coloured with the ggplot2 default discrete palette
# (scales::hue_pal, alphabetical order: border, core, stroma), so the mask image
# matches the plotSpatial compartment figures in 07_spatial_analysis, which use
# those default colours.
compartment_cols <- setNames(scales::hue_pal()(3), c("border", "core", "stroma"))

# Same Set3 palette / neighborhood numbering as Fig4f in chapter 07.
cn_levels <- levels(factor(spe$cn_celltypes_2))
cn_cols   <- setNames(brewer.pal(max(3, length(cn_levels)), "Set3")[seq_along(cn_levels)],
                      cn_levels)

# Evaluate a plotPixels/plotCells call lazily; on success save + show it and
# return TRUE, on any error report it, skip, and return FALSE - so one bad panel
# never aborts the whole render. A flat channel (a marker with no signal in this
# view) makes EBImage::normalize fail with "clipping range is 0"; such markers
# fail here and are simply left out (also of the merge). `plot_call` is a
# promise, so the plotting only runs inside the tryCatch below.
render_and_save <- function(plot_call, filename, width, height) {
  po <- tryCatch(plot_call,
                 error = function(e) {
                   message("Skipped ", filename, ": ", conditionMessage(e)); NULL
                 })
  if (is.null(po)) return(invisible(FALSE))
  p <- ggdraw(po$plot)
  save_plot(p, filename, width = width, height = height)
  print(p)
  invisible(TRUE)
}

# Colour for each marker (used for BOTH its individual image and the merge),
# keyed by the short marker name (token before the first "_"). Markers without an
# explicit colour fall back to spare hues so they stay distinct in a merge.
.marker_col_map <- local({
  fallback <- c("white", "orange", "yellow", "magenta", "pink", "brown")
  tok  <- sub("_.*$", "", markers)
  cols <- unname(params$marker_colours[tok])   # NA where the short name is unmapped
  miss <- which(is.na(cols))
  if (length(miss))
    cols[miss] <- fallback[((seq_along(miss) - 1) %% length(fallback)) + 1]
  setNames(cols, markers)
})

# Render every panel (individual markers, merge, celltype, compartment,
# neighborhood) for one view (whole image or a crop). tag goes into the filename.
# No legends are drawn - each panel is just the image.
plot_region <- function(img_v, masks_v, tmp, markers, tag) {

  # --- individual markers, each in its assigned merge colour; keep the ones
  #     that render (a marker absent from this view is a flat channel, skipped) ---
  usable <- character(0)
  for (m in markers) {
    ok <- render_and_save(
      plotPixels(img_v, colour_by = m,
                 colour = setNames(list(c("black", .marker_col_map[[m]])), m),
                 bcg = setNames(list(c(0, 1, 1)), m),
                 legend = NULL, return_plot = TRUE),
      paste0(tag, "_marker_", m, ".png"), 8, 8)
    if (isTRUE(ok)) usable <- c(usable, m)
  }

  # --- merge of the markers that rendered, same per-marker colours ---
  if (length(usable)) {
    colour_list <- setNames(lapply(usable, function(mm) c("black", .marker_col_map[[mm]])), usable)
    bcg_list    <- setNames(lapply(usable, function(x) c(0, 1, 1)), usable)
    render_and_save(
      plotPixels(img_v, colour_by = usable, colour = colour_list,
                 bcg = bcg_list, legend = NULL, return_plot = TRUE),
      paste0(tag, "_markers_merge.png"), 8, 8)
  }

  # --- cell types (on the masks) ---
  render_and_save(
    plotCells(masks_v, object = tmp, cell_id = "ObjectNumber", img_id = "sample_id",
              colour_by = "celltype",
              colour = list(celltype = metadata(spe)$color_vectors$col_celltype),
              missing_colour = "white", legend = NULL, return_plot = TRUE),
    paste0(tag, "_celltype.png"), 10, 8)

  # --- spatial compartment (on the masks) ---
  render_and_save(
    plotCells(masks_v, object = tmp, cell_id = "ObjectNumber", img_id = "sample_id",
              colour_by = "compartment",
              colour = list(compartment = compartment_cols),
              missing_colour = "white", legend = NULL, return_plot = TRUE),
    paste0(tag, "_compartment.png"), 10, 8)

  # --- cellular neighborhood (on the masks) ---
  render_and_save(
    plotCells(masks_v, object = tmp, cell_id = "ObjectNumber", img_id = "sample_id",
              colour_by = "cn_celltypes_2",
              colour = list(cn_celltypes_2 = cn_cols),
              missing_colour = "white", legend = NULL, return_plot = TRUE),
    paste0(tag, "_cellular_neighborhood.png"), 10, 8)
}
```

## Whole image and zoom
Both the whole image and (when `params$crop` is set) the zoomed region are
rendered; every output filename carries the image id and `_whole` / `_zoom`.

```{r generate}
# Always the whole image; add the zoom when crop coordinates are provided.
views <- list(whole = NULL)
if (!is.null(params$crop)) views[["zoom"]] <- params$crop

for (vn in names(views)) {
  cr      <- views[[vn]]
  img_v   <- if (is.null(cr)) img   else crop_cil(img,   cr)
  masks_v <- if (is.null(cr)) masks else crop_cil(masks, cr)
  message("=== view: ", vn, " (", paste(dim(img_v[[1]])[1:2], collapse = " x "), " px) ===")
  plot_region(img_v, masks_v, tmp, markers, paste0(image_id, "_", vn))
}
```
