---
title: "Spatial analysis in primary tumor samples"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/data/input_all"
  input_docker: "/mnt/scratch/rstudio/mapmetsc"
  output: "/mnt/scratch/rstudio_spatial"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/10_spatial_analysis")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
    
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {
    
    # save as PNG instead of PDF if you want consistency
    png(full_path, width = width * dpi, height = height * dpi, res = dpi)
    draw(plot_obj)
    dev.off()
    
  } else {
    
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Spatial analysis in PT {#sp-pt}

Herein, we explore spatial interactions in the primary tumor. 
Due to a [bug](https://github.com/BodenmillerGroup/imcRtools) in the `testInteractions` function in imcRtools versions < 1.5.5, we provide another docker image with a newer imcRtools version for spatial analysis. Please pull *lazdaria/mapmetsc_spatial:v1.0* from [docker hub]() and for the present RMD file.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(imcRtools)
library(scales)
library(BiocParallel)
library(RColorBrewer)
library(circlize)
library(patchwork)
library(ggplot2)
library(lisaClust)
library(tidyr)
library(diffcyt)
library(dplyr)
library(CATALYST)
library(ComplexHeatmap)
library(gtools)
library(ggpubr)
library(gridExtra)
library(ggridges)
library(forcats)
library(pheatmap)
library(tidyverse)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). We perform spatial analysis only in the primary tumor samples, since the bone marrow samples are aspirates, where the spatial context is lost. Therefore, we first subset the SPE object to the PT samples and then load cell neighbors that were computed in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP). Therein, we have extracted neighbors by expanding individual high-resolution cell masks by 15 pixels. To be able to load the cell neighbors using the function `read_steinbock`, we will temporarily copy intensities for PT samples into a new folder `intensities-0px-TU`.

```{r temp-folder}
#dir.create(file.path(params$input, "intensities-0px-TU"))
#files <- list.files(file.path(params$input, "intensities-0px"), pattern = "TU", full.names = TRUE)
#file.copy(files, file.path(params$input, "intensities-0px-TU"))
```

We can now read in the neighbors.
For the spatial interaction analysis, we will use the metaclusters for non-tumor cells, since these represent the minority in the PT samples, and the celltype level annotations for the tumor cells (majority). 
Once the data is read into a SPE object, the folder `intensities-0px-TU` can be removed. 

```{r read-data}
spe <- readRDS(file.path(params$input_docker,"spe_final_communities.rds"))
spe <- spe[,spe$tissue=="PT"]

spe_neighbors <- read_steinbock(path=file.path(params$input), image_file=NULL, panel_file = NULL, intensities_folder = "intensities-0px-TU",regionprops_folder = NULL, graphs_folder="neighbors") 
colnames(spe_neighbors) <- paste0(spe_neighbors$sample_id, "_", spe_neighbors$ObjectNumber)
colPair(spe, "neighborhood") <- colPair(spe_neighbors[,colnames(spe)], "neighborhood")

#celltype 2
spe$celltype_2 <- spe$metacluster
spe[,spe$metacluster=="tumor"]$celltype_2 <- spe[,spe$metacluster=="tumor"]$celltype

# generate new color vector
tumor_names <- unique(spe[,spe$metacluster=="tumor"]$celltype)
metadata(spe)$color_vectors$col_celltype_2 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_2 <- c(metadata(spe)$color_vectors$col_celltype_2, metadata(spe)$color_vectors$col_celltype[tumor_names])

#celltype 3
spe$celltype_3 <- spe$metacluster
spe[,spe$metacluster=="T cells" | spe$metacluster=="B cells" | spe$metacluster=="granulocytes" | spe$metacluster=="MO/DC/NK"]$celltype_3 <- "immune"

# generate new color vector
metadata(spe)$color_vectors$col_celltype_3 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_3 <- c(metadata(spe)$color_vectors$col_celltype_3, "immune" = "red")

# remove temporary folder
#unlink(file.path(params$input, "MapMetIP_ProcessedDataset/intensities-0px-TU"), recursive = TRUE)
```

## Interaction analysis

Next, we determine the interactions between cells on the celltype_2 level. For each image, the `testInteractions` function computes the averaged cell type interactions and compares this number against an empirical null distribution generated by permuting all cell labels, while maintaining tissue structure. 

```{r interaction-analysis}
out <- testInteractions(spe,
                        group_by = "sample_id",
                        label = "metacluster",
                        colPairName = "neighborhood",
                        BPPARAM = SerialParam(RNGseed = 221029))
```

These interactions can be shown in a heatmap.

```{r interaction-heatmap, fig.width=5, fig.height=5}

data <- out %>% as_tibble() %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

# Reshaping the data to a wide format suitable for heatmap
data_wide <- data %>%
  pivot_wider(
    names_from = to_label,
    values_from = sum_sigval
  )

data_wide <- tibble::column_to_rownames(data_wide, var = "from_label")

# Define maximum range value for symmetric color scale
#max_range <- max(abs(max(data_wide)), abs(min(data_wide)))  
max_range <- max(abs(50), abs(50))

# Define breaks, making sure 0 is exactly in the middle
breaks <- seq(-max_range, max_range, length.out = 101)  # From -50 to 50

# Create a color palette from blue (negative) to white (zero) to red (positive)
colors <- colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1)

p_heatmap <- pheatmap(data_wide,
         color = colors,
         breaks = breaks,
         cellwidth = 20,
         cellheight = 20,
         fontsize = 10,
         display_numbers = TRUE)  # Optionally display the values in cells

save_plot(p_heatmap, "Fig4b_cell-cell_interactions.png", width = 6, height = 6)
```

## Cellular neighborhoods

In a next step, we detect cellular neighborhoods (CNs). Cells are grouped into neighborhoods based on celltypes in their proximity. For each cell, the fraction of cells of a celltype among its neighbors, is detected. Based on these metrics, the cells can then be aggregated into neighborhoods. The number of neighborhoods has to be predefined. 

```{r c-neighborhoods cn-analysis, fig.width=20, fig.height=50}
spe <- aggregateNeighbors(spe, 
                          colPairName = "neighborhood", 
                          aggregate_by = "metadata", 
                          count_by = "celltype_2")

set.seed(220705)

cn_1 <- kmeans(spe$aggregatedNeighbors, centers = 6)
spe$cn_celltypes_2 <- as.factor(cn_1$cluster)

p_spatial <- plotSpatial(spe, 
            node_color_by = "cn_celltypes_2", 
            img_id = "sample_id", 
            ncols=6,
            node_size_fix = 0.5) +
    scale_color_brewer(palette = "Set3")

save_plot(p_spatial, "Fig4f_spatial_neighborhood_images.png", width = 20, height = 49)

p_spatial
```

Lastly, we can visualize the enrichment of cell types within cellular neighborhoods
using the `regionMap` function of the `lisaClust` package.

```{r regionmap, fig.height=6, fig.width=5}
library(lisaClust)
p_map <- regionMap(spe,
          cellType = "celltype_2",
          region = "cn_celltypes_2")

p_map <- p_map + ggplot2::coord_flip() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_map, "Fig4c_spatial_neighborhood_map.png", width = 8, height = 4)

p_map
```
         
### Differential abundance of CNs

Finally, we will generate a barplot showing the proportions of CNs per sample. To be able to use the diffcyt package, we will first transform the SPE object into a catalyst-compatible object.

```{r transform-data, fig.height=6, fig.width=10}
sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts"), exprs=assay(spe, "counts")))

#recode metadata on progression
sce$progress <- factor(spe$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe$patient_id)
sce$fm_id <- spe$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe$cn_celltypes_2)
sce$metacluster <- factor(spe$metacluster)

sce$condition <- 'na' 

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe$cn_celltypes_2))
```

Next, we will generate a barplot showing the proportion of cells assigned to each of the 6 CNs per sample.

```{r cohort-overview}
set.seed(20231103)

p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=brewer.pal(6, "Set3"), 
                    linkage="ward.D", distance="manhattan") 

save_plot(p, "Fig4d1_spatial_neighborhood_clustering.png", width = 6, height = 4)

p
```

We generate a dendogram for the hierarchical clustering of barplots.

```{r dendogram, fig.height=8, fig.width=7}
#Dendogram
set.seed(20231103)
wide_df <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

wide_df <- data.frame(wide_df)
rownames(wide_df) <- wide_df$sample_id
wide_df <- wide_df %>% select(-sample_id)

dist_matrix <- dist(wide_df, method = "manhattan")
hc <- hclust(dist_matrix, method = "ward.D")

p2 <- plot(hc)

save_plot(p2, "Fig4d2_spatial_neighborhood_clustering_dendogram.png", width = 6, height = 6)
p2
```

In addition, we plot the metadata of these samples in the respective order.

```{r plot-metadata, fig.height=3}
clustered_samples <- levels(p$data$sample_id)
sce$sample_id <- factor(sce$sample_id, levels=clustered_samples)

#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


#assign color codes
col_code <- c("MNA"=rgb(202,0,32, maxColorValue = 255),
              "het"=rgb(146,197,222, maxColorValue = 255),
              "nMNA"=rgb(5,113,176, maxColorValue = 255),
              "unknown"="white",
              "Prog"=rgb(202,0,32, maxColorValue = 255),
              "NoProg"=rgb(5,113,176, maxColorValue = 255)
              )


# Plot metadata
p_anno<- columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) 

save_plot(p_anno, "Fig4d3_spatial_neighborhood_clustering_metadata.png", width = 8, height = 4)

p_anno
```

We then compare the abundance of cellular communities between nMNA and MNA samples.
Therefore, we first prepare data for DA analysis using the `diffcyt` package. 

```{r transform-data2, fig.height=6, fig.width=10}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$MYCN_amp==0 | spe$MYCN_amp==1
sce <- SingleCellExperiment(assays=list(counts=assay(spe[keep_rows,keep_cols], "counts"), exprs=assay(spe[keep_rows,keep_cols], "counts")))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows,keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows,keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows,keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows,keep_cols]$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe[keep_rows,keep_cols]$cn_celltypes_2)
sce$metacluster <- factor(spe[keep_rows,keep_cols]$metacluster)

sce$condition <- sce$mna 

# Add experiment_info
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows,keep_cols]$cn_celltypes_2))

# Define cell type and state markers
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")
```

```{r prepare-data}
#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe[keep_rows,]) %in% type_markers, "type",
                                    ifelse(rownames(spe[keep_rows,]) %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between to genetic subgroups. 

```{r da-analysis}
# Create design matrix
design <- createDesignMatrix(experiment_info, cols_design = "condition")

# Create contrast matrix
contrast <- createContrast(c(0, 1)) 
nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_voom(d_counts,design, contrast
                       )

FDR_cutoff = 0.1
rowData(res_DA)[rowData(res_DA)$p_adj<FDR_cutoff,]
res<- data.frame(rowData(res_DA))
```

Finally, we visualize DA results in boxplots and barplots showing the logFC.

```{r visualize-da2, fig.width=12, fig.height=3}
p <- plotAbundances(sce, k = "celltype", by = "sample") 
comparisons <- c("nMNAvsMNA")

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  color <- brewer.pal(6, "Set3")[cluster_name]
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("nMNA", "MNA"),
              color = "condition", palette =c(color, color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

p_grid <- grid.arrange(grobs = plot_list_fc, ncol = 6)

save_plot(p_anno, "Fig4e2_spatial_neighborhood_DA_mycn_boxplot.png", width = 8, height = 4)

p_grid


```

and in dodge barplots.

```{r barplot-logFC2}
df <- res[, c("logFC", "p_adj")]
df$p_adj_signif <- stars.pval(df$p_adj)
df$p_adj_signif[df$p_adj_signif==" "] <- "ns"
df$cluster <- rownames(df)

p <- ggplot(df, aes(fill=cluster, y=logFC, x=cluster)) + 
  ggtitle("DA - CNs") +
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = F) + 
  scale_fill_manual(values = brewer.pal(6, "Set3")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_text(aes(label = p_adj_signif), vjust = 0)

save_plot(p, "Fig4e1_spatial_neighborhood_DA_mycn_barplot.png", width = 8, height = 4)

p
```

## Patch detection

Subsequently, we detect tumor patches. We first detect connected sets of tumor cells with a predefined min. number of cells making up a patch and then we expand the concave hull around the patch by 1 pixel.

```{r patchDetection-1, fig.height=20, fig.width=20}
spe <- patchDetection(spe, 
                      patch_cells = spe$metacluster == "tumor",
                      img_id = "sample_id",
                      min_patch_size=20,
                      expand_by=1,
                      colPairName = "neighborhood")
```

We can now plot the tumor patch in the neighborhood graph of individual samples.

```{r plot-patch}
p_s1<- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "GD2_Gd155_mean", 
            assay_type = "counts",
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "neighborhood", 
            nodes_first = FALSE, 
            node_size_by = "area", 
            directed = FALSE,
            edge_color_fix = "grey") + 
    scale_size_continuous(range = c(0.1, 2)) +
    ggtitle("GD2")


save_plot(p_s1, "Fig4g1_tumor_patch.png", width = 6, height = 6)

p_s2 <- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "patch_id", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    theme(legend.position = "none")

save_plot(p_s2, "Fig4g2_tumor_patch.png", width = 6, height = 6)

(p_s1 | p_s2)
```
```{r}
p <-plotSpatial(spe[,spe$sample_id == "20220927_20220810_05-009_TU_002"], 
            node_color_by = "GD2_Gd155_mean",
            assay_type = "counts",
            node_size_by = "area",
            img_id = "sample_id") +    scale_size_continuous(range = c(1, 10)) + theme(legend.position = "none")

p2 <- plotSpatial(spe[,spe$sample_id == "20220927_20220810_05-009_TU_002"], 
            node_color_by = "celltype", 
            assay_type = "counts",
            node_size_by = "area",
            img_id = "sample_id") 


p + coord_cartesian(xlim = c(1000, 1500), ylim = c(0, 1000))
```
```{r}
p2 +  scale_color_manual(values = c(
    "marker-lo Tu" = "red",
    "other" = "grey80"
  )) + theme(legend.position = "none") #+ coord_cartesian(xlim = c(1000, 1500), ylim = c(0, 1000))
```

Then we determine the frequency of cells within the patch belonging to a metacluster per sample with respect to progression status of the respective patient.

```{r patch-freq, fig.width=14, fig.height=7}
variable <- "MYCN_amp"
plot_list <- list()
celltypes <- unique(spe$metacluster)

for(i in 1:length(unique(celltypes))){
  
  c <- celltypes[i]
  
  cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
      group_by(patch_id, sample) %>%
      summarize(cell_count = sum(metacluster == c),
                patch_size = n(),
                cell_freq = cell_count / patch_size) %>%
      filter(!is.na(patch_id)) 
    
  mean_cell_freq_per_sample <- cell_freq_per_patch %>%
    group_by(sample) %>%
    summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
    arrange(desc(mean_cell_freq))
  
  df <- data.frame(
    sample = spe$sample,
    condition = colData(spe)[,variable]
  )
  
  progression_per_sample <- df %>%
    group_by(sample) %>%
    summarise(condition = first(condition))
  
  combined_data <- mean_cell_freq_per_sample %>%
    left_join(progression_per_sample, by = "sample")
  
  # Calculate log2FC
  log2fc_df <- combined_data %>%
    group_by(condition) %>%
    summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE)) %>%
    spread(condition, mean_freq) %>%
    mutate(log2FC = log2(`1` / `0`))
  
  stats_df <- compare_means(mean_cell_freq ~ condition,  
                            method = "wilcox.test",  
                            data = combined_data) 
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                            " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_metacluster[as.character(c)])
  
  plot_list[[c]] <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                  color = "condition", palette = rep(color, length(unique(combined_data$condition))),
                  add.params = list(size = 3),
                  add = "jitter",
                  title=c) + 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                  theme(aspect.ratio = 1.6, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
  
}

p_box <- grid.arrange(grobs = plot_list, ncol = 6)

save_plot(p_box, paste0("Fig4i_DA_",variable,".png"), width = 16, height = 12)

p_box

```

```{r patch-freq, fig.width=14, fig.height=7}
variable <- "progression"
plot_list <- list()
celltypes <- unique(spe$metacluster)

for(i in 1:length(unique(celltypes))){
  
  c <- celltypes[i]
  
  cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
      group_by(patch_id, sample) %>%
      summarize(cell_count = sum(metacluster == c),
                patch_size = n(),
                cell_freq = cell_count / patch_size) %>%
      filter(!is.na(patch_id)) 
    
  mean_cell_freq_per_sample <- cell_freq_per_patch %>%
    group_by(sample) %>%
    summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
    arrange(desc(mean_cell_freq))
  
  df <- data.frame(
    sample = spe$sample,
    condition = colData(spe)[,variable]
  )
  
  progression_per_sample <- df %>%
    group_by(sample) %>%
    summarise(condition = first(condition))
  
  combined_data <- mean_cell_freq_per_sample %>%
    left_join(progression_per_sample, by = "sample")
  
  # Calculate log2FC
  log2fc_df <- combined_data %>%
    group_by(condition) %>%
    summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE)) %>%
    spread(condition, mean_freq) %>%
    mutate(log2FC = log2(`1` / `0`))
  
  stats_df <- compare_means(mean_cell_freq ~ condition,  
                            method = "wilcox.test",  
                            data = combined_data) 
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                            " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_metacluster[as.character(c)])
  
  plot_list[[c]] <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                  color = "condition", palette = rep(color, length(unique(combined_data$condition))),
                  add.params = list(size = 3),
                  add = "jitter",
                  title=c) + 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                  theme(aspect.ratio = 1.6, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
  
}

p_box <- grid.arrange(grobs = plot_list, ncol = 6)

save_plot(p_box, paste0("Fig4h_DA_",variable,".png"), width = 16, height = 12)

p_box

```

We repeat the same for immune cells overall. 

```{r patch-freq-immune}
variable <- "progression"
  
c <- "immune"

cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
    group_by(patch_id, sample) %>%
    summarize(cell_count = sum(celltype_3 == c),
              patch_size = n(),
              cell_freq = cell_count / patch_size) %>%
    filter(!is.na(patch_id)) 
  
mean_cell_freq_per_sample <- cell_freq_per_patch %>%
  group_by(sample) %>%
  summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
  arrange(desc(mean_cell_freq))

df <- data.frame(
  sample = spe$sample,
  condition = colData(spe)[,variable]
)

progression_per_sample <- df %>%
  group_by(sample) %>%
  summarise(condition = first(condition))

combined_data <- mean_cell_freq_per_sample %>%
  left_join(progression_per_sample, by = "sample")

# Calculate log2FC
log2fc_df <- combined_data %>%
  group_by(condition) %>%
  summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE)) %>%
  spread(condition, mean_freq) %>%
  mutate(log2FC = log2(`1` / `0`))

stats_df <- compare_means(mean_cell_freq ~ condition,  
                          method = "wilcox.test",  
                          data = combined_data) 
  
stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                          " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")

color <- unname(metadata(spe)$color_vectors$col_celltype_3[as.character(c)])

p <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                color = "condition", palette =c(color, color),
                add.params = list(size = 3),
                add = "jitter",
                title=c) + 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                theme(aspect.ratio = 1.6, legend.position="none") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
p
```

### Distance to tumor-stroma border

Based on detected patches, the distance to the tumor stroma border can be computed. Therefore the [minDistToCells](https://bodenmillergroup.github.io/imcRtools/reference/minDistToCells.html) can be used. Negative values indicate cells within patches and positive values cells outside of patches.

```{r minDistCells}
spe <- minDistToCells(spe, 
                      x_cells = !is.na(spe$patch_id), 
                      img_id = "sample_id")

plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "distToCells", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    scale_color_gradient2(low = "dark blue", mid = "white", high = "dark red")
```

The distance distribution per celltype can be shown in ridge plots.

```{r ridges-distance, fig.height=10,fig.width=8}
p_dist <- ggplot(as.data.frame(colData(spe)) %>%
         mutate(celltype_2 = fct_reorder(celltype_2, distToCells, .fun = median))) + 
  stat_density_ridges(aes(distToCells, celltype_2, fill = celltype_2), 
                      quantiles = c(0.5), quantile_lines = TRUE, scale = 3) +
  geom_vline(xintercept = 0, color = "dark red", linewidth = 0.5) +
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_2) +
  theme_ridges()

save_plot(p_dist, "Fig4j_border_distance.png", width = 6, height = 10)

p_dist
```

Based on the calculated distance for each cell, three compartments - tumor core, tumor/stroma border and stroma - can be defined using a distance threshold. Image `20221018_20220906_03-018_TU_001` lacks tumor patches, therefore all cells are defined as `stroma` for this image.

```{r spatial-compartments}
threshold <- 100

spe$compartment <- "stroma"
spe[, abs(spe$distToCells) <= threshold & spe$sample_id != "20221018_20220906_03-018_TU_001"]$compartment <- "border"
spe[,spe$compartment != "border" & spe$distToCells<=threshold & spe$sample_id != "20221018_20220906_03-018_TU_001"]$compartment <- "core"
```

### DA analysis in spatial compartments

Next, we will plot the differential abundance of metaclusters with respect to spatial compartment on a sample level. Let's again convert the SPE object into a catalyst-compatible object. Decide whether you want to compare `tumor` celltypes or `immune` metaclusters by setting the variable `cells`. Furthermore, select which compartments to compare, by setting `compare` to `core-border`, `core-stroma` or `border-stroma`. 

```{r define markers}

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

```





```{r}
comparisons_list <- c("core-border", "core-stroma", "border-stroma")
cells <- "immune" #"tumor"  # or "immune"

for (compare in comparisons_list) {

  message("Running comparison: ", compare)

  comps <- strsplit(compare, "-")[[1]]

  # ---- Setup ----
  if (cells =="immune"){
    order <- c("T cells", "B cells", "MO/DC/NK", "granulocytes")
    keep_cols <- spe$metacluster %in% order
    color <- metadata(spe)$color_vectors$col_celltype_2
  } else {
    order <- c("marker-lo Tu","GD2lo Tu","CD44+ Tu","CXCR4hi Tu","Ki67hi Tu",
               "GATA3hi Tu","early NB Tu")
    keep_cols <- spe$metacluster == "tumor"
    color <- metadata(spe)$color_vectors$col_celltype_2
  }

  keep_cols <- keep_cols & (spe$compartment %in% comps)
  keep_rows <- rowData(spe)$use_channel 

  sce <- SingleCellExperiment(
    assays = list(
      counts = assay(spe, "counts")[keep_rows, keep_cols],
      exprs  = assay(spe, "counts")[keep_rows, keep_cols]
    )
  )

  # ---- Metadata ----
  sce$fm_id <- spe[keep_rows,keep_cols]$fm_id
  sce$compartment <- factor(spe[keep_rows,keep_cols]$compartment, levels = comps)
  sce$sample_id <- factor(paste0(sce$fm_id, "_", sce$compartment))
  sce$sample <- factor(spe[keep_rows,keep_cols]$sample)
  sce$condition <- factor(sce$compartment, levels = comps)
  sce$cluster_id <- factor(spe[keep_rows,keep_cols]$celltype_2, levels=order)
  sce$metacluster <- factor(spe[keep_rows,keep_cols]$metacluster)

  metadata(sce)$cluster_codes <- data.frame(
    celltype = factor(spe[keep_rows,keep_cols]$celltype_2, levels=order)
  )

  metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>%
    dplyr::group_by(sample_id, condition, sample) %>%
    dplyr::summarise(n_cells=n(), .groups = 'drop') %>%
    as.data.frame()

  # =========================
  # ---- ABUNDANCE PLOT ----
  # =========================
  
  color <- metadata(spe)$color_vectors$col_celltype_2
  
  # IMPORTANT: get the real cluster order used in sce
  cluster_levels <- levels(factor(sce$cluster_id))
  
  # align color vector to cluster levels
  color <- color[cluster_levels]
  
  # enforce factor ordering (THIS fixes mismatch + ordering)
  sce$celltype <- factor(sce$cluster_id, levels = cluster_levels)

  p_abund <- plotAbundances(sce, k = "celltype", by = "sample")
  abundance_data <- p_abund$data %>%
    dplyr::mutate(
      cluster_id = factor(cluster_id, levels = cluster_levels),
      condition = factor(condition, levels = comps),
      Freq_pct = Freq
    )

  abundance_summary <- abundance_data %>%
    dplyr::group_by(condition, cluster_id) %>%
    dplyr::summarise(
      mean_pct = mean(Freq_pct),
      sd_pct = sd(Freq_pct),
      n = dplyr::n(),
      se_pct = dplyr::if_else(n > 1 & !is.na(sd_pct), sd_pct / sqrt(n), 0),
      .groups = "drop"
    )

  p_abund_bar <- ggplot(abundance_summary, aes(x = cluster_id, y = mean_pct, fill = cluster_id)) +
    geom_col(width = 0.75, alpha = 0.45, show.legend = FALSE) +
    geom_errorbar(
      aes(ymin = pmax(mean_pct - se_pct, 0), ymax = pmin(mean_pct + se_pct, 100)),
      width = 0.2
    ) +
    geom_jitter(
      data = abundance_data,
      aes(x = cluster_id, y = Freq_pct, color = cluster_id),
      inherit.aes = FALSE,
      width = 0.15,
      alpha = 0.8,
      size = 1.5,
      show.legend = FALSE
    ) +
    facet_wrap(~condition, nrow = 1) +
    scale_fill_manual(values = color) +
    scale_color_manual(values = color) +
    ylab("% abundance") +
    xlab("celltype") +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold")
    )

  save_plot(
    p_abund_bar,
    paste0("Fig4k_abundance_celltype_barplot_", compare, "_", cells, ".png"),
    width = 12,
    height = 6
  )

  p_abund_box <- ggplot(abundance_data, aes(x = cluster_id, y = Freq_pct, fill = cluster_id, color = cluster_id)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.45, width = 0.7, show.legend = FALSE) +
    geom_jitter(width = 0.15, alpha = 0.8, size = 1.5, show.legend = FALSE) +
    facet_wrap(~condition, nrow = 1) +
    scale_fill_manual(values = color) +
    scale_color_manual(values = color) +
    ylab("% abundance") +
    xlab("celltype") +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold")
    )

  save_plot(
    p_abund_box,
    paste0("Fig4k_abundance_celltype_boxplot_", compare, "_", cells, ".png"),
    width = 12,
    height = 6
  )

  # ---- diffcyt prep ----
  df_list <- lapply(unique(sce$sample_id), function(x){
    as.data.frame(t(counts(sce[,sce$sample_id==x])))
  })
  names(df_list) <- unique(sce$sample_id)

  experiment_info <- metadata(sce)$experiment_info
  experiment_info <- experiment_info[
    match(names(df_list), experiment_info$sample_id),]

  experiment_info$condition <- factor(experiment_info$condition, levels = comps)

  channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
  marker_name  <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)

  marker_class <- ifelse(rownames(rowData(spe)[keep_rows,]) %in% type_markers, "type",
                  ifelse(rownames(rowData(spe)[keep_rows,]) %in% state_markers, "state","other"))

  marker_info <- data.frame(channel_name, marker_name,
                            marker_class = factor(marker_class),
                            row.names = marker_name)

  d_se <- prepareData(df_list, experiment_info, marker_info)
  rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id
  metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

  d_se <- transformData(d_se)
  d_counts <- calcCounts(d_se)

  # ---- DA analysis ----
  design <- createDesignMatrix(experiment_info, cols_design = c("condition", "sample"))

  # logFC = comps[2] - comps[1]
  contrast <- createContrast(c(0, 1, rep(0, length(unique(experiment_info$sample))-1)))

  res_DA <- testDA_edgeR(
    d_counts, design, contrast,
    min_cells = 3,
    min_samples = round(dim(d_counts)[2]*0.2)
  )

  res <- data.frame(rowData(res_DA))

  # =========================
  # ---- BOXPLOTS ----------
  # =========================
  ord <- comps
  plot_list_fc <- list()

  for(i in seq_along(unique(p_abund$data$cluster_id))){
    cluster_name <- unique(p_abund$data$cluster_id)[i]

    proportions <- data.frame(
      p_abund$data[p_abund$data$cluster_id == cluster_name,
                   c("sample_id", "condition", "Freq")]
    )

    p_adj <- round(res[rownames(res)==cluster_name, "p_adj"], 3)
    logFC <- round(res[rownames(res)==cluster_name, "logFC"], 2)

    stats_df <- data.frame(
      p_adj = p_adj,
      group1 = comps[1],
      group2 = comps[2],
      logFC = logFC
    )

    stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
    stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
    stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ") (", stats_df$p_adj_signif, ")")

    color_cluster <- unname(metadata(spe)$color_vectors$col_celltype_2[as.character(cluster_name)]) 

    plot_list_fc[[i]] <- ggboxplot(
      proportions, x="condition", y="Freq", order=ord,
      color="condition", palette=c(color_cluster, color_cluster),
      add="jitter"
    ) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq) * 1.1,
        label="p_fc"
      ) +
      ggtitle(cluster_name) +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
      theme(
        aspect.ratio = 1.5,
        legend.position = "none",
        plot.title = element_text(margin = margin(b = 10))
      )
  }

  p_box <- gridExtra::grid.arrange(grobs = plot_list_fc, ncol = 5)

  save_plot(
    p_box,
    paste0("Fig4k_boxplot_", compare, "_", cells, ".png"),
    width = 12,
    height = 8
  )

  # =========================
  # ---- BARPLOT ----------
  # =========================
  df <- res[, c("logFC", "p_adj")]
  df$p_adj_signif <- stars.pval(df$p_adj)
  df$p_adj_signif[df$p_adj_signif==" "] <- "ns"
  df$cluster <- rownames(df)

  direction_label <- paste0(comps[2], " vs ", comps[1])

  ymax <- max(df$logFC, na.rm = TRUE)
  ymin <- min(df$logFC, na.rm = TRUE)

  p_bar <- ggplot(df, aes(fill = cluster, y = logFC, x = cluster)) + 
    geom_bar(stat = "identity", show.legend = FALSE) + 
    scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_2) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1),
      plot.title = element_text(hjust = 0.5)
    ) +
    geom_text(aes(label = p_adj_signif), vjust = 0) +
    geom_hline(yintercept = 0) +
    ylab("logFC") +
    ggtitle(paste0("Differential abundance (", direction_label, ")")) +
    annotate("text", x = 1, y = ymax*0.9, 
             label = paste0(comps[2], " enriched"), hjust = 0) +
    annotate("text", x = 1, y = ymin*0.9, 
             label = paste0(comps[1], " enriched"), hjust = 0)

  save_plot(
    p_bar,
    paste0("Fig4k_barplot_", compare, "_", cells, ".png"),
    width = 12,
    height = 6
  )
}
```

```{r}
# =========================
# PARAMETERS
# =========================
cells <- "mmune"

if (cells == "immune") {
  order <- c("T cells", "B cells", "MO/DC", "granulocytes")
  keep_meta <- spe$metacluster %in% order
} else {
  order <- c("marker-lo Tu","GD2lo Tu","CD44+ Tu","CXCR4hi Tu","Ki67hi Tu",
               "GATA3hi Tu","early NB Tu")
  keep_meta <- spe$metacluster == "tumor"
}

# =========================
# BUILD CELL TABLE
# =========================
df <- data.frame(
  sample     = spe$sample,
  condition  = spe$compartment,
  cluster_id = spe$celltype_2,
  keep       = keep_meta
)

df <- df[df$keep, ]

# =========================
# COUNT CELLS
# =========================
df_counts <- df %>%
  dplyr::group_by(condition, sample, cluster_id) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop")

# ensure full structure (prevents missing slices)
df_counts <- tidyr::complete(
  df_counts,
  sample, condition, cluster_id,
  fill = list(n = 0)
)

# =========================
# GLOBAL NORMALIZATION (PER CONDITION)
# =========================
df_normalized <- df_counts %>%
  dplyr::group_by(condition, cluster_id) %>%
  dplyr::summarise(
    n = sum(n),
    .groups = "drop"
  ) %>%
  dplyr::group_by(condition) %>%
  dplyr::mutate(
    normalized_count = n / sum(n)
  ) %>%
  dplyr::ungroup()

# =========================
# FACTORS
# =========================
df_normalized$cluster_id <- factor(df_normalized$cluster_id, levels = order)
df_normalized$condition  <- factor(df_normalized$condition,
                                   levels = c("core", "border", "stroma"))

# =========================
# COLORS (SAFE ALIGNMENT)
# =========================
color_vector <- metadata(spe)$color_vectors$col_celltype_2
color_vector <- color_vector[order]

stopifnot(all(!is.na(color_vector)))

# =========================
# LABELS (CONSISTENT n + %)
# =========================
df_normalized <- df_normalized %>%
  dplyr::group_by(condition) %>%
  dplyr::mutate(
    label = ifelse(
      normalized_count > 0,
      paste0(n, "\n(", round(normalized_count * 100, 1), "%)"),
      ""
    )
  ) %>%
  dplyr::ungroup()

# =========================
# PLOT
# =========================
p_pie <- ggplot(df_normalized,
       aes(x = "", y = normalized_count, fill = cluster_id)) +

  geom_bar(stat = "identity", width = 1, color = "white") +

  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +

  scale_fill_manual(values = color_vector, drop = FALSE) +
  coord_polar("y", start = 0) +
  facet_grid(. ~ condition) +
  theme_void() +
  ggtitle("Cell type composition across compartments") +

  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

# =========================
# SAVE
# =========================
save_plot(
  p_pie,
  paste0("Fig4_suppl_piecharts_core-border-stroma_", cells, ".png"),
  width = 10,
  height = 4
)
```


```{r,fig.width=20, fig.height=5}
library(dplyr)
library(ggplot2)

df_counts <- colData(spe) %>%
  as_tibble() %>%
  filter(tissue == "BM") %>%
  group_by(fm_id) %>%
  summarise(
    n_cells = sum(metacluster == "tumor"),
    control = first(control),
    .groups = "drop"
  )

ggplot(df_counts, aes(x = fm_id, y = n_cells)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = n_cells), vjust = -0.3) +
  facet_wrap(~control, scales = "free_x") +   # 🔑 key fix
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  theme_classic() +
  labs(
    x = "fm_id",
    y = "Number of tumor cells (BM only)",
    title = "Tumor cells per BM fm_id (by control)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
p_s2 <- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "compartment", 
            img_id = "sample_id", 
            node_size_fix = 0.5) 

save_plot(p_s2, "Fig4g3_tumor_patch.png", width = 6, height = 6)
```

## Save R object

Finally, we save the R object.

```{r save-rds}
saveRDS(spe, file.path(params$output,"spe_spatial.rds"))
```


## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
