---
title: "Spatial analysis in primary tumor samples"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/data/input_all"
  input_docker: "/mnt/scratch/rstudio/mapmetsc"
  output: "/mnt/scratch/rstudio_spatial"
---

```{r setup, include=FALSE}
# ================================
# Create plots directory
# ================================
plot_dir <- file.path(params$output, "plots/10_spatial_analysis")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}


# Date prefix for descriptive filenames
date_prefix <- format(Sys.Date(), "%Y%m%d")

# ================================
# Helper function to save plots
# ================================
save_plot <- function(plot_obj, filename, width = 8, height = 6, dpi = 300) {
  full_path <- file.path(plot_dir, paste0(date_prefix, "_", filename))
  
  if (inherits(plot_obj, "ggplot")) {
    
    ggsave(full_path, plot = plot_obj, width = width, height = height, dpi = dpi)
    
  } else if (inherits(plot_obj, c("Heatmap", "HeatmapList", "HeatmapAnnotation"))) {
    
    # save as PNG instead of PDF if you want consistency
    png(full_path, width = width * dpi, height = height * dpi, res = dpi)
    draw(plot_obj)
    dev.off()
    
  } else {
    
    dev.copy(png, full_path, width = width * dpi, height = height * dpi, res = dpi)
    dev.off()
  }
  
  message("Saved: ", full_path)
}

# ================================
# Global chunk options - no automatic plot saving
# ================================
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

# Spatial analysis in PT {#sp-pt}

Herein, we explore spatial interactions in the primary tumor. 
Due to a [bug](https://github.com/BodenmillerGroup/imcRtools) in the `testInteractions` function in imcRtools versions < 1.5.5, we provide another docker image with a newer imcRtools version for spatial analysis. Please pull *lazdaria/mapmetsc_spatial:v1.0* from [docker hub]() and for the present RMD file.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(imcRtools)
library(scales)
library(BiocParallel)
library(RColorBrewer)
library(circlize)
library(patchwork)
library(ggplot2)
library(lisaClust)
library(tidyr)
library(diffcyt)
library(dplyr)
library(CATALYST)
library(ComplexHeatmap)
library(gtools)
library(ggpubr)
library(gridExtra)
library(ggridges)
library(forcats)
library(pheatmap)
library(tidyverse)
library(ggdendro)
library(lme4)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). We perform spatial analysis only in the primary tumor samples, since the bone marrow samples are aspirates, where the spatial context is lost. Therefore, we first subset the SPE object to the PT samples and then load cell neighbors that were computed in [MapMetIP](https://github.com/TaschnerMandlGroup/MapMetIP). Therein, we have extracted neighbors by expanding individual high-resolution cell masks by 15 pixels. To be able to load the cell neighbors using the function `read_steinbock`, we will temporarily copy intensities for PT samples into a new folder `intensities-0px-TU`.

```{r temp-folder}
#dir.create(file.path(params$input, "intensities-0px-TU"))
#files <- list.files(file.path(params$input, "intensities-0px"), pattern = "TU", full.names = TRUE)
#file.copy(files, file.path(params$input, "intensities-0px-TU"))
```

We can now read in the neighbors.
For the spatial interaction analysis, we will use the metaclusters for non-tumor cells, since these represent the minority in the PT samples, and the celltype level annotations for the tumor cells (majority). 
Once the data is read into a SPE object, the folder `intensities-0px-TU` can be removed. 

```{r read-data}
spe <- readRDS(file.path(params$input_docker,"spe_final_communities.rds"))
spe <- spe[,spe$tissue=="PT"]

spe_neighbors <- read_steinbock(path=file.path(params$input), image_file=NULL, panel_file = NULL, intensities_folder = "intensities-0px-TU",regionprops_folder = NULL, graphs_folder="neighbors") 
colnames(spe_neighbors) <- paste0(spe_neighbors$sample_id, "_", spe_neighbors$ObjectNumber)
colPair(spe, "neighborhood") <- colPair(spe_neighbors[,colnames(spe)], "neighborhood")

#celltype 2
spe$celltype_2 <- spe$metacluster
spe[,spe$metacluster=="tumor"]$celltype_2 <- spe[,spe$metacluster=="tumor"]$celltype

# generate new color vector
tumor_names <- unique(spe[,spe$metacluster=="tumor"]$celltype)
metadata(spe)$color_vectors$col_celltype_2 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_2 <- c(metadata(spe)$color_vectors$col_celltype_2, metadata(spe)$color_vectors$col_celltype[tumor_names])

#celltype 3
spe$celltype_3 <- spe$metacluster
spe[,spe$metacluster=="T cells" | spe$metacluster=="B cells" | spe$metacluster=="granulocytes" | spe$metacluster=="MO/DC/NK"]$celltype_3 <- "immune"

# generate new color vector
metadata(spe)$color_vectors$col_celltype_3 <- metadata(spe)$color_vectors$col_metacluster
metadata(spe)$color_vectors$col_celltype_3 <- c(metadata(spe)$color_vectors$col_celltype_3, "immune" = "red")

# remove temporary folder
#unlink(file.path(params$input, "MapMetIP_ProcessedDataset/intensities-0px-TU"), recursive = TRUE)
```

## Interaction analysis

Next, we determine the interactions between cells on the celltype_2 level. For each image, the `testInteractions` function computes the averaged cell type interactions and compares this number against an empirical null distribution generated by permuting all cell labels, while maintaining tissue structure. 

```{r interaction-analysis}
out <- testInteractions(spe,
                        group_by = "sample_id",
                        label = "metacluster",
                        colPairName = "neighborhood",
                        BPPARAM = SerialParam(RNGseed = 221029))
```

These interactions can be shown in a heatmap.

```{r interaction-heatmap, fig.width=5, fig.height=5}

data <- out %>% as_tibble() %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE))

# Reshaping the data to a wide format suitable for heatmap
data_wide <- data %>%
  pivot_wider(
    names_from = to_label,
    values_from = sum_sigval
  )

data_wide <- tibble::column_to_rownames(data_wide, var = "from_label")

# Define maximum range value for symmetric color scale
#max_range <- max(abs(max(data_wide)), abs(min(data_wide)))  
max_range <- max(abs(50), abs(50))

# Define breaks, making sure 0 is exactly in the middle
breaks <- seq(-max_range, max_range, length.out = 101)  # From -50 to 50

# Create a color palette from blue (negative) to white (zero) to red (positive)
colors <- colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1)

p_heatmap <- pheatmap(data_wide,
         color = colors,
         breaks = breaks,
         cellwidth = 20,
         cellheight = 20,
         fontsize = 10,
         display_numbers = TRUE)  # Optionally display the values in cells

save_plot(p_heatmap, "Fig4b_cell-cell_interactions.png", width = 6, height = 6)
```

We next repeat the interaction analysis and heatmap visualization separately for nMNA and MNA samples.

```{r interaction-analysis-mycn}
if (any(is.na(spe$MYCN_amp))) {
  warning("Missing MYCN_amp values detected; NA cells are excluded from MYCN-stratified interaction analysis.")
}

valid_mycn <- !is.na(spe$MYCN_amp)
spe_known_mycn <- spe[, valid_mycn]
spe_mycn <- spe_known_mycn[, spe_known_mycn$MYCN_amp == 1]
spe_non_mycn <- spe_known_mycn[, spe_known_mycn$MYCN_amp == 0]

run_interactions_subset <- function(spe_subset, empty_warning) {
  if (ncol(spe_subset) == 0) {
    warning(empty_warning)
    return(NULL)
  }
  
  testInteractions(spe_subset,
                   group_by = "sample_id",
                   label = "metacluster",
                   colPairName = "neighborhood",
                   BPPARAM = SerialParam(RNGseed = 221029))
}

out_non_mycn <- run_interactions_subset(
  spe_non_mycn,
  "No non-MYCN-amplified cells detected; out_non_mycn is set to NULL."
)
out_mycn <- run_interactions_subset(
  spe_mycn,
  "No MYCN-amplified cells detected; out_mycn is set to NULL."
)
```

```{r interaction-heatmap-mycn, fig.width=10, fig.height=5}
make_interaction_heatmap <- function(interaction_tbl) {
  data <- interaction_tbl %>% as_tibble() %>%
    group_by(from_label, to_label) %>%
    summarize(sum_sigval = sum(sigval, na.rm = TRUE), .groups = "drop")
  
  data_wide <- data %>%
    pivot_wider(
      names_from = to_label,
      values_from = sum_sigval
    )
  
  data_wide <- tibble::column_to_rownames(data_wide, var = "from_label")
  
  max_range <- max(abs(50), abs(50))
  breaks <- seq(-max_range, max_range, length.out = 101)
  colors <- colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1)
  
  pheatmap(data_wide,
           color = colors,
           breaks = breaks,
           cellwidth = 20,
           cellheight = 20,
           fontsize = 10,
           display_numbers = TRUE)
}

plot_interaction_heatmap <- function(interaction_tbl, filename) {
  if (is.null(interaction_tbl)) {
    return(invisible(NULL))
  }
  
  p_heatmap <- make_interaction_heatmap(interaction_tbl)
  save_plot(p_heatmap, filename, width = 6, height = 6)
  p_heatmap
}

p_heatmap_non_mycn <- plot_interaction_heatmap(out_non_mycn, "Fig4b_cell-cell_interactions_nMNA.png")
p_heatmap_mycn <- plot_interaction_heatmap(out_mycn, "Fig4b_cell-cell_interactions_MNA.png")

heatmap_grobs <- Filter(
  Negate(is.null),
  list(
    if (is.null(p_heatmap_non_mycn)) NULL else p_heatmap_non_mycn$gtable,
    if (is.null(p_heatmap_mycn)) NULL else p_heatmap_mycn$gtable
  )
)

if (length(heatmap_grobs) > 0) {
  gridExtra::grid.arrange(
    grobs = heatmap_grobs,
    ncol = 2,
    top = "Cell-cell interactions: nMNA and MNA"
  )
} else {
  warning("No MYCN-stratified interaction heatmaps were generated.")
}
```

## Cellular neighborhoods

In a next step, we detect cellular neighborhoods (CNs). Cells are grouped into neighborhoods based on celltypes in their proximity. For each cell, the fraction of cells of a celltype among its neighbors, is detected. Based on these metrics, the cells can then be aggregated into neighborhoods. The number of neighborhoods has to be predefined. 

```{r c-neighborhoods cn-analysis, fig.width=20, fig.height=50}
spe <- aggregateNeighbors(spe, 
                          colPairName = "neighborhood", 
                          aggregate_by = "metadata", 
                          count_by = "celltype_2")

set.seed(220705)

n_neighborhoods <- 6

cn_1 <- kmeans(spe$aggregatedNeighbors, centers = n_neighborhoods)
spe$cn_celltypes_2 <- as.factor(cn_1$cluster)

p_spatial <- plotSpatial(spe, 
            node_color_by = "cn_celltypes_2", 
            img_id = "sample_id", 
            ncols=6,
            node_size_fix = 0.5) +
    scale_color_brewer(palette = "Set3")

save_plot(p_spatial, "Fig4f_spatial_neighborhood_images.png", width = 20, height = 49)

p_spatial
```

Lastly, we can visualize the enrichment of cell types within cellular neighborhoods
using the `regionMap` function of the `lisaClust` package.

```{r regionmap, fig.height=6, fig.width=5}
library(lisaClust)
p_map <- regionMap(spe,
          cellType = "celltype_2",
          region = "cn_celltypes_2")

p_map <- p_map + ggplot2::coord_flip() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_map, "Fig4c_spatial_neighborhood_map.png", width = 8, height = 4)

p_map
```
         
### Differential abundance of CNs

Finally, we will generate a barplot showing the proportions of CNs per sample. To be able to use the diffcyt package, we will first transform the SPE object into a catalyst-compatible object.

```{r transform-data, fig.height=6, fig.width=10}
sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts"), exprs=assay(spe, "counts")))

#recode metadata on progression
sce$progress <- factor(spe$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$tumor_community <- factor(spe$tumor_community)

sce$patient <- factor(spe$patient_id)
sce$fm_id <- spe$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe$cn_celltypes_2)
sce$metacluster <- factor(spe$metacluster)

sce$condition <- 'na' 

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe$cn_celltypes_2))
```

Next, we will generate a barplot showing the proportion of cells assigned to each of the 6 CNs per sample.

```{r cohort-overview}
set.seed(20231103)

distance <- "manhattan"
linkage <- "complete"

p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=brewer.pal(n_neighborhoods, "Set3"), 
                    linkage = linkage, distance= distance) 

save_plot(p, "Fig4d1_spatial_neighborhood_clustering.png", width = 6, height = 4)

p
```

We generate a dendogram for the hierarchical clustering of barplots.

```{r dendrogram, fig.height=8, fig.width=7}
#Dendrogram
set.seed(20231103)
wide_df <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

wide_df <- data.frame(wide_df)
rownames(wide_df) <- wide_df$sample_id
wide_df <- wide_df %>% select(-sample_id)

dist_matrix <- dist(wide_df, method = distance)
hc <- hclust(dist_matrix, method = linkage)

p2 <- plot(hc)

save_plot(p2, "Fig4d2_spatial_neighborhood_clustering_dendogram.png", width = 6, height = 6)
p2
```

In addition, we plot the metadata of these samples in the respective order.

```{r plot-metadata, fig.height=3}
clustered_samples <- levels(p$data$sample_id)
sce$sample_id <- factor(sce$sample_id, levels=clustered_samples)

#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
            tumor_community=first(tumor_community),
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


#assign color codes
col_code <- c("MNA"=rgb(202,0,32, maxColorValue = 255),
              "het"=rgb(146,197,222, maxColorValue = 255),
              "nMNA"=rgb(5,113,176, maxColorValue = 255),
              "unknown"="white",
              "Prog"=rgb(202,0,32, maxColorValue = 255),
              "NoProg"=rgb(5,113,176, maxColorValue = 255),
              "1"="blue","2"="red","3"="green"
              )


# Plot metadata
p_anno<- columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(patient_group=sample_summary$tumor_community, col=list(patientgroup=col_code))

save_plot(p_anno, "Fig4d3_spatial_neighborhood_clustering_metadata.png", width = 8, height = 4)

p_anno
```

```{r fig4d-combined-stacked, fig.height=12, fig.width=8}

ddata <- dendro_data(hc, type = "rectangle")
segs <- segment(ddata)

p_dendro <- ggplot(segs) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  scale_x_continuous(expand = expansion(add = 0.5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    plot.margin = margin(2, 2, 0, 2)
  ) +
  labs(title = "Sample clustering")

meta_df <- sample_summary[clustered_samples, c("mna", "progress", "tumor_community")]
meta_df$sample_id <- factor(rownames(meta_df), levels = clustered_samples)

meta_long <- meta_df %>%
  pivot_longer(cols = c(mna, progress, tumor_community), names_to = "annotation", values_to = "value") %>%
  mutate(
    annotation = factor(annotation, levels = c("mna", "progress", "tumor_community"),
                        labels = c("MYCN", "Progression", "Patient Group"))
  )

p_meta <- ggplot(meta_long, aes(x = sample_id, y = annotation, fill = as.character(value))) +
  geom_tile() +
  scale_fill_manual(values = col_code, na.value = "white") +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "right",
    legend.key.size = unit(0.5, "cm"),
    panel.grid = element_blank(),
    plot.margin = margin(0, 2, 2, 2)
  )

p_abund <- p +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    plot.title = element_blank(),
    plot.margin = margin(0, 2, 0, 2)
  )

# Empirically tuned panel proportions:
# - dendrogram = 1.2 keeps branch structure readable without dominating the figure
# - abundance = 4 prioritizes the main abundance signal (largest information density)
# - metadata = 1.5 keeps annotation strips legible while compact
panel_heights <- c(dendrogram = 1.2, abundance = 4, metadata = 1.5)
p_fig4d <- (p_dendro / p_abund / p_meta) +
  plot_layout(ncol = 1, heights = panel_heights, guides = "collect")

p_fig4d

save_plot(p_fig4d, "Fig4d_combined_stacked.png", width = 8, height = 12)
```

We then compare the abundance of cellular communities between nMNA and MNA samples.
Therefore, we first prepare data for DA analysis using the `diffcyt` package. 

```{r transform-data2, fig.height=6, fig.width=10}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$MYCN_amp==0 | spe$MYCN_amp==1
sce <- SingleCellExperiment(assays=list(counts=assay(spe[keep_rows,keep_cols], "counts"), exprs=assay(spe[keep_rows,keep_cols], "counts")))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows,keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows,keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows,keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows,keep_cols]$fm_id

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id))

sce$cluster_id <- factor(spe[keep_rows,keep_cols]$cn_celltypes_2)
sce$metacluster <- factor(spe[keep_rows,keep_cols]$metacluster)

sce$condition <- sce$mna 

# Add experiment_info
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows,keep_cols]$cn_celltypes_2))

# Define cell type and state markers
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")
```

```{r prepare-data}
#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe[keep_rows,])), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe[keep_rows,]) %in% type_markers, "type",
                                    ifelse(rownames(spe[keep_rows,]) %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between to genetic subgroups. 

```{r da-analysis}
# Create design matrix
design <- createDesignMatrix(experiment_info, cols_design = "condition")

# Create contrast matrix
contrast <- createContrast(c(0, 1)) 
nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_voom(d_counts,design, contrast
                       )

FDR_cutoff = 0.1
rowData(res_DA)[rowData(res_DA)$p_adj<FDR_cutoff,]
res<- data.frame(rowData(res_DA))
```

Finally, we visualize DA results in boxplots and barplots showing the logFC.

```{r visualize-da2, fig.width=12, fig.height=3}
p <- plotAbundances(sce, k = "celltype", by = "sample") 
comparisons <- c("nMNAvsMNA")

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  color <- brewer.pal(n_neighborhoods, "Set3")[cluster_name]
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("nMNA", "MNA"),
              color = "condition", palette =c(color, color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

p_grid <- grid.arrange(grobs = plot_list_fc, ncol = 6)

save_plot(p_anno, "Fig4e2_spatial_neighborhood_DA_mycn_boxplot.png", width = 8, height = 4)

p_grid


```

and in dodge barplots.

```{r barplot-logFC2}
df <- res[, c("logFC", "p_adj")]
df$p_adj_signif <- stars.pval(df$p_adj)
df$p_adj_signif[df$p_adj_signif==" "] <- "ns"
df$cluster <- rownames(df)

p <- ggplot(df, aes(fill=cluster, y=logFC, x=cluster)) + 
  ggtitle("DA - CNs") +
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = F) + 
  scale_fill_manual(values = brewer.pal(6, "Set3")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_text(aes(label = p_adj_signif), vjust = 0)

save_plot(p, "Fig4e1_spatial_neighborhood_DA_mycn_barplot.png", width = 8, height = 4)

p
```

## Patch detection

Subsequently, we detect tumor patches. We first detect connected sets of tumor cells with a predefined min. number of cells making up a patch and then we expand the concave hull around the patch by 1 pixel.

```{r patchDetection-1, fig.height=20, fig.width=20}
spe <- patchDetection(spe, 
                      patch_cells = spe$metacluster == "tumor",
                      img_id = "sample_id",
                      min_patch_size=20,
                      expand_by=1,
                      colPairName = "neighborhood")
```

We can now plot the tumor patch in the neighborhood graph of individual samples.

```{r plot-patch}
p_s1<- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "GD2_Gd155_mean", 
            assay_type = "counts",
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "neighborhood", 
            nodes_first = FALSE, 
            node_size_by = "area", 
            directed = FALSE,
            edge_color_fix = "grey") + 
    scale_size_continuous(range = c(0.1, 2)) +
    ggtitle("GD2")


save_plot(p_s1, "Fig4g1_tumor_patch.png", width = 6, height = 6)

p_s2 <- plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "patch_id", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    theme(legend.position = "none")

save_plot(p_s2, "Fig4g2_tumor_patch.png", width = 6, height = 6)

(p_s1 | p_s2)
```


Then we determine the frequency of cells within the patch belonging to a metacluster per sample with respect to progression status of the respective patient.

```{r patch-freq-mycn, fig.width=14, fig.height=7}
variable <- "MYCN_amp"
plot_list <- list()
celltypes <- unique(spe$metacluster)

for(i in 1:length(unique(celltypes))){
  
  c <- celltypes[i]
  
  cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
      group_by(patch_id, sample) %>%
      summarize(cell_count = sum(metacluster == c),
                patch_size = n(),
                cell_freq = cell_count / patch_size) %>%
      filter(!is.na(patch_id)) 
    
  mean_cell_freq_per_sample <- cell_freq_per_patch %>%
    group_by(sample) %>%
    summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
    arrange(desc(mean_cell_freq))
  
  df <- data.frame(
    sample = spe$sample,
    condition = colData(spe)[,variable]
  )
  
  progression_per_sample <- df %>%
    group_by(sample) %>%
    summarise(condition = first(condition))
  
  combined_data <- mean_cell_freq_per_sample %>%
    left_join(progression_per_sample, by = "sample")
  
  # Calculate log2FC
  log2fc_df <- combined_data %>%
    group_by(condition) %>%
    summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE)) %>%
    spread(condition, mean_freq) %>%
    mutate(log2FC = log2(`1` / `0`))
  
  stats_df <- compare_means(mean_cell_freq ~ condition,  
                            method = "wilcox.test",  
                            data = combined_data) 
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                            " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_metacluster[as.character(c)])
  
  plot_list[[c]] <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                  color = "condition", palette = rep(color, length(unique(combined_data$condition))),
                  add.params = list(size = 3),
                  add = "jitter",
                  title=c) + 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                  theme(aspect.ratio = 1.6, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
  
}

p_box <- grid.arrange(grobs = plot_list, ncol = 6)

save_plot(p_box, paste0("Fig4i_DA_",variable,".png"), width = 16, height = 12)

p_box

```

```{r patch-freq-comm, fig.width=14, fig.height=7}
variable <- "tumor_community"
plot_list <- list()
celltypes <- unique(spe$metacluster)

for(i in 1:length(unique(celltypes))){
  
  c <- celltypes[i]
  
  cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
      group_by(patch_id, sample) %>%
      summarize(cell_count = sum(metacluster == c),
                patch_size = n(),
                cell_freq = cell_count / patch_size) %>%
      filter(!is.na(patch_id)) 
    
  mean_cell_freq_per_sample <- cell_freq_per_patch %>%
    group_by(sample) %>%
    summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
    arrange(desc(mean_cell_freq))
  
  df <- data.frame(
    sample = spe$sample,
    condition = colData(spe)[,variable]
  )
  
  progression_per_sample <- df %>%
    group_by(sample) %>%
    summarise(condition = first(condition))
  
  combined_data <- mean_cell_freq_per_sample %>%
    left_join(progression_per_sample, by = "sample")
  
  # Calculate log2FC
  log2fc_df <- combined_data %>%
    group_by(condition) %>%
    summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE)) %>%
    spread(condition, mean_freq) %>%
    mutate(log2FC = log2(`1` / `0`))
  
  stats_df <- compare_means(mean_cell_freq ~ condition,  
                            method = "wilcox.test",  
                            data = combined_data) 
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                            " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_metacluster[as.character(c)])
  
  plot_list[[c]] <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                  color = "condition", palette = rep(color, length(unique(combined_data$condition))),
                  add.params = list(size = 3),
                  add = "jitter",
                  title=c) + 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                  theme(aspect.ratio = 1.6, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
  
}

p_box <- grid.arrange(grobs = plot_list, ncol = 6)

save_plot(p_box, paste0("Fig4h_DA_",variable,".png"), width = 16, height = 12)

p_box

```

We repeat the same for immune cells overall. 

```{r patch-freq-immune}
variable <- "progression"
  
c <- "immune"

cell_freq_per_patch <- colData(spe) %>% as_tibble() %>%
    group_by(patch_id, sample) %>%
    summarize(cell_count = sum(celltype_3 == c),
              patch_size = n(),
              cell_freq = cell_count / patch_size) %>%
    filter(!is.na(patch_id)) 
  
mean_cell_freq_per_sample <- cell_freq_per_patch %>%
  group_by(sample) %>%
  summarise(mean_cell_freq = mean(cell_freq, na.rm = TRUE))%>%
  arrange(desc(mean_cell_freq))

df <- data.frame(
  sample = spe$sample,
  condition = colData(spe)[,variable]
)

progression_per_sample <- df %>%
  group_by(sample) %>%
  summarise(condition = first(condition))

combined_data <- mean_cell_freq_per_sample %>%
  left_join(progression_per_sample, by = "sample")

# Calculate log2FC
log2fc_df <- combined_data %>%
  group_by(condition) %>%
  summarise(mean_freq = mean(mean_cell_freq, na.rm = TRUE)) %>%
  spread(condition, mean_freq) %>%
  mutate(log2FC = log2(`1` / `0`))

stats_df <- compare_means(mean_cell_freq ~ condition,  
                          method = "wilcox.test",  
                          data = combined_data) 
  
stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
stats_df$p_long <- paste0("log2FC: ", round(log2fc_df$log2FC, 2), 
                          " (", stats_df$p.adj, " ", stats_df$p_adj_signif, ")")

color <- unname(metadata(spe)$color_vectors$col_celltype_3[as.character(c)])

p <- ggboxplot(combined_data, x = "condition", y = "mean_cell_freq",
                color = "condition", palette =c(color, color),
                add.params = list(size = 3),
                add = "jitter",
                title=c) + 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(combined_data$mean_cell_freq), step.increase = 0.4,) +
                theme(aspect.ratio = 1.6, legend.position="none") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
                  
p
```

### Distance to tumor-stroma border

Based on detected patches, the distance to the tumor stroma border can be computed. Therefore the [minDistToCells](https://bodenmillergroup.github.io/imcRtools/reference/minDistToCells.html) can be used. Negative values indicate cells within patches and positive values cells outside of patches.

```{r minDistCells}
spe <- minDistToCells(spe, 
                      x_cells = !is.na(spe$patch_id), 
                      img_id = "sample_id")

plotSpatial(spe[,spe$sample_id == "20220926_20220809_16-006_TU_003"], 
            node_color_by = "distToCells", 
            img_id = "sample_id", 
            node_size_fix = 0.5) +
    scale_color_gradient2(low = "dark blue", mid = "white", high = "dark red")
```

The distance distribution per celltype can be shown in ridge plots.

```{r ridges-distance, fig.height=10,fig.width=8}
p_dist <- ggplot(as.data.frame(colData(spe)) %>%
         mutate(celltype_2 = fct_reorder(celltype_2, distToCells, .fun = median))) + 
  stat_density_ridges(aes(distToCells, celltype_2, fill = celltype_2), 
                      quantiles = c(0.5), quantile_lines = TRUE, scale = 3) +
  geom_vline(xintercept = 0, color = "dark red", linewidth = 0.5) +
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_2) +
  theme_ridges()

save_plot(p_dist, "Fig4j_border_distance.png", width = 6, height = 10)

p_dist
```

Based on the calculated distance for each cell, three compartments - tumor core, tumor/stroma border and stroma - can be defined using a distance threshold. Image `20221018_20220906_03-018_TU_001` lacks tumor patches, therefore all cells are defined as `stroma` for this image.

```{r spatial-compartments}
thresholds <- seq(100, 300, by = 50)
panel_sample_id_1 <- "20220926_20220809_16-006_TU_003"
panel_sample_id_2 <- "20221018_20220906_03-018_TU_004"
# sanitize sample IDs for file-safe output tags
sanitize_sample_tag <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
panel_sample_tag_1 <- sanitize_sample_tag(panel_sample_id_1)
panel_sample_tag_2 <- sanitize_sample_tag(panel_sample_id_2)
y_axis_upper_expand <- 0.1

plot_list_img1 <- list()
plot_list_img2 <- list()
results_list <- list()

for (th in thresholds) {

  # reset compartment
  spe$compartment <- "stroma"

  # border
  spe[
    ,
    abs(spe$distToCells) <= th &
      spe$sample_id != "20221018_20220906_03-018_TU_001"
  ]$compartment <- "border"

  # core
  spe[
    ,
    spe$compartment != "border" &
      spe$distToCells <= th &
      spe$sample_id != "20221018_20220906_03-018_TU_001"
  ]$compartment <- "core"

  # -----------------------------
  # PLOTS
  # -----------------------------

  p_s2 <- plotSpatial(
    spe[, spe$sample_id == panel_sample_id_1],
    node_color_by = "compartment",
    img_id = "sample_id",
    node_size_fix = 0.5
  ) +
    ggtitle(paste("Border size =", th))

  save_plot(
    p_s2,
    paste0("Fig4g3_tumor_patch_", th, ".png"),
    width = 6,
    height = 6
  )

  p_s3 <- plotSpatial(
    spe[, spe$sample_id == panel_sample_id_2],
    node_color_by = "compartment",
    img_id = "sample_id",
    node_size_fix = 0.5
  ) +
    ggtitle(paste("Border size =", th))

  save_plot(
    p_s3,
    paste0("Fig4g3_tumor_patch2_", th, ".png"),
    width = 6,
    height = 6
  )

  plot_list_img1[[as.character(th)]] <- p_s2
  plot_list_img2[[as.character(th)]] <- p_s3
  
  # -----------------------------
  # IMMUNE COUNTS
  # -----------------------------
  
  immune_df <- as.data.frame(
    table(spe$compartment[spe$celltype_3 == "immune"])
  )

  colnames(immune_df) <- c("compartment", "count")

  immune_df$threshold <- th
  immune_df$cell_class <- "immune"

  # -----------------------------
  # TUMOR COUNTS
  # -----------------------------

  tumor_df <- as.data.frame(
    table(spe$compartment[spe$metacluster == "tumor"])
  )

  colnames(tumor_df) <- c("compartment", "count")

  tumor_df$threshold <- th
  tumor_df$cell_class <- "tumor"

  # store
  results_list[[paste0("immune_", th)]] <- immune_df
  results_list[[paste0("tumor_", th)]] <- tumor_df
}

# combine all
results_df <- bind_rows(results_list)

panel_img1 <- wrap_plots(plot_list_img1, ncol = length(thresholds)) +
  plot_annotation(title = panel_sample_id_1)

save_plot(
  panel_img1,
  paste0("Fig4g3_tumor_patch_panel_all_thresholds_", panel_sample_tag_1, ".png"),
  width = 30,
  height = 6
)

panel_img2 <- wrap_plots(plot_list_img2, ncol = length(thresholds)) +
  plot_annotation(title = panel_sample_id_2)

save_plot(
  panel_img2,
  paste0("Fig4g3_tumor_patch_panel_all_thresholds_", panel_sample_tag_2, ".png"),
  width = 30,
  height = 6
)

panel_img1
panel_img2

# order compartments
results_df$compartment <- factor(
  results_df$compartment,
  levels = c("core", "border", "stroma")
)

# -----------------------------
# IMMUNE PLOT
# -----------------------------

immune_plot <- ggplot(
  results_df %>% filter(cell_class == "immune"),
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  labs(
    title = "Immune cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    immune_plot,
    paste0("Fig4g3_tumor_bordersize_immune_barplot.png"),
    width = 6,
    height = 6
  )

immune_plot

# -----------------------------
# TUMOR PLOT
# -----------------------------

tumor_plot <- ggplot(
  results_df %>% filter(cell_class == "tumor"),
  aes(
    x = factor(threshold),
    y = count,
    fill = compartment
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    size = 3
  ) +
  labs(
    title = "Tumor cell counts by compartment",
    x = "Border size",
    y = "Cell count"
  ) +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, y_axis_upper_expand)))

save_plot(
    tumor_plot,
    paste0("Fig4g3_tumor_bordersize_tumor_barplot.png"),
    width = 6,
    height = 6
  )

tumor_plot
```


### DA analysis in spatial compartments

Next, we will plot the differential abundance of metaclusters with respect to spatial compartment on a sample level. We analyse both tumor celltypes and immune metaclusters, looping over all three compartment comparison pairs (`core-border`, `core-stroma`, `border-stroma`). Differential abundance is tested with `diffcyt` (`testDA_edgeR`) controlling for sample-level pairing.

```{r define-markers}

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

```

```{r cluster-order-colors}
# Canonical cluster orders used consistently throughout compartment analyses
tumor_order  <- c("marker-lo Tu", "GD2lo Tu", "CD44+ Tu", "CXCR4+ Tu",
                  "Ki67+ Tu", "bridge-like Tu", "early NB Tu")
immune_order <- c("T cells", "B cells", "MO/DC/NK", "granulocytes")
```






```{r compartment-boxplots-all-comparisons}
# Runs diffcyt (testDA_edgeR) for every compartment pair × cell type,
# then draws combined boxplots with logFC + BH-adjusted p-value labels.

cells_list       <- c("tumor", "immune")
comp_levels      <- c("core", "border", "stroma")
comparison_pairs <- list(c("core","border"), c("core","stroma"), c("border","stroma"))

y_offset_multipliers <- 1 + seq_along(comparison_pairs) * 0.20
y_base_offset        <- 0.01
y_expand_upper       <- (max(y_offset_multipliers) - 1) + 0.25
min_annotation_step  <- 0.05
max_plot_columns     <- 5

keep_rows <- rowData(spe)$use_channel

# helper to build marker_info from the global marker vectors
.make_marker_info <- function(spe, keep_rows) {
  channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 2)
  marker_name  <- sapply(strsplit(rownames(rowData(spe)[keep_rows, ]), "_"), "[[", 1)
  marker_class <- ifelse(rownames(rowData(spe)[keep_rows, ]) %in% type_markers, "type",
                  ifelse(rownames(rowData(spe)[keep_rows, ]) %in% state_markers, "state", "other"))
  data.frame(channel_name, marker_name,
             marker_class = factor(marker_class),
             row.names = marker_name)
}

marker_info_da <- .make_marker_info(spe, keep_rows)

for (cells in cells_list) {

  order     <- if (cells == "immune") immune_order else tumor_order
  keep_meta <- if (cells == "immune") spe$metacluster %in% immune_order else spe$metacluster == "tumor"

  # ------------------------------------------------------------------
  # 1. Precompute diffcyt DA results for every compartment pair
  # ------------------------------------------------------------------
  diffcyt_results <- list()

  for (pair in comparison_pairs) {
    comps     <- pair
    key       <- paste(comps, collapse = "-")
    keep_cols <- keep_meta & spe$compartment %in% comps

    if (sum(keep_cols) < 10) { diffcyt_results[[key]] <- NULL; next }

    sce_da <- SingleCellExperiment(
      assays = list(
        counts = assay(spe, "counts")[keep_rows, keep_cols],
        exprs  = assay(spe, "counts")[keep_rows, keep_cols]
      )
    )
    sce_da$fm_id       <- spe[keep_rows, keep_cols]$fm_id
    sce_da$sample      <- factor(spe[keep_rows, keep_cols]$sample)
    sce_da$compartment <- factor(spe[keep_rows, keep_cols]$compartment, levels = comps)
    sce_da$sample_id   <- factor(paste0(sce_da$fm_id, "_", sce_da$compartment))
    sce_da$condition   <- factor(sce_da$compartment, levels = comps)
    sce_da$cluster_id  <- factor(spe[keep_rows, keep_cols]$celltype_2, levels = order)

    metadata(sce_da)$cluster_codes   <- data.frame(
      celltype = factor(spe[keep_rows, keep_cols]$celltype_2, levels = order)
    )
    metadata(sce_da)$experiment_info <- as.data.frame(colData(sce_da)) %>%
      dplyr::group_by(sample_id, condition, sample) %>%
      dplyr::summarise(n_cells = dplyr::n(), .groups = "drop") %>%
      as.data.frame()

    exp_info           <- metadata(sce_da)$experiment_info
    exp_info           <- exp_info[match(unique(sce_da$sample_id), exp_info$sample_id), ]
    exp_info$condition <- factor(exp_info$condition, levels = comps)

    df_list_da <- lapply(unique(sce_da$sample_id), function(x)
      as.data.frame(t(counts(sce_da[, sce_da$sample_id == x]))))
    names(df_list_da) <- unique(sce_da$sample_id)

    d_se_da <- prepareData(df_list_da, exp_info, marker_info_da)
    rowData(d_se_da)$cluster_id     <- sce_da[, rownames(d_se_da)]$cluster_id
    metadata(d_se_da)$cluster_codes <- metadata(sce_da)$cluster_codes
    d_se_da     <- transformData(d_se_da)
    d_counts_da <- calcCounts(d_se_da)

    design_da   <- createDesignMatrix(exp_info, cols_design = c("condition", "sample"))
    contrast_da <- createContrast(c(0, 1, rep(0, length(unique(exp_info$sample)) - 1)))

    res_DA_da <- tryCatch(
      testDA_edgeR(d_counts_da, design_da, contrast_da,
                   min_cells = 3,
                   min_samples = round(ncol(d_counts_da) * 0.2)),
      error = function(e) { message("diffcyt failed for ", key, ": ", e$message); NULL }
    )
    diffcyt_results[[key]] <- if (!is.null(res_DA_da)) data.frame(rowData(res_DA_da)) else NULL
  }

  # ------------------------------------------------------------------
  # 2. Sample-level proportions (% within compartment) for plotting
  # ------------------------------------------------------------------
  comp_df <- data.frame(
    fm_id       = spe$fm_id,
    compartment = spe$compartment,
    celltype_2  = spe$celltype_2,
    keep        = keep_meta
  )
  comp_df <- comp_df[comp_df$keep & comp_df$compartment %in% comp_levels, ]
  comp_df$compartment <- factor(comp_df$compartment, levels = comp_levels)

  sample_totals <- comp_df %>%
    dplyr::group_by(fm_id, compartment) %>%
    dplyr::summarise(total = dplyr::n(), .groups = "drop")

  plot_data_list <- list()
  for (cluster_name in order) {
    cluster_counts <- comp_df %>%
      dplyr::filter(celltype_2 == cluster_name) %>%
      dplyr::group_by(fm_id, compartment) %>%
      dplyr::summarise(n = dplyr::n(), .groups = "drop")

    plot_data_list[[cluster_name]] <- dplyr::left_join(
      sample_totals, cluster_counts, by = c("fm_id", "compartment")
    ) %>%
      dplyr::mutate(
        n         = dplyr::if_else(is.na(n), 0L, n),
        Freq      = 100 * n / total,
        condition = factor(compartment, levels = comp_levels),
        sample_id = fm_id
      )
  }

  # ------------------------------------------------------------------
  # 3. Collate diffcyt stats for all pairs into one table
  # ------------------------------------------------------------------
  all_stats_list <- list()
  for (cluster_name in order) {
    for (pair in comparison_pairs) {
      key      <- paste(pair, collapse = "-")
      res_pair <- diffcyt_results[[key]]
      if (!is.null(res_pair) && cluster_name %in% rownames(res_pair)) {
        p_adj <- res_pair[cluster_name, "p_adj"]
        logFC <- res_pair[cluster_name, "logFC"]
      } else {
        p_adj <- NA_real_; logFC <- NA_real_
      }
      p_signif <- if (!is.na(p_adj)) gtools::stars.pval(p_adj) else "NA"
      if (p_signif == " ") p_signif <- "ns"
      all_stats_list <- c(all_stats_list, list(data.frame(
        cluster_name = cluster_name,
        group1       = pair[1],
        group2       = pair[2],
        p_adj        = p_adj,
        logFC        = round(logFC, 2),
        label        = paste0(round(logFC, 2), " (",
                              if (!is.na(p_adj)) sprintf("%.3f", p_adj) else "NA",
                              " ", p_signif, ")")
      )))
    }
  }
  all_stats <- dplyr::bind_rows(all_stats_list)

  # ------------------------------------------------------------------
  # 4. Boxplots with diffcyt annotation
  # ------------------------------------------------------------------
  plot_list_comp <- list()
  for (cluster_name in order) {
    proportions <- plot_data_list[[cluster_name]]
    stats_df    <- all_stats[all_stats$cluster_name == cluster_name, ]

    y_max <- max(proportions$Freq, na.rm = TRUE)
    if (!is.finite(y_max)) y_max <- 0
    stats_df$y.position <- if (y_max > 0)
      y_max * y_offset_multipliers[seq_len(nrow(stats_df))] + y_base_offset
    else
      seq_len(nrow(stats_df)) * min_annotation_step

    color_cluster <- unname(
      metadata(spe)$color_vectors$col_celltype_2[as.character(cluster_name)]
    )

    plot_list_comp[[cluster_name]] <- ggboxplot(
      proportions, x = "condition", y = "Freq", order = comp_levels,
      color = "condition", palette = rep(color_cluster, 3),
      add = "jitter"
    ) +
      stat_pvalue_manual(stats_df, y.position = "y.position", label = "label") +
      ggtitle(cluster_name) +
      scale_y_continuous(expand = expansion(mult = c(0.05, y_expand_upper))) +
      theme(aspect.ratio = 1.2, legend.position = "none",
            plot.title = element_text(margin = margin(b = 10)))
  }

  p_comp <- gridExtra::grid.arrange(
    grobs = plot_list_comp,
    ncol  = min(length(plot_list_comp), max_plot_columns)
  )

  save_plot(p_comp,
            paste0("Fig4m_boxplot_compartments_", cells, ".png"),
            width = 12, height = 8)
}
```

```{r composition-stacked-bars}
# Stacked-bar charts showing cell-type composition per compartment,
# with total n displayed on top of each bar.

comp_levels <- c("core", "border", "stroma")

for (cells in c("tumor", "immune")) {

  order     <- if (cells == "immune") immune_order else tumor_order
  keep_meta <- if (cells == "immune") spe$metacluster %in% immune_order else spe$metacluster == "tumor"

  df_bar <- data.frame(
    compartment = spe$compartment,
    cluster_id  = spe$celltype_2,
    keep        = keep_meta
  )
  df_bar <- df_bar[df_bar$keep & df_bar$compartment %in% comp_levels, ]

  # totals per compartment (for the top-of-bar label)
  totals <- df_bar %>%
    dplyr::group_by(compartment) %>%
    dplyr::summarise(N = dplyr::n(), .groups = "drop")

  df_prop <- df_bar %>%
    dplyr::group_by(compartment, cluster_id) %>%
    dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
    dplyr::group_by(compartment) %>%
    dplyr::mutate(pct = 100 * n / sum(n)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      compartment = factor(compartment, levels = comp_levels),
      cluster_id  = factor(cluster_id,  levels = order)
    )

  color_vec <- metadata(spe)$color_vectors$col_celltype_2[order]

  p_stack <- ggplot(df_prop, aes(x = compartment, y = pct, fill = cluster_id)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_text(
      data = totals %>% dplyr::mutate(compartment = factor(compartment, levels = comp_levels)),
      aes(x = compartment, y = 103, label = paste0("n=", N)),
      inherit.aes = FALSE,
      vjust = 0, size = 3.5
    ) +
    scale_fill_manual(values = color_vec, drop = FALSE, name = "Cell type") +
    scale_y_continuous(limits = c(0, 115), breaks = seq(0, 100, 25),
                       expand = expansion(mult = c(0, 0))) +
    labs(title = paste0("Compartment composition (", cells, ")"),
         x = "Compartment", y = "% cells") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.key.size = unit(0.4, "cm"))

  save_plot(p_stack,
            paste0("Fig4_suppl_stackedbar_compartments_", cells, ".png"),
            width = 6, height = 5)
  print(p_stack)
}
```

### MYCN-stratified compartment DA

Here we repeat the compartment DA analysis (diffcyt) separately within MNA and nMNA samples (het samples are excluded). We additionally test whether the MNA vs nMNA difference differs across compartments using a Binomial GLMM with a condition × compartment interaction, overall baseline abundance as a covariate, and a per-patient random intercept (BH-corrected). The model fits all three compartments simultaneously; the interaction terms test whether the MNA vs nMNA difference at border or stroma deviates from the reference compartment (core). Compartment main effects (shifts between compartments within the nMNA reference group) are included in the model but not reported separately here.

```{r mycn-stratified-compartment-da}
comp_levels      <- c("core", "border", "stroma")
comparison_pairs <- list(c("core","border"), c("core","stroma"), c("border","stroma"))
mycn_levels      <- c("nMNA", "MNA")   # het excluded

y_offset_multipliers <- 1 + seq_along(comparison_pairs) * 0.20
y_base_offset        <- 0.01
y_expand_upper       <- (max(y_offset_multipliers) - 1) + 0.25
min_annotation_step  <- 0.05
max_plot_columns     <- 5

keep_rows <- rowData(spe)$use_channel

for (cells in c("tumor", "immune")) {

  order     <- if (cells == "immune") immune_order else tumor_order
  keep_meta <- if (cells == "immune") spe$metacluster %in% immune_order else spe$metacluster == "tumor"

  # ----------------------------------------------------------------
  # A. Per-MYCN-group compartment DA (diffcyt)
  # ----------------------------------------------------------------
  for (mna_grp in mycn_levels) {
    mna_code  <- if (mna_grp == "MNA") 1L else 0L
    keep_mna  <- spe$MYCN_amp == mna_code

    diffcyt_res_mna <- list()
    for (pair in comparison_pairs) {
      comps     <- pair
      key       <- paste(comps, collapse = "-")
      keep_cols <- keep_meta & keep_mna & spe$compartment %in% comps

      if (sum(keep_cols) < 10) { diffcyt_res_mna[[key]] <- NULL; next }

      sce_m <- SingleCellExperiment(
        assays = list(
          counts = assay(spe, "counts")[keep_rows, keep_cols],
          exprs  = assay(spe, "counts")[keep_rows, keep_cols]
        )
      )
      sce_m$fm_id       <- spe[keep_rows, keep_cols]$fm_id
      sce_m$sample      <- factor(spe[keep_rows, keep_cols]$sample)
      sce_m$compartment <- factor(spe[keep_rows, keep_cols]$compartment, levels = comps)
      sce_m$sample_id   <- factor(paste0(sce_m$fm_id, "_", sce_m$compartment))
      sce_m$condition   <- factor(sce_m$compartment, levels = comps)
      sce_m$cluster_id  <- factor(spe[keep_rows, keep_cols]$celltype_2, levels = order)

      metadata(sce_m)$cluster_codes   <- data.frame(
        celltype = factor(spe[keep_rows, keep_cols]$celltype_2, levels = order)
      )
      metadata(sce_m)$experiment_info <- as.data.frame(colData(sce_m)) %>%
        dplyr::group_by(sample_id, condition, sample) %>%
        dplyr::summarise(n_cells = dplyr::n(), .groups = "drop") %>%
        as.data.frame()

      exp_info_m           <- metadata(sce_m)$experiment_info
      exp_info_m           <- exp_info_m[match(unique(sce_m$sample_id), exp_info_m$sample_id), ]
      exp_info_m$condition <- factor(exp_info_m$condition, levels = comps)

      df_list_m <- lapply(unique(sce_m$sample_id), function(x)
        as.data.frame(t(counts(sce_m[, sce_m$sample_id == x]))))
      names(df_list_m) <- unique(sce_m$sample_id)

      d_se_m <- prepareData(df_list_m, exp_info_m, marker_info_da)
      rowData(d_se_m)$cluster_id     <- sce_m[, rownames(d_se_m)]$cluster_id
      metadata(d_se_m)$cluster_codes <- metadata(sce_m)$cluster_codes
      d_se_m     <- transformData(d_se_m)
      d_counts_m <- calcCounts(d_se_m)

      design_m   <- createDesignMatrix(exp_info_m, cols_design = c("condition", "sample"))
      contrast_m <- createContrast(c(0, 1, rep(0, length(unique(exp_info_m$sample)) - 1)))

      res_m <- tryCatch(
        testDA_edgeR(d_counts_m, design_m, contrast_m,
                     min_cells = 3,
                     min_samples = round(ncol(d_counts_m) * 0.2)),
        error = function(e) { message("diffcyt failed: ", e$message); NULL }
      )
      diffcyt_res_mna[[key]] <- if (!is.null(res_m)) data.frame(rowData(res_m)) else NULL
    }

    # sample-level proportions
    comp_df_m <- data.frame(
      fm_id       = spe$fm_id,
      compartment = spe$compartment,
      celltype_2  = spe$celltype_2,
      keep        = keep_meta & keep_mna
    )
    comp_df_m <- comp_df_m[comp_df_m$keep & comp_df_m$compartment %in% comp_levels, ]
    comp_df_m$compartment <- factor(comp_df_m$compartment, levels = comp_levels)

    sample_totals_m <- comp_df_m %>%
      dplyr::group_by(fm_id, compartment) %>%
      dplyr::summarise(total = dplyr::n(), .groups = "drop")

    plot_data_m <- list()
    for (cluster_name in order) {
      cc <- comp_df_m %>%
        dplyr::filter(celltype_2 == cluster_name) %>%
        dplyr::group_by(fm_id, compartment) %>%
        dplyr::summarise(n = dplyr::n(), .groups = "drop")
      plot_data_m[[cluster_name]] <- dplyr::left_join(
        sample_totals_m, cc, by = c("fm_id", "compartment")
      ) %>% dplyr::mutate(
        n = dplyr::if_else(is.na(n), 0L, n),
        Freq = 100 * n / total,
        condition = factor(compartment, levels = comp_levels),
        sample_id = fm_id
      )
    }

    # stats from diffcyt
    stats_m_list <- list()
    for (cluster_name in order) {
      for (pair in comparison_pairs) {
        key      <- paste(pair, collapse = "-")
        res_pair <- diffcyt_res_mna[[key]]
        if (!is.null(res_pair) && cluster_name %in% rownames(res_pair)) {
          p_adj <- res_pair[cluster_name, "p_adj"]
          logFC <- res_pair[cluster_name, "logFC"]
        } else { p_adj <- NA_real_; logFC <- NA_real_ }
        p_signif <- if (!is.na(p_adj)) gtools::stars.pval(p_adj) else "NA"
        if (p_signif == " ") p_signif <- "ns"
        stats_m_list <- c(stats_m_list, list(data.frame(
          cluster_name = cluster_name,
          group1 = pair[1], group2 = pair[2],
          p_adj = p_adj, logFC = round(logFC, 2),
          label = paste0(round(logFC, 2), " (",
                         if (!is.na(p_adj)) sprintf("%.3f", p_adj) else "NA",
                         " ", p_signif, ")")
        )))
      }
    }
    all_stats_m <- dplyr::bind_rows(stats_m_list)

    plot_list_m <- list()
    for (cluster_name in order) {
      proportions_m <- plot_data_m[[cluster_name]]
      stats_df_m    <- all_stats_m[all_stats_m$cluster_name == cluster_name, ]
      y_max_m <- max(proportions_m$Freq, na.rm = TRUE)
      if (!is.finite(y_max_m)) y_max_m <- 0
      stats_df_m$y.position <- if (y_max_m > 0)
        y_max_m * y_offset_multipliers[seq_len(nrow(stats_df_m))] + y_base_offset
      else seq_len(nrow(stats_df_m)) * min_annotation_step

      color_cluster <- unname(
        metadata(spe)$color_vectors$col_celltype_2[as.character(cluster_name)]
      )
      plot_list_m[[cluster_name]] <- ggboxplot(
        proportions_m, x = "condition", y = "Freq", order = comp_levels,
        color = "condition", palette = rep(color_cluster, 3), add = "jitter"
      ) +
        stat_pvalue_manual(stats_df_m, y.position = "y.position", label = "label") +
        ggtitle(cluster_name) +
        scale_y_continuous(expand = expansion(mult = c(0.05, y_expand_upper))) +
        theme(aspect.ratio = 1.2, legend.position = "none",
              plot.title = element_text(margin = margin(b = 10)))
    }

    p_m <- gridExtra::grid.arrange(
      grobs = plot_list_m,
      ncol  = min(length(plot_list_m), max_plot_columns)
    )
    save_plot(p_m,
              paste0("Fig4n_boxplot_compartments_", mna_grp, "_", cells, ".png"),
              width = 12, height = 8)
  }

  # ----------------------------------------------------------------
  # B. Condition × compartment interaction: Binomial GLMM
  #    For each cell type, fit:
  #      cbind(n, total-n) ~ MYCN_amp * compartment + baseline_prop + (1|fm_id)
  #    The main MYCN_amp effect captures MNA vs nMNA in the reference
  #    compartment (core); interaction terms capture whether that difference
  #    changes at border/stroma, controlling for each patient's overall
  #    baseline abundance of the cell type.  BH-correction across all
  #    tests per cell type.
  # ----------------------------------------------------------------
  glmm_input <- data.frame(
    fm_id       = spe$fm_id,
    compartment = spe$compartment,
    celltype_2  = spe$celltype_2,
    MYCN_amp    = spe$MYCN_amp,
    keep        = keep_meta
  )
  glmm_input <- glmm_input[glmm_input$keep &
                              glmm_input$compartment %in% comp_levels &
                              glmm_input$MYCN_amp %in% c(0L, 1L), ]

  glmm_totals <- glmm_input %>%
    dplyr::group_by(fm_id, compartment) %>%
    dplyr::summarise(total = dplyr::n(), .groups = "drop")

  glmm_total_all <- glmm_input %>%
    dplyr::group_by(fm_id) %>%
    dplyr::summarise(total_all = dplyr::n(), .groups = "drop")

  sample_mycn <- glmm_input %>%
    dplyr::select(fm_id, MYCN_amp) %>%
    dplyr::distinct()

  glmm_results_list <- list()

  for (cluster_name in order) {
    ct_counts <- glmm_input %>%
      dplyr::filter(celltype_2 == cluster_name) %>%
      dplyr::group_by(fm_id, compartment) %>%
      dplyr::summarise(n = dplyr::n(), .groups = "drop")

    ct_baseline <- glmm_input %>%
      dplyr::filter(celltype_2 == cluster_name) %>%
      dplyr::group_by(fm_id) %>%
      dplyr::summarise(n_all = dplyr::n(), .groups = "drop") %>%
      dplyr::left_join(glmm_total_all, by = "fm_id") %>%
      dplyr::mutate(baseline_prop = n_all / total_all) %>%
      dplyr::select(fm_id, baseline_prop)

    dat_ct <- dplyr::left_join(glmm_totals, ct_counts,
                               by = c("fm_id", "compartment")) %>%
      dplyr::mutate(n = dplyr::if_else(is.na(n), 0L, n)) %>%
      dplyr::left_join(sample_mycn, by = "fm_id") %>%
      dplyr::left_join(ct_baseline,  by = "fm_id") %>%
      dplyr::mutate(
        compartment   = factor(compartment, levels = comp_levels),
        MYCN_amp      = factor(
          dplyr::recode(as.character(MYCN_amp), `0` = "nMNA", `1` = "MNA"),
          levels = mycn_levels),
        baseline_prop = dplyr::if_else(is.na(baseline_prop), 0, baseline_prop)
      )

    fit <- tryCatch(
      suppressWarnings(
        lme4::glmer(
          cbind(n, total - n) ~ MYCN_amp * compartment + baseline_prop + (1 | fm_id),
          data    = dat_ct,
          family  = binomial,
          # bobyqa is a derivative-free optimizer that is more robust to
          # near-boundary variance estimates common in small GLMMs
          control = lme4::glmerControl(optimizer = "bobyqa")
        )
      ),
      error = function(e) {
        message("GLMM failed for ", cluster_name, ": ", e$message)
        NULL
      }
    )

    if (!is.null(fit)) {
      coefs    <- summary(fit)$coefficients
      # Extract the MNA vs nMNA main effect (at core, the reference compartment)
      # and the interaction terms (whether that difference changes at border/stroma).
      # Compartment main effects (differences across compartments within nMNA)
      # are part of the fitted model but are not separately reported here.
      main_row <- grep("^MYCN_ampMNA$", rownames(coefs))
      # Interaction: MNA × (border or stroma vs core)
      int_rows <- grep("MYCN_ampMNA:compartment", rownames(coefs))

      if (length(main_row) > 0) {
        glmm_results_list <- c(glmm_results_list, list(data.frame(
          cluster_name = cluster_name,
          compartment  = comp_levels[1],
          estimate     = coefs[main_row, "Estimate"],
          p_raw        = coefs[main_row, "Pr(>|z|)"],
          test_type    = "main effect (MNA vs nMNA in core)",
          stringsAsFactors = FALSE
        )))
      }
      for (ir in int_rows) {
        comp_name <- gsub("MYCN_ampMNA:compartment", "", rownames(coefs)[ir])
        glmm_results_list <- c(glmm_results_list, list(data.frame(
          cluster_name = cluster_name,
          compartment  = comp_name,
          estimate     = coefs[ir, "Estimate"],
          p_raw        = coefs[ir, "Pr(>|z|)"],
          test_type    = paste0("interaction (MNA effect change at ", comp_name, " vs core)"),
          stringsAsFactors = FALSE
        )))
      }
    }
  }

  if (length(glmm_results_list) > 0) {
    glmm_stats <- dplyr::bind_rows(glmm_results_list)
    glmm_stats$p_adj    <- p.adjust(glmm_stats$p_raw, method = "BH")
    glmm_stats$p_signif <- gtools::stars.pval(glmm_stats$p_adj)
    glmm_stats$p_signif[glmm_stats$p_signif == " "] <- "ns"

    # Per-compartment boxplots: MNA vs nMNA, annotated with GLMM p-values.
    # For core: main effect p-value.
    # For border/stroma: interaction p-value (change in MNA effect vs core).
    for (comp_b in comp_levels) {
      comp_b_input <- glmm_input[glmm_input$compartment == comp_b, ]

      comp_b_totals <- comp_b_input %>%
        dplyr::group_by(fm_id, MYCN_amp) %>%
        dplyr::summarise(total = dplyr::n(), .groups = "drop")

      comp_b_props_list <- list()
      for (cluster_name in order) {
        cc <- comp_b_input %>%
          dplyr::filter(celltype_2 == cluster_name) %>%
          dplyr::group_by(fm_id, MYCN_amp) %>%
          dplyr::summarise(n = dplyr::n(), .groups = "drop")

        comp_b_props_list[[cluster_name]] <-
          dplyr::left_join(comp_b_totals, cc, by = c("fm_id", "MYCN_amp")) %>%
          dplyr::mutate(
            n         = dplyr::if_else(is.na(n), 0L, n),
            Freq      = 100 * n / total,
            condition = factor(
              dplyr::recode(as.character(MYCN_amp), `0` = "nMNA", `1` = "MNA"),
              levels = mycn_levels)
          )
      }

      plot_list_b <- list()
      for (cluster_name in order) {
        props_b <- comp_b_props_list[[cluster_name]]
        s       <- glmm_stats[glmm_stats$cluster_name == cluster_name &
                                glmm_stats$compartment  == comp_b, ]
        y_max_b <- max(props_b$Freq, na.rm = TRUE)
        if (!is.finite(y_max_b)) y_max_b <- 0

        p_label <- if (nrow(s) > 0)
          paste0(if (comp_b == comp_levels[1]) "GLMM: " else "GLMM int: ",
                 sprintf("%.3f", s$p_adj), " (", s$p_signif, ")")
        else "NA"

        stats_manual_b <- data.frame(
          group1     = "nMNA", group2 = "MNA",
          label      = p_label,
          y.position = max(y_max_b * 1.15, min_annotation_step)
        )
        color_cluster <- unname(
          metadata(spe)$color_vectors$col_celltype_2[as.character(cluster_name)]
        )
        plot_list_b[[cluster_name]] <- ggboxplot(
          props_b, x = "condition", y = "Freq", order = mycn_levels,
          color = "condition", palette = rep(color_cluster, 2), add = "jitter"
        ) +
          stat_pvalue_manual(stats_manual_b, y.position = "y.position", label = "label") +
          ggtitle(cluster_name) +
          scale_y_continuous(expand = expansion(mult = c(0.05, 0.20))) +
          theme(aspect.ratio = 1.5, legend.position = "none",
                plot.title = element_text(margin = margin(b = 10)))
      }

      p_b <- gridExtra::grid.arrange(
        grobs = plot_list_b,
        ncol  = min(length(plot_list_b), max_plot_columns)
      )
      save_plot(p_b,
                paste0("Fig4n_", comp_b, "_MNA_vs_nMNA_GLMM_", cells, ".png"),
                width = 12, height = 6)
    }
  } # end GLMM section B
}
```


## Save R object

Finally, we save the R object.

```{r save-rds}
saveRDS(spe, file.path(params$output,"spe_spatial.rds"))
```


## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
